---
title: "Gene set analysis and GRN inference on PBMCs"
vignette: >
  %\VignetteIndexEntry{Gene set analysis and GRN inference on PBMCs}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

This vignette demonstrates the gene set scoring, spatial autocorrelation, and
gene regulatory network (GRN) inference methods available in `bixverse`. It
picks up where the
[PBMC vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_pbmc.html)
left off: QC, normalisation, PCA, nearest neighbours, clustering and UMAP
are all assumed to be done already. If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the [introductory vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_intro.html),
please do so first.

The methods fall into two broad categories:

- The first, module scores, AUCell, and VISION, takes *pre-defined* gene sets 
  and scores them per cell. 
- The second, Hotspot and SCENIC, *discovers* gene programmes and regulatory
  relationships from the data itself.

We use the PBMC3k dataset throughout. At 2,700 cells this is likely too small 
for the GRN methods to produce biologically meaningful results, but it is large
enough to show the API and the workflow end to end. On real datasets with tens
of thousands of cells the same code applies unchanged. Also, as the 

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(magrittr)
```

::: {.callout-note}
The vignette was built on a GitHub runner which is a 2-core machine with ~8 GB
of memory. That this works is incredible enough, but take this into account when 
looking at this vignette.
:::

### Rebuilding the processed object

We reconstruct the processed `SingleCells` object from the PBMC vignette. The
chunk is hidden for brevity: it downloads the data, runs QC, HVG selection,
PCA, neighbour computation, Leiden clustering and UMAP.

```{r}
#| label: rebuild-object
#| code-fold: true
#| code-summary: "Rebuild the PBMC3k object (click to expand)"

pbmc3k_path <- bixverse:::download_pbmc3k()
tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)
mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")

var <- get_sc_var(sc_object)
ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)

# gene set proportions for QC
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = FALSE
)

# MAD QC
qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]
metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)
qc <- run_cell_qc(metrics, directions, threshold = 3)
sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

# HVG, PCA, neighbours, clustering, UMAP
sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, sparse_svd = TRUE)
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
sc_object <- find_clusters_sc(sc_object, res = 1.5, name = "leiden_clusters")
sc_object <- umap_sc(sc_object, knn_method = "annoy")

# UMAP coordinates for reuse
umap_dt <- as.data.table(
  get_embedding(sc_object, "umap"),
  keep.rownames = "cell_id"
)
```

## Gene set scoring

### Module scores

The simplest approach to gene set activity is the module score from
[Tirosh et al. (2016)](https://www.science.org/doi/10.1126/science.aad0501):
for each cell, compute the mean expression of the gene set and subtract the
mean expression of a size-matched control set drawn from the same expression
bins. We will just use some lineage-specific genes to have pretty 
visualisations.

```{r}
#| label: module-scores
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Module scores for lineage genes projected onto UMAP"

lineage_sets <- list(
  `T cell` = symbol_to_ensembl[c(
    "CD3D",
    "CD3E",
    "CD3G",
    "CD2",
    "IL7R",
    "CD7",
    "LEF1",
    "TCF7",
    "LTB",
    "TRAC"
  )],
  `Cytotoxic NK` = symbol_to_ensembl[c(
    "NKG7",
    "GNLY",
    "GZMA",
    "GZMB",
    "GZMH",
    "PRF1",
    "CST7",
    "KLRD1",
    "KLRB1",
    "FGFBP2"
  )],
  `B cell` = symbol_to_ensembl[c(
    "CD79A",
    "CD79B",
    "MS4A1",
    "BANK1",
    "IGHM",
    "IGHD",
    "CD74",
    "HLA-DQA1",
    "TCL1A",
    "VPREB3"
  )],
  `Monocyte` = symbol_to_ensembl[c(
    "CD14",
    "LYZ",
    "S100A8",
    "S100A9",
    "S100A12",
    "CST3",
    "FCN1",
    "VCAN",
    "MNDA",
    "TYROBP"
  )]
)

# drop any NAs
lineage_sets <- lapply(lineage_sets, function(x) x[!is.na(x)])

module_scores <- module_scores_sc(
  object = sc_object,
  gs_list = lineage_sets,
  .verbose = FALSE
)

ms_dt <- as.data.table(module_scores, keep.rownames = "cell_id")
ms_dt <- merge(ms_dt, umap_dt, by = "cell_id")

ms_long <- melt(
  ms_dt,
  id.vars = c("cell_id", "umap_1", "umap_2"),
  measure.vars = names(lineage_sets),
  variable.name = "phase",
  value.name = "score"
)

ggplot(ms_long, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = score), size = 0.3) +
  scale_colour_viridis_c() +
  facet_wrap(~phase, ncol = 2) +
  theme_bw() +
  labs(colour = "Module\nscore")
```

### AUCell

[AUCell](https://www.nature.com/articles/nmeth.4463) ranks genes within each
cell by expression and computes the area under the recovery curve for the gene
set. This is more robust to outliers than a simple mean. `bixverse` provides
two flavours: a Wilcoxon-statistic-based AUC and a standard AUROC. The two are
highly correlated; we show the Wilcoxon variant here.

```{r}
#| label: aucell
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "AUCell scores for lineage genes projected onto UMAP"

auc_scores <- aucell_sc(
  object = sc_object,
  gs_list = lineage_sets,
  auc_type = "wilcox",
  .verbose = FALSE
)

auc_dt <- as.data.table(auc_scores, keep.rownames = "cell_id")
auc_dt <- merge(auc_dt, umap_dt, by = "cell_id")

auc_long <- melt(
  auc_dt,
  id.vars = c("cell_id", "umap_1", "umap_2"),
  measure.vars = names(lineage_sets),
  variable.name = "phase",
  value.name = "auc"
)

ggplot(auc_long, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = auc), size = 0.3) +
  scale_colour_viridis_c() +
  facet_wrap(~phase) +
  theme_bw() +
  labs(colour = "AUC")

```

### VISION with spatial autocorrelation

Scoring gene sets is useful, but a natural follow-up question is: *are these
scores spatially structured on the cell neighbourhood graph, or just noise?*
[VISION](https://www.nature.com/articles/s41467-019-12235-0) provides a
permutation-based test of spatial autocorrelation (Geary's C) on the kNN graph.
Gene sets with significant autocorrelation are those whose activity varies
smoothly across cell neighbourhoods -- a strong indication that the signal is
biologically meaningful rather than random.

VISION expects gene sets in a signed format with `pos` and (optionally) `neg`
components. In the cause of our cell lineage genes, we could use the other 
cell type's markers as negatives, but we will just leave it blank here.

```{r}
#| label: vision
#| fig-cap: "Spatial autocorrelation test for cell cycle gene sets"

vision_gs <- lapply(lineage_sets, function(genes) list(pos = genes))

vision_res <- vision_w_autocor_sc(
  object = sc_object,
  gs_list = vision_gs,
  embd_to_use = "pca",
  vision_params = params_sc_vision(n_perm = 500L),
  .verbose = TRUE
)

head(vision_res$auto_cor_dt)
```

Unsurprisingly, all of these gene sets show highly significant spatial 
correlation.

## Hotspot

Where the methods above test *pre-defined* gene sets, Hotspot
([DeTomaso & Yosef, 2021](https://www.cell.com/cell-systems/fulltext/S2405-4712(21)00114-9))
discovers gene programmes *de novo* by testing each gene for spatial
autocorrelation on the kNN graph and then grouping the significant genes by
their local correlation structure.

### Gene autocorrelation

The first step computes Geary's C for every gene against the neighbour graph.
Genes with significant autocorrelation are those whose expression varies
smoothly across neighbouring cells, i.e., genes that mark spatially coherent
programmes.

```{r}
#| label: hotspot-autocor

hotspot_autocor <- hotspot_autocor_sc(
  object = sc_object,
  .verbose = FALSE
)

hotspot_autocor[, gene_symbol := ensembl_to_symbol[gene_id]]

head(hotspot_autocor[order(fdr)], 20L)
```

Known PBMC markers (LYZ, CD3D, NKG7, etc.) should appear among the top hits.

### Gene-gene local correlations and modules

Taking the significant genes forward, we compute pairwise local correlations
and cluster them into modules. Each module represents a group of genes that are
not only individually autocorrelated but also co-vary locally - a much
stronger signal than global correlation alone.

```{r}
#| label: hotspot - gene correlations

sig_genes <- hotspot_autocor[fdr <= 0.05, gene_id]

hotspot_cor <- hotspot_gene_cor_sc(
  object = sc_object,
  genes_to_take = sig_genes,
  .verbose = TRUE
)

hotspot_cor
```

This returns a HotSpot result. Let's add the membership and plot it.

```{r}
#| label: hotspot - plot modules

hotspot_cor <- generate_hotspot_membership(hotspot_cor)

# this will only plot a subsample of 500 genes for speed (stratified by
# module). You can control this via a parameter in the plotting function
plot(hotspot_cor)
```

Let's pull out the genes:

```{r}
#| label: hotspot - module membership

membership <- get_hotspot_membership(hotspot_cor)
membership[, gene_symbol := ensembl_to_symbol[gene_id]]

membership <- membership[!is.na(cluster_member)]

head(membership[order(cluster_member)], 20L)
```

The resulting modules should roughly correspond to the major cell-type
programmes in PBMCs (T cell, monocyte, B cell, NK cell signatures, etc.). Let's
visualise this with AUCell check it out.

```{r}
#| label: hotspot - plot the gene sets

hotspot_gene_sets <- membership %$% split(gene_id, cluster_member)

hotspot_gene_sets <- lapply(hotspot_gene_sets, function(genes) {
  list(pos = genes)
})

vision_scores_hotspot <- vision_sc(
  object = sc_object,
  gs_list = hotspot_gene_sets,
  .verbose = FALSE
)

vision_scores_dt <- as.data.table(
  vision_scores_hotspot,
  keep.rownames = "cell_id"
)
vision_scores_dt <- merge(vision_scores_dt, umap_dt, by = "cell_id")


vision_scores_long <- melt(
  vision_scores_dt,
  id.vars = c("cell_id", "umap_1", "umap_2"),
  measure.vars = names(hotspot_gene_sets),
  variable.name = "gene_set",
  value.name = "vision_score"
)

# min max individually for prettier visualisations
vision_scores_long[,
  vision_score_scaled := (vision_score - min(vision_score)) /
    (max(vision_score) - min(vision_score)),
  by = gene_set
]

ggplot(vision_scores_long, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = vision_score_scaled), size = 0.3) +
  scale_colour_viridis_c() +
  facet_wrap(~gene_set) +
  theme_bw() +
  labs(colour = "Vision (scaled)")
```

## SCENIC

[SCENIC](https://www.nature.com/articles/nmeth.4463) infers gene regulatory
networks by asking: for each gene, which transcription factors (TFs) best
predict its expression? The original implementation uses either random forests
or GRNBoost2 (gradient boosted trees) to regress each target gene on all TFs,
then extracts feature importances as a proxy for regulatory strength.

The `bixverse` implementation re-implements this in Rust with several
optimisations that make the RF and ExtraTrees paths substantially faster than
the single-target-at-a-time approach used by GRNBoost2:

- **Quantisation**: TF expression values are discretised into 256 bins (u8),
  so histogram construction during tree building operates on single bytes and
  fits comfortably in L1 cache.
- **Multi-output batching**: up to 64 target genes share a single tree
  structure. The histogram construction cost is paid once per node per feature,
  but the split score aggregates variance reduction across all targets in the
  batch. This amortises the dominant cost of tree building across targets.
- **Correlated gene batching**: rather than assigning genes to batches
  randomly, an optional SVD + k-means step groups co-expressed genes together
  so that the shared tree structure is more informative for each target in the
  batch.
- **GRNBoost2 with histogram subtraction**: for the gradient boosted path
  (single-target), the code builds full-feature histograms at each node and
  derives the larger child via parent-minus-smaller subtraction, avoiding a
  redundant scan of the larger child's samples. OOB early stopping prevents
  overfitting and is the main source of speedup.

On the 2,700-cell PBMC3k dataset these optimisations are not that noticeable...
The data is too small for any method to take long. On datasets with tens of
thousands of cells and thousands of TFs, the RF/ET multi-output path
comfortably outperforms GRNBoost2. You need to just decide if you are okay with
the batching of genes. Generally speaking, the big signals will be recovered
again and again (from empirical testing).

### Gene filtering

SCENIC first filters genes by minimum total counts and minimum proportion of
expressing cells to remove uninformative targets.

```{r}
#| label: scenic-gene-filter

scenic_genes <- scenic_gene_filter_sc(
  object = sc_object,
  scenic_params = params_scenic(min_counts = 50L),
  .verbose = FALSE
)

length(scenic_genes)
```

### Transcription factor list

We need a list of known TFs. The Aerts lab provides a curated list for human
which we can download and map to Ensembl IDs.

```{r}
#| label: scenic-tfs

tf_dt <- data.table::fread(
  "https://resources.aertslab.org/cistarget/tf_lists/allTFs_hg38.txt",
  header = FALSE,
  col.names = "tf"
)
tf_dt[, gene_id := symbol_to_ensembl[tf]]
tf_dt <- tf_dt[!is.na(gene_id)]

nrow(tf_dt)
```

### GRN inference

With genes and TFs in hand, we run the GRN inference step. Here we use
random forests with a batch size of 64 to illustrate the multi-output path.

```{r}
#| label: scenic-grn

scenic_res <- scenic_grn_sc(
  object = sc_object,
  tf_ids = tf_dt$gene_id,
  genes_to_take = scenic_genes,
  scenic_params = params_scenic(
    learner_type = "randomforest",
    gene_batch_size = 64L
  ),
  .verbose = TRUE
)

scenic_res
```

If you are bored, you can run both methods (`learner_type = "grnboost2"`) and 
compare the importance scores. You will see very high correlations here (and
can explore the speed differences...)

### TF-to-gene refinement

The raw importance matrix is large and noisy. The next steps winnow it down to
a candidate regulatory network by keeping only the top TF-gene pairs and
filtering by pairwise expression correlation. The approach used here is that
per gene only TFs with an 
`importance score ≥ mean(importance_score) + n_sd * sd(importance_score)` 
per given gene are kept. Note that `n_sd` should be adjusted depending on the
learner: RF and ExtraTrees spread importance more evenly across TFs than
GRNBoost2, so a lower threshold (e.g. `n_sd = 1`) is appropriate, whereas
GRNBoost2 concentrates importance onto fewer TFs and tolerates `n_sd = 1.5` or
even `n_sd = 2`.

```{r}
#| label: scenic-tf-to-gene

scenic_res <- identify_tf_to_genes(
  scenic_res,
  n_sd = 2,
  .verbose = TRUE
)

scenic_res <- tf_to_genes_correlations(
  x = scenic_res,
  object = sc_object,
  cor_filter = 0.01,
  .verbose = TRUE
)

tf_gene_dt <- get_tf_to_gene(scenic_res)
tf_gene_dt[, tf_symbol := ensembl_to_symbol[tf]]
tf_gene_dt[, gene_symbol := ensembl_to_symbol[gene]]

head(tf_gene_dt[order(-importance)], 5L)
```

### CisTarget motif enrichment

The final (and optional) step in the SCENIC pipeline validates the predicted
TF-gene links against known motif binding sites. This requires downloading
the CisTarget motif ranking database and motif-to-TF annotation files from
the [Aerts lab resources](https://resources.aertslab.org/cistarget/), which
together are roughly 1.5 GB. We therefore show the code but do not evaluate
it by default.

```{r}
#| label: scenic-cistarget
#| eval: false

# Download the CisTarget reference files (cached after first download)
paths <- download_cistarget_hg38()

rankings <- read_motif_ranking(paths$rankings)
annotations <- read_motif_annotation_file(paths$motif_annotations)

# Run motif enrichment on the TF-gene links
scenic_res <- tf_to_genes_motif_enrichment(
  x = scenic_res,
  motif_rankings = rankings,
  annot_data = annotations,
  # we store everything as Ensembl... The SCENIC data has gene symbols, hence,
  # the need to provide conversion
  gene_id_to_symbol = ensembl_to_symbol,
  cis_target_params = params_cistarget(
    nes_threshold = 3.0
  ),
  only_high_conf_tf = FALSE,
  .verbose = TRUE
)

# CisTarget results per TF
cistarget_dt <- get_cistarget_res(scenic_res)
head(cistarget_dt[order(-nes)])

# The TF-to-gene table now has a leading_edge column indicating which
# genes fall within the motif recovery curve
tf_gene_final <- get_tf_to_gene(scenic_res)

head(tf_gene_final[(in_leading_edge)][order(-importance)])
```

This is how you work with all types of `"bag of genes"` analyses for single
cell in `bixverse`.

## Clean up
```{r}
#| label: clean-up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```