---
title: "Cell trees with Bonsai"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Cell trees with Bonsai}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

UMAP and t-SNE squash a kNN graph into two dimensions and lose most of the
global structure on the way. Bonsai (de Groot, et al., Nat. Biotechnol., 2026)
takes a different route: it reconstructs a tree over the cells, every cell a
leaf and every internal node an inferred ancestral state, with branch lengths
that carry how much changed between them. The tree is the representation; a 2D
layout is just a way to draw it.

What makes it work is that Bonsai uses the error bar on every measurement, not
only the measurement. Those come from Sanity (Breda, et al., Nat. Biotechnol.,
2021), which turns raw UMI counts into posterior log fold changes with error
bars, gene by gene. `bonsai_sc()` runs both: counts from disk, Sanity, Bonsai,
layout. Both are Rust ports under the hood,
[`sanity-sc-rs`](https://github.com/GregorLueg/sanity-sc-rs) and
[`bonsai-rs`](https://github.com/GregorLueg/bonsai-rs).

We use the PBMC3k data here. If you have not worked through the
[PBMC vignette](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html),
do that first; the preprocessing below is the same, just condensed.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(bixverse.plots)
library(magrittr)
```

## Preparing the data

Loading, QC and the usual HVG, PCA, neighbours and clustering chain. Bonsai
itself needs none of the last four, but the clusters and cell types make the
tree readable, and the UMAP gives us something to compare against.

```{r}
#| label: load and qc
pbmc3k_path <- download_pbmc3k()
tempdir_pbmc <- tempdir()

# load
sc_object <- load_mtx(
  object = SingleCells(dir_data = tempdir_pbmc),
  sc_mtx_io_param = get_cell_ranger_params(pbmc3k_path),
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")
var <- get_sc_var(sc_object)

# doublets
scrublet_res <- scrublet_sc(sc_object)

sc_object <- add_sc_new_obs(
  object = sc_object,
  obs_data = get_data(scrublet_res)
)

cells_without_doublets <- sc_object[[c("doublet", "cell_id")]][
  !(doublet),
  cell_id
]

sc_object <- set_cells_to_keep(
  x = sc_object,
  cells_to_keep = cells_without_doublets
)

# qc
sc_object <- gene_set_proportions_sc(
  sc_object,
  list(MT = var[grepl("^MT-", gene_symbol), gene_id]),
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]
qc <- run_cell_qc(
  metrics = list(
    log10_lib_size = log10(qc_df$lib_size),
    log10_nnz = log10(qc_df$nnz),
    MT = qc_df$MT
  ),
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = c(
    log10_lib_size = "twosided",
    log10_nnz = "twosided",
    MT = "above"
  ),
  threshold = 3
)
sc_object <- set_cells_to_keep(sc_object, qc_df[!qc$combined, cell_id])

# hvg, pca, knn, clustering, umap
sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = TRUE)
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  .verbose = FALSE
)
sc_object <- find_clusters_sc(sc_object, res = 1, name = "leiden_clusters")
sc_object <- umap_sc(sc_object, .verbose = FALSE)

sc_object
```

Cell types via scType, with the same markers as the PBMC vignette.

```{r}
#| label: cell types
cell_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  IL7R = "CD4+ T",
  CD4 = "CD4+ T",
  CD8A = "CD8+ T",
  CD8B = "CD8+ T",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD14 = "CD14+ Mono",
  LYZ = "CD14+ Mono",
  FCGR3A = "CD16+ Mono",
  CDKN1C = "CD16+ Mono",
  GNLY = "NK",
  NKG7 = "NK",
  FCER1A = "mDC",
  CD1C = "mDC",
  LILRA4 = "pDC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)

cell_markers_dt <- stack(cell_markers) %>%
  as.data.table() %>%
  setnames(c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var$gene_id[match(gene_symbol, var$gene_symbol)]] %>%
  .[!is.na(gene_id)]

sctype_scores <- calc_sc_type_scores(
  object = sc_object,
  cell_marker_list = prepare_cell_markers(sc_object, cell_markers_dt)
)
obs <- get_sc_obs(sc_object, filtered = TRUE)
cell_type_anno <- score_clusters(sctype_scores, obs$leiden_clusters)
sc_object[["cell_type"]] <- cell_type_anno$cell_type[
  match(obs$leiden_clusters, cell_type_anno$cluster_id)
]
```

## Running Bonsai

One call. No HVG selection needed: by default every gene is a candidate, and
Sanity streams them through in chunks of 1,024, keeping only the genes with
enough signal over their own noise (`min_signal_to_noise`, 1 by default). That
is the gene selection of the Bonsai paper. Memory stays at one chunk of counts
plus the survivors, whatever the number of candidates.

```{r}
#| label: bonsai
tree <- bonsai_sc(sc_object, .verbose = TRUE)

tree
```

Where did the time go? Every stage is timed, and `total` is the whole call as R
saw it, so a gap between `total` and the stages would be time spent somewhere
nobody measured. At this size Sanity is most of the run: it fits every one of
the candidate genes, while the search only sees the few hundred that survive.

```{r}
#| label: timings
tree$timings
```

The search reports its steps too: the loglikelihood after each and what it
took. Steps 4 and 7 re-fit the branch lengths, 5 and 6 rearrange the tree
(subtree pruning and regrafting, then nearest-neighbour interchanges), and 8
collapses the zero-length edges the rest leave behind.

```{r}
#| label: steps
tree$steps
```

### Which genes made it?

Most genes carry no usable signal over their Poisson noise and are dropped. The
ones that stay overlap with, but are not, the HVGs.

```{r}
#| label: genes
hvg_ids <- get_gene_names_from_idx(sc_object, get_hvg(sc_object))

data.table(
  genes_used = length(tree$genes_used),
  genes_dropped = length(tree$genes_dropped),
  also_hvg = sum(tree$genes_used %in% hvg_ids)
)
```

Want the tree on your own gene set instead? Pass `hvg` (1-indexed), e.g.
`bonsai_sc(sc_object, hvg = get_hvg(sc_object) + 1L)`. The signal-to-noise
filter still applies on top.

## Plotting

`plot()` draws every branch and a point per cell. `colour_by` takes one value
per cell, in the order of the object's cells.

```{r}
#| label: plot equal angle
#| fig-width: 7
#| fig-height: 7
cell_type <- get_sc_obs(sc_object, filtered = TRUE)$cell_type

plot(tree, colour_by = cell_type, point_size = 0.8) +
  guides(colour = guide_legend(override.aes = list(size = 3)))
```

Every cell type gets its own subtree, and none of that was told to Bonsai: the
annotation only colours the leaves. The CD16+ monocytes sit inside the monocyte
subtree rather than next to it, and the NK cells branch off close to the T
cells, which is what you would expect from their biology. Branch length is
change, so the long isolated branches are cells that look like nothing else in
the data.

The default is the equal-angle layout: every subtree gets a wedge proportional
to its size. Two alternatives, recomputed from the stored tree without
searching again. The dendrogram draws the branch lengths along x:

```{r}
#| label: plot dendrogram
#| fig-width: 7
#| fig-height: 7
plot(tree, colour_by = cell_type, layout = "dendrogram", point_size = 0.5)
```

The hyperbolic projection maps the same layout onto the unit disk. How much it
changes the picture depends on how far the tips sit from the root; on PBMC3k
the branches are short and it barely does, on deeper trees it gives the crowded
centre more room:

```{r}
#| label: plot hyperbolic
#| fig-width: 7
#| fig-height: 7
plot(tree, colour_by = cell_type, hyperbolic = TRUE, point_size = 0.8)
```

`relayout_bonsai()` does the same but keeps the result, which matters for the
next step.

## Bonsai next to UMAP

The leaf coordinates can go into the object as an embedding, so everything that
plots embeddings works on them.

```{r}
#| label: embedding
#| fig-width: 12
#| fig-height: 5
sc_object <- set_bonsai_embedding(sc_object, tree)

patchwork::wrap_plots(
  embedding_plot_sc(
    sc_object,
    embedding = "umap",
    colour_by = "cell_type",
    discrete = TRUE
  ) +
    ggtitle("UMAP"),
  embedding_plot_sc(
    sc_object,
    embedding = "bonsai",
    colour_by = "cell_type",
    discrete = TRUE
  ) +
    ggtitle("Bonsai, equal angle")
)
```

## Bonsai on metacells

At 100k cells a tree over every cell is slow and hard to look at. Metacells
compress first. Their raw counts are sums over their cells, and a sum of
Poisson counts is Poisson, so Sanity treats a metacell exactly as it treats a
cell, with its total counts as the library size. A metacell of 20 cells has
about 20 times the counts, so its error bars are much tighter, and Bonsai
weights by them. The same `bonsai_sc()` call works on `MetaCells`; every
metacell becomes a leaf.

Here SEACells compresses the PBMCs to 100 metacells, and each metacell gets
the cell type most of its cells carry.

```{r}
#| label: supercells
supercells <- generate_supercells_sc(
  object = sc_object,
  sc_supercell_params = params_sc_supercell(),
  .verbose = FALSE
)

supercells <- calc_meta_cell_purity(
  supercells,
  original_cell_type = get_sc_obs(sc_object, filtered = TRUE)$cell_type,
  add_additional_info = "top_label"
)

tree_mc <- bonsai_sc(supercells, .verbose = TRUE)

tree_mc
```

The same PBMCs as 100 metacells: a few seconds instead of the minutes the
cell-level tree took, most of it Sanity. And the summed counts are deep enough
that thousands of genes clear the signal-to-noise bar, against a few hundred
for single cells, so the tree sees far more of the transcriptome.

`size_by_cells = TRUE` scales every leaf by the cells behind it.

```{r}
#| label: plot metacells
#| fig-width: 7
#| fig-height: 7
plot(
  tree_mc,
  colour_by = supercells[[]]$mc_top_label,
  size_by_cells = TRUE,
  point_size = 1
)
```

The subtrees of the cell-level tree survive the compression, and the
monocytes come out sharper: CD14+ and CD16+ split into two branches of one
monocyte subtree. A few T-cell metacells sit on the NK branch; T and NK cells
share much of their cytotoxic programme, so that is where the tree is least
sure of the labels. What the metacell tree cannot show
is anything inside a metacell, so pick the compression to match the resolution
you care about.

## Scaling

The search grows a little faster than linearly with the cell number. In
`bonsai-rs`'s own benchmarks on ten cores it took about 70 seconds at 10,000
cells and 210 seconds at 25,000. Sanity comes on top and grows with the number
of candidate genes, and on the CPU it is the bigger share once every gene goes
in. The GPU fixes that: `bonsai_gpu_sc()` in `bixverse.gpu` takes the same
arguments, returns the same `BonsaiTree`, and runs Sanity on the device.

On 10,000 simulated cells with all 16,974 genes as candidates (M1 Max, ten
cores):

| | Sanity | Bonsai search | total |
|---|---|---|---|
| `bonsai_sc()` | 570 s | 80 s | 650 s |
| `bonsai_gpu_sc()` | 8 s | 74 s | 82 s |

Both kept the same 2,751 genes. With Sanity on the GPU the tree search is the
cost, and that is where 100k cells will hurt. Metacells are the way around it:
compress first and the tree has a few thousand leaves instead of 100k. Without
a GPU, restricting the candidates with `hvg` also helps.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```
