---
title: "Bulk co-expression modules"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Bulk co-expression modules}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

Bulk co-expression module detection tries to find groups of genes that vary
together across samples. `bixverse` ships five approaches for this on a shared
`BulkCoExp` class:

- **CoReMo** ([Srivastava, et al.](https://www.nature.com/articles/s41467-018-06008-4):
  hierarchical clustering on an RBF-transformed correlation matrix. Fast,
  deterministic, gives you signed modules and per-sample eigengenes.
- **Graph-based Leiden** ([Barrio-Hernandez, et al.](https://www.nature.com/articles/s41588-023-01327-9))
  — build a sparse affinity graph from correlations, then cluster with Leiden.
  Great when you want tunable resolution and hierarchical sub-clustering.
- **ICA** — independent component analysis with stability across random
  initialisations. Each component picks up an orthogonal signal; genes with
  the largest absolute loadings define the module.
- **DGRDL** ([Pan, et al.](https://doi.org/10.1016/j.cels.2022.09.005))
  — dual graph-regularised dictionary learning. Sparse decomposition
  regularised by feature- and sample-side Laplacians.
- **NMF (HALS)** — non-negative matrix factorisation with hierarchical
  alternating least squares. Non-negative loadings give a natural
  "parts-based" module interpretation and per-sample module activity.

None of these is uniformly best. This vignette walks through all five on the
same synthetic dataset with planted ground-truth modules, so you can see the
inputs, the outputs, and how well each recovers the true structure.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(ggplot2)
```

## Data

`synthetic_bulk_cor_matrix()` generates a bulk RNA-seq count matrix with
heteroskedasticity plus a user-specified number of planted correlation modules.
The `$module_data` table gives us ground truth for a recall/precision check
at the end.

```{r}
#| label: synth-data
synth <- synthetic_bulk_cor_matrix(
  num_samples = 100L,
  num_genes = 1000L,
  add_modules = TRUE,
  module_sizes = c(100L, 100L, 100L),
  seed = 123L
)

truth <- synth$module_data
truth[membership > 0, .N, by = membership]
```

Three planted modules of 100 genes each. The remaining 700 genes have no
structure. For the correlation-based methods we work in log-CPM space
(negative values are fine); NMF needs non-negative input so we'll build a
separate object from the raw counts for it.

```{r}
#| label: normalise
norm_counts <- edgeR::cpm(synth$counts, log = TRUE)
raw_cpm <- edgeR::cpm(synth$counts, log = FALSE)

data_log <- t(norm_counts)
data_raw <- t(raw_cpm)

meta_data <- data.table(sample_id = rownames(data_log))

coexp <- BulkCoExp(raw_data = data_log, meta_data = meta_data)
coexp <- preprocess_bulk_coexp(coexp, hvg = 0.5, .verbose = FALSE)
coexp
```

Preprocessing keeps the top 50% of genes by MAD — 500 features carried through
to the module-detection methods below. The `BulkCoExp` object mutates in place
as each method runs, so we'll make copies for each branch.

## Method 1: CoReMo

CoReMo shrinks weak correlations via an RBF, hierarchically clusters the
result, and picks the cut that maximises the within-module weighted median
R². The epsilon parameter controls how aggressively the RBF suppresses weak
correlations — you usually tune it against a scale-free topology fit.

```{r}
#| label: coremo-workflow
coexp_coremo <- cor_module_processing(
  coexp,
  cor_method = "spearman",
  .verbose = TRUE
)

coexp_coremo@processed_data$correlation_res$get_data_table() %$% hist(sim)

coexp_coremo <- cor_module_check_epsilon(
  coexp_coremo,
  rbf_func = "gaussian",
  .verbose = TRUE
)

plot_epsilon_res(coexp_coremo)

epsilon_pick <- get_epsilon_res(coexp_coremo)[r2_vals == max(r2_vals), epsilon][
  1
]

epsilon_pick <- 2

coremo_params <- params_coremo(epsilon = epsilon_pick, rbf_func = "gaussian")

coexp_coremo <- cor_module_coremo_clustering(
  coexp_coremo,
  coremo_params = coremo_params,
  .verbose = TRUE
)
coexp_coremo <- cor_module_coremo_stability(coexp_coremo, .verbose = FALSE)
coexp_coremo <- cor_module_coremo_cor_sign(coexp_coremo, .verbose = FALSE)
coexp_coremo <- cor_module_coremo_eigengene(coexp_coremo, .verbose = FALSE)

coremo_modules <- get_outputs(coexp_coremo)$final_modules
coremo_modules[, .N, by = cluster_id]
```

Each module has a sign suffix (`_pos` / `_neg`) reflecting whether the
member genes are positively or negatively correlated with the module
eigengene, and a per-gene stability score from the leave-one-out resampling.

## Method 2: Graph-based Leiden

The Leiden branch keeps the same RBF-based affinity but treats the sparse
non-zero pattern as an igraph. Modularity is optimised over a Leiden
resolution grid; large communities can be sub-clustered recursively.

```{r}
#| label: leiden-workflow
coexp_leiden <- cor_module_processing(
  coexp,
  cor_method = "spearman",
  .verbose = TRUE
)

coexp_leiden <- cor_module_check_epsilon(
  coexp_leiden,
  rbf_func = "bump",
  .verbose = TRUE
)

plot_epsilon_res(coexp_leiden)

graph_params <- params_cor_graph(
  epsilon = 2
)

coexp_leiden <- cor_module_graph_check_res(
  coexp_leiden,
  resolution_params = params_graph_resolution(),
  graph_params = graph_params,
  parallel = FALSE,
  .verbose = TRUE
)

plot_resolution_res(coexp_leiden)

coexp_leiden <- cor_module_graph_final_modules(
  coexp_leiden,
  min_size = 10L,
  max_size = 500L,
  subclustering = TRUE,
  .graph_params = graph_params,
  .verbose = FALSE
)

leiden_modules <- get_modules(get_results(coexp_leiden))
leiden_modules[, .N, by = module_id]
```

## Method 3: ICA

ICA whitens the data, then finds independent components across many random
initialisations. The stability across restarts, the fraction of converged
runs, and the mutual information between components combine into a
`combined_score` — the loess inflection point of this score against the
number of components gives you an "optimal" `ncomp`.

```{r}
#| label: ica-workflow
coexp_ica <- ica_processing(coexp, fast_svd = TRUE, .verbose = FALSE)

ncomp_params <- params_ica_ncomp(max_no_comp = 20L, steps = 5L)
iter_params <- params_ica_randomisation(
  cross_validate = FALSE,
  random_init = 20L
)

coexp_ica <- ica_evaluate_comp(
  coexp_ica,
  ica_type = "logcosh",
  iter_params = iter_params,
  ncomp_params = ncomp_params,
  .verbose = FALSE
)

coexp_ica <- ica_optimal_ncomp(
  coexp_ica,
  span = 0.4,
  show_plot = FALSE,
  .verbose = FALSE
)
coexp_ica <- ica_stabilised_results(
  coexp_ica,
  iter_params = iter_params,
  .verbose = FALSE
)

ica_res <- get_results(coexp_ica)
ica_gene_loadings <- get_factors(ica_res, which = "gene_loadings")
dim(ica_gene_loadings)
```

`gene_loadings` is the `genes x k` matrix (transposed from the raw ICA `S`
convention of `k x genes`). Module membership comes straight from
`get_modules()`:

```{r}
#| label: ica-modules
ica_modules <- get_modules(ica_res)
ica_modules[, .N, by = module_id][order(-N)]
```

## Method 4: DGRDL

DGRDL learns a sparse dictionary of features constrained by a k-nearest-
neighbour Laplacian on both the feature and the sample side. The grid
search helps pick `dict_size` and `k_neighbours` before the final fit.

```{r}
#| label: dgrdl-workflow
coexp_dgrdl <- dgrdl_grid_search(
  coexp,
  neighbours_vec = c(5L, 10L),
  dict_size_vec = c(4L, 6L),
  seed_vec = c(1L, 2L),
  dgrdl_params = params_dgrdl(),
  .verbose = TRUE
)

grid_res <- get_grid_search_res(coexp_dgrdl)
setorder(grid_res, reconstruction_errs)
head(grid_res)
```

Pick the top row and fit the final DGRDL model:

```{r}
#| label: dgrdl-final
top <- grid_res[1L]
dgrdl_params_final <- params_dgrdl(
  dict_size = as.integer(top$dict_size),
  k_neighbours = as.integer(top$k_neighbours)
)

coexp_dgrdl <- dgrdl_result(
  coexp_dgrdl,
  dgrdl_params = dgrdl_params_final,
  seed = as.integer(top$seed),
  .verbose = FALSE
)

dgrdl_res <- get_results(coexp_dgrdl)
dim(get_factors(dgrdl_res, which = "dictionary"))
dim(get_factors(dgrdl_res, which = "loadings"))
```

`get_modules()` handles the "gene to strongest atom" assignment:

```{r}
#| label: dgrdl-modules
dgrdl_modules <- get_modules(dgrdl_res)
dgrdl_modules[, .N, by = module_id][order(-N)]
```

## Method 5: NMF (HALS)

NMF needs non-negative input, so we build a fresh `BulkCoExp` from CPM
(without the log transform). The single-run version is deterministic given
NNDSVD initialisation; the stabilised version runs multiple random restarts
and returns the run with the lowest reconstruction loss.

```{r}
#| label: nmf-workflow
coexp_nmf <- BulkCoExp(raw_data = data_raw, meta_data = meta_data)
coexp_nmf <- preprocess_bulk_coexp(coexp_nmf, hvg = 0.5, .verbose = FALSE)

coexp_nmf <- nmf_bulk(
  coexp_nmf,
  k = 4L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 300L, tol = 1e-4),
  seed = 42L,
  .verbose = FALSE
)

nmf_modules <- get_nmf_modules(coexp_nmf)
nmf_modules[, .N, by = module_id]
```

`get_nmf_gene_loadings()` gives the features x k matrix, and
`get_nmf_sample_activity()` gives the samples x k matrix. A quick heatmap
of gene loadings shows the parts-based decomposition:

```{r}
#| label: nmf-loadings-heatmap
#| fig-cap: "NMF gene loadings, top 40 genes per component."
loadings <- get_nmf_gene_loadings(coexp_nmf)

top_genes <- unique(unlist(lapply(seq_len(ncol(loadings)), \(j) {
  names(sort(loadings[, j], decreasing = TRUE))[1:40]
})))

heat_dt <- as.data.table(loadings[top_genes, , keep.rownames = "gene"])
heat_long <- melt(
  heat_dt,
  id.vars = "gene",
  variable.name = "component",
  value.name = "loading"
)

ggplot(heat_long, aes(component, gene, fill = loading)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

For a more robust fit, `stabilised_nmf_bulk()` runs several restarts and
returns per-run losses via `get_nmf_stability()`.

```{r}
#| label: nmf-stabilised
coexp_nmf_stab <- stabilised_nmf_bulk(
  coexp_nmf,
  k = 4L,
  n_runs = 5L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 200L, tol = 1e-3),
  seed = 7L,
  .verbose = FALSE
)

stab <- get_nmf_stability(coexp_nmf_stab)
stab
```

## Recovery of planted modules

With ground truth in hand we can quantify how well each method recovered the
three planted modules. For each method we compute the best-matching planted
module per detected cluster (max Jaccard), then take the mean Jaccard across
the three planted modules — a simple "recall" score.

```{r}
#| label: recovery
truth_split <- split(truth$gene, truth$membership)
truth_split <- truth_split[names(truth_split) != "0"]

best_jaccard_per_planted <- function(assignments) {
  # assignments: data.table with columns `gene`, `module_id`
  detected <- split(assignments$gene, assignments$module_id)
  purrr::map_dbl(truth_split, \(planted) {
    max(purrr::map_dbl(detected, \(det) {
      length(intersect(planted, det)) / length(union(planted, det))
    }))
  })
}

recovery <- data.table(
  method = c("CoReMo", "Leiden", "ICA", "DGRDL", "NMF"),
  score = c(
    mean(best_jaccard_per_planted(
      coremo_modules[, .(gene, module_id = cluster_id)]
    )),
    mean(best_jaccard_per_planted(leiden_modules)),
    mean(best_jaccard_per_planted(ica_modules)),
    mean(best_jaccard_per_planted(dgrdl_modules)),
    mean(best_jaccard_per_planted(nmf_modules))
  )
)

setorder(recovery, -score)
recovery
```

```{r}
#| label: recovery-plot
#| fig-cap: "Mean best-Jaccard across the three planted modules."
ggplot(recovery, aes(x = reorder(method, score), y = score)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  ylim(0, 1) +
  theme_minimal() +
  xlab("") +
  ylab("Mean best Jaccard vs. planted modules")
```

## Wrap-up

Different methods suit different questions:

- **CoReMo** if you want signed modules with per-sample eigengenes and a
  fast, deterministic pipeline. Solid default for standard bulk data.
- **Graph-based Leiden** if you want tunable resolution or hierarchical
  sub-clustering. Handles very heterogeneous module sizes better than CoReMo.
- **ICA** if you want components that decompose independent, potentially
  overlapping biological signals. Interpretation is per-component, not
  per-gene.
- **DGRDL** if you want a sparse dictionary regularised by feature and sample
  neighbour graphs. Sparsity is a tunable constraint, not a byproduct.
- **NMF** if you want non-negative, parts-based decompositions with a natural
  interpretation as "how strongly is each sample expressing each module".

For a real dataset, run more than one and cross-check. The synthetic setup
here has a clean signal to noise regime — real bulk data (GTEx, TCGA,
recount3) will happily show larger gaps between methods.
