---
title: "Bulk co-expression modules"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Bulk co-expression modules}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Intro

Bulk co-expression module detection tries to find groups of genes that vary
together across samples. `bixverse` ships five approaches for this on a shared
`BulkCoExp` class:

- **CoReMo** ([Srivastava, et al.](https://www.nature.com/articles/s41467-018-06008-4):
  hierarchical clustering on an RBF-transformed correlation matrix. Fast,
  deterministic, gives you signed modules and per-sample eigengenes.
- **Graph-based Leiden** ([Barrio-Hernandez, et al.](https://www.nature.com/articles/s41588-023-01327-9))
  — build a sparse affinity graph from correlations, then cluster with Leiden.
  Great when you want tunable resolution and hierarchical sub-clustering.
- **ICA** — independent component analysis with stability across random
  initialisations. Each component picks up an independent signal; the genes in
  the tails of a component's loading distribution define its module.
- **DGRDL** ([Pan, et al.](https://doi.org/10.1016/j.cels.2022.09.005))
  — dual graph-regularised dictionary learning. Sparse decomposition
  regularised by feature- and sample-side Laplacians.
- **NMF (HALS)** — non-negative matrix factorisation with hierarchical
  alternating least squares. Non-negative loadings give a natural
  "parts-based" module interpretation and per-sample module activity. Comes in
  three flavours: one run, best-of-many restarts, or consensus across restarts
  (cNMF), which also hands you a way to choose `k`.

One structural difference matters throughout. CoReMo and Leiden partition genes:
one gene, one module. The three factorisation methods do not, and should not. A
gene can load on several components, so `get_modules()` returns one row per
(gene, component) pair for ICA, NMF and DGRDL, and genes in no component's tail
appear nowhere. `params_module_membership()` controls that thresholding. Do not
assume `gene` is a unique key.

None of these is uniformly best. This vignette walks through all five on the
same synthetic dataset with planted ground-truth modules, so you can see the
inputs, the outputs, and how well each recovers the true structure.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(magrittr)
library(ggplot2)
```

## Data

`synthetic_bulk_cor_matrix()` generates a bulk RNA-seq count matrix with
heteroskedasticity plus a user-specified number of planted co-expression
modules. Counts come from a negative binomial with a mean-dispersion trend, and
a module is built by putting its genes on a shared latent factor.

`params_synthetic_bulk_rnaseq()` controls the whole thing. The parameter that
matters most here is `generator`, which picks how loadings and factors are
drawn:

- **`hub_modular`** — LogNormal loadings on a Normal factor. Some genes end up
  far more connected than others, which is the WGCNA-style default and our
  starting point.
- **`modular`** — Beta(5, 2) loadings on a Normal factor. Homogeneous
  within-module correlation, no hubs.
- **`non_negative_factor`** — LogNormal loadings on a Gamma factor. The activity
  matrix is non-negative by construction, so NMF has a ground truth it can
  reach.
- **`non_gaussian_factor`** — LogNormal loadings on a Laplace factor.
  Non-Gaussian sources satisfy ICA identifiability.

We'll start with `hub_modular` for the head-to-head, then come back to the
generator choice at the end. It turns out to matter more than you'd guess.

```{r}
#| label: synth-data
synth_params <- params_synthetic_bulk_rnaseq(
  num_samples = 100L,
  num_genes = 1000L,
  module_sizes = c(100L, 100L, 100L),
  generator = "hub_modular",
  seed = 123L
)

synth <- synthetic_bulk_cor_matrix(synth_params)

truth <- synth$module_data
truth[membership > 0, .N, by = membership]
```

Three planted modules of 100 genes each. The remaining 700 genes are background: 
no loading, never a hub.

The ground truth comes back with the counts, so we can score a recovered module
against what was actually simulated rather than eyeballing a heatmap. Per gene
we get its module, its loading on that module's factor, and whether it ranked as
a hub:

```{r}
#| label: ground-truth
truth[
  membership > 0,
  .(
    n_genes = .N,
    n_hubs = sum(is_hub),
    mean_loading = round(mean(loading), 3)
  ),
  by = membership
]
```

Hubs are ranked on loading globally across all module genes rather than per
module, so the counts per module don't have to be equal. On top of that,
`$module_factors` holds the latent factors themselves, one row per module:

```{r}
#| label: true-factors
dim(synth$module_factors)
```

That matrix is what a module eigengene, an ICA component or an NMF activity
column is trying to recover, and we use it for a second scoring pass later on.

For the correlation-based methods we work in log-CPM space (negative values are
fine); NMF needs non-negative input so we'll build a separate object from the
raw counts for it.

```{r}
#| label: normalise
norm_counts <- edgeR::cpm(synth$counts, log = TRUE)
raw_cpm <- edgeR::cpm(synth$counts, log = FALSE)

data_log <- t(norm_counts)
data_raw <- t(raw_cpm)

meta_data <- data.table(sample_id = rownames(data_log))

coexp <- BulkCoExp(raw_data = data_log, meta_data = meta_data)
coexp <- preprocess_bulk_coexp(coexp, hvg = 0.5, .verbose = FALSE)
coexp
```

Preprocessing keeps the top 50% of genes by MAD; 500 features carried through
to the module-detection methods below. The `BulkCoExp` object mutates in place
as each method runs, so we'll make copies for each branch.

## Method 1: CoReMo

CoReMo shrinks weak correlations via an RBF, hierarchically clusters the
result, and picks the cut that maximises the within-module weighted median
R². The epsilon parameter controls how aggressively the RBF suppresses weak
correlations — you usually tune it against a scale-free topology fit.

```{r}
#| label: coremo-workflow
coexp_coremo <- cor_module_processing(
  coexp,
  cor_method = "spearman",
  .verbose = TRUE
)

# show correlation structure
coexp_coremo@processed_data$correlation_res$get_data_table() %$% hist(sim)

coexp_coremo <- cor_module_check_epsilon(
  coexp_coremo,
  rbf_func = "gaussian",
  .verbose = TRUE
)

plot_epsilon_res(coexp_coremo)

epsilon_sf <- get_epsilon_res(coexp_coremo)[r2_vals == max(r2_vals), epsilon][1]
epsilon_sf
```

Here is a wrinkle worth knowing about, and one the ground truth lets us actually
diagnose. The scale-free R² criterion picks a high epsilon, but epsilon controls
how aggressively the RBF suppresses weak correlations, so a high value throws
genes out of modules entirely. Counting how many of the 500 features survive
into a module tells the story:

```{r}
#| label: coremo-epsilon-tradeoff
epsilon_scan <- rbindlist(lapply(c(1, 2, 3, 4, 6, epsilon_sf), \(eps) {
  fitted <- coexp_coremo %>%
    cor_module_coremo_clustering(
      coremo_params = params_coremo(epsilon = eps, rbf_func = "gaussian"),
      .verbose = FALSE
    ) %>%
    cor_module_coremo_stability(.verbose = FALSE) %>%
    cor_module_coremo_cor_sign(.verbose = FALSE)

  mods <- get_outputs(fitted)$final_modules
  data.table(
    epsilon = eps,
    n_modules = uniqueN(mods$cluster_id),
    genes_assigned = nrow(mods)
  )
}))

epsilon_scan
```

The scale-free fit and "keep as many real module genes as possible" pull in
opposite directions on this data. We go with a low epsilon deliberately, because
the planted modules are what we are trying to find:

```{r}
#| label: coremo-workflow-2
coremo_params <- params_coremo(epsilon = 2, rbf_func = "gaussian")

coexp_coremo <- cor_module_coremo_clustering(
  coexp_coremo,
  coremo_params = coremo_params,
  .verbose = TRUE
)
coexp_coremo <- cor_module_coremo_stability(coexp_coremo, .verbose = FALSE)
coexp_coremo <- cor_module_coremo_cor_sign(coexp_coremo, .verbose = FALSE)
coexp_coremo <- cor_module_coremo_eigengene(coexp_coremo, .verbose = FALSE)

coremo_modules <- get_outputs(coexp_coremo)$final_modules
coremo_modules[, .N, by = cluster_id]
```

Each module has a sign suffix (`_pos` / `_neg`) reflecting whether the member 
genes are positively or negatively correlated with the module eigengene, and a 
per-gene stability score from the leave-one-out resampling. Since the generator 
draws strictly positive loadings, essentially everything lands in a `_pos` 
module; the odd `_neg` singleton is an anti-correlated background gene getting 
split off, which is correct behaviour.

## Method 2: Graph-based Leiden

The Leiden branch keeps the same RBF-based affinity but treats the sparse 
non-zero pattern as an igraph. Modularity is optimised over a Leiden
resolution grid; large communities can be sub-clustered recursively.

```{r}
#| label: leiden-workflow
coexp_leiden <- cor_module_processing(
  coexp,
  cor_method = "spearman",
  .verbose = TRUE
)

coexp_leiden <- cor_module_check_epsilon(
  coexp_leiden,
  rbf_func = "bump",
  .verbose = TRUE
)

plot_epsilon_res(coexp_leiden)

graph_params <- params_cor_graph(
  epsilon = 2
)

coexp_leiden <- cor_module_graph_check_res(
  coexp_leiden,
  resolution_params = params_graph_resolution(),
  graph_params = graph_params,
  parallel = FALSE,
  .verbose = TRUE
)

plot_resolution_res(coexp_leiden)

coexp_leiden <- cor_module_graph_final_modules(
  coexp_leiden,
  min_size = 10L,
  max_size = 500L,
  subclustering = TRUE,
  .graph_params = graph_params,
  .verbose = FALSE
)

leiden_modules <- get_modules(get_results(coexp_leiden))
leiden_modules[, .N, by = module_id]
```

## Method 3: ICA

ICA whitens the data, then finds independent components across many random
initialisations. The stability across restarts, the fraction of converged
runs, and the mutual information between components combine into a
`combined_score`: the loess inflection point of this score against the
number of components gives you an "optimal" `ncomp`. (Please explore your data
and see how it looks...)

```{r}
#| label: ica-workflow
coexp_ica <- ica_processing(coexp, fast_svd = TRUE, .verbose = FALSE)

# A dense enough grid that the loess in ica_optimal_ncomp() has something to
# fit. On a handful of points it gives up and warns instead.
ncomp_params <- params_ica_ncomp(max_no_comp = 30L, steps = 2L)
iter_params <- params_ica_randomisation(
  cross_validate = FALSE,
  random_init = 20L
)

coexp_ica <- ica_evaluate_comp(
  coexp_ica,
  ica_type = "logcosh",
  iter_params = iter_params,
  ncomp_params = ncomp_params,
  .verbose = FALSE
)

coexp_ica <- ica_optimal_ncomp(
  coexp_ica,
  span = 0.4,
  show_plot = FALSE,
  .verbose = FALSE
)
coexp_ica <- ica_stabilised_results(
  coexp_ica,
  iter_params = iter_params,
  .verbose = FALSE
)

ica_res <- get_results(coexp_ica)
ica_gene_loadings <- get_factors(ica_res, which = "gene_loadings")
dim(ica_gene_loadings)
```

`gene_loadings` is the `genes x k` matrix (transposed from the raw ICA `S`
convention of `k x genes`). Module membership keeps the tails of each component,
so a gene may appear more than once and many genes not at all:

```{r}
#| label: ica-modules
ica_modules <- get_modules(ica_res)
ica_modules[, .N, by = module_id][order(-N)]

# Sparse and overlapping, which is the whole point of a factorisation.
data.table(
  rows = nrow(ica_modules),
  unique_genes = uniqueN(ica_modules$gene),
  in_multiple_modules = ica_modules[, .N, by = gene][N > 1, .N]
)
```

## Method 4: DGRDL

DGRDL learns a sparse dictionary of features constrained by a k-nearest-
neighbour Laplacian on both the feature and the sample side. The grid search 
helps pick `dict_size` and `k_neighbours` before the final fit.

Unlike the other four, DGRDL needs **centred** input. Handed unscaled log-CPM it
spends its dictionary explaining the per-gene mean offset instead of the
covariance: atoms come out as near-duplicates of each other, and the loadings
become a large constant plus noise. So we build a scaled object for it.

```{r}
#| label: dgrdl-workflow
coexp_dgrdl <- BulkCoExp(raw_data = data_log, meta_data = meta_data) %>%
  preprocess_bulk_coexp(
    hvg = 0.5,
    scaling = TRUE,
    scaling_type = "normal",
    .verbose = FALSE
  )

coexp_dgrdl <- dgrdl_grid_search(
  coexp_dgrdl,
  neighbours_vec = c(5L, 10L),
  dict_size_vec = c(4L, 6L),
  seed_vec = c(1L, 2L),
  dgrdl_params = params_dgrdl(),
  .verbose = TRUE
)

grid_res <- get_grid_search_res(coexp_dgrdl)
setorder(grid_res, reconstruction_errs)
head(grid_res)
```

Pick the top row and fit the final DGRDL model:

```{r}
#| label: dgrdl-final
top <- grid_res[1L]

dgrdl_params_final <- params_dgrdl(
  dict_size = as.integer(top$dict_size),
  k_neighbours = as.integer(top$k_neighbours)
)

coexp_dgrdl <- dgrdl_result(
  coexp_dgrdl,
  dgrdl_params = dgrdl_params_final,
  seed = as.integer(top$seed),
  .verbose = FALSE
)

dgrdl_res <- get_results(coexp_dgrdl)
dim(get_factors(dgrdl_res, which = "dictionary"))
dim(get_factors(dgrdl_res, which = "loadings"))
```

`get_modules()` keeps the tails of each dictionary atom, so a gene active in
several atoms shows up in several modules:

```{r}
#| label: dgrdl-modules
dgrdl_modules <- get_modules(dgrdl_res)
dgrdl_modules[, .N, by = module_id][order(-N)]
```

## Method 5: NMF (HALS)

NMF needs non-negative input, so we build a fresh `BulkCoExp` from CPM (without 
the log transform). The single-run version is deterministic given NNDSVD 
initialisation; the stabilised version runs multiple random restarts and returns 
the run with the lowest reconstruction loss.

```{r}
#| label: nmf-workflow
coexp_nmf <- BulkCoExp(raw_data = data_raw, meta_data = meta_data)
coexp_nmf <- preprocess_bulk_coexp(coexp_nmf, hvg = 0.5, .verbose = FALSE)

coexp_nmf <- nmf_bulk(
  coexp_nmf,
  k = 4L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 300L, tol = 1e-4),
  seed = 42L,
  .verbose = FALSE
)

nmf_modules <- get_nmf_modules(coexp_nmf)
nmf_modules[, .N, by = module_id]
```

`get_nmf_gene_loadings()` gives the features x k matrix, and
`get_nmf_sample_activity()` gives the samples x k matrix. A quick heatmap
of gene loadings shows the parts-based decomposition:

```{r}
#| label: nmf-loadings-heatmap
#| fig-cap: "NMF gene loadings, top 40 genes per component."
loadings <- get_nmf_gene_loadings(coexp_nmf)

top_genes <- unique(unlist(lapply(seq_len(ncol(loadings)), \(j) {
  names(sort(loadings[, j], decreasing = TRUE))[1:40]
})))

heat_dt <- as.data.table(
  loadings[top_genes, , drop = FALSE],
  keep.rownames = "gene"
)
heat_long <- melt(
  heat_dt,
  id.vars = "gene",
  variable.name = "component",
  value.name = "loading"
)

ggplot(heat_long, aes(component, gene, fill = loading)) +
  geom_tile() +
  scale_fill_viridis_c() +
  theme_minimal() +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())
```

For a more robust fit, `stabilised_nmf_bulk()` runs several restarts and
returns per-run losses via `get_nmf_stability()`.

```{r}
#| label: nmf-stabilised
coexp_nmf_stab <- stabilised_nmf_bulk(
  coexp_nmf,
  k = 4L,
  n_runs = 5L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 200L, tol = 1e-3),
  seed = 7L,
  .verbose = FALSE
)

stab <- get_nmf_stability(coexp_nmf_stab)
stab
```

### Consensus NMF

`stabilised_nmf_bulk()` picks the restart with the lowest loss, which tells you
nothing about whether that restart is reproducible. Consensus NMF asks the
better question: pool the components of every restart, throw away the ones that
sit on their own, cluster what is left into `k` groups, and take the median of
each group as the answer. What survives is the structure the restarts agree on.
That agreement gets a number, the mean silhouette of those clusters, and it is
the honest way to judge whether `k` is right.

Which brings us to picking `k`. `nmf_k_sweep_bulk()` runs the whole thing across
a range and reports stability against reconstruction error, no factors kept, so
it stays cheap.

```{r}
#| label: nmf-k-sweep
k_sweep <- nmf_k_sweep_bulk(
  coexp_nmf,
  k_range = 2:8,
  n_runs = 10L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 200L, tol = 1e-3),
  nmf_consensus_params = params_nmf_consensus(density_threshold = 2),
  seed = 42L,
  .verbose = FALSE
)

k_sweep
```

```{r}
#| label: nmf-k-sweep-plot
#| fig-cap: "Consensus NMF k sweep: stability against reconstruction error."
plot(k_sweep)
```

Error falls monotonically with `k`, so on its own it would push you to the
largest `k` you can afford. Stability is the counterweight. You want the last
`k` before it drops off, where the error curve is still coming down.

::: {.callout-note}
`density_threshold = 2` switches the outlier filter off, because cosine distance
cannot exceed 2. With the filter live, components whose neighbours sit far away
get dropped, and if that leaves fewer than `k` of them the fit errors rather
than handing back something degenerate. That is the right behaviour on real
data, but with only 10 restarts on a small synthetic matrix it fires more often
than it should. Plot `local_density` from `get_nmf_stability()` to pick a real
threshold for your data.
:::

Now fit at the chosen `k`:

```{r}
#| label: nmf-consensus
coexp_nmf_consensus <- consensus_nmf_bulk(
  coexp_nmf,
  k = 4L,
  n_runs = 10L,
  preprocessing = "none",
  nmf_hals_params = params_nmf_hals(max_iter = 300L, tol = 1e-4),
  nmf_consensus_params = params_nmf_consensus(density_threshold = 2),
  seed = 42L,
  .verbose = FALSE
)

consensus_stab <- get_nmf_stability(coexp_nmf_consensus)
consensus_stab$stability
consensus_stab$cluster_sizes
```

The per-component table shows where every restart's components ended up.
`component_id` uses the same `run_XX.comp_YY` naming as the `w_all` columns of
a stabilised fit, so at matching `k` and `n_runs` the two line up directly.

```{r}
#| label: nmf-consensus-clusters
head(consensus_stab$clusters)
```

```{r}
#| label: nmf-consensus-modules
nmf_consensus_modules <- get_nmf_modules(coexp_nmf_consensus)
nmf_consensus_modules[, .N, by = module_id]
```

## Recovery of planted modules

With ground truth in hand we can quantify how well each method recovered the
three planted modules. For each method we compute the best-matching planted
module per detected cluster (max Jaccard), then take the mean Jaccard across
the three planted modules.... A simple "recall" score.

```{r}
#| label: recovery
# assignments: data.table with columns `gene` and `module_id`.
# Returns one score per planted module: the best Jaccard any detected cluster
# manages against it.
best_jaccard_vs <- function(assignments, truth_dt) {
  planted <- split(truth_dt$gene, truth_dt$membership)
  planted <- planted[names(planted) != "0"]
  detected <- split(assignments$gene, assignments$module_id)
  purrr::map_dbl(planted, \(p) {
    max(purrr::map_dbl(detected, \(d) {
      length(intersect(p, d)) / length(union(p, d))
    }))
  })
}

recovery <- data.table(
  method = c("CoReMo", "Leiden", "ICA", "DGRDL", "NMF", "Consensus NMF"),
  score = c(
    mean(best_jaccard_vs(
      coremo_modules[, .(gene, module_id = cluster_id)],
      truth
    )),
    mean(best_jaccard_vs(leiden_modules, truth)),
    mean(best_jaccard_vs(ica_modules, truth)),
    mean(best_jaccard_vs(dgrdl_modules, truth)),
    mean(best_jaccard_vs(nmf_modules, truth)),
    mean(best_jaccard_vs(nmf_consensus_modules, truth))
  )
)

setorder(recovery, -score)
recovery
```

```{r}
#| label: recovery-plot
#| fig-cap: "Mean best-Jaccard across the three planted modules."
ggplot(recovery, aes(x = reorder(method, score), y = score)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  ylim(0, 1) +
  theme_minimal() +
  xlab("") +
  ylab("Mean best Jaccard vs. planted modules")
```

Worth being blunt about one result there: consensus NMF scores *below* plain
NMF. That is not a bug, and it is not an argument against consensus NMF. Two
things are going on. This data was generated with `hub_modular`, a Normal
latent factor, so there is no non-negative ground truth for NMF to reach in the
first place, and both NMF variants are fighting the data. On top of that,
consensus optimises for reproducibility, not recall: the median across restarts
is deliberately conservative, and the components it keeps are the ones every
restart found rather than the ones that happen to match a planted module best
on one lucky seed. A recall number on simulated data rewards the lucky seed.
Real data does not.

## Scoring against the latent factors

Jaccard on gene sets only asks whether a method found the right genes. The
latent factors let us ask a sharper question: did it recover the right
*signal*? For CoReMo we can correlate each per-sample eigengene against the
factor the module was actually built on.

```{r}
#| label: factor-recovery
# activity: samples x k matrix of per-sample scores (eigengenes, ICA sample
# activity, NMF sample activity). Returns one score per true factor: the best
# absolute correlation any recovered column manages against it.
#
# Index the factor matrix by rownames(activity), not by column position: these
# per-sample matrices come back in lexicographic sample order (sample_1,
# sample_10, sample_100, ...) while module_factors is in numeric order. Line them
# up positionally and you get correlations near zero for no good reason.
best_factor_cor <- function(activity, module_factors) {
  purrr::map_dbl(seq_len(nrow(module_factors)), \(k) {
    true_factor <- module_factors[k, rownames(activity)]
    max(apply(activity, 2, \(a) {
      abs(cor(a, true_factor, method = "spearman"))
    }))
  })
}

eigengenes <- get_factors(get_results(coexp_coremo))$module_eigengenes

data.table(
  module = rownames(synth$module_factors),
  best_abs_cor = round(best_factor_cor(eigengenes, synth$module_factors), 3)
)
```

Those sit far higher than the Jaccard scores, and the gap is the interesting
part: CoReMo reconstructs the module's *signal* almost exactly even where it
disagrees with the planted gene list at the edges. If what you care about is a
per-sample module score to regress against a phenotype, the eigengene is doing
its job well before the gene membership is perfect.

## Does the generator matter?

A lot, and the answer depends on what you measure. `non_negative_factor`
advertises itself as NMF-friendly and `non_gaussian_factor` as ICA-friendly, so
the obvious guess is that each method wins on its matched generator. We score 
two ways to see whether that holds:

- **Jaccard**, as above: a hard partition of genes into modules, compared 
  against the planted gene sets.
- **Factor recovery**: the best absolute correlation any recovered per-sample
  activity column achieves against a true latent factor.

The distinction matters because "ICA-friendly" is a claim about *identifiability
of the factorisation*, that is, recovering the sources up to permutation and
scaling. It says nothing about whether assigning each gene to its
strongest-loading component reproduces a planted block. Three methods rather than
five here, to keep the render time honest: Leiden and DGRDL both carry a
parameter search on top.

```{r}
#| label: generator-grid
#| cache: true
run_methods <- function(generator) {
  params <- params_synthetic_bulk_rnaseq(
    num_samples = 100L,
    num_genes = 1000L,
    module_sizes = c(100L, 100L, 100L),
    generator = generator,
    seed = 123L
  )
  synth_g <- synthetic_bulk_cor_matrix(params)
  meta_g <- data.table(sample_id = colnames(synth_g$counts))

  log_obj <- BulkCoExp(
    raw_data = t(edgeR::cpm(synth_g$counts, log = TRUE)),
    meta_data = meta_g
  ) %>%
    preprocess_bulk_coexp(hvg = 0.5, .verbose = FALSE)

  raw_obj <- BulkCoExp(
    raw_data = t(edgeR::cpm(synth_g$counts, log = FALSE)),
    meta_data = meta_g
  ) %>%
    preprocess_bulk_coexp(hvg = 0.5, .verbose = FALSE)

  coremo_g <- log_obj %>%
    cor_module_processing(cor_method = "spearman", .verbose = FALSE) %>%
    cor_module_coremo_clustering(
      coremo_params = params_coremo(epsilon = 2),
      .verbose = FALSE
    ) %>%
    cor_module_coremo_stability(.verbose = FALSE) %>%
    cor_module_coremo_cor_sign(.verbose = FALSE) %>%
    cor_module_coremo_eigengene(.verbose = FALSE)

  ica_g <- log_obj %>%
    ica_processing(fast_svd = TRUE, .verbose = FALSE) %>%
    ica_evaluate_comp(
      ica_type = "logcosh",
      iter_params = iter_params,
      ncomp_params = ncomp_params,
      .verbose = FALSE
    ) %>%
    ica_optimal_ncomp(span = 0.4, show_plot = FALSE, .verbose = FALSE) %>%
    ica_stabilised_results(iter_params = iter_params, .verbose = FALSE)

  nmf_g <- nmf_bulk(
    raw_obj,
    k = 4L,
    preprocessing = "none",
    seed = 42L,
    .verbose = FALSE
  )

  truth_g <- synth_g$module_data
  coremo_mods <- get_outputs(coremo_g)$final_modules
  coremo_res_g <- get_results(coremo_g)
  ica_res_g <- get_results(ica_g)

  jacc <- \(assign) round(mean(best_jaccard_vs(assign, truth_g)), 3)
  fcor <- \(act) round(mean(best_factor_cor(act, synth_g$module_factors)), 3)

  data.table(
    generator = generator,
    metric = c("jaccard", "factor_cor"),
    CoReMo = c(
      jacc(coremo_mods[, .(gene, module_id = cluster_id)]),
      fcor(get_factors(coremo_res_g)$module_eigengenes)
    ),
    ICA = c(
      jacc(get_modules(ica_res_g)),
      fcor(get_factors(ica_res_g, which = "sample_activity"))
    ),
    NMF = c(
      jacc(get_nmf_modules(nmf_g)),
      fcor(get_nmf_sample_activity(nmf_g))
    )
  )
}

grid_res <- rbindlist(lapply(
  c("hub_modular", "modular", "non_negative_factor", "non_gaussian_factor"),
  run_methods
))

grid_res
```

```{r}
#| label: generator-grid-plot
#| fig-cap: "Recovery by generator, method and metric. Note how the two panels disagree."
grid_long <- melt(
  grid_res,
  id.vars = c("generator", "metric"),
  variable.name = "method",
  value.name = "score"
)

ggplot(grid_long, aes(x = generator, y = method, fill = score)) +
  geom_tile() +
  geom_text(aes(label = sprintf("%.2f", score)), colour = "white") +
  scale_fill_viridis_c(limits = c(0, 1)) +
  facet_wrap(~metric) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  xlab("") +
  ylab("")
```

Two things to read out of that.

**ICA's factor recovery is best on `non_gaussian_factor`, exactly as 
advertised.** The Laplace-factor generator is the one built for ICA and it wins
on the metric identifiability is actually about. NMF's factor recovery sits 
above 0.96 on all four, so that metric does not discriminate for NMF at all: it
recovers per-sample activity almost perfectly regardless.

**ICA's Jaccard on `modular` collapses, and the reason is instructive.**
Membership for the factorisation methods comes from keeping the tails of each
component's loading distribution, which quietly assumes there *is* a tail. That
assumption depends entirely on how the loadings were drawn:

```{r}
#| label: loading-kurtosis
#| cache: true
excess_kurtosis <- function(x) {
  centred <- x - mean(x)
  mean(centred^4) / (mean(centred^2)^2) - 3
}

rbindlist(lapply(
  c("hub_modular", "modular", "non_negative_factor", "non_gaussian_factor"),
  \(gen) {
    synth_g <- synthetic_bulk_cor_matrix(params_synthetic_bulk_rnaseq(
      num_samples = 100L,
      num_genes = 1000L,
      module_sizes = c(100L, 100L, 100L),
      generator = gen,
      seed = 123L
    ))
    loadings_g <- synth_g$module_data[membership > 0, loading]
    data.table(
      generator = gen,
      loading_cv = round(sd(loadings_g) / mean(loadings_g), 2),
      loading_kurtosis = round(excess_kurtosis(loadings_g), 2)
    )
  }
))
```

`modular` draws Beta(5, 2), so module-gene loadings have a coefficient of
variation around 0.23: a tight, homogeneous block. Push that through ICA and the
resulting component loadings come out *sub*-Gaussian, with negative excess
kurtosis, and essentially nothing clears `abs(z) > 3`. There is no tail, so a
tail-based rule finds nothing and Jaccard bottoms out. The three LogNormal
generators have a loading CV around 0.8 and produce strongly leptokurtic
components, where the rule works as designed.

So `modular` is a good benchmark for correlation and graph clustering, which
threshold a similarity, and a poor one for matrix factorisation, which thresholds
a loading tail. That is a property of the *generator and metric pairing*, not a
verdict on any method.

Two practical lessons. Pick the generator that matches the property you are
measuring and state which one you used, because a method that looks twice as good
as another may simply have been handed a friendlier dataset. And if a
factorisation returns almost no modules, check the kurtosis of its loadings before
blaming the algorithm: you may just need a different `cutoff` in
`params_module_membership()`, or a generator whose loadings have a tail at all.

## Wrap-up

Different methods suit different questions:

- **CoReMo** if you want signed modules with per-sample eigengenes and a fast, 
  deterministic pipeline. Solid default for standard bulk data.
- **Graph-based Leiden** if you want tunable resolution or hierarchical
  sub-clustering. Handles very heterogeneous module sizes better than CoReMo.
- **ICA** if you want components that decompose independent, potentially
  overlapping biological signals. Interpretation is per-component, not
  per-gene.
- **DGRDL** if you want a sparse dictionary regularised by feature and sample
  neighbour graphs. Sparsity is a tunable constraint, not a byproduct. It comes
  last on recovery here, and note that `dgrdl_grid_search()` ranks by
  reconstruction error, which is not the same objective as recovering planted
  modules. Remember to centre the input!
- **NMF** if you want non-negative, parts-based decompositions with a natural
  interpretation as "how strongly is each sample expressing each module".

For a real dataset, run more than one and cross-check. The synthetic setup here
has a clean signal to noise regime; real bulk data (GTEx, TCGA, recount3) will
happily show larger gaps between methods. `simulate_dropouts()` is the next
knob to turn if you want to see how far each method degrades once you thin the
library sizes towards single-cell-like sparsity.
