---
title: "Bulk differential expression with BulkDge"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Bulk differential expression with BulkDge}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

`BulkDge` is the class for bulk RNA-seq differential expression. You hand it a
count matrix and the sample metadata, and it walks you through the usual
steps: sample QC, gene filtering, normalisation, a PCA to see what drives the
data, optional batch correction, and then the actual tests. Every step stores
its output and its plots on the object, so you can come back to them later.

```
BulkDge()
  -> qc_bulk_dge()              # outlier samples, filterByExpr
  -> normalise_bulk_dge()       # TMM + voom
  -> calculate_pca_bulk_dge()   # what separates the samples?
  -> batch_correction_bulk_dge() # optional, removeBatchEffect
  -> calculate_dge_limma()      # limma-voom
  -> calculate_dge_hedges()     # effect sizes
```

None of this needs limma or edgeR. The numerics run in Rust via
[edge-rs](https://crates.io/crates/edge-rs), a port of the edgeR and limma
stack that is checked against both. Towards the end we put the two side by
side, and go through where the results match, where they differ on purpose, and
what edge-rs does not do (yet).

```{r}
#| label: setup
#| message: false
#| warning: false
library(bixverse)
library(data.table)
library(ggplot2)
library(magrittr)

# the comparison section at the end needs the reference implementations
has_reference <- requireNamespace("limma", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE)
```

## The data

The classic: [airway](https://bioconductor.org/packages/airway/) from
[Himes, et al.](https://doi.org/10.1371/journal.pone.0099625). Four airway
smooth muscle cell lines, each treated with dexamethasone or left untreated.
Eight samples, and a paired design: the cell line is a nuisance factor we want
to account for, the treatment is what we care about.

```{r}
#| label: load data
data("airway", package = "airway")

counts <- SummarizedExperiment::assay(airway, "counts")

meta_data <- as.data.table(
  as.data.frame(SummarizedExperiment::colData(airway)),
  keep.rownames = "sample_id"
)[, .(sample_id, cell, dex)]

meta_data

dim(counts)
```

Setting up the class needs the counts (genes x samples, with names) and a
data.table with a `sample_id` column matching the column names of the counts.

```{r}
#| label: class
dge_obj <- BulkDge(raw_counts = counts, meta_data = meta_data)

dge_obj
```

## QC

`qc_bulk_dge()` does two things. It flags samples that detect far fewer genes
than the rest (more than `outlier_threshold` standard deviations below the
mean) and drops them, and it removes lowly expressed genes with edgeR's
`filterByExpr()` logic within the groups of `group_col`.

```{r}
#| label: qc
dge_obj <- qc_bulk_dge(dge_obj, group_col = "dex")
```

The QC plots live on the object. Plot 1 shows the number of detected genes per
sample, plot 2 the outlier thresholds.

```{r}
#| label: qc plots
#| fig-width: 7
#| fig-height: 4
get_dge_qc_plot(dge_obj, plot_choice = 1L)
get_dge_qc_plot(dge_obj, plot_choice = 2L)
```

## Normalisation

`normalise_bulk_dge()` calculates the normalisation factors (TMM by default)
and applies voom on top, which gives you log2-CPM values to work with for
PCAs, plotting and effect sizes. The mean-variance trend and the per-sample
distributions end up as plots 3 and 4.

```{r}
#| label: normalisation
#| fig-width: 7
#| fig-height: 4
dge_obj <- normalise_bulk_dge(dge_obj, group_col = "dex")

get_dge_qc_plot(dge_obj, plot_choice = 3L)
get_dge_qc_plot(dge_obj, plot_choice = 4L)
```

The factors are stored alongside the counts:

```{r}
#| label: norm factors
get_outputs(dge_obj)$norm_factors
```

Want the edgeR `DGEList` anyway, say for a method bixverse does not wrap?
`get_dge_list()` builds one from the stored counts, library sizes and factors.
This is the one function in the class that needs edgeR installed.

```{r}
#| label: dge list
#| eval: !expr has_reference
get_dge_list(dge_obj)
```

## PCA

Before testing anything, check what actually separates the samples.

```{r}
#| label: pca
#| fig-width: 8
#| fig-height: 4
dge_obj <- calculate_pca_bulk_dge(dge_obj)

plot_pca_res(dge_obj, cols_to_plot = c("dex", "cell"))
```

The class also runs a quick ANOVA of PC1 and PC2 against the groups:

```{r}
#| label: pca anova
get_outputs(dge_obj)$pca_anova
```

## Batch correction

Cell line is a nuisance factor here. For the linear model we'll put it into the
design (see below), which is the right way to deal with it when testing. For
plotting and for effect sizes you want it gone from the expression values
themselves. `batch_correction_bulk_dge()` does that via limma's
`removeBatchEffect()` logic, while protecting the contrast of interest.

```{r}
#| label: batch correction
#| fig-width: 8
#| fig-height: 4
dge_obj <- batch_correction_bulk_dge(
  dge_obj,
  contrast_column = "dex",
  batch_col = "cell"
)

get_dge_qc_plot(dge_obj, plot_choice = "p6_batch_correction_plot")
```

## Differential expression

### limma-voom

`calculate_dge_limma()` fits `~ 0 + dex + cell` and tests every pairwise
contrast between the levels of `contrast_column`. With two levels that is one
contrast, treated versus untreated. The knobs sit in `params_limma_voom()`: the
route (`"voom"` or `"trend"`), the normalisation method, robust empirical Bayes
and friends.

```{r}
#| label: limma
dge_obj <- calculate_dge_limma(
  dge_obj,
  contrast_column = "dex",
  co_variates = "cell",
  limma_params = params_limma_voom(route = "voom", robust = FALSE)
)

limma_res <- get_dge_limma_voom(dge_obj)
head(limma_res)
```

The columns are the ones from limma's `topTable(confint = TRUE)`. Here's the
volcano:

```{r}
#| label: volcano
#| fig-width: 6
#| fig-height: 5
ggplot(
  limma_res,
  aes(x = logFC, y = -log10(P.Value), colour = adj.P.Val < 0.05)
) +
  geom_point(size = 0.5, alpha = 0.5) +
  scale_colour_manual(values = c("grey70", "#3C4B6D")) +
  theme_minimal() +
  labs(
    title = "Dexamethasone vs untreated",
    colour = "FDR < 0.05"
  )
```

```{r}
#| label: n significant
limma_res[, .(
  up = sum(adj.P.Val < 0.05 & logFC > 0),
  down = sum(adj.P.Val < 0.05 & logFC < 0)
)]
```

### Effect sizes

p-values tell you how sure you are, not how big the change is. Hedges' g gives
a standardised effect size per gene, and since we ran the batch correction
above, it is calculated on the corrected values.

```{r}
#| label: hedges
dge_obj <- calculate_dge_hedges(dge_obj, contrast_column = "dex")

head(get_dge_effect_sizes(dge_obj))
```

Watch the direction: the combination here is `untrt_vs_trt`, the opposite way
round to the limma contrast, so the signs flip.

### Outside the class

Both linear model chains are also available as plain functions on a count
matrix. `run_limma_voom()` is what `calculate_dge_limma()` calls under the
hood. `run_edger_ql()` gives you the edgeR quasi-likelihood test instead, which
is the one to reach for when you have few replicates or very low counts.

```{r}
#| label: edger ql
dge_counts <- get_outputs(dge_obj)$dge_counts
# copy: the data.table on the object would otherwise be modified by reference
sample_info <- copy(get_outputs(dge_obj)$sample_info)
sample_info[, dex := factor(dex, levels = c("untrt", "trt"))]

design <- model.matrix(~ cell + dex, data = sample_info)

edger_res <- run_edger_ql(
  counts = dge_counts[, sample_info$sample_id],
  design = design,
  coef = "dextrt",
  edger_params = params_edger_ql(filter = FALSE)
)

head(edger_res[order(p_value)])
```

## edge-rs versus edgeR and limma

Swapping out the R implementations for a Rust port raises the obvious question:
do you get the same answer? For the parts bixverse uses, yes. The package tests
check `filterByExpr()`, TMM factors, `cpm()`, voom's log-CPM values and
weights, the full `voomLmFit()` -> `eBayes()` -> `topTable()` table (log fold
changes, confidence intervals, moderated t, p-values, B) and
`removeBatchEffect()` against edgeR 4.8.2 and limma 3.66.0, to 1e-8 or better.

Let's check it on airway. We rebuild exactly what `calculate_dge_limma()` ran,
on the same filtered counts, with edgeR and limma.

```{r}
#| label: reference fit
#| eval: !expr has_reference
design_ref <- model.matrix(~ 0 + dex + cell, data = sample_info)
colnames(design_ref) <- gsub("dex", "", colnames(design_ref))

# the class keeps the pre-filter library sizes, as a subset DGEList does
y <- edgeR::normLibSizes(
  edgeR::DGEList(
    dge_counts[, sample_info$sample_id],
    lib.size = get_outputs(dge_obj)$lib_size[sample_info$sample_id]
  )
)

fit <- edgeR::voomLmFit(y, design_ref, sample.weights = FALSE)
fit <- limma::contrasts.fit(
  fit,
  limma::makeContrasts(trt - untrt, levels = design_ref)
)
fit <- limma::eBayes(fit)

ref_res <- as.data.table(
  limma::topTable(fit, number = Inf, sort.by = "none", confint = TRUE),
  keep.rownames = "gene_id"
)

comparison <- merge(
  limma_res[, .(gene_id, logFC, t, P.Value)],
  ref_res[, .(gene_id, logFC, t, P.Value)],
  by = "gene_id",
  suffixes = c("_rs", "_r")
)

comparison[, .(
  max_abs_diff_logfc = max(abs(logFC_rs - logFC_r)),
  max_abs_diff_t = max(abs(t_rs - t_r)),
  max_abs_diff_log10p = max(abs(log10(P.Value_rs) - log10(P.Value_r)))
)]
```

Or as plots:

```{r}
#| label: reference fit plots
#| eval: !expr has_reference

p1 <- ggplot(data = comparison, mapping = aes(x = logFC_rs, y = logFC_r)) +
  geom_point() +
  xlab("LFC (Rust)") +
  ylab("LFC (R)") +
  ggtitle(label = waiver(), subtitle = "LFC comparison") +
  theme_minimal() +
  geom_abline(slope = 1, intercept = 0, linesize = 0.25)

p2 <- ggplot(data = comparison, mapping = aes(x = t_rs, y = t_r)) +
  geom_point() +
  xlab("t stat (Rust)") +
  ylab("t stat (R)") +
  ggtitle(label = waiver(), subtitle = "t stat comparison") +
  theme_minimal() +
  geom_abline(slope = 1, intercept = 0, linesize = 0.25)

p1 + p2 + plot_annotation(title = "Rust vs R limma-voom")
```

And voom's normalised values against the ones `normalise_bulk_dge()` stored:

```{r}
#| label: reference voom
#| eval: !expr has_reference
voom_ref <- limma::voom(
  y,
  model.matrix(~ 0 + dex, data = sample_info)
)

max(abs(get_outputs(dge_obj)$normalised_counts - voom_ref$E))
```

Same for the batch correction: `removeBatchEffect()` on the stored voom values,
cell line as batch, treatment protected, against what
`batch_correction_bulk_dge()` stored. Again, this is for plotting and effect
sizes. The test itself had the cell line in the design.

```{r}
#| label: reference batch correction
#| eval: !expr has_reference
batch_info <- get_outputs(dge_obj)$sample_info

corrected_ref <- limma::removeBatchEffect(
  get_outputs(dge_obj)$normalised_counts,
  batch = batch_info$cell,
  design = model.matrix(~ 0 + factor(batch_info$dex))
)

max(abs(get_outputs(dge_obj)$normalised_counts_corrected - corrected_ref))
```

### Where the results differ on purpose

Numbers matching is only half the story. A few defaults differ from what a
classic limma script does, and they will move your results a bit:

- **`voomLmFit()`, not `voom()` + `lmFit()`.** The DGE chain follows edgeR's
  `voomLmFit()`. It masks structural zeros, i.e. genes that are all zero in a
  group, before fitting the mean-variance trend, and gives those genes their
  own residual degrees of freedom. On well-filtered data the two agree
  closely, on sparse data they don't.
- **No quantile normalisation.** Older versions of bixverse quantile-normalised
  the voom output in `normalise_bulk_dge()`. Voom on top of TMM is limma's own
  default, and the one we stick with now.
- **The empirical Bayes trend follows the route.** voom carries the
  mean-variance relationship in its weights, so the prior is not trended;
  limma-trend has nothing else to absorb it, so it is. limma lets you mix and
  match, bixverse doesn't.

Here's how much the first point matters on airway: the classic `voom()` +
`lmFit()` script next to what bixverse runs.

```{r}
#| label: classic limma
#| eval: !expr has_reference
v <- limma::voom(y, design_ref)
fit_classic <- limma::lmFit(v, design_ref)
fit_classic <- limma::contrasts.fit(
  fit_classic,
  limma::makeContrasts(trt - untrt, levels = design_ref)
)
fit_classic <- limma::eBayes(fit_classic)

classic_res <- as.data.table(
  limma::topTable(fit_classic, number = Inf, sort.by = "none"),
  keep.rownames = "gene_id"
)

classic <- merge(
  limma_res[, .(gene_id, logFC, P.Value, adj.P.Val)],
  classic_res[, .(gene_id, logFC, P.Value, adj.P.Val)],
  by = "gene_id",
  suffixes = c("_bixverse", "_classic")
)

classic[, .(
  cor_logfc = cor(logFC_bixverse, logFC_classic),
  cor_log10p = cor(log10(P.Value_bixverse), log10(P.Value_classic)),
  sig_bixverse = sum(adj.P.Val_bixverse < 0.05),
  sig_classic = sum(adj.P.Val_classic < 0.05),
  sig_both = sum(adj.P.Val_bixverse < 0.05 & adj.P.Val_classic < 0.05)
)]
```

Fold changes are identical to the last digit and the p-values barely move. The
classic script calls a few dozen more genes at 5% FDR, and nearly everything
bixverse calls, it calls too. Different defaults, same biology.

### What edge-rs does not do (yet)

- **Sample or array weights.** No `voomWithQualityWeights()` or
  `arrayWeights()` in the bixverse chain.
- **Blocking.** No `duplicateCorrelation()` and no `block` argument. Repeated
  measures have to go into the design as a fixed effect, as the cell line did
  above.
- **The moderated F-test.** One coefficient or contrast at a time. Several at
  once errors rather than silently testing the first.
- **Arbitrary contrasts.** `contrast_list` takes `"a-b"` strings only, not the
  full expression syntax of `makeContrasts()`.
- **`treat()`** and the other limma extras.

If you need any of those, `get_dge_list()` hands you a `DGEList` and you can
carry on in limma or edgeR directly.

## Where next

The DGE results feed straight into the gene set enrichment methods, see
`vignette("gse_methods")`. For co-expression rather than differential
expression, `vignette("bulk_coexpression_modules")`. For single cell
pseudobulk, NEBULA and friends, `vignette("differential_expression")`.
