---
title: "Ambient RNA removal with CellSweep"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Ambient RNA removal with CellSweep}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Cells lyse during dissociation and their transcripts end up in the buffer. Every
droplet then gets loaded with a bit of that soup on top of whatever cell it
caught, which is why you find monocyte genes in T cells, haemoglobin everywhere,
and lineage markers that refuse to be clean. That background is the ambient RNA.

[CellSweep](https://github.com/pachterlab/cellsweep) attacks it as a mixture
problem. Every observed count in a barcode comes from one of three places:
the ambient profile of that emulsion, a global bulk profile, or the barcode's
own cell type profile. An EM fit splits the counts three ways per barcode, and
subtracting the first two components leaves the denoised matrix. The per-barcode
ambient fraction, `alpha`, falls out of the same fit, which doubles as a quality
metric: a barcode that is 60% soup is not a cell you want in your clustering.

Two things about where this sits in a workflow, because both are easy to get
wrong.

**It runs after annotation, not before.** The model subtracts against cell type
profiles, so it needs the labels on input. The chain is: ingest the raw
barcodes, cluster and annotate as usual, then `cellsweep_sc()`, then redo the
feature selection and reduction on the clean counts.

**It needs the empty droplets.** The ambient profile is estimated from their
pooled counts, and they stay in the EM with their contamination fraction pinned
at 1. So the object has to be ingested permissively. The load-time cutoffs in
`bixverse` are irreversible, and the defaults delete exactly the barcodes the
model trains on.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(data.table)
library(ggplot2)
```

## The raw matrix

We use the 10x Genomics 5' PBMC 1k run, the same data the CellSweep reference
notebook works through. This is the unfiltered matrix, so every barcode the
sequencer saw is in there.

```{r}
#| label: download the raw matrix
h5_path <- download_pbmc_1k_5p(quiet = FALSE)

meta <- read_tenx_h5_metadata(h5_path)
meta$dims
meta$feature_types
```

737,280 barcodes and 36,620 features, of which 19 are Antibody Capture. Only the
gene expression ones are wanted here, which is what `feature_type` selects.

Now the permissive load. `min_lib_size = 1L` throws away the barcodes that saw
no molecule at all, which is most of them, and keeps everything else.

```{r}
#| label: load the raw barcodes
tempdir_cellsweep <- file.path(tempdir(), "cellsweep_bixverse")
dir_raw <- file.path(tempdir_cellsweep, "raw")
dir.create(dir_raw, showWarnings = FALSE, recursive = TRUE)

raw_object <- load_tenx_h5(
  object = SingleCells(dir_data = dir_raw),
  h5_path = h5_path,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 0L,
    min_lib_size = 1L,
    min_cells = 0L
  ),
  feature_type = "Gene Expression",
  .verbose = TRUE
)

raw_object
```

98k barcodes carrying at least one UMI. The vast majority are empty droplets and
that is the point: they are the training data for the soup.

## Calling the empty droplets

`params_sc_empty_droplets()` offers four ways to decide which barcode is empty,
in descending order of how much you should trust them. `"supplied"` takes an
existing logical column from the obs table, and if you have Cell Ranger's
filtered barcode list you already know the answer. The other three infer it from
the library sizes alone.

The barcode rank curve is the thing they are all trying to read.

```{r}
#| label: barcode rank plot
#| fig-width: 7
#| fig-height: 4.5
obs_raw <- get_sc_obs(raw_object, filtered = FALSE)

rank_dt <- data.table(lib_size = obs_raw$lib_size)
setorder(rank_dt, -lib_size)
rank_dt[, rank := .I]

ggplot(rank_dt, aes(x = rank, y = lib_size)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 257, colour = "firebrick", linetype = "dashed") +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "Barcode rank",
    y = "Library size",
    title = "Barcode rank curve",
    subtitle = "Dashed line: the 257 UMI cutoff the reference notebook uses"
  ) +
  theme_minimal()
```

The cliff is obvious. Roughly a thousand barcodes sit well above it and
everything else falls off to single-digit libraries.

`rs_sc_infer_empty_droplets()` is the Rust side of the three inference methods,
exposed so the calls can be compared before committing to one.

```{r}
#| label: compare the empty droplet methods
methods <- list(
  umi_cutoff = params_sc_empty_droplets(
    method = "umi_cutoff",
    umi_cutoff = 257L
  ),
  expected_cells = params_sc_empty_droplets(
    method = "expected_cells",
    expected_cells = 1000L
  ),
  knee = params_sc_empty_droplets(method = "knee")
)

comparison <- rbindlist(purrr::imap(methods, \(params, name) {
  call <- rs_sc_infer_empty_droplets(
    lib_size = as.integer(obs_raw$lib_size),
    empty_params = unclass(params)
  )
  data.table(
    method = name,
    empty = sum(call),
    kept = sum(!call),
    min_kept_lib_size = min(obs_raw$lib_size[!call])
  )
}))

comparison
```

`umi_cutoff` and `expected_cells` land on the same thousand barcodes, which they
should: the cutoff the notebook quotes is the one you get by asking for a
thousand cells. The knee detector stops at 120 UMIs instead of 257 and hands
back 19 extra barcodes. It is experimental in the reference too, so treat it as
a starting point rather than an answer.

We go with the cutoff.

```{r}
#| label: pick the empty droplet call
empty_params <- params_sc_empty_droplets(
  method = "umi_cutoff",
  umi_cutoff = 257L
)
```

## Annotating the real cells

CellSweep needs cell type labels, so the real barcodes have to go through a
normal pipeline first. The raw object is not the place to do that, since 97k of
its barcodes are soup. Load the same file a second time with QC that means
something, and cluster there.

```{r}
#| label: load the cells
dir_cells <- file.path(tempdir_cellsweep, "cells")

dir.create(dir_cells, showWarnings = FALSE, recursive = TRUE)

cell_object <- load_tenx_h5(
  object = SingleCells(dir_data = dir_cells),
  h5_path = h5_path,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 200L,
    min_lib_size = 257L,
    min_cells = 3L
  ),
  feature_type = "Gene Expression",
  .verbose = TRUE
)

cell_object
```

Standard QC on top: mitochondrial fraction, then MAD-based outlier detection
across the usual three metrics.

```{r}
#| label: quality control
var_cells <- get_sc_var(cell_object)

cell_object <- gene_set_proportions_sc(
  cell_object,
  gene_set_list = list(MT = var_cells[grepl("^MT-", gene_name), gene_id]),
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- cell_object[[c("cell_id", "lib_size", "nnz", "MT")]]

qc <- run_cell_qc(
  metrics = list(
    log10_lib_size = log10(qc_df$lib_size),
    log10_nnz = log10(qc_df$nnz),
    MT = qc_df$MT
  ),
  cells_to_keep = get_cells_to_keep(cell_object),
  directions = c(
    log10_lib_size = "twosided",
    log10_nnz = "twosided",
    MT = "above"
  ),
  threshold = 3
)

cell_object <- set_cells_to_keep(cell_object, qc_df[!qc$combined, cell_id])

sum(qc$combined)
```

Then the pipeline, unchanged from any other single cell workflow.

```{r}
#| label: cluster the cells
cell_object <- find_hvg_sc(cell_object, hvg_no = 2000L, .verbose = FALSE)

cell_object <- calculate_pca_sc(
  cell_object,
  no_pcs = 30L,
  sparse_svd = TRUE,
  .verbose = FALSE
)

cell_object <- find_neighbours_sc(
  cell_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  .verbose = FALSE
)

cell_object <- find_clusters_sc(
  cell_object,
  res = 1,
  name = "leiden_clusters"
)

obs_cells <- get_sc_obs(cell_object)

table(obs_cells$leiden_clusters)
```

762 cells over eight clusters. Good enough to subtract against; you would name
them with `calc_sc_type_scores()` or a reference in a real analysis, but the
model only cares that barcodes with the same label share a profile.

## Putting the labels back

The labels live on the cell object and CellSweep reads them off the raw one, so
they have to be joined across by barcode. Every barcode that is not a clustered
cell keeps an `NA`, which is exactly how `cellsweep_sc()` finds the barcodes it
should ignore.

```{r}
#| label: join the labels
label <- obs_cells$leiden_clusters[
  match(obs_raw$cell_id, obs_cells$cell_id)
]

cell_type <- rep(NA_character_, nrow(obs_raw))
cell_type[!is.na(label)] <- sprintf("cluster_%s", label[!is.na(label)])

raw_object[["cell_type"]] <- cell_type
raw_object[["sample_id"]] <- rep("pbmc_1k", nrow(obs_raw))

table(cell_type, useNA = "ifany")
```

Watch the `NA` handling there. `sprintf("cluster_%s", NA)` gives you the string
`"cluster_NA"`, and CellSweep will happily treat those barcodes as a ninth cell
type and fit a profile to them. Keep the `NA` an `NA`.

`sample_id` is a constant here because this is one emulsion. It is still
required: the ambient profile is a property of a single run, and pooling samples
into one profile is wrong even when it runs.

## Running CellSweep

```{r}
#| label: run cellsweep
dir_clean <- file.path(tempdir_cellsweep, "clean")
dir.create(dir_clean, showWarnings = FALSE, recursive = TRUE)

clean_object <- cellsweep_sc(
  target = SingleCells(dir_data = dir_clean),
  input = raw_object,
  celltype_column = "cell_type",
  sample_column = "sample_id",
  empty_params = empty_params,
  .verbose = TRUE
)

clean_object
```

Three numbers in that output are worth reading. 96,998 barcodes went into the
ambient profile. 762 got denoised. And 238 barcodes were neither: they cleared
the UMI cutoff but failed QC or the outlier check, so they have no label to
subtract against. Those are dropped from the fit and from the output, which is
why the clean object is 762 cells and not 1000.

The denoised counts land in a new directory. The raw layer takes stochastically
rounded values so the negative binomial methods downstream still see integers,
and the normalised layer keeps the float magnitudes.

## What came out

The ambient profile is the first thing to look at, because it is the one part of
the fit you can sanity check against what you know about the tissue.

```{r}
#| label: the ambient profile
#| fig-width: 7
#| fig-height: 5
var_clean <- get_sc_var(clean_object)

top_ambient <- var_clean[order(-cellsweep_ambient)][1:20]

ggplot(
  top_ambient,
  aes(x = cellsweep_ambient, y = reorder(gene_name, cellsweep_ambient))
) +
  geom_col(fill = "steelblue") +
  labs(
    x = "Fraction of the ambient profile",
    y = NULL,
    title = "What the soup is made of"
  ) +
  theme_minimal()
```

MALAT1, elongation factors, ribosomal proteins, B2M, mitochondrial genes. That
is the soup of a lysing PBMC and nothing else, which is a good sign.

Next, the per-barcode contamination.

```{r}
#| label: the alpha distribution
#| fig-width: 7
#| fig-height: 4.5
obs_clean <- get_sc_obs(clean_object, filtered = FALSE)

ggplot(obs_clean, aes(x = cellsweep_alpha)) +
  stat_ecdf(linewidth = 0.7) +
  geom_vline(xintercept = 0.3, colour = "firebrick", linetype = "dashed") +
  labs(
    x = "Estimated ambient fraction",
    y = "Cumulative fraction of cells",
    title = "How much soup each barcode carries"
  ) +
  theme_minimal()
```

```{r}
#| label: alpha summary
summary(obs_clean$cellsweep_alpha)

obs_clean[,
  .(n = .N, median_alpha = median(cellsweep_alpha)),
  by = cell_type
][order(-median_alpha)]
```

This run is clean. The median barcode is essentially uncontaminated and only a
handful sit above the 0.3 the reference notebook filters at. That is not always
what you get, and it is worth knowing before you assume your lineage markers are
real.

### Does the subtraction do anything?

The honest test is not whether alpha looks plausible but whether the counts
change in the direction they should. Monocyte genes appearing in T cells are
soup by construction, so track a couple of them across clusters, before and
after.

```{r}
#| label: soup markers before and after
markers <- c("LYZ" = "monocyte", "S100A8" = "monocyte", "PPBP" = "platelet")
marker_ids <- var_clean[gene_name %in% names(markers), gene_id]
names(marker_ids) <- var_clean[gene_name %in% names(markers), gene_name]

detection_rate <- function(object, layer_name) {
  expr <- extract_gene_expression(
    object,
    features = marker_ids,
    obs_cols = c("cell_id", "cell_type")
  )
  setDT(expr)
  expr <- expr[!is.na(cell_type)]
  long <- melt(
    expr,
    id.vars = c("cell_id", "cell_type"),
    variable.name = "gene_id",
    value.name = "expr"
  )
  long[, gene := names(marker_ids)[match(gene_id, marker_ids)]]
  long[,
    .(detected = mean(expr > 0), layer = layer_name),
    by = .(gene, cell_type)
  ]
}

rates <- rbind(
  detection_rate(raw_object, "raw"),
  detection_rate(clean_object, "denoised")
)

dcast(rates, gene + cell_type ~ layer, value.var = "detected")
```

```{r}
#| label: soup marker plot
#| fig-width: 10
#| fig-height: 4
ggplot(rates, aes(x = cell_type, y = detected, fill = layer)) +
  geom_col(position = "dodge") +
  facet_wrap(~gene) +
  scale_fill_manual(values = c(raw = "grey60", denoised = "steelblue")) +
  labs(x = NULL, y = "Fraction of cells with a non-zero count", fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Cluster 1 is the monocytes and it keeps LYZ and S100A8 in essentially every
cell. Everywhere else those genes were showing up in a few percent of barcodes
before the sweep and are gone afterwards. PPBP does the same thing in reverse:
platelet transcripts leaking into the monocytes disappear. The subtraction went
after the right counts.

## Reprocessing the clean counts

Feature selection and reduction were computed on contaminated counts, so they
have to be redone. Drop the barcodes CellSweep says are mostly soup while you
are at it.

```{r}
#| label: reprocess
clean_object <- set_cells_to_keep(
  clean_object,
  obs_clean[cellsweep_alpha <= 0.3, cell_id]
)

clean_object <- find_hvg_sc(clean_object, hvg_no = 2000L, .verbose = FALSE)

clean_object <- calculate_pca_sc(
  clean_object,
  no_pcs = 30L,
  sparse_svd = TRUE,
  .verbose = FALSE
)

clean_object <- find_neighbours_sc(
  clean_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  .verbose = FALSE
)

clean_object <- umap_sc(clean_object, .verbose = FALSE)

clean_object
```

```{r}
#| label: umap on the clean counts
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  clean_object,
  embedding = "umap",
  colour_by = "cell_type",
  label_by = "cell_type",
  discrete = TRUE
)
```

The clusters survive the sweep, which is what you want. CellSweep is not
supposed to rearrange your populations, it is supposed to stop the space between
them being filled with borrowed transcripts. If your structure falls apart after
denoising, the labels you handed it were wrong.

## Where this goes next

The obvious extension is multi-sample data, where the ambient profile is fitted
per emulsion and `sample_column` starts earning its keep. A soup that differs
between runs is a batch effect that no correction method will name for you, and
having `cellsweep_alpha` and the per-sample profiles in hand is a much better
starting point than pretending the contamination is uniform.

`cellsweep_alpha` is also worth keeping in the obs table after the fact. It is a
QC metric in its own right, and a cluster with a suspiciously high median is
usually debris rather than biology.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_cellsweep, recursive = TRUE, force = TRUE)
```
