---
title: "Multicellular programmes with DIALOGUE"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Multicellular programmes with DIALOGUE}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Most programme discovery methods ask what varies together *within* a cell type.
NMF on the T cells gives you T cell programmes; Hotspot on the macrophages
gives you macrophage programmes. Neither tells you whether the T cell
programme and the macrophage programme are the same thing seen from two sides.

[DIALOGUE](https://doi.org/10.1038/s41587-022-01288-0) asks the other question.
It looks for multicellular programmes (MCPs): axes of variation that show up in
several cell types *at once*, sample by sample. If macrophages in a biopsy sit
high on some axis and the epithelial cells in the same biopsy also sit high,
and that holds across enough biopsies, then whatever those cells are doing is
coordinated rather than independent.

Three stages, and it is worth knowing the shape before you run it:

1. Each cell type's features are averaged to one row per sample and put through
   a sparse multi-CCA. That gives every programme a weight vector per cell type
   and a provisional gene signature.
2. For every ordered pair of cell types and every candidate gene, a mixed model
   asks whether a cell's own programme score tracks the partner cell type's
   expression of that gene in the same sample. The random intercept is over
   samples.
3. The partners are meta-analysed, and the scores are refit onto the surviving
   genes by non-negative least squares.

If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first; this one assumes you know how `SingleCells`, subsets and
on-disk storage work.

::: {.callout-important}
## DIALOGUE needs a study design, not just a lot of cells

Because stage one works on per-sample averages, DIALOGUE needs **multiple
samples and multiple cell types, with every cell type captured in every
sample**. `bixverse` hard-errors below five samples present in all cell types,
and drops samples that do not clear `abn_c` cells for a given cell type.

This is the thing that catches people out. A beautiful 500k cell atlas from one
donor is useless here. Ten scrappy biopsies are not.
:::

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(ggplot2)
library(data.table)
library(magrittr)
```

## The data

We use the ulcerative colitis subset from the DIALOGUE paper, originally from
[Smillie, et al.](https://doi.org/10.1016/j.cell.2019.06.029). 5,374 cells
across five cell subtypes and 30 donors: twelve healthy, eighteen with UC who
each contributed an inflamed and a non-inflamed biopsy.

One provenance note. The published matrix holds `log2(TPM / 10 + 1)` and the
raw counts were never released, so the file served here has been linearised
back to the TPM/10 scale and rounded to integers. It therefore loads like any
count matrix and picks up `bixverse`'s own log CPM on the way in. That is fine
for DIALOGUE, which only ever reads the normalised layer.

```{r}
#| label: load-data
dialogue_h5ad <- download_dialogue_uc()

tempdir_dlg <- file.path(tempdir(), "dialogue_uc")
dir.create(tempdir_dlg, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = tempdir_dlg)
sc_object <- load_h5ad(
  object = sc_object,
  h5_path = dialogue_h5ad,
  .verbose = FALSE
)

sc_object
```

Before anything else, look at the design. This is the table that decides
whether DIALOGUE can run at all.

```{r}
#| label: design
obs <- get_sc_obs(sc_object, filtered = TRUE)

table(obs$cell_subtype, obs$clinical_status)
```

`CD8+ IL17+` is the problem: 99 cells in total, and completely absent from
twelve of the thirty donors. A cell type missing from a sample means that
sample cannot contribute to any pair involving it, so the sensible move is to
drop the cell type rather than lose the samples. pertpy's DIALOGUE tutorial
does the same thing on the same data.

```{r}
#| label: drop-rare
sc_object <- set_cells_to_keep(
  x = sc_object,
  cells_to_keep = obs[cell_subtype != "CD8+ IL17+", cell_id]
)

sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)

obs <- get_sc_obs(sc_object, filtered = TRUE)
cell_types <- sort(unique(obs$cell_subtype))

cell_types
min(table(obs$sample, obs$cell_subtype))
```

Four cell types, thirty donors, and the sparsest donor-by-cell-type cell still
has cells in it. That is a design DIALOGUE can work with.

## Per-cell-type features

This is the step to get right, and the one the documentation warns about: the
`features` you hand DIALOGUE must be computed **per cell type**, not sliced out
of a global PCA. A global PCA's leading components mostly separate the cell
types from each other, so slicing it hands every cell type the same axis and
DIALOGUE dutifully reports that the cell types agree with themselves.

So: subset, find variable genes within the subset, run a PCA within the subset.

```{r}
#| label: features
features <- purrr::map(cell_types, \(ct) {
  sub <- SingleCellsSubset(
    sc_object = sc_object,
    grouping_column = "cell_subtype",
    group = ct
  )
  sub <- find_hvg_sc(sub, hvg_no = 2000L, .verbose = FALSE)
  sub <- calculate_pca_sc(sub, no_pcs = 10L, .verbose = FALSE)

  get_pca_factors(sub)
})
names(features) <- cell_types

purrr::map(features, dim)
```

`features` is a named list of matrices, names matching the cell type labels,
row names matching cell ids. Rows are matched by name rather than position, so
the order does not matter, but the coverage does: every cell of that type has
to be in there.

## Running it

Three parameter bundles, one per stage. The defaults follow upstream's
`DLG.get.param`, and the two knobs worth thinking about live in
`params_dialogue_pmd()`: `k` is how many programmes you are asking for, with no
sweep to help you pick it, and `n_permutations` sets the resolution of the
empirical p-value.

We ask for three programmes, matching what pertpy reports on this data.

```{r}
#| label: run-dialogue
dlg_res <- dialogue_sc(
  object = sc_object,
  cell_type_col = "cell_subtype",
  sample_col = "sample",
  features = features,
  quality_col = "cell_q",
  pmd_params = params_dialogue_pmd(k = 3L, n_permutations = 100L)
)

dlg_res
```

`quality_col = "cell_q"` is the per-cell quality score the original paper
shipped with this data. Leave it `NULL` and `bixverse` falls back to the
z-scored log library size instead. That choice matters more than it looks, and
we come back to it below.

## Reading the output

`get_results()` gives the programme table: one row per programme per ordered
pair of cell types, with the empirical p-value and the correlation between the
two cell types' sample-level scores.

```{r}
#| label: programmes
get_results(dlg_res)
```

A programme is only interesting if it holds up across *every* pair, so the
number to look at is the worst p in each programme, which is what `print()`
reports. With `n_permutations = 100` the p-value cannot go below roughly 0.01,
so do not read too much into the exact values; raise it for anything you intend
to believe.

`refit_fidelity` says how well each cell type's score survived stage three,
i.e. how much of the programme is really carried by the genes that passed the
mixed model.

```{r}
#| label: fidelity
round(dlg_res$refit_fidelity, 3)
```

The signatures come in two flavours. `permissive` and `strict` are nested by
evidence rather than by threshold, and `strict` is the one to quote.

```{r}
#| label: signatures
dcast(
  dlg_res$signatures[
    list == "strict",
    .N,
    by = .(cell_type, programme, direction)
  ],
  cell_type + programme ~ direction,
  value.var = "N"
)
```

## Does it track the disease?

The scores are a named list, one cells-by-programme matrix per cell type.
Average them up to the donor and compare across UC status.

```{r}
#| label: scores
score_dt <- rbindlist(purrr::imap(dlg_res$scores, \(mat, ct) {
  data.table(cell_id = rownames(mat), cell_type = ct, as.data.table(mat))
}))

score_dt <- merge(
  score_dt,
  obs[, .(cell_id, sample, pathology, clinical_status, lib_size)],
  by = "cell_id"
)

per_sample <- score_dt[,
  lapply(.SD, mean),
  by = .(sample, cell_type, pathology),
  .SDcols = patterns("^mcp_")
]

per_sample[,
  .(p = signif(wilcox.test(mcp_01 ~ pathology)$p.value, 3)),
  by = cell_type
]
```

`mcp_01` separates UC donors from healthy ones in all four cell types at once,
which is the thing DIALOGUE exists to find. Plotted:

```{r}
#| label: mcp-plot
#| fig-cap: "Donor-level mcp_01 score by cell type and disease status."
ggplot(
  per_sample,
  aes(x = pathology, y = mcp_01, fill = pathology)
) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1.2, alpha = 0.7) +
  facet_wrap(~cell_type, nrow = 1) +
  labs(x = "Ulcerative colitis", y = "mcp_01 (donor mean)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Before you believe it: check the confounders

Here is where a vignette that wanted to look good would stop. Do not stop here.

A programme that shows up in every cell type at once is exactly what a
*technical* axis looks like. Depth, capture efficiency and ambient RNA hit all
cell types in a sample together, which is precisely the signature DIALOGUE is
built to detect. So before reading any biology into `mcp_01`, correlate it
against the quality covariates.

```{r}
#| label: confound-check
score_dt[,
  .(rho = round(cor(mcp_01, lib_size, method = "spearman"), 3)),
  by = cell_type
]
```

That is not a small correlation. `mcp_01` is tracking library size at
`rho ~ 0.5` in every cell type, and library size itself differs between the UC
and healthy donors on this subset. So a good chunk of what we just called a
disease programme is a depth axis wearing a lab coat.

The fix is to condition on the right covariate. `cell_q` is the paper's own
quality score and it barely correlates with `mcp_01` here, so it was never
going to absorb this. The library-size covariate will.

```{r}
#| label: rerun-lib
dlg_lib <- dialogue_sc(
  object = sc_object,
  cell_type_col = "cell_subtype",
  sample_col = "sample",
  features = features,
  quality_col = NULL,
  pmd_params = params_dialogue_pmd(k = 3L, n_permutations = 100L),
  .verbose = FALSE
)

score_lib <- rbindlist(purrr::imap(dlg_lib$scores, \(mat, ct) {
  data.table(cell_id = rownames(mat), cell_type = ct, as.data.table(mat))
}))
score_lib <- merge(
  score_lib,
  obs[, .(cell_id, sample, pathology, lib_size)],
  by = "cell_id"
)

score_lib[,
  .(rho = round(cor(mcp_01, lib_size, method = "spearman"), 3)),
  by = cell_type
]
```

Much better. Two things to notice about what changed and what did not.

The programme table and the p-values are **identical** to the first run. That
is not a bug: the quality covariate only enters stage two, so stage one's
decomposition and its permutation null never saw it.

What did change is the signature. Conditioning on depth drops the strict gene
count by more than half, because the genes that were only there to track
library size no longer clear the mixed model.

```{r}
#| label: signature-shrink
c(
  cell_q = nrow(dlg_res$signatures[list == "strict"]),
  lib_size = nrow(dlg_lib$signatures[list == "strict"])
)
```

And the disease association survives.

```{r}
#| label: survives
score_lib[,
  lapply(.SD, mean),
  by = .(sample, cell_type, pathology),
  .SDcols = patterns("^mcp_")
][,
  .(p = signif(wilcox.test(mcp_01 ~ pathology)$p.value, 3)),
  by = cell_type
]
```

So there is a real coordinated programme here, it is just a good deal smaller
than the first run advertised. That is the normal outcome of doing this check,
and it is why you do it.

## Caveats and what is next

A few things worth knowing before you point this at your own data.

- **`k` has no sweep.** Unlike `nmf_k_sweep_sc()`, there is no diagnostic to
  pick the number of programmes. Run a few values and look at whether the extra
  programmes hold up across every pair.
- **`satterthwaite` dominates the runtime.** Turning it off in
  `params_dialogue_hlm()` falls back to the residual count for the degrees of
  freedom, which is faster and usually close enough for a first pass.
- **The random intercept level is a real modelling choice.** We used the donor,
  matching the original analysis. Using the biopsy instead
  (`paste(sample, clinical_status)`) would let the inflamed and non-inflamed
  samples from one patient separate, at the cost of samples that then fall
  below `abn_c`.
- **Meta cells have to be built within samples.** Otherwise the random
  intercept has no well-defined level. `meta_cells_per_group()` with
  `group_col` set to your sample column does exactly that, and `dialogue_sc()`
  dispatches on `MetaCells` the same way it does on `SingleCells`. You have to
  aggregate the features over each meta cell's membership yourself.
- **This is a four-cell-type slice of a fifty-one-cell-type atlas.** The
  published UC programme spanned cell types that are simply not in this file,
  so partial agreement with the paper is the expected result rather than a
  failure.

## Clean up

```{r}
#| label: clean-up
unlink(tempdir_dlg, recursive = TRUE, force = TRUE)
```
