---
title: "Differential expression with donor structure"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Differential expression with donor structure}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

You have 3,000 monocytes from eight donors and you want to know which genes
respond to a treatment. The obvious move is a Wilcoxon test over cells. It is
also the wrong one, and the reason is worth being precise about.

Those 3,000 cells are not 3,000 independent observations. They came from eight
people. Cells from the same donor are correlated, because donors differ in
genotype, in batch, in how the sample was handled, in everything. A test that
treats each cell as independent has an effective sample size of eight and
thinks it has 3,000, so its p-values are far too small. This is
pseudoreplication, and in single cell data it is the default failure mode
rather than an exotic one.

There are two honest ways out.

**Pseudobulk.** Sum the counts per donor per condition and run a bulk method.
The unit of replication becomes the donor, which is correct. You throw away
everything the cell-level data told you about variability within a donor, and
with few donors you end up with very little power.

**A mixed model.** Keep the cells, but model the donor as a random effect.
[NEBULA](https://doi.org/10.1038/s42003-021-02146-6) fits a negative binomial
gamma mixed model per gene, splitting the variance into a subject-level term
and a cell-level term. You keep the cell-level information and still pay the
right price for the donor structure.

`bixverse` gives you all three, and this vignette runs them side by side, first
on a real effect and then on no effect at all. The second one is the
interesting bit.

```{r}
#| label: setup
#| message: false
#| warning: false
library(bixverse)
library(data.table)
library(SingleCellExperiment)
```

## The data

[Kang, et al.](https://doi.org/10.1038/nbt.4042) pooled PBMCs from eight lupus
donors, split each sample in two, and stimulated one half with IFN-β. Every
donor contributes to both arms, so the design is paired and the donor is a
genuine random effect. Roughly 29k cells before quality control.

The object is a Bioconductor `SingleCellExperiment`, which
[`load_sce()`](https://gregorlueg.github.io/bixverse/reference/load_sce.html)
takes directly: `colData` becomes obs, `rowData` becomes var, and the counts go
through the usual Rust quality control and normalisation.

Drop the doublets first. The `multiplets` column is the demultiplexing call,
and a doublet is two cells' counts added together, which is not something any
of these models should see.

```{r}
#| label: load-data
sce <- qs2::qs_read(download_kang_pbmc())
sce <- sce[, sce$multiplets == "singlet" & !is.na(sce$cell)]

dir_kang <- file.path(tempdir(), "kang")
dir.create(dir_kang, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = dir_kang)
sc_object <- load_sce(
  object = sc_object,
  sce = sce,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 200L,
    min_lib_size = 500L,
    min_cells = 20L,
    target_size = 1e4
  ),
  .verbose = TRUE
)

sc_object
```

Look at the design before anything else.

```{r}
#| label: design
obs <- sc_object[[]]

table(obs$ind, obs$stim)
```

Eight donors, both arms, hundreds to thousands of cells each. That is what
NEBULA needs. It also needs enough cells per donor: below thirty it quietly
downgrades from its LN method to HL, as the original package does.

## One cell type at a time

Differential expression across a mixed population mostly measures composition.
If the treatment shifts the monocyte fraction, every monocyte gene comes out
"differential" whether or not any monocyte changed. Work within a cell type.

```{r}
#| label: subset
mono <- SingleCellsSubset(
  sc_object,
  grouping_column = "cell",
  group = "CD14+ Monocytes"
)

mono <- find_hvg_sc(mono, hvg_no = 500L, .verbose = FALSE)
hvgs <- get_gene_names_from_idx(mono, get_hvg(mono), rust_based = TRUE)

mono
```

Restricting to the highly variable genes is not just for speed. NEBULA fits an
optimiser per gene, and running it over 12,000 genes to find the 300 that
matter is a poor trade.

## The real contrast

IFN-β stimulation is about as strong a perturbation as single cell data offers.
All three approaches should find it. Same gene set for all three, so the counts
are comparable.

```{r}
#| label: real-wilcox
mono_obs <- mono[[]]

wilcox_res <- find_markers_sc(
  mono,
  cells_1 = mono_obs[stim == "stim", cell_id],
  cells_2 = mono_obs[stim == "ctrl", cell_id],
  .verbose = FALSE
)

wilcox_res[gene_id %in% hvgs & fdr <= 0.05, .N]
```

For pseudobulk, the samples are donor by condition, so sixteen of them. The
design blocks on the donor, which is what makes it a paired test.

```{r}
#| label: real-pseudobulk
cell_list <- split(
  mono_obs$cell_id,
  paste(mono_obs$ind, mono_obs$stim, sep = "_")
)

samples <- data.table(sample = names(cell_list))
samples[, c("ind", "stim") := tstrsplit(sample, "_", fixed = TRUE)]
design_pb <- stats::model.matrix(~ ind + stim, data = samples)

pb_res <- pseudobulk_dge_sc(
  object = mono,
  cell_list = cell_list,
  design = design_pb,
  coef = "stimstim",
  edger_params = params_edger_ql(filter = FALSE),
  .verbose = FALSE
)

pb_res[feature_id %in% hvgs & fdr <= 0.05, .N]
```

NEBULA takes the design as a formula against the obs table and the donor column
separately. The donor is not a term in the design; it is the random effect.

```{r}
#| label: real-nebula
nebula_res <- nebula_sc(
  object = mono,
  subject_col = "ind",
  design = ~stim,
  genes_to_use = hvgs,
  .verbose = TRUE
)

nebula_res
```

```{r}
#| label: real-nebula-top
head(
  nebula_res$results[order(p_value), .(gene_id, log_fc, z, p_value, fdr)],
  8
)
```

CXCL10, CCL8 and APOBEC3B at the top, IL8 down. That is the interferon
response, and it is reassuring rather than surprising.

Note that the Wilcoxon test finds *fewer* genes here than the other two, not
more. Pseudoreplication inflates significance, but the Wilcoxon test is also
a rank test with a expression-proportion filter, and it is simply less powerful
for a large shift in mean. On a real, strong effect all three agree the effect
is there. The difference shows up somewhere else.

## No contrast at all

Here is the test that matters. Take the control cells only, so there is no
treatment anywhere in the data, and split the eight donors into two arbitrary
groups of four. Any gene called differential is a false positive by
construction.

```{r}
#| label: null-setup
ctrl <- mono_obs[stim == "ctrl"]
donors <- sort(unique(ctrl$ind))
group_a <- donors[c(1, 3, 5, 7)]

ctrl[, fake := ifelse(ind %in% group_a, "a", "b")]

table(ctrl$ind, ctrl$fake)
```

```{r}
#| label: null-wilcox
wilcox_null <- find_markers_sc(
  mono,
  cells_1 = ctrl[fake == "a", cell_id],
  cells_2 = ctrl[fake == "b", cell_id],
  .verbose = FALSE
)

wilcox_null[gene_id %in% hvgs & fdr <= 0.05, .N]
```

```{r}
#| label: null-pseudobulk
cell_list_null <- split(ctrl$cell_id, ctrl$ind)

samples_null <- data.table(ind = names(cell_list_null))
samples_null[, fake := ifelse(ind %in% group_a, "a", "b")]
design_null <- stats::model.matrix(~fake, data = samples_null)

pb_null <- pseudobulk_dge_sc(
  object = mono,
  cell_list = cell_list_null,
  design = design_null,
  edger_params = params_edger_ql(filter = FALSE),
  .verbose = FALSE
)

pb_null[feature_id %in% hvgs & fdr <= 0.05, .N]
```

`SingleCellsSubset()` subsets a `SingleCells`, not another subset, so fold both
restrictions into one grouping column on the parent.

```{r}
#| label: null-nebula
parent_obs <- sc_object[[]]
sc_object[["cell_stim"]] <- paste(parent_obs$cell, parent_obs$stim, sep = "__")
sc_object[["fake"]] <- ifelse(parent_obs$ind %in% group_a, "a", "b")

mono_ctrl <- SingleCellsSubset(
  sc_object,
  grouping_column = "cell_stim",
  group = "CD14+ Monocytes__ctrl"
)

nebula_null <- nebula_sc(
  object = mono_ctrl,
  subject_col = "ind",
  design = ~fake,
  genes_to_use = hvgs,
  .verbose = TRUE
)

nebula_null$results[fdr <= 0.05, .N]
```

The cell-level test calls dozens of genes differential between two groups of
donors that differ in nothing. The donor-aware tests call almost none. That gap
is the entire argument, and it does not go away with more cells: sequencing
more cells per donor makes the Wilcoxon test *more* confident about a
difference that is not there.

NEBULA is not exactly zero, and it should not be advertised as such. Eight
donors split four against four is a small experiment, real donors genuinely
differ, and an arbitrary split can land partway along some real axis of donor
variation. A handful of calls out of several hundred genes is the honest
expectation.

## Reading the rest of the output

The Wald test is the headline, but the fit carries the variance decomposition
that motivated the model.

```{r}
#| label: overdispersions
head(
  nebula_res$results[
    order(-subject_overdispersion),
    .(
      gene_id,
      subject_overdispersion,
      cell_overdispersion,
      sigma_at_bound,
      convergence
    )
  ],
  6
)
```

`subject_overdispersion` is how much the gene's expression moves between
donors, and it is exactly the term a cell-level test pretends is zero. Genes
with a large value are the ones a Wilcoxon test would most confidently get
wrong.

Two columns are diagnostics rather than results. `convergence` at or below
`-20` marks a fit that likely failed, and `sigma_at_bound` marks a gene where
the subject-level variance finished pinned on its lower bound, meaning the
mixed model collapsed to a plain negative binomial. Neither is fatal, but a
sweep where most genes show either is telling you something about the design.

```{r}
#| label: diagnostics
nebula_res$results[, .(
  n = .N,
  failed = sum(convergence <= -20L),
  collapsed = sum(sigma_at_bound)
)]
```

The fixed effects and their standard errors are on the object as matrices of
genes by coefficients, so the full model is available without refitting.

```{r}
#| label: coefficients
head(nebula_res$coefficients, 4)
```

## Which one to reach for

Use pseudobulk when the number of donors is decent and you want something
conservative, well understood and fast. It is a bulk method with decades of
scrutiny behind it, and `run_edger_ql()` gives you the edgeR quasi-likelihood
chain in Rust.

Use NEBULA when the cell-level information is worth keeping: few donors, uneven
cell counts, or when you care about the variance decomposition rather than only
the fold change.

Use the Wilcoxon test for marker detection, which is what it is good at. Asking
"which genes distinguish this cluster" is a description of the data in front of
you, not an inference about a population of donors, so pseudoreplication does
not apply. Asking "does this treatment change expression" is an inference, and
then it does.

## Cost

NEBULA fits an optimiser per gene and the fits are independent, so it
parallelises well and scales linearly in genes. On this data 500 genes over
5,000 cells takes a couple of seconds. Something with 50,000 cells and 5,000
genes will take minutes rather than seconds, which is fine for a considered
analysis and painful in a loop over twenty cell types. Restrict the genes.

`gene_batch_size` in
[`params_nebula()`](https://gregorlueg.github.io/bixverse/reference/params_nebula.html)
bounds how much of the store is resident at once. It changes nothing about the
answer, since NEBULA is gene-independent, so leave it alone unless memory is
tight.

## Where next

The differential *abundance* question is the other half of this, and it has its
own [vignette](https://gregorlueg.github.io/bixverse/articles/differential_abundance.html):
whether the treatment changed which cells are there, rather than what those
cells are expressing. The same donor structure argument applies, and Milo
handles it the same way.
