---
title: "Differential abundance with Milo and MELD"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Differential abundance with Milo and MELD}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Differential expression asks whether cells changed what they express.
Differential abundance asks whether the cells that are there changed at all.
Both matter, and the second is the one people usually answer badly.

The obvious approach is to cluster, count cells per cluster per sample, and
test the proportions. It works, and it inherits every problem the clustering
had. If a condition shifts a subpopulation that your resolution parameter
merged into a bigger cluster, you see nothing. If it shifts half a cluster, the
proportion barely moves. The answer depends on a choice you made before you
asked the question.

Two methods here avoid that, in different ways.

[Milo](https://doi.org/10.1038/s41587-021-01033-z) tests overlapping
neighbourhoods of the kNN graph instead of clusters. Each neighbourhood is a
cell and its neighbours, so the resolution is the graph's rather than the
clustering's, and a shift that spans part of a cluster shows up. Counts per
neighbourhood per sample go through the same negative binomial machinery bulk
RNA-seq uses.

[MELD](https://doi.org/10.1038/s41587-020-00803-5) does not test anything. It
smooths the condition labels over the graph and returns, per cell, how likely
that cell is under each condition. It is a continuous score rather than a set
of calls, which suits a graded response better than a discrete one.

They answer the same question and they should agree. At the end of this
vignette they do, which is a better check on both than either provides alone.

```{r}
#| label: setup
#| message: false
#| warning: false
library(bixverse)
library(data.table)
library(SingleCellExperiment)
```

## The data

[Baran-Gale, et al.](https://doi.org/10.1242/dev.183996) profiled mouse thymic
epithelial cells at one, four and sixteen weeks. The thymus involutes with age,
so the cell type proportions genuinely move, which is what makes this a
sensible differential abundance example.

::: {.callout-important}
## A stimulation experiment is a bad Milo example

It is tempting to reuse a treatment dataset here. Do not. In vitro stimulation
moves cells a long way in embedding space without moving the cell type
proportions much, so stimulated and control cells separate into different
neighbourhoods and almost every neighbourhood comes out significant. The result
looks spectacular and says nothing beyond "the treatment did something", which
you knew.

Differential abundance needs a design where the *composition* plausibly
changed.
:::

The object carries no gene names at all, with the identifiers sitting in the
`rowData` instead.
[`load_sce()`](https://gregorlueg.github.io/bixverse/reference/load_sce.html)
falls back to the first metadata column and tells you it did. 336 genes have no
annotation whatsoever and get a generated identifier, which it also tells you.
Both warnings are expected here.

```{r}
#| label: load-data
sce <- qs2::qs_read(download_thymus_ageing())

dir_thymus <- file.path(tempdir(), "thymus")
dir.create(dir_thymus, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = dir_thymus)
sc_object <- load_sce(
  object = sc_object,
  sce = sce,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 200L,
    min_lib_size = 500L,
    min_cells = 20L,
    target_size = 1e4
  ),
  .verbose = TRUE
)

sc_object
```

Now the design, and there is a trap in it. `samp_id` looks like the sample. It
is not: it is the sequencing run, and every run carries all three ages, because
the ages were hashed together and demultiplexed.

```{r}
#| label: design-trap
obs <- sc_object[[]]

table(obs$samp_id, obs$age)
```

The sample is the run and the age together, which gives eighteen. Using
`samp_id` as the sample would ask Milo whether abundance differs between
sequencing runs, and every run contains every age, so the answer would be a
flat no.

```{r}
#| label: design
sc_object[["sample"]] <- paste(obs$samp_id, obs$age, sep = "_")
obs <- sc_object[[]]

table(obs$sample, obs$age)[1:6, ]
```

## The embedding

Milo and MELD both run on the kNN graph, so the usual pipeline comes first.

```{r}
#| label: embedding
sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = FALSE)
```

The default `k` of 15 is fine for clustering and too small for Milo. A
neighbourhood holds `k + 1` cells, and those get spread over the samples. With
eighteen samples and `k = 15` that is under one cell per sample per
neighbourhood, and a negative binomial fitted to counts that are mostly zero or
one has very little to work with.

Rough rule: pick `k` so the average neighbourhood holds at least a few cells
per sample. Eighteen samples and `k = 60` gives around three, which is enough
here.

```{r}
#| label: knn
sc_object <- find_neighbours_sc(
  sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(k = 60L)
  ),
  .verbose = TRUE
)
```

## Milo

Sampling the neighbourhoods and counting the cells per sample is one call. The
counts come back as neighbourhoods by samples, with the graph-overlap weighting
and the k-th neighbour distances taken at the same time, since all three are
functions of the neighbourhood matrix alone.

```{r}
#| label: milo-nhoods
milo_obj <- get_miloR_abundances_sc(
  object = sc_object,
  sample_id_col = "sample",
  miloR_params = params_sc_miloR(prop = 0.1),
  .verbose = FALSE
)

dim(milo_obj$sample_counts)
mean(rowSums(milo_obj$sample_counts))
```

The test needs a design table with one row per sample, and its rownames have to
cover the sample names on the counts. Blocking on the run is worth it: every
run contributes all three ages, so run-to-run variation is a nuisance term the
model can remove.

```{r}
#| label: milo-design
samp_design <- unique(obs[, .(sample, samp_id, age)])

design_df <- data.frame(
  age = samp_design$age,
  run = samp_design$samp_id,
  row.names = samp_design$sample
)
design_df <- design_df[colnames(milo_obj$sample_counts), , drop = FALSE]

table(design_df$age, design_df$run)
```

::: {.callout-warning}
## Check which coefficient you are testing

`test_nhoods()` defaults to the last column of the design, the way edgeR and
limma do. With `age` as a factor, R orders the levels alphabetically:
`Wk1`, `Wk16`, `Wk4`. The last coefficient is therefore **Wk4 against Wk1**,
not the sixteen week contrast you probably wanted.

Name the coefficient rather than trusting the default.
:::

```{r}
#| label: milo-test
milo_obj <- test_nhoods(
  milo_obj,
  design = ~ run + age,
  design_df = design_df,
  coef = "ageWk16"
)

da_res <- get_differential_abundance_res(milo_obj)

da_res[SpatialFDR <= 0.1, .N]
```

```{r}
#| label: milo-top
head(da_res[order(PValue), .(Nhood, logFC, F, PValue, FDR, SpatialFDR)], 6)
```

A neighbourhood is not a cell type, so on its own a list of neighbourhood
indices is not much use. `add_nhoods_info()` tags each with the cell type its
cells mostly belong to.

```{r}
#| label: milo-annotate
milo_obj <- add_nhoods_info(milo_obj, cell_info = obs$cluster_annot)
da_res <- get_differential_abundance_res(milo_obj)

sig <- da_res[SpatialFDR <= 0.1]
sig[, direction := ifelse(logFC > 0, "up at Wk16", "down at Wk16")]

dcast(
  sig[, .N, by = .(majority_celltype, direction)],
  majority_celltype ~ direction,
  value.var = "N",
  fill = 0
)[order(-`up at Wk16`)]
```

The two proliferating populations are the most strongly depleted of anything
here, which is the clearest ageing signal the tissue has: an involuting thymus
stops making new cells. The intertypical TECs dominate the enriched side.

The medullary subsets split both ways, so "mTECs go up" would be the wrong
summary to take away. That is the point of testing neighbourhoods rather than
clusters: a cluster level test would have averaged those subsets into one
number and reported whichever direction happened to win.

### On the spatial FDR

Neighbourhoods overlap, so their tests are not independent and a plain
Benjamini-Hochberg adjustment is anti-conservative. Milo's spatial FDR weights
each p-value by the reciprocal of its connectivity and runs the step-up on
those weights.

```{r}
#| label: spatial-fdr
data.table(
  plain_fdr = da_res[FDR <= 0.1, .N],
  spatial_fdr = da_res[SpatialFDR <= 0.1, .N]
)
```

On this data the two barely differ. That is worth knowing rather than
disappointing: the correction bites when neighbourhoods overlap heavily and the
connectivity varies a lot between them, and here it does not. Do not assume it
is always doing work.

The weighting scheme is a choice. `"k-distance"` weights by the distance to the
k-th neighbour, `"graph-overlap"` by how many cells a neighbourhood shares with
the others. Both are precomputed, so switching is cheap.

## MELD

MELD needs no design and no neighbourhood sampling. Hand it the condition
column and it smooths the indicator over the graph.

```{r}
#| label: meld
meld_res <- meld_sc(
  object = sc_object,
  sample_id_col = "age",
  .verbose = FALSE
)

dim(meld_res$norm_scores)
head(round(meld_res$norm_scores, 3), 4)
```

The normalised scores are clamped at zero and L1 normalised per cell, so each
row is a likelihood over the three ages. A cell sitting at 0.6 for Wk16 lives
in a part of the manifold that is enriched for sixteen week old thymus.

Averaging per cell type turns that into something readable.

```{r}
#| label: meld-per-celltype
meld_dt <- data.table(
  cluster = obs$cluster_annot,
  as.data.table(meld_res$norm_scores)
)

per_celltype <- meld_dt[, lapply(.SD, mean), by = cluster]

head(per_celltype[order(-Wk16)], 8)
```

## Do they agree

They are different methods on the same graph, so this is a real check rather
than a formality. Milo's mean log fold change per cell type against MELD's mean
Wk16 likelihood per cell type.

```{r}
#| label: agreement
milo_per_ct <- sig[,
  .(milo_logfc = mean(logFC)),
  by = .(cluster = majority_celltype)
]
comparison <- merge(per_celltype, milo_per_ct, by = "cluster")

comparison[order(-milo_logfc), .(cluster, Wk1, Wk16, milo_logfc)]
```

```{r}
#| label: correlation
cor(comparison$Wk16, comparison$milo_logfc, method = "spearman")
```

Strong agreement, and the disagreements are informative rather than
embarrassing. Milo tests neighbourhoods and reports only the ones that clear a
threshold, so a cell type that shifted a little everywhere can score low on the
Milo axis while MELD still registers the shift. The two are measuring related
but not identical things.

## Which one to reach for

Milo when you want calls with error control: a list of regions that changed,
with p-values you can defend and a multiple testing correction that accounts
for the overlap. It needs a real design, enough samples to fit a negative
binomial, and a `k` large enough that the counts are not mostly zero.

MELD when the response is graded, when you want a per-cell score to carry into
downstream analysis, or when the design is too thin for a test. It gives you no
p-values, which is a feature when you have four samples and would only have
been fooling yourself.

Running both costs almost nothing once the graph is built, and their agreement
is the most useful diagnostic either one has.

## Where next

The other half of this is differential *expression* with the same donor
structure, which has its own
[vignette](https://gregorlueg.github.io/bixverse/articles/differential_expression.html).
Neighbourhood level expression testing, rather than abundance, is not wired up
yet.
