---
title: "Doublet detection with bixverse"
vignette: >
  %\VignetteIndexEntry{Doublet detection with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Gregor Lueg"
---

## Intro

Doublets - droplets containing two or more cells - are a common artefact in
droplet-based single cell experiments. If left undetected, they masquerade as
intermediate cell states or novel populations and quietly corrupt downstream
analyses. This vignette demonstrates three doublet detection approaches 
available in `bixverse` and benchmarks them against ground truth calls from
[demuxlet](https://github.com/statgen/demuxlet), which identifies doublets
via genetic variation across pooled donors.

`bixverse` provides three computational doublet detection methods:

- **Scrublet** is a simulation-based approach
([Wolock, et al., 2019](https://doi.org/10.1016/j.cels.2018.11.005)). It
generates synthetic doublets by averaging pairs of randomly selected cells,
embeds both real and synthetic cells together, and scores each real cell by the
proportion of its nearest neighbours that are synthetic. A bimodal score
distribution is expected: singlets cluster near zero, doublets near one. A
threshold on this score determines the final calls.

- **Boosted doublet detection** takes a different angle. Rather than simulating
doublets, it trains a gradient-boosted tree ensemble to distinguish real cells
from synthetic doublets generated in the same way as Scrublet. The classifier
produces a probability per cell, and doublet calls are derived from that
probability. This can be more robust when the score distribution from Scrublet
is not cleanly bimodal, though it introduces its own hyperparameters (and is
substantially slower).

- **scDblFinder** is a port of the
[scDblFinder](https://bioconductor.org/packages/release/bioc/html/scDblFinder.html)
method from [Germain et al.](https://pmc.ncbi.nlm.nih.gov/articles/PMC9204188/),
with one modification: it uses lightGBM with the histogram trick on the feature
space instead of xgboost, which tends to be faster to fit. Like the boosted
approach it trains a classifier on real versus synthetic doublets, but it
combines the classifier output with cxds co-expression scores
([Bais and Kostka](https://pmc.ncbi.nlm.nih.gov/articles/PMC7703774/)) into a
weighted final score. Across published benchmarks it tends to be the strongest
of the three.

All three methods are purely computational and work without genetic or hashing
information. They complement experimental approaches like demuxlet, which
require pooled donors or hashtag antibodies. They also have a failure mode
if a large number of doublets dominates the original data.

```{r}
#| label: setup
library(bixverse)
library(data.table)
```

## Loading the data

We use a PBMC data set with demuxlet ground truth calls. The demuxlet output
classifies each barcode as a singlet (`SNG`) or doublet (`DBL`) (or ambiguous,
but irrelevant for this situation), giving us something to benchmark against.

```{r}
#| label: load data
doublet_path <- download_demuxlet_pbmc()

tempdir_doublet <- file.path(tempdir(), "demuxlet_bixverse")
dir.create(tempdir_doublet, showWarnings = FALSE, recursive = TRUE)

demuxlet_data <- fread(file.path(doublet_path, "demuxlet_calls.tsv"))

table(demuxlet_data$Call)
```

Load in the data into bixverse

```{r}
#| label: load into bixverse
sc_object <- SingleCells(dir_data = tempdir_doublet)

mtx_io_params <- get_cell_ranger_params(doublet_path)
mtx_io_params$cells_as_rows <- TRUE

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  streaming = FALSE
)

sc_object
```

We also define a small helper to compute precision, recall and F1 against the
demuxlet calls.

```{r}
#| label: metrics helper
doublet_metrics <- function(
  predicted,
  actual,
  pos_predicted = TRUE,
  pos_actual = "DBL"
) {
  tp <- sum(predicted == pos_predicted & actual == pos_actual)
  fp <- sum(predicted == pos_predicted & actual != pos_actual)
  fn <- sum(predicted != pos_predicted & actual == pos_actual)
  precision <- tp / (tp + fp)
  recall <- tp / (tp + fn)
  f1 <- 2 * (precision * recall) / (precision + recall)
  list(precision = precision, recall = recall, f1 = f1)
}
```

## Doublet detection methods

Let's check the two methods out and see how they behave

### Scrublet

```{r}
#| label: run scrublet
scrublet <- scrublet_sc(
  object = sc_object,
  scrublet_params = params_scrublet(expected_doublet_rate = 0.12)
)

scrublet
```

The `plot` method shows the score distribution for observed and simulated cells.
A clean separation between the two modes indicates the threshold is sensible.

```{r}
#| label: scrublet plot
#| fig-width: 7
#| fig-height: 4
plot(scrublet)
```

How does the automatic thresholding perform against the demuxlet ground truth?

```{r}
#| label: scrublet metrics
scrublet_dt <- get_obs_data(scrublet)
scrublet_dt[, Barcode := get_cell_names(sc_object)]
scrublet_dt <- merge(scrublet_dt, demuxlet_data, by = "Barcode")

doublet_metrics(predicted = scrublet_dt$doublet, actual = scrublet_dt$Call)
```

#### Manual threshold adjustment

The automatic threshold does not always land in the right place. If the score
distribution suggests a different cut-off, `call_doublets_manual` lets you
override it.

```{r}
#| label: scrublet manual threshold
#| fig-width: 7
#| fig-height: 4
scrublet_adj <- call_doublets_manual(scrublet, threshold = 0.25)

plot(scrublet_adj)
```

```{r}
#| label: scrublet adjusted metrics
scrublet_adj_dt <- get_obs_data(scrublet_adj)
scrublet_adj_dt[, Barcode := get_cell_names(sc_object)]
scrublet_adj_dt <- merge(scrublet_adj_dt, demuxlet_data, by = "Barcode")

doublet_metrics(
  predicted = scrublet_adj_dt$doublet,
  actual = scrublet_adj_dt$Call
)
```

We have made it worse here, but maybe there are cases where the scores are not
very good. Compared to the original version, Otsu's method is used to identify
the perfect threshold, but it might fail in some situations.

### Boosted doublet detection

The boosted approach trains a classifier to separate real cells from synthetic
doublets. This can work better when the Scrublet score distribution is messy or
unimodal. However, in this case we are running 25 iterations of the algorithm,
which means 20x generation of synthetic doublets. The initial PCA will be stored
in the object for subsequent reprojection - nonetheless, this makes this usually 
slower than the other methods in this vignette.

```{r}
#| label: run boosted
boosted_doublets <- doublet_detection_boost_sc(
  object = sc_object,
  boost_params = params_boost()
)
```

Let's check how boosted works?

```{r}
#| label: boosted metrics
boosted_dt <- get_obs_data(boosted_doublets)
boosted_dt[, Barcode := get_cell_names(sc_object)]
boosted_dt <- merge(boosted_dt, demuxlet_data, by = "Barcode")

doublet_metrics(predicted = boosted_dt$doublet, actual = boosted_dt$Call)
```

Overall, it takes longer and has worse performance (in this data set). If you
wish to have a partially orthogonal method, it can be worthwhile nonetheless.
There is also an option to accelerate the algorithm via fast clustering. This
will run k-means clustering with `sqrt(n_cells) * 4` centroids by default, run 
the kNN on these subset of centroids, Louvain on the smaller kNN and project 
back the Louvain membership for final scoring. However, this comes at cost of 
Recall, as you can see below. Generally speaking, you need to see how many
centroids you need to represent your data properly. On smaller data sets, you
can get away with `n_cells / 10` and compress 10 cells into one centroid... On
larger data sets, you might want to compress further.

```{r}
#| label: run boosted (fast)
boosted_doublets_fast <- doublet_detection_boost_sc(
  object = sc_object,
  boost_params = params_boost(
    fast_cluster = TRUE,
    # in this case, we set the n_centroids to N_cells / 10
    # you want this to not be too low; also the data is so small, we just
    # use standard k-means over mini batch
    fast_cluster_params = list(n_centroids = 1500L, km_type = "standard")
  )
)
```

```{r}
#| label: boosted metrics (fast)
boosted_fast_dt <- get_obs_data(boosted_doublets_fast)
boosted_fast_dt[, Barcode := get_cell_names(sc_object)]
boosted_fast_dt <- merge(boosted_fast_dt, demuxlet_data, by = "Barcode")

doublet_metrics(
  predicted = boosted_fast_dt$doublet,
  actual = boosted_fast_dt$Call
)
```

### scDblFinder

Lastly, `bixverse` also provides an implementation of the 
[scDblFinder](https://bioconductor.org/packages/release/bioc/html/scDblFinder.html) 
from [Germain et al.](https://pmc.ncbi.nlm.nih.gov/articles/PMC9204188/) with 
some modifications: instead of xgboost, we will be using an implementation of
lightGBM using the histogram trick on the feature space. This usually yields
faster fitting compared to the original xgboost approach. To run it, you can 
just do the following:

```{r}
#| label: run scrublet
scdblfinder <- scdblfinder_sc(
  object = sc_object,
  scdblfinder_params = params_scdblfinder(
    expected_doublet_rate = 0.12
  ),
  # if you wish to get the features that were used to train the classifier
  return_features = FALSE
)
```

The scores are similar to the other methods.

```{r}
#| label: scdblfinder metrics
scdblfinder_dt <- get_obs_data(scdblfinder)

scdblfinder_dt[, Barcode := get_cell_names(sc_object)]
scdblfinder_dt <- merge(scdblfinder_dt, demuxlet_data, by = "Barcode")

doublet_metrics(
  predicted = scdblfinder_dt$predicted_doublets,
  actual = scdblfinder_dt$Call
)
```

We can also extract the other scores from scDblFinder. We can for example
extract the weighted score, identify the threshold via Otsu's method and check
the doublets like this.

```{r}
#| label: weighted scores from scdblfinder
weighted_scores <- get_scores(scdblfinder)

weighted_scores_threshold <- find_threshold_otsu(x = weighted_scores)

weighted_doublets <- weighted_scores >= weighted_scores_threshold

table(
  lightgbm = scdblfinder_dt$predicted_doublets,
  weighted = weighted_doublets
)
```

Or alternatively, the cxds scores can also be extracted and used. These scores
were originally developed by 
[Bais and Kostka](https://pmc.ncbi.nlm.nih.gov/articles/PMC7703774/) and are
used as an input into the classifier.

```{r}
#| label: cxds scores from scdblfinder
cxds_scores <- get_scores(scdblfinder, score_type = "cxds_scores")

cxds_scores_threshold <- find_threshold_otsu(x = cxds_scores)

cxds_doublets <- cxds_scores >= cxds_scores_threshold

table(
  lightgbm = scdblfinder_dt$predicted_doublets,
  cxds = cxds_doublets
)
```

Should you observe that this is very flat or you only have a few data points
(this can occur in low complexity experiments in which the highest expressing
genes are VERY similar across all the cells for example in cell culture
experiments), you can bump `cxds_genes` up. The default here is 500L, but you
might want to increase this pending your experiment.

### Conclusion

All methods can help to identify in a data-driven way doublets in your data.
In this example, Scrublet performed better compared to the other two methods;
in a comparison paper scDblFinder is the best across different benchmark data 
sets, see [here](https://www.cell.com/cell-systems/fulltext/S2405-4712(20)30459-2).

### Future outlook

While Rust makes these methods blazingly fast (usually faster than the 
corresponding R or Python versions), they start feeling quite slow on large
data sets. A potential approach via GPU acceleration, see 
[bixverse.gpu](https://github.com/GregorLueg/bixverse.gpu) could be an idea 
here to improve the speed

## Clean up

```{r}
#| label: clean up
unlink(tempdir_doublet, recursive = TRUE, force = TRUE)
```