---
title: "Gene Set Enrichment Methods"
vignette: >
  %\VignetteIndexEntry{gse_methods}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Gene Set Enrichment Methods in bixverse

This vignette shows you how to use the different implement gene set enrichment
methods in `bixverse`.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(magrittr)
```

### Hypergeometric tests

`bixverse` has implementations of the standard hypergeometric tests to identify
gene set enrichment for a given set of target genes. In this example, we will
be loading the Hallmark gene sets from `msigdbr`. Let's first load these in and
create a list of gene sets (the expected format of most functions in `bixverse`
for these type of analysis.)

```{r}
#| label: loading data from msigdbr
#| eval: !expr requireNamespace("msigdbr", quietly = TRUE)
h_gene_sets <- msigdbr::msigdbr(species = "human", collection = "H")

h_gene_sets_ls <- split(h_gene_sets$ensembl_gene, h_gene_sets$gs_name)

set.seed(123L)

target_genes_1 <- c(
  sample(h_gene_sets_ls[["HALLMARK_MYC_TARGETS_V1"]], 25),
  sample(h_gene_sets_ls[["HALLMARK_MYC_TARGETS_V2"]], 25)
)
```

To run the hypergeometric test against one target set, you can just use 
`gse_hypergeometric`. The function has parameters to define a minimum FDR 
threshold and minimum overlap science. Additionally, the gene universe will be 
set (if not provided) to all of the genes represented in the `gene_set_list`. 
You can provide your own gene universe if you wish to.

```{r}
#| label: run hypergeom test single
results <- gse_hypergeometric(
  target_genes = target_genes_1,
  gene_set_list = h_gene_sets_ls
)

head(results)
```

To run the hypergeometric test over a list of target gene sets, `bixverse`
provides `gse_hypergeometric_list` that leverages 
[rayon](https://github.com/rayon-rs/rayon) in Rust for multi-threading. This 
makes this function very fast even on large gene set libraries with lots of 
target gene sets.

```{r}
#| label: run hypergeom test multiple
target_genes_2 <- c(
  sample(h_gene_sets_ls[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]], 20),
  sample(h_gene_sets_ls[["HALLMARK_IL6_JAK_STAT3_SIGNALING"]], 25)
)

target_list <- list(
  set_1 = target_genes_1,
  set_2 = target_genes_2
)

results_multiple <- gse_hypergeometric_list(
  target_genes_list = target_list,
  gene_set_list = h_gene_sets_ls
)

head(results_multiple)
```

To exemplify the speed in Rust, in this case we run for 250 target gene sets
over-enrichment tests against 5000 gene sets sampled from 20k genes. This totals
1.25m hypergeometric tests. Pending on your system, you can expect this to be
done in a few seconds.

```{r}
#| label: rust speed hypergeom multiple
#| eval: !expr requireNamespace("tictoc", quietly = TRUE)

seed = 10101L

set.seed(seed)

# Define the parameters
universe <- sprintf("gene_%i", 1:20000)
gene_sets_no <- 5000
target_gene_sets_no <- 250

# Generate random gene sets
gene_sets <- purrr::map(
  1:gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    size <- sample(20:100, 1)
    sample(universe, size, replace = FALSE)
  }
)

names(gene_sets) <- purrr::map_chr(
  1:gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    paste(sample(LETTERS, 3), collapse = "")
  }
)

# Generate random target sets
target_gene_sets <- purrr::map(
  1:target_gene_sets_no,
  ~ {
    set.seed(.x * seed)
    size <- sample(50:100, 1)
    sample(universe, size, replace = FALSE)
  }
)

names(target_gene_sets) <- purrr::map_chr(
  1:target_gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    paste(sample(letters, 3), collapse = "")
  }
)

# Run the Rust function
tictoc::tic()
rs_results_example <- gse_hypergeometric_list(
  target_genes_list = target_gene_sets,
  gene_set_list = gene_sets
)
tictoc::toc()
```

### Gene ontology aware enrichment tests (for sets)

Additionally, `bixverse` provides Rust-accelerated functions to run an ontology-
aware enrichment test for the gene ontology in particular - the concept is based
on [Adrian et al.](https://academic.oup.com/bioinformatics/article/22/13/1600/193669).
Briefly, the function will start in the leafs of the ontology (lowest level)
and calculate the enrichment statistic for that level. If an enrichment test
reaches a threshold, the genes from that term are removed from all of its 
ancestor terms, ensuring that the more specific terms have stronger test
statistics. To make it easy for the users, we provide the human gene ontology as
data in the package. This is a data.table with 5 columns:

- go_id: the gene ontology term identifier.
- go_name: the name gene ontology term name.
- ensembl_id: the ensembl identifiers belonging to this term.
- ancestors: in the ontology DAG all identified ancestors of this specific
term
- depth: the depth of that term (identified via BFS) for this term. Higher
values indicate further distance from the top terms.

These columns are expected by the `gene_ontology_data` class that stores the
data and exposes it in Rust.

```{r}
#| label: load human go data

# Load in the package provided data
go_data_dt <- get_go_human_data()

# Generate the S7 object for subsequent usage
go_data <- gene_ontology_data(go_data_dt, min_genes = 3L)
```

Similar to the more generic hypergeometric tests, you can provide either a 
single set of a target genes or a list of target genes. Leveraging Rusts
borrowing, the overhead for multiple target gene sets is reduced and hence
very fast.
```{r}
#| label: run single go-aware gse

go_aware_res <- gse_go_elim_method(
  object = go_data,
  target_genes = target_genes_1
)

head(go_aware_res)
```

Here is the example on how to run this over a list.

```{r}
#| label: run multiple go-aware gse

go_aware_res_2 <- gse_go_elim_method_list(
  object = go_data,
  target_gene_list = target_list
)

head(go_aware_res_2)
```

### GSEA

`bixverse` also has implementations of the GSEA based on
[Subramanian et al](https://www.pnas.org/doi/10.1073/pnas.0506580102) to test
against continuous scores. The package provides also the fgsea implementations
of [Korotkevich et al.](https://www.biorxiv.org/content/10.1101/060012v3).
These can be run like:

```{r}
#| label: fgsea run
#| eval: !expr requireNamespace("fgsea", quietly = TRUE)

# This is the example of fgsea

library("fgsea")

data(examplePathways)
data(exampleRanks)

set.seed(42L)

fgsea_res <- fgsea(
  pathways = examplePathways, 
  stats = exampleRanks,
  minSize = 15,
  maxSize = 500
) %>% setorder(pathway)

head(fgsea_res)
```

And this is how you run it via `bixverse`.

```{r}
#| label: bixverse version of fgsea
bixverse_fgsea <- calc_fgsea(
  stats = exampleRanks,
  pathways = examplePathways,
  gsea_params = params_gsea(min_size = 15L)
) %>% setorder(pathway_name)

head(bixverse_fgsea)
```

We can appreciate the (more or less) same p-values estimated by both methods.

```{r}
#| label: p-value comparison between fgsea and bixverse fgsea
plot(
  x = -log10(fgsea_res$pval),
  y = -log10(bixverse_fgsea$pvals),
  xlab = "-log10(pval) fgsea",
  ylab = "-log10(pval) bixverse",
  main = "fgsea and bixverse"
)
```

The speed is very comparable with the Rcpp implementation of fgsea, indicating
no issues in terms of performance.

```{r}
#| label: comparison of speed fgsea vs. bixverse version
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  fgsea = fgsea(
    pathways = examplePathways, 
    stats = exampleRanks,
    minSize = 15,
    maxSize = 500
  ),
  rust = calc_fgsea(
    stats = exampleRanks,
    pathways = examplePathways,
    gsea_params = params_gsea(min_size = 15L)
  ),
  times = 10L
)
```

### GSEA gene ontology aware

`bixverse` also implements GSEA in a gene-ontology aware form. In this case,
you use the same object as for the hypergeometric test, but instead provide
the ranked list. Similar to the hyper geometric version, the ontology is
traversed from the lowest level first and the permutation-based p-values are
calculated for a given term. Should this p-value be below the elimination
threshold, the genes from that given gene ontology term are removed from all
of its ancestors, yielding more specific GO terms. Once this has been done
with the full ontology, the multi-level method from fgsea is used to estimate
lower p-values than just possible via permutation for terms that show 
significance.

```{r}
#| label: gsea gene ontology aware

gene_universe_go <- unique(
  unlist(go_data_dt[, "ensembl_id"], use.names = FALSE)
)

random_stats <- rnorm(length(gene_universe_go))
names(random_stats) <- gene_universe_go

go_gsea_res <- fgsea_go_elim(
  object = go_data,
  stats = random_stats
)

head(go_gsea_res)
```

