---
title: "Gene Set Enrichment Methods"
vignette: >
  %\VignetteIndexEntry{gse_methods}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Gene Set Enrichment Methods in bixverse

This vignette shows you how to use the gene set enrichment methods implemented
in `bixverse`.

```{r}
#| label: setup
if (!requireNamespace("fgsea", quietly = TRUE)) {
  BiocManager::install("fgsea")
}
if (!requireNamespace("msigdbr", quietly = TRUE)) {
  install.packages("msigdbr")
}
library(bixverse)
library(data.table)
library(magrittr)
```

## Intro

Gene set enrichment analysis sits at the heart of most functional genomics
workflows. Given a set of genes of interest (or a ranked list of all measured
genes) you want to know which biological processes, pathways, or molecular
functions are over-represented relative to what you would expect by chance.
`bixverse` provides Rust-accelerated implementations of the two dominant
paradigms for this, along with gene ontology-aware variants of each.

**Over-representation analysis (ORA)** via the hypergeometric test asks: given
that I have drawn $k$ genes from a universe of $N$, and $K$ of those belong to
a particular gene set, is the overlap of $k \cap K$ larger than expected? This
is fast and interpretable, but requires a discrete gene list — typically genes
passing some fold-change and significance threshold. The choice of threshold
and universe can substantially affect results, which is a known limitation of
the approach.

**Gene Set Enrichment Analysis (GSEA)** sidesteps this by working with a
continuously ranked list of all measured genes, asking whether the members of a
gene set are systematically concentrated towards the top or bottom of the
ranking. This avoids arbitrary thresholding and tends to be more sensitive to
coordinated but modest shifts in expression, at the cost of being more
computationally demanding and less straightforward to interpret.

A practical complication with both approaches is the **redundancy of the Gene
Ontology (GO)**. The GO is a directed acyclic graph: a significant hit on a
specific term (e.g., "mitochondrial complex I assembly") will typically also
produce significant hits on all of its ancestor terms (e.g., "oxidative
phosphorylation", "metabolic process"), most of which add no new information.
`bixverse` provides two strategies for dealing with this: the **elimination
method** of [Adrian et al.](https://academic.oup.com/bioinformatics/article/22/13/1600/193669),
which propagates signal up the DAG while removing explained genes, and a
**post-hoc simplification** approach based on Wang semantic similarity, which
collapses redundant terms after running a standard enrichment.

All methods are implemented in Rust and exposed via `extendr`. The performance
gains relative to existing R implementations range from modest (~20% faster for
GSEA relative to `fgsea`'s Rcpp implementation) to substantial: the
parallelised hypergeometric test over large gene set libraries can be orders of
magnitude faster than a naive R loop.

## Hypergeometric tests

`bixverse` implements the standard hypergeometric over-representation test. In
this example we load the Hallmark gene sets from `msigdbr` and construct a
named list, which is the expected input format throughout `bixverse` for gene
set analyses.

```{r}
#| label: loading data from msigdbr
#| eval: !expr requireNamespace("msigdbr", quietly = TRUE)
h_gene_sets <- msigdbr::msigdbr(species = "human", collection = "H")

h_gene_sets_ls <- split(h_gene_sets$ensembl_gene, h_gene_sets$gs_name)

set.seed(123L)

target_genes_1 <- c(
  sample(h_gene_sets_ls[["HALLMARK_MYC_TARGETS_V1"]], 25),
  sample(h_gene_sets_ls[["HALLMARK_MYC_TARGETS_V2"]], 25)
)
```

To test a single target gene set, use `gse_hypergeometric`. The gene universe
defaults to the union of all genes represented in the `gene_set_list` if not
provided explicitly. This is usually reasonable but worth checking if your
assay covers a non-standard set of genes. FDR threshold and minimum overlap are
also configurable.

```{r}
#| label: run hypergeom test single
results <- gse_hypergeometric(
  target_genes = target_genes_1,
  gene_set_list = h_gene_sets_ls
)

head(results)
```

When you have many target gene sets to test simultaneously, use
`gse_hypergeometric_list`. Under the hood this runs all tests in parallel via
[rayon](https://github.com/rayon-rs/rayon) and optimiser 
[FxHashSets](https://docs.rs/rustc-hash/latest/rustc_hash/index.html), making it 
very fast even at scale.

```{r}
#| label: run hypergeom test multiple
target_genes_2 <- c(
  sample(h_gene_sets_ls[["HALLMARK_TNFA_SIGNALING_VIA_NFKB"]], 20),
  sample(h_gene_sets_ls[["HALLMARK_IL6_JAK_STAT3_SIGNALING"]], 25)
)

target_list <- list(
  set_1 = target_genes_1,
  set_2 = target_genes_2,
  set_3 = c("random gene 1", "random gene 2", "random gene 3")
)

results_multiple <- gse_hypergeometric_list(
  target_genes_list = target_list,
  gene_set_list = h_gene_sets_ls
)

head(results_multiple)
```

To give a sense of the scale this handles comfortably, the following runs 250
target gene sets against 5,000 gene sets drawn from a universe of 20,000 genes
— 1.25 million hypergeometric tests in total. On most systems this completes
in a few seconds.

```{r}
#| label: rust speed hypergeom multiple
#| eval: !expr requireNamespace("tictoc", quietly = TRUE)

seed = 10101L

set.seed(seed)

universe <- sprintf("gene_%i", 1:20000)
gene_sets_no <- 5000
target_gene_sets_no <- 250

gene_sets <- purrr::map(
  1:gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    size <- sample(20:100, 1)
    sample(universe, size, replace = FALSE)
  }
)

names(gene_sets) <- purrr::map_chr(
  1:gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    paste(sample(LETTERS, 3), collapse = "")
  }
)

target_gene_sets <- purrr::map(
  1:target_gene_sets_no,
  ~ {
    set.seed(.x * seed)
    size <- sample(50:100, 1)
    sample(universe, size, replace = FALSE)
  }
)

names(target_gene_sets) <- purrr::map_chr(
  1:target_gene_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    paste(sample(letters, 3), collapse = "")
  }
)

tictoc::tic()
rs_results_example <- gse_hypergeometric_list(
  target_genes_list = target_gene_sets,
  gene_set_list = gene_sets
)
tictoc::toc()
```

## Gene Ontology-aware enrichment: the elimination method

The naive approach of running a hypergeometric test independently for each GO
term produces heavily redundant results, because significant child terms
mechanically inflate their ancestors. The elimination method from
[Adrian et al.](https://academic.oup.com/bioinformatics/article/22/13/1600/193669)
addresses this by traversing the ontology from the most specific terms (leaves)
upwards. When a term reaches the significance threshold, the genes annotated to
that term are removed from all of its ancestor terms before their tests are
computed. This means the ancestor terms are tested on the residual signal not
already explained by their more specific descendants, yielding results that are
both less redundant and more interpretable.

`bixverse` ships with the human GO data and exposes it via a dedicated S7 class
that transfers the ontology structure into Rust for efficient traversal.

```{r}
#| label: load human go data

go_data_dt <- get_go_data_human()

go_data_s7 <- GeneOntologyElim(go_data_dt, min_genes = 3L)
```

The interface mirrors the generic hypergeometric functions: a single target set
or a list, with Rust's ownership model ensuring the ontology structure is shared
across all tests without copying, keeping memory overhead low.

```{r}
#| label: run single go-aware gse

go_aware_res <- gse_go_elim_method(
  object = go_data_s7,
  target_genes = target_genes_1
)

head(go_aware_res)
```

And the version for lists:

```{r}
#| label: run multiple go-aware gse

go_aware_res_2 <- gse_go_elim_method_list(
  object = go_data_s7,
  target_gene_list = target_list
)

head(go_aware_res_2)
```

The DAG traversal is inherently sequential per gene set (each level depends on
the one below), but tests within a level are parallelised via rayon. Running 100
target gene sets against the full human GO completes in a matter of seconds:
```{r}
#| label: speed go-aware gse
#| eval: !expr requireNamespace("tictoc", quietly = TRUE)

go_gene_universe <- unique(unlist(go_data_dt$ensembl_id))

go_target_sets_no <- 100L
seed <- 246L

go_target_gene_sets <- purrr::map(
  1:go_target_sets_no,
  ~ {
    set.seed(.x * seed)
    size <- sample(50:100, 1)
    sample(go_gene_universe, size, replace = FALSE)
  }
)

names(go_target_gene_sets) <- purrr::map_chr(
  1:go_target_sets_no,
  ~ {
    set.seed(seed + .x + 1)
    paste(sample(letters, 3), collapse = "")
  }
)

tictoc::tic()
rs_results_example <- gse_go_elim_method_list(
  object = go_data_s7,
  target_gene_list = go_target_gene_sets
)
tictoc::toc()
```

## Alternative: post-hoc simplification of GO results

An alternative strategy is to run an unconstrained hypergeometric test over all
GO terms and then collapse redundant results afterwards using semantic
similarity. This approach is more flexible — you can apply it to any enrichment
results, not just those produced by `bixverse` — but it does not adjust the
test statistics themselves the way the elimination method does. Whether you
prefer pre-hoc elimination or post-hoc simplification depends on your use case;
for exploratory analyses the simplification approach can be more convenient,
while for rigorous GO reporting the elimination method is generally preferable.

```{r}
#| label: prepare go data for simplification approach

go_data <- load_go_human_data()

min_genes <- 3L

go_genes <- go_data$go_to_genes
go_genes_ls <- split(go_genes$ensembl_id, go_genes$go_id)
go_genes_ls <- purrr::keep(go_genes_ls, \(x) length(x) > min_genes)

go_results_unfiltered <- gse_hypergeometric(
  target_genes = target_genes_1,
  gene_set_list = go_genes_ls
)

head(go_results_unfiltered)
```

The simplification uses Wang semantic similarity to group related terms, then
retains the term with the best test statistic within each cluster. Ties on FDR
are broken by ontology depth, preferring the more specific term.

```{r}
#| label: run simplification with the gene ontology info

go_parent_child_dt <- go_data$gene_ontology[
  relationship %in% c("is_a", "part_of")
] %>%
  setnames(
    old = c("from", "to", "relationship"),
    new = c("parent", "child", "type")
  )

go_results_simplified <- simplify_hypergeom_res(
  res = go_results_unfiltered,
  parent_child_dt = go_parent_child_dt,
  weights = setNames(c(0.8, 0.6), c("is_a", "part_of"))
)

head(go_results_simplified)
```

## GSEA

`bixverse` implements GSEA following
[Subramanian et al.](https://www.pnas.org/doi/10.1073/pnas.0506580102) for
testing against a continuously ranked gene list. The package also wraps the
`fgsea` multilevel method from
[Korotkevich et al.](https://www.biorxiv.org/content/10.1101/060012v3), which
uses an adaptive permutation scheme to estimate precise p-values efficiently.
Both are available for comparison.
```{r}
#| label: fgsea run
#| eval: !expr requireNamespace("fgsea", quietly = TRUE)

library("fgsea")

data(examplePathways)
data(exampleRanks)

set.seed(42L)

fgsea_res <- fgsea(
  pathways = examplePathways,
  stats = exampleRanks,
  minSize = 15,
  maxSize = 500
) %>%
  setorder(pathway)

head(fgsea_res)
```

How does bixverse look in comparison?

```{r}
#| label: bixverse version of fgsea
bixverse_fgsea <- calc_fgsea(
  stats = exampleRanks,
  pathways = examplePathways,
  gsea_params = params_gsea(min_size = 15L)
) %>%
  setorder(pathway_name)

head(bixverse_fgsea)
```

The p-values from both implementations are in close agreement, as expected:

```{r}
#| label: p-value comparison between fgsea and bixverse fgsea
plot(
  x = -log10(fgsea_res$pval),
  y = -log10(bixverse_fgsea$pvals),
  xlab = "-log10(pval) fgsea",
  ylab = "-log10(pval) bixverse",
  main = "fgsea and bixverse"
)
```

The Rust implementation is consistently faster than `fgsea`'s Rcpp backend —
not by an enormous margin on small data (roughly 20% in typical benchmarks),
but the gap widens with the number of pathways and permutations:

```{r}
#| label: comparison of speed fgsea vs. bixverse version
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  fgsea = fgsea(
    pathways = examplePathways,
    stats = exampleRanks,
    minSize = 15,
    maxSize = 500
  ),
  rust = calc_fgsea(
    stats = exampleRanks,
    pathways = examplePathways,
    gsea_params = params_gsea(min_size = 15L)
  ),
  times = 5L
)
```

## GO-aware GSEA: the elimination method

`bixverse` combines the elimination method with GSEA, using the same
`GeneOntologyElim` object as the hypergeometric variant. The ontology is
traversed leaf-first; for each term a permutation-based p-value is computed and
if it falls below the elimination threshold, genes from that term are removed
from all ancestors before they are tested. Once the full traversal is complete,
the fgsea multilevel method is applied to terms that reached nominal
significance to obtain more precise p-values than pure permutation alone can
provide.

```{r}
#| label: gsea gene ontology aware

gene_universe_go <- unique(
  unlist(go_data_dt[, "ensembl_id"], use.names = FALSE)
)

set.seed(42L)

random_stats <- rnorm(length(gene_universe_go))
names(random_stats) <- gene_universe_go

go_gsea_res <- fgsea_go_elim(
  object = go_data_s7,
  stats = random_stats
)

head(go_gsea_res)
```