---
title: "Meta cells with bixverse"
vignette: >
  %\VignetteIndexEntry{Meta cells with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Gregor Lueg"
---

## Intro

Meta cells are small groups of transcriptionally similar cells aggregated into
a single representative profile. The motivation is twofold: firstly, single cell 
counts are sparse and noisy, and many downstream methods (correlation networks, 
GRN inference, archetype-based modelling) behave poorly on raw single cell data.
Aggregating into meta cells reduces sparsity while preserving the heterogeneity
that bulk pseudo-replicates would average out. Secondly, they reduce the
computational burden substantially.

`bixverse` provides three meta cell algorithms with different underlying ideas:

- **hdWGCNA-style meta cells** ([Morabito, et al., 2023](https://doi.org/10.1016/j.crmeth.2023.100498))
iterate over a kNN graph, picking seed cells and aggregating their neighbours,
with a constraint on how many cells two meta cells are allowed to share. Simple,
fast and reasonably effective. Works directly on a kNN graph.

- **SEACells** ([Persad, et al., 2023](https://doi.org/10.1038/s41587-023-01716-9))
uses kernel archetypal analysis. The algorithm finds a set of archetypes (the
SEACells) such that every cell can be expressed as a convex combination of
nearby archetypes. Tends to produce purer aggregations but is the most
computationally expensive of the three.

- **SuperCells** ([Bilous, et al., 2022](https://doi.org/10.1186/s12859-022-04861-1))
applies the walktrap community detection algorithm to the kNN graph and treats
each community as a meta cell. Granularity is controlled indirectly through a
graining factor (the average number of cells per meta cell).

One implementation detail worth flagging upfront. In `bixverse`, the
`SingleCells` class keeps counts on disk in DuckDB and binary Rust-based files 
and streams them in as needed. The `MetaCells` class does not. After aggregation 
the count matrix is small enough to hold entirely in memory as a sparse matrix, 
and all downstream operations (HVG, PCA, neighbours, clustering, embeddings) run
against this in-memory representation (most of the code still makes usage of
Rust's incredible performance nonetheless). Practically that means the meta cell 
pipeline is substantially faster than the equivalent operations on the parent
`SingleCells` object for two reasons: data is already in memory and more
importantly, it is simply less data.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
```

## Preparing the single cell data

We use the same CD34 cells from the [SEACells vignette](https://github.com/dpeerlab/SEACells/blob/main/notebooks/SEACell_computation.ipynb) to set up the parent `SingleCells` object: 
load (QC not needed - it's already filtered data), HVG selection, PCA, 
kNN graph. We can use the provided cell type labels to check purity.

```{r}
#| label: load and qc
cd34_path <- "~/Desktop/zenodo_data/cd34_multiome_rna.h5ad"

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

sc_object <- load_h5ad(object = sc_object, h5_path = cd34_path)
```

Let's run HVG detection, PCA and kNN (+ sNN) generation.

```{r}
#| label: preprocess
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = FALSE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = FALSE
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "kmknn")
  )
)
```

## Generating meta cells

All three generators take a `SingleCells` object, read counts from the binary
files as needed, and return a `MetaCells` object. They share a common output
structure: an obs table with one row per meta cell (recording which original
cells went into it), a var table mirrored from the parent object, and an
in-memory sparse count matrix.

### hdWGCNA-style (bootstrapped metacells)

The hdWGCNA-style algorithm reuses the kNN graph already on the object. The
two parameters that matter most are `target_no_metacells` and `max_shared`,
which controls how many cells two meta cells may have in common.

```{r}
#| label: hdwgcna
hdwgcna <- generate_bt_meta_cells_sc(
  object = sc_object,
  sc_meta_cell_params = params_sc_bt_metacells(
    target_no_metacells = 250L,
    max_shared = 15L,
    knn = list(k = 25L)
  ),
  regenerate_knn = TRUE,
  .verbose = TRUE
)

hdwgcna
```

Importantly, not all cells HAVE to be assigned to a meta cell. It's a fast,
bootstrapped method.

```{r}
#| label: hdwgcna - cells per meta cell
no_cells_hdwgcna <- purrr::map_dbl(hdwgcna[[]]$original_cell_idx, length)

hist(
  no_cells_hdwgcna,
  xlab = "No cells per meta cell",
  main = "Cells per hdWGCNA meta cell"
)
```

### SEACells

SEACells operates on the PCA embedding (or any batch corrected embedding) and 
the kNN graph and runs an iterative kernel archetypal analysis. `n_sea_cells` 
sets the number of archetypes; `min_iter`/`max_iter` bound the optimisation; 
`convergence_epsilon` sets the early stopping threshold relative to the initial 
RSS. The algorithm was aggressively optimised to be faster. With 
`pruning = TRUE` small values are set to zero, reducing calculations... This can 
impact quality, but usually it is very minor.

```{r}
#| label: seacells
seacells <- generate_seacells_sc(
  object = sc_object,
  seacell_params = params_sc_seacells(
    n_sea_cells = 250L,
    min_iter = 10L,
    convergence_epsilon = 0.001,
    pruning = TRUE
  ),
  .verbose = TRUE
)

seacells
```

Compared to hdWGCNA methods, more cells can be included in a given "SEACell"

```{r}
#| label: seacells - cells per meta cell
no_cells_seacells <- purrr::map_dbl(seacells[[]]$original_cell_idx, length)

hist(
  no_cells_seacells,
  xlab = "No cells per meta cell",
  main = "Cells per SEACells meta cell"
)
```

### SuperCells

SuperCells runs walktrap on the kNN graph. The number of meta cells is set
indirectly through `graining_factor`: with a factor of 20 over ~6800 cells
you should get roughly ~350 meta cells.

```{r}
#| label: supercells
supercells <- generate_supercells_sc(
  object = sc_object,
  sc_supercell_params = params_sc_supercell(
    graining_factor = 15
  ),
  .verbose = TRUE
)

supercells
```

Let's plot the number of cells per SEACell.

```{r}
#| label: supercells - cells per meta cell
no_cells_supercells <- purrr::map_dbl(supercells[[]]$original_cell_idx, length)

hist(
  no_cells_supercells,
  xlab = "No cells per meta cell",
  main = "Cells per Supercell"
)
```

### Purity

A simple sanity check is to ask, for each meta cell, what fraction of its
constituent cells share the same underlying label. We use the Leiden clusters
from the parent `SingleCells` object as a proxy for cell type identity. The
labels themselves are imperfect, so absolute numbers should be read with that
caveat, but relative differences between methods are reasonably indicative.

```{r}
#| label: purity
cell_labels <- unlist(sc_object[["celltype"]]) %>% as.character()

hdwgcna <- calc_meta_cell_purity(hdwgcna, original_cell_type = cell_labels)
seacells <- calc_meta_cell_purity(seacells, original_cell_type = cell_labels)
supercells <- calc_meta_cell_purity(
  supercells,
  original_cell_type = cell_labels
)

purity_dt <- rbind(
  data.table(method = "hdWGCNA", purity = hdwgcna[[]]$mc_purity),
  data.table(method = "SEACells", purity = seacells[[]]$mc_purity),
  data.table(method = "SuperCells", purity = supercells[[]]$mc_purity)
)

purity_dt[, .(mean = mean(purity), median = median(purity)), by = method]
```

```{r}
#| label: purity plot
#| fig-width: 6
#| fig-height: 4
ggplot(purity_dt, aes(x = method, y = purity)) +
  geom_violin(aes(fill = method), alpha = 0.6) +
  geom_boxplot(width = 0.05, outlier.size = 0.5) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "", y = "Meta cell purity (Leiden)") +
  ylim(0, 1)
```

## Working with the MetaCells class

This is where the in-memory representation pays off. HVG selection, PCA,
neighbour graphs and embeddings on a few hundred meta cells take seconds and
do not touch disk.

We continue with the SEACells output as it had the highest purity, but the
same code works on any of the three.

### HVGs and PCA

```{r}
#| label: mc hvg pca
seacells <- find_hvg_sc(
  object = seacells,
  hvg_no = 2000L,
  .verbose = FALSE
)

seacells <- calculate_pca_sc(
  object = seacells,
  no_pcs = 30L
)
```

### Neighbours, clustering and UMAP

```{r}
#| label: mc neighbours clustering umap
seacells <- find_neighbours_sc(
  object = seacells,
  neighbours_params = params_sc_neighbours(
    knn = list(k = 5L, knn_method = "exhaustive")
  )
)

seacells <- find_clusters_sc(seacells, res = 1.0, name = "leiden_clusters")

seacells <- umap_sc(seacells, k = 10L, knn_method = "exhaustive")
```

```{r}
#| label: mc umap plot
#| fig-width: 7
#| fig-height: 5
umap_dt <- as.data.table(
  get_embedding(seacells, "umap"),
  keep.rownames = "meta_cell_id"
)[, leiden_clusters := seacells[["leiden_clusters"]]]

ggplot(umap_dt, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = as.factor(leiden_clusters)), size = 1.5) +
  theme_bw() +
  labs(colour = "MetaCell cluster")
```

The plot is much sparser than the equivalent UMAP on the original cells,
which is the point: each dot is a denoised aggregate of ~15-20 cells, and
the structure that survives aggregation is the structure that's robust to
sparsity.

## Pseudo-bulking

Pseudo-bulks are conceptually adjacent but solve a different problem. Where
meta cells are dense aggregates designed to feed into co-expression and
archetype methods, pseudo-bulks are coarse summaries (typically one bulk per
sample-by-cluster combination) used as input to bulk-style DGE tools such as
limma-voom, edgeR or DESeq2.

`get_pseudobulked_sc` takes a named list of cell IDs and returns either a
dense matrix or a sparse `dgRMatrix`. With `assay = "raw"` it sums raw counts
(what bulk DGE tools expect); with `assay = "norm"` it averages normalised
counts.

```{r}
#| label: pseudobulk
cl_dt <- sc_object[[c("leiden_clusters", "cell_id")]]
cell_list <- split(cl_dt$cell_id, cl_dt$leiden_clusters)

pb_counts <- get_pseudobulked_sc(
  object = sc_object,
  cell_list = cell_list,
  return_format = "sparse",
  assay = "raw",
  .verbose = FALSE
)

dim(pb_counts)
```

The returned matrix has one row per group and one column per gene. From here
the standard bulk DGE pipeline applies; `bixverse` does not wrap the DGE call
itself.

## Coming up

A few things sit naturally on top of `MetaCells` and will be added in
upcoming releases:

- A SCENIC implementation for `MetaCells`. Meta cells suit SCENIC well
because the higher counts per observation make the GRN inference step less
noisy.
- An AUCell port for scoring gene sets on meta cells.

Both will dispatch on the `MetaCells` class with the in-memory speed
advantages described above.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```