---
title: "Multi-modal single cell analysis with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Multi-modal single cell analysis with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Gregor Lueg"
---

## Intro

This vignette walks through a multi-modal single cell workflow on a CITE-seq
data set (10x's 10k Human PBMC TotalSeq-B). The general `bixverse` philosophy
— do not keep things in memory if you do not have to, keep heavy lifting in
Rust, only round-trip to R when the user actually needs to look at something —
still applies. The bit that changes is that there is now a second modality
hanging off the main class: Antibody-Derived Tags (ADT). ATAC is on the roadmap
but not yet wired up.

Before going further, please read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html)
if you have not already. This vignette assumes you are comfortable with the
`SingleCells` class, the on-disk count storage, and the cells-to-keep logic.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(patchwork)
library(ggplot2)
library(data.table)
```

## Loading the data

A 10x Cell Ranger H5 with multiple feature types contains both the RNA and
the ADT counts in one file, separated by `feature_type`. We peek at the
metadata first to confirm what is in there before committing to anything.

```{r}
#| label: paths
# NEEDS UPDATE
h5_10x_path <- "~/Desktop/zenodo_data/multi-omics-data/10k_Human_PBMC_TotalSeqB.h5"

tempdir_10x_total_seq <- tempdir()
```

### EDA

```{r}
#| label: data exploration
h5_metadata <- read_tenx_h5_metadata(h5_10x_path)

h5_metadata$feature_types
```

Two feature types: gene expression and antibody capture. Good, we have both
modalities to play with.

### Loading in the RNA

The RNA goes through `SingleCellsMultiModal`, which is essentially the same
class as `SingleCells` with extra slots for `adt_counts` (and, later, an
`atac` connection). RNA loading and storage are identical to the single-modal
case: counts get pre-filtered, normalised, transposed, and written to the two
binary files on disk.

```{r}
#| label: load the RNA in 
sc_object <- SingleCellsMultiModal(dir_data = tempdir_10x_total_seq)

sc_object <- load_tenx_h5(object = sc_object, h5_path = h5_10x_path)

head(sc_object)
```

### Loading in the ADT counts

The ADT design philosophy differs from RNA. Protein panels are tiny: dozens of
antibodies, sometimes a couple of hundred for the larger commercial panels.
The cost-benefit calculus that justifies the on-disk binary format for RNA
just does not hold here. Even a million cells times 200 proteins fits in
memory without breaking a sweat. So the whole `ADTCounts` object — raw counts,
normalised counts, a bit of metadata — lives in memory as a plain matrix. The
saving is in code complexity rather than RAM.

For normalisation, you currently have two options:

- **CLR** (centred log-ratio): the Seurat default. Comes in two flavours via
  `seurat_clr`: the original CLR (allows negatives) and the modified
  Seurat-style version (does not).
- **DSB** (denoised and scaled by background): from
  [Mulè et al., 2022](https://www.nature.com/articles/s41467-022-29356-8).
  More principled if you have empty droplets to estimate the protein
  background; falls back to a 2-component k-means on log counts if you do
  not (the "ModelNegativeADTnorm" path).

Isotype controls help DSB tease apart per-cell technical noise from true
biological signal. `detect_adt_isotypes()` regexes the obvious patterns
(`IgG.*`, `Ig.*ctrl`, etc.) and is fine for standard TotalSeq panels; for
anything unusual just pass the column names directly via `isotype_names`.

```{r}
#| label: load the ADT counts 
adt_counts <- read_tenx_h5_adt(f_path = h5_10x_path)

iso <- detect_adt_isotypes(colnames(adt_counts))

sc_object <- add_adt_counts_sc(
  object = sc_object,
  adt_counts = adt_counts,
  method = "dsb",
  dsb_params = params_sc_dsb(
    use_isotype_controls = TRUE,
    quantile_low = 0.01,
    quantile_high = 1.0
  ),
  isotype_names = iso
)
```

## QC

### Low quality cells

QC runs jointly across modalities. The RNA loader already populated
`lib_size` and `nnz` on the obs table; `add_adt_counts_sc()` added
`adt_lib_size` and `adt_nnz` on top. We feed all four into MAD-based outlier
detection. Two-sided is sensible across the board — extremely high ADT
library size can also flag doublets or aggregates, not just dead cells.

```{r}
#| label: cell qc
qc_df <- sc_object[[c(
  "cell_id",
  "lib_size",
  "nnz",
  "adt_lib_size",
  "adt_nnz"
)]]
metrics <- list(
  log10_rna_lib_size = log10(qc_df$lib_size),
  log10_rna_nnz = log10(qc_df$nnz),
  log10_adt_lib_size = log10(qc_df$adt_lib_size),
  log10_adt_nnz = log10(qc_df$adt_nnz)
)

directions <- c(
  log10_rna_lib_size = "twosided",
  log10_rna_nnz = "twosided",
  log10_adt_lib_size = "twosided",
  log10_adt_nnz = "twosided"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)

qc_obs <- get_data(qc)

plots <- violin_plot_sc(qc)

plots$log10_rna_lib_size + plots$log10_adt_lib_size
```

```{r}
#| label: set cells to keep
sc_object[["outlier"]] <- qc$combined

cells_to_keep <- qc_df[!qc$combined, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)
```

## Standard workflows

For the initial steps, the two modalities are treated independently. Each
gets its own dimensionality reduction, kNN graph, and embeddings. The
`modality` argument routes to the relevant Rust kernel and stores results in
the right cache slot: `sc_cache` for RNA, `adt_cache` for ADT. Cross-modal
integration comes later, in the WNN section.

### RNA

Standard recipe: HVGs, PCA, neighbours, Leiden, tSNE.

```{r}
#| label: hvg, pca, knn for RNA
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  no_embd_to_use = 16L,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
```

```{r}
#| label: leiden clustering for RNA
sc_object <- find_clusters_sc(sc_object)

head(sc_object)
```

```{r}
#| label: tSNE for RNA
sc_object <- tsne_sc(
  sc_object,
  modality = "rna"
)
```

```{r}
#| label: embedding plots with leiden clustering
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "leiden_clustering",
  discrete = TRUE,
  point_size = 0.5
)
```

### ADT

For ADT we skip HVG selection — there is nothing to select from in a panel
of thirty-odd antibodies — and feed the panel directly into PCA, minus the
isotype controls (they only add background noise). Otherwise the recipe is
the same as for RNA, with `modality = "adt"` doing the routing.

```{r}
#| label: pca and knn for ADT
sc_object <- calculate_pca_adt_sc(
  object = sc_object,
  no_pcs = 15L,
  features = remove_adt_isotypes(get_adt_names(sc_object))
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  modality = "adt"
)
```

```{r}
#| label: tsne and feature plot for RNA
sc_object <- tsne_sc(
  sc_object,
  modality = "adt"
)

feature_plot_sc(
  sc_object,
  features = "CD3D",
  embedding = "tsne",
  expr_modality = "adt",
  embd_modality = "adt",
  point_size = 1
) +
  theme_bx()
```

## Plotting

`bixverse.plots` has a few helpers tuned to multi-modal data. The main
mechanism is that the plotting functions take an `expr_modality` (where the
values come from) and an `embd_modality` (which embedding to overlay them
on) separately. That split is what lets you colour an ADT-derived UMAP by
RNA expression, or any other combination.

### Features against each other

The classic CITE-seq sanity check: does the protein signal match the mRNA
signal where you would expect it to? `feature_scatter_plot_sc()` takes two
features with optional `_rna` / `_adt` suffixes, so you can mix modalities
freely. mRNA tends to be sparse and noisy; ADT tends to be smoother but with
more background. A scatter against each other surfaces both effects in one
picture.

```{r}
#| label: cd3 scatter plot
feature_scatter_plot_sc(
  object = sc_object,
  feature_1 = "ENSG00000167286_rna",
  feature_2 = "CD3D_adt"
) +
  theme_bx() +
  xlab("CD3D mRNA") +
  ylab("CD3D ADT")
```

```{r}
#| label: cd19 scatter plot
feature_scatter_plot_sc(
  object = sc_object,
  feature_1 = "ENSG00000177455_rna",
  feature_2 = "CD19_adt"
) +
  theme_bx() +
  xlab("CD19 mRNA") +
  ylab("CD19 ADT")
```

### Embeddings

Same idea on the embedding side. The cell below paints RNA expression of CD3D
onto an ADT-derived tSNE. `feature_labels` is the cleanest way to render a
human-readable label without renaming columns upstream.

```{r}
#| label: cd3 tsne plot - rna on adt tsne
feature_plot_sc(
  sc_object,
  features = "ENSG00000167286",
  feature_labels = c(ENSG00000167286 = "CD3D_rna"),
  embedding = "tsne",
  expr_modality = "rna",
  embd_modality = "adt",
  point_size = 1
) +
  theme_bx()
```

For the reverse — ADT expression on the same ADT embedding — features just
go by their panel name:

```{r}
#| label: cd19 tsne plot - adt on adt tsne
feature_plot_sc(
  sc_object,
  features = "CD19",
  embedding = "tsne",
  expr_modality = "adt",
  embd_modality = "adt",
  point_size = 1
)
```

`embedding_plot_sc()` accepts the same `embd_modality` argument when you want
to overlay obs columns (clusters, donor IDs, cell type calls) on a particular
modality's embedding:

```{r}
embedding_plot_sc(
  object = sc_object,
  embedding = "tsne",
  embd_modality = "adt",
  colour_by = "leiden_clustering",
  discrete = TRUE,
  point_size = 1
)
```

## WNN

So far we have looked at each modality in isolation. For a single cell-type
assignment that uses both, you want a joint representation. The current
implementation in `bixverse` is Weighted Nearest Neighbours (WNN) from
[Hao et al., 2021](https://www.cell.com/cell/fulltext/S0092-8674(21)00583-3),
the same method Seurat v4 introduced.

The intuition is that different modalities carry different information for
different cells. For something like monocyte versus B cell, the RNA signal
is very strong. For CD4 versus CD8 T cell subsets, the protein signal is
much cleaner — the mRNAs are barely differential at that resolution. WNN
formalises this: for each cell, it asks how well that cell's neighbours in
one modality predict its profile in the *other* modality, and uses the
answer to weight the two modalities per cell. The weighted neighbours are
then merged into a single joint graph.

The whole thing runs in Rust — per-cell weights, neighbour selection, graph
merging — so it scales to hundreds of thousands of cells comfortably. The
resulting graph lands under `other_data$wnn` and can be embedded just like
any other modality.

```{r}
sc_object <- generate_wnn_graph_sc(
  sc_object,
  wnn_params = params_sc_wnn(sd_scale = 1)
)
```

```{r}
sc_object <- tsne_sc(
  sc_object,
  modality = "wnn"
)
```

A nice sanity check is to look at the same marker from both modalities,
projected onto the WNN embedding. If WNN is doing its job, the two pictures
should largely agree — even though they come from completely independent
measurements.

```{r}
p1 <- feature_plot_sc(
  sc_object,
  features = "ENSG00000167286",
  feature_labels = c(ENSG00000167286 = "CD3D_rna"),
  embedding = "tsne",
  expr_modality = "rna",
  embd_modality = "wnn",
  point_size = 1
)

p2 <- feature_plot_sc(
  sc_object,
  features = "CD3D",
  feature_labels = c(CD3D = "CD3D_adt"),
  embedding = "tsne",
  expr_modality = "adt",
  embd_modality = "wnn",
  point_size = 1
)

p1 + p2
```

From here, the rest of the pipeline (clustering, marker detection, cell type
annotation) works exactly as in the standard vignette — just point the
relevant functions at the WNN graph instead of the RNA one.