---
title: "NicheNet ligand-receptor analysis with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{NicheNet ligand-receptor analysis with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

This vignette walks through a [NicheNet-style](https://pubmed.ncbi.nlm.nih.gov/31819264/) 
ligand-receptor analysis on the PBMC3k data set using `bixverse`. It assumes 
familiarity with the `SingleCells` class and the basic processing pipeline; if 
you have not read the
[PBMC processing vignette](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html),
please do so first.

NicheNet ranks upstream ligands by how well their predicted target gene sets
align with a gene set of interest in a receiver cell type. The pipeline here
is:

1. Build a ligand-target regulatory potential matrix from a signalling
   network (PPI) and a gene regulatory network (GRN).
2. Define a gene set of interest in the receiver (here, T-cell marker
   genes).
3. Score each ligand by how well its targets overlap with that gene set
   (AUROC, AUPR, correlations).
4. Combine ligand activity with sender/receiver expression and DE into a
   single prioritisation score.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(ggplot2)
library(magrittr)
```

## Loading and processing PBMC3k

We reuse the standard processing pipeline. See the PBMC processing vignette
for commentary on each step.

```{r}
#| label: load and process
pbmc3k_path <- download_pbmc3k()
tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)
mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(sc_object, table = "var", old = "column1", new = "gene_symbol")
var <- get_sc_var(sc_object)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)

# QC and processing
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  .verbose = FALSE
)

# minimal cell filtering
cells_to_keep <- sc_object[[]][MT < 0.2, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)
sc_object <- calculate_pca_sc(sc_object, no_pcs = 30L, .verbose = FALSE)
sc_object <- find_neighbours_sc(sc_object, .verbose = FALSE)
sc_object <- find_clusters_sc(sc_object, res = 0.5)
```

For demonstration purposes we use the Leiden cluster IDs directly as cell
types. In a real analysis you would annotate clusters via marker genes
(e.g. with `find_all_markers_sc` and a canonical marker panel) before
running prioritisation; senders and receivers are biological concepts.

```{r}
#| label: label clusters
sc_object[["celltype"]] <- paste0(
  "cluster_",
  unlist(sc_object[["leiden_clustering"]])
)

celltype_counts <- sc_object[[]][!is.na(celltype), .N, by = celltype][order(-N)]

celltype_counts
```

## Building the NicheNet networks

A production analysis would load the curated NicheNet networks (signalling
+ GRN) from [Zenodo](https://zenodo.org/records/7074291). For a self-contained 
vignette we construct a small network covering a handful of canonical 
immune-signalling pathways. The shape of the inputs is what matters here... 
Also, this allows you to feed in your own networks if desired.

```{r}
#| label: build networks
# Ligand-receptor pairs (canonical immune signalling)
lr_network <- data.table(
  # ligands
  from = c("CCL5", "CXCL10", "IL15", "TNF", "TGFB1", "IL10", "ICAM1"),
  # receptors
  to = c(
    "CCR5",
    "CXCR3",
    "IL2RG",
    "TNFRSF1A",
    "TGFBR2",
    "IL10RA",
    "ITGAL"
  ),
  weight = 1
)

# Signalling (PPI) layer: ligand -> receptor -> intracellular signalling -> TFs
ppi_network <- data.table(
  from = c(
    "CCL5",
    "CXCL10",
    "IL15",
    "TNF",
    "TGFB1",
    "IL10",
    "ICAM1",
    "CCR5",
    "CXCR3",
    "IL2RG",
    "TNFRSF1A",
    "TGFBR2",
    "IL10RA",
    "ITGAL",
    "JAK1",
    "JAK2",
    "JAK3",
    "STAT1",
    "STAT3",
    "STAT5A",
    "NFKB1",
    "SMAD2"
  ),
  to = c(
    "CCR5",
    "CXCR3",
    "IL2RG",
    "TNFRSF1A",
    "TGFBR2",
    "IL10RA",
    "ITGAL",
    "JAK1",
    "JAK1",
    "JAK3",
    "NFKB1",
    "SMAD2",
    "JAK1",
    "NFKB1",
    "STAT1",
    "STAT3",
    "STAT5A",
    "STAT3",
    "STAT1",
    "STAT3",
    "STAT3",
    "STAT3"
  ),
  weight = 1
)

# GRN layer: TF -> targets
grn_network <- data.table(
  from = c(
    rep("STAT1", 6),
    rep("STAT3", 6),
    rep("STAT5A", 4),
    rep("NFKB1", 6),
    rep("SMAD2", 4)
  ),
  to = c(
    "IRF1",
    "MX1",
    "ISG15",
    "OAS1",
    "GBP1",
    "CXCL10",
    "SOCS3",
    "IL6",
    "BCL2",
    "VEGFA",
    "MYC",
    "FOS",
    "CCND1",
    "BCL2L1",
    "IL2RA",
    "FOXP3",
    "TNF",
    "IL1B",
    "IL6",
    "NFKBIA",
    "CCL5",
    "ICAM1",
    "SERPINE1",
    "CDKN1A",
    "JUNB",
    "ID1"
  ),
  weight = 1
)
```

Map gene symbols to the Ensembl IDs used internally by `sc_object`, dropping
any genes not present in the data set.

```{r}
#| label: symbol to ensembl
symbol_to_id <- function(syms) {
  ids <- symbol_to_ensembl[syms]
  ids[!is.na(ids)]
}

# Restrict the networks to genes present in PBMC3k
keep_ppi <- ppi_network$from %in%
  names(symbol_to_ensembl) &
  ppi_network$to %in% names(symbol_to_ensembl)
keep_grn <- grn_network$from %in%
  names(symbol_to_ensembl) &
  grn_network$to %in% names(symbol_to_ensembl)
keep_lr <- lr_network$ligand %in%
  names(symbol_to_ensembl) &
  lr_network$receptor %in% names(symbol_to_ensembl)

ppi_network <- ppi_network[keep_ppi]
grn_network <- grn_network[keep_grn]
lr_network <- lr_network[keep_lr]
```

## Ligand-target regulatory potential

Construct the ligand-target influence matrix. Each row is one ligand and
each column is a gene from the union of both networks; entries are
NicheNet-style regulatory potential scores.

```{r}
#| label: ligand target influence
ligand_seeds <- setNames(
  lapply(lr_network$from, identity),
  lr_network$from
)

ligand_influence <- generate_ligand_target_influence(
  ligand_seeds = ligand_seeds,
  ppi_network = rbind(lr_network, ppi_network),
  grn_network = grn_network,
  # because of the tiny network, the original ltf_cutoff of 0.99 would be way
  # too aggressive
  params = params_ligand_target(ltf_cutoff = 0)
)

ligand_influence
```

This returns a `LigandTargetInfluence` class. If you want to access the actual
scores, you can use:

```{r}
#| label: check the ligand to target influence matrix
get_influence(ligand_influence)
```

## Defining the gene set of interest

In a case/control experiment, the gene set of interest would be the DEGs
between conditions in the receiver cell type. PBMC3k is a single condition,
so we use cluster-vs-rest marker genes as a stand-in. Pick the largest
cluster as receiver and a few others as senders.

```{r}
#| label: receiver markers
receiver_id <- celltype_counts[1, celltype]
senders_oi <- celltype_counts[2:min(5, .N), celltype]

receiver_cells <- sc_object[[]][celltype == receiver_id, cell_id]
other_cells <- sc_object[[]][
  celltype != receiver_id & !is.na(celltype),
  cell_id
]

receiver_de <- find_markers_sc(
  object = sc_object,
  cells_1 = receiver_cells,
  cells_2 = other_cells,
  .verbose = FALSE
)

# upregulated genes with FDR <= 0.05 form the gene set
geneset_oi <- receiver_de[lfc > 0 & fdr <= 0.05, gene_id]
length(geneset_oi)
```

## Ligand activity scoring

Score each ligand by how well its target potential vector aligns with
`geneset_oi`. The background is the full set of genes in the influence
matrix.

```{r}
#| label: ligand activity
activity <- ligand_activity_scores(
  ligand_influence = ligand_influence,
  # need to transform to gene symbols
  gene_sets = list(receiver_degs = ensembl_to_symbol[geneset_oi])
)

activity[order(-aupr_corrected)]
```

## Sender and receiver tables

Compute the inputs needed for prioritisation: DE of ligands across senders,
DE of receptors in the receiver, and average expression of ligands /
receptors per cluster.

```{r}
#| label: senders and DE
celltype_de <- find_all_markers_sc(
  object = sc_object,
  column_of_interest = "celltype",
  .verbose = FALSE
)

# `find_*_markers_sc` returns `gene_id` and `cluster`; rename to match the
# prioritisation function's expected schema.
setnames(
  celltype_de,
  c("gene_id", "grp", "p_values"),
  c("gene", "cluster_id", "pval"),
  skip_absent = TRUE
)
celltype_de <- celltype_de[, .(cluster_id, gene, lfc, pval)]
```

Compute average expression per cluster for the ligands and receptors of
interest. This calls into the streaming engine via
`compute_expression_info_sc`.

```{r}
#| label: expression info
genes_of_interest <- union(lr_network$to, lr_network$from)
genes_of_interest_ids <- symbol_to_id(genes_of_interest)

expression_info <- compute_expression_info_sc(
  object = sc_object,
  celltype_colname = "celltype",
  genes = genes_of_interest_ids
)
```

## Prioritisation

Combine all of the above into a single weighted score. PBMC3k has no
condition contrast, so we use the `"one_condition"` scenario, which sets
the condition-specificity weights to zero.

```{r}
#| label: prioritisation
# Reduce activity table to one row per ligand (we ran one gene set)
la_input <- activity[, .(ligand, aupr_corrected)]
la_input[, ligand := symbol_to_ensembl[ligand]]

lr_network_ensembl <- data.table(
  ligand = symbol_to_ensembl[lr_network$from],
  receptor = symbol_to_ensembl[lr_network$to]
) %>%
  na.omit()

result <- prioritise_interactions(
  celltype_de = celltype_de,
  expression_info = expression_info,
  ligand_activities = na.omit(la_input),
  lr_network = lr_network_ensembl,
  senders_oi = senders_oi,
  receivers_oi = receiver_id,
  scenario = "one_condition"
)
```

Add symbol columns for readability.

```{r}
#| label: annotate result
result[, ligand_symbol := ensembl_to_symbol[ligand]]
result[, receptor_symbol := ensembl_to_symbol[receptor]]

head(
  result[, .(
    sender,
    receiver,
    ligand_symbol,
    receptor_symbol,
    prioritisation_score,
    prioritisation_rank
  )],
  10
)
```

## Top interactions

A quick dotplot of the top prioritised interactions.

```{r}
#| label: dotplot
#| fig.width: 7
#| fig.height: 4
top_n <- 15L

plot_dt <- head(result, top_n)[, `:=`(
  interaction = paste(ligand_symbol, "->", receptor_symbol)
)]

ggplot(
  plot_dt,
  aes(
    x = sender,
    y = interaction,
    size = prioritisation_score,
    colour = prioritisation_score
  )
) +
  geom_point() +
  scale_colour_viridis_c() +
  theme_minimal() +
  labs(
    x = "Sender",
    y = "Ligand -> Receptor",
    size = "Score",
    colour = "Score"
  )
```

## Notes on DE methods

`find_markers_sc` runs a Wilcoxon-style test, which is convenient but suffers
from well-documented p-value inflation in single-cell data (because each
cell is treated as an independent observation). For statistical inference
that matters, prefer pseudo-bulk + `DESeq2` or `edgeR`:

1. Aggregate counts per (sample, cluster) into pseudo-bulk.
2. Run a standard bulk DE pipeline.
3. Feed the resulting `data.table` into `prioritise_interactions` after
   renaming columns to `cluster_id`, `gene`, `lfc`, `pval`.

The prioritisation function does not care which method produced the DE
table.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```
