---
title: "Ontologies"
vignette: >
  %\VignetteIndexEntry{Ontologies}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Ontologies

`bixverse` provides methods for computing semantic similarities over
ontologies — disease ontologies, gene ontologies, phenotype ontologies, and so
on. The heavy lifting is done yet again in Rust with heavy parallelism being
exploited. This makes this (usually) faster than the alternative packages.

This vignette walks through the core workflow: building the ancestry from a
parent–child table, deriving information content, and computing different
flavours of semantic similarity.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(magrittr)
library(zeallot)
```

## Data

We will use the Human Phenotype Ontology (HPO) bundled with the
`ontologyIndex` package. The only thing `bixverse` needs is a `data.table`
with three columns: `parent`, `child`, and `type`. These are describing the 
directed edges of the ontology graph.

```{r}
#| label: loading-hpo

library("ontologyIndex")

data("hpo")

hpo_data <- as.data.table(stack(hpo$children)) %>%
  setnames(old = c("values", "ind"), new = c("child", "parent")) %>%
  .[, c("parent", "child"), with = FALSE] %>%
  .[, type := "parent_of"]
```

## Semantic similarities

### Ancestry and information content

Before computing any similarity we need two ingredients: the full set of
ancestors for every term, and the information content (IC) derived from the
number of descendants each term has. Both are obtained in one step:

```{r}
#| label: ancestry-and-ic

c(ancestors, descendants) %<-% get_ontology_ancestry(hpo_data)

information_content <- calculate_information_content(descendants)
```

### Full similarity matrices

Armed with ancestors and IC values we can build pairwise similarity matrices
over the entire ontology. `bixverse` supports three IC-based measures:

- **Resnik** similarity — the IC of the most informative common ancestor.
- **Lin** similarity — Resnik similarity normalised by the IC of the two terms.
- **Combined** — the average of the (normalised) Resnik and Lin scores.

All matrices are rescaled to the $[0, 1]$ interval.

::: {.callout-note}
The full matrix generations are not run during Vignette generation. While fast,
the GitHub runners are VERY memory limited making this impossible.
:::

```{r}
#| label: resnik-similarity
#| eval: false

resnik_similarity <- calculate_semantic_sim_mat(
  similarity_type = "resnik",
  ancestor_list = ancestors,
  ic_list = information_content
)

resnik_similarity[1:5, 1:5]
```

Switching to the Lin similarity is just a parameter change:

```{r}
#| label: lin-similarity
#| eval: false

lin_similarity <- calculate_semantic_sim_mat(
  similarity_type = "lin",
  ancestor_list = ancestors,
  ic_list = information_content
)

lin_similarity[1:5, 1:5]
```

### Similarities for a subset of terms

Computing the full matrix is not always necessary or desirable. If you only
care about a handful of terms you can calculate pairwise similarities directly,
which returns a long-format `data.table` instead of a dense matrix. Here we
draw a random sample of ten terms and use the combined similarity:

```{r}
#| label: subset-similarity

set.seed(10101L)

random_term_sample <- sort(sample(names(information_content), 10L))

combined_similarity <- calculate_semantic_sim(
  terms = random_term_sample,
  similarity_type = "combined",
  ancestor_list = ancestors,
  ic_list = information_content
)

head(combined_similarity)
```

## Wang similarities

In addition to the IC-based measures above, `bixverse` implements the
[Wang similarity](https://academic.oup.com/bioinformatics/article/23/10/1274/197095).
This approach operates directly on the DAG structure of the ontology and lets
you assign an edge weight to each relationship type. For ontologies with
multiple relationship types — the Gene Ontology, for instance — this gives you
fine-grained control over how different edges contribute to similarity.

```{r}
#| label: wang-similarity-matrix
#| eval: false

# The HPO only has one relationship type; for the GO you would specify
# several, e.g. c(is_a = 0.8, part_of = 0.6)

wang_weights <- setNames(0.8, "parent_of")

wang_similarity <- calculate_wang_sim_mat(
  parent_child_dt = hpo_data,
  weight = wang_weights
)

wang_similarity[1:5, 1:5]
```

As with the IC-based measures, you can restrict the computation to a subset of
terms:

```{r}
#| label: wang-similarity-subset

# needs to be provided as previous chunk is not evaluated
wang_weights <- setNames(0.8, "parent_of")

wang_similarity_subset <- calculate_wang_sim(
  terms = random_term_sample,
  parent_child_dt = hpo_data,
  weight = wang_weights
)

head(wang_similarity_subset)
```