---
title: "Pathway activity functions"
vignette: >
  %\VignetteIndexEntry{Pathway activity functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: Gregor Lueg
---

## Single-sample pathway activity

A common question in transcriptomics is whether the genes belonging to a given
pathway are concordantly up- or down-regulated in each sample. Several
algorithms tackle this by collapsing a gene-by-sample expression matrix into a
pathway-by-sample score matrix, the most widely used being
[GSVA](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-14-7)
and [ssGSEA](https://www.nature.com/articles/nature08460). `biverse` provides
fast versions of both with some tricks under the hood to make them faster than
the original implementations.
Moreover, if you have multiple contrasts (different drugs vs. DMSO for example).
`bixverse` also provides Rust-accelerated implementations from
[mitch](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-06856-9).

```{r}
#| label: setup
library(bixverse)
```

## Data

Following the GSVA vignette convention, we generate some synthetic expression
data (Gaussian-distributed to mimic normalised microarray or RNA-seq log
counts, and Poisson-distributed to mimic raw counts) together with a
collection of random gene sets.

```{r}
#| label: random-data-generation

p <- 10000L # genes
n <- 100L # samples

# Gaussian expression matrix
X <- matrix(
  rnorm(p * n),
  nrow = p,
  dimnames = list(paste0("g", seq_len(p)), paste0("s", seq_len(n)))
)

# Poisson count matrix
X_counts <- matrix(
  rpois(p * n, lambda = 10),
  nrow = p,
  dimnames = list(paste0("g", seq_len(p)), paste0("s", seq_len(n)))
)
storage.mode(X_counts) <- "numeric"

# Random gene sets of varying size
gs <- as.list(sample(10:100, size = 250, replace = TRUE))
gs <- lapply(
  gs,
  function(n, p) paste0("g", sample(seq_len(p), size = n, replace = FALSE)),
  p
)
names(gs) <- paste0("gs", seq_along(gs))
```

## GSVA

### Gaussian kernel

The Gaussian kernel version of GSVA is appropriate for continuous expression
values (log-CPM, microarray intensities, etc.). Running it in `bixverse` is a
single call:

```{r}
#| label: bixverse-gsva-gaussian

bixverse_res_gaussian <- calc_gsva(
  exp = X,
  pathways = gs,
  gaussian = TRUE
)

bixverse_res_gaussian[1:5, 1:5]
```

If the original `GSVA` Bioconductor package is available we can verify that
the results are essentially identical. Minor differences in numerical precision
arise from optimisations on the Rust side, but per-pathway correlations between
the two implementations are consistently above 0.99.

```{r}
#| label: gsva-gaussian-comparison
#| eval: !expr requireNamespace("GSVA", quietly = TRUE)

library(GSVA)

gsvaPar <- gsvaParam(X, gs)
gsva_res_gaussian <- as.matrix(gsva(gsvaPar, verbose = FALSE))

correlations <- diag(cor(bixverse_res_gaussian, gsva_res_gaussian))

print(sprintf(
  "All pathway correlations >= 0.99: %s (min: %.4f)",
  all(correlations >= 0.99),
  min(correlations)
))
```

Let's compare speed (differences will be more pronounced, the more 
computational fire power your system has.)

```{r}
#| label: gsva-gaussian-speed
#| eval: !expr requireNamespace(c("GSVA", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  gsva = {
    gsvaPar <- gsvaParam(X, gs)
    gsva(gsvaPar, verbose = FALSE)
  },
  bixverse = calc_gsva(exp = X, pathways = gs, gaussian = TRUE),
  times = 3L
)
```

### Poisson kernel

For count data a Poisson kernel is more appropriate. The only change is
setting `gaussian = FALSE`:

```{r}
#| label: bixverse-gsva-poisson

bixverse_res_poisson <- calc_gsva(
  exp = X_counts,
  pathways = gs,
  gaussian = FALSE
)

bixverse_res_poisson[1:5, 1:5]
```

And vs. the original

```{r}
#| label: gsva-poisson-comparison
#| eval: !expr requireNamespace("GSVA", quietly = TRUE)

gsvaParPoisson <- gsvaParam(X_counts, gs, kcdf = "Poisson")
gsva_res_poisson <- as.matrix(gsva(gsvaParPoisson, verbose = FALSE))

correlations <- diag(cor(bixverse_res_poisson, gsva_res_poisson))

print(sprintf(
  "All pathway correlations >= 0.99: %s (min: %.4f)",
  all(correlations >= 0.99),
  min(correlations)
))
```

And speed:

```{r}
#| label: gsva-poisson-speed
#| eval: !expr requireNamespace(c("GSVA", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  gsva = {
    gsvaPar <- gsvaParam(X_counts, gs, kcdf = "Poisson")
    gsva(gsvaPar, verbose = FALSE)
  },
  bixverse = calc_gsva(exp = X_counts, pathways = gs, gaussian = FALSE),
  times = 3L
)
```

## ssGSEA

ssGSEA, first described in
[Barbie et al.](https://www.nature.com/articles/nature08460), takes a similar
approach but does not apply a kernel-based normalisation across samples.
`bixverse` provides a Rust-optimised version here as well:

```{r}
#| label: bixverse-ssgsea

bixverse_res_ssgsea <- calc_ssgsea(
  exp = X,
  pathways = gs
)

bixverse_res_ssgsea[1:5, 1:5]
```

Let's compare again against the GSVA version:

```{r}
#| label: ssgsea-comparison
#| eval: !expr requireNamespace("GSVA", quietly = TRUE)

ssgseaPar <- ssgseaParam(X, gs)
ssgsea_res <- as.matrix(gsva(ssgseaPar, verbose = FALSE))

correlations <- diag(cor(bixverse_res_ssgsea, ssgsea_res))

print(sprintf(
  "All pathway correlations >= 0.99: %s (min: %.4f)",
  all(correlations >= 0.99),
  min(correlations)
))
```

And check the underlying speed differences:
```{r}
#| label: ssgsea-speed
#| eval: !expr requireNamespace(c("GSVA", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  gsva = {
    ssgseaPar <- ssgseaParam(X, gs)
    gsva(ssgseaPar, verbose = FALSE)
  },
  bixverse = calc_ssgsea(exp = X, pathways = gs),
  times = 3L
)
```

## Multi-contrast enrichment (mitch)

When you have multiple contrasts - say, differential expression results from
several comparisons or different omics layers - you often want to know which
pathways show coordinated changes *across* contrasts. The
[mitch](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-06856-9)
method does exactly this: it ranks genes within each contrast, computes mean
ranks per pathway, and then uses a MANOVA test to identify pathways that are
enriched in one or more contrasts simultaneously. `bixverse` re-implements the
core algorithm so that it slots into the same workflow as the other pathway
activity functions... Also, heavily multi-threaded and designed to go 
**brrrrrr** in terms of speed.

`calc_mitch` expects a numeric matrix of contrast statistics (genes in rows,
contrasts in columns) and a named list of gene sets. It returns a `data.table`
containing per-pathway MANOVA p-values, FDR-adjusted p-values, and the
individual contrast-level enrichment scores and p-values.

```{r}
#| label: mitch-example

set.seed(42L)

contrast_data <- matrix(rnorm(3 * 26), nrow = 26)
colnames(contrast_data) <- sprintf("contrast_%i", 1:3)
rownames(contrast_data) <- letters

gene_sets <- list(
  pathway_A = sample(letters, 4),
  pathway_B = sample(letters, 5),
  pathway_C = sample(letters, 6),
  pathway_D = sample(letters, 7)
)

res <- calc_mitch(
  contrast_mat = contrast_data,
  gene_set_list = gene_sets
)

res
```

Note that pathways smaller than the internal minimum set size are dropped from
the output, and the results are sorted by significance. The function also
validates its input: passing a matrix containing `NA` values will throw an
error rather than silently producing nonsense.

### Comparison with the mitch package

When the original `mitch` package is available we can verify that the two
implementations produce identical results. Here we use the example data
bundled with `mitch`:

```{r}
#| label: mitch-comparison
#| eval: !expr requireNamespace("mitch", quietly = TRUE)

data(myImportedData, genesetsExample, package = "mitch")

mitch_res <- suppressMessages(mitch::mitch_calc(
  myImportedData,
  genesetsExample,
  priority = "significance",
  minsetsize = 5,
  cores = 2
))

bixverse_res <- calc_mitch(
  contrast_mat = as.matrix(myImportedData),
  gene_set_list = genesetsExample
)

# Pathway ordering and FDR values should match exactly
all(bixverse_res$pathway_names == mitch_res$enrichment_result$set) &&
  all.equal(bixverse_res$manova_fdr, mitch_res$enrichment_result$p.adjustMANOVA)
```

And let's check the speed differences:

```{r}
#| label: mitch-speed
#| eval: !expr requireNamespace(c("mitch", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  mitch = suppressMessages(mitch::mitch_calc(
    myImportedData,
    genesetsExample,
    priority = "significance",
    minsetsize = 5,
    cores = 2
  )),
  bixverse = calc_mitch(
    contrast_mat = as.matrix(myImportedData),
    gene_set_list = genesetsExample
  ),
  times = 5L
)
```

As with the GSVA and ssGSEA implementations, you should observe meaningful
speed improvements from Rust, particularly as the number of gene sets or 
contrasts grows and the more cores/oomph your system has.