---
title: "Analysing PBMCs with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Analysing PBMCs with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Grant Neilson, Gregor Lueg, Liesbeth François"
---

## Intro

This vignette walks through a standard single cell analysis on the PBMC3k data
set using `bixverse`. If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first; this vignette assumes familiarity with how the
`SingleCells` class, on-disk storage and cells-to-keep logic work.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(bixverse.plots)
library(magrittr)
```

## Loading the data

We start by downloading the PBMC3k data set bundled with the package and
loading it via Cell Ranger-style MTX I/O.

```{r}
#| label: load pbmc3k
pbmc3k_path <- download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = TRUE
)

sc_object
```

Let's have a quick look at the variable table. The column names from the MTX
files are not always informative, so we rename the gene symbol column to
something sensible and set up mappings between Ensembl IDs and symbols.

```{r}
#| label: var table and renaming
var <- get_sc_var(sc_object)

head(var)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
```

## Quality control

### Gene set proportions

A typical first step is computing the proportion of counts mapping to
mitochondrial and ribosomal genes per cell. These are added directly to the
obs table in the DuckDB.

```{r}
#| label: gene set proportions
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

sc_object[[1:5L]]
```

As we can see we now have MT and Ribo as columns in the obs table.

### MAD outlier detection

We use per-cell QC metrics and MAD-based outlier detection to flag problematic
cells. The `run_cell_qc` function returns a `CellQc` object that carries the
metrics, per-metric outlier calls and a combined outlier vector.

```{r}
#| label: qc outlier detection
qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3,
  groups = qc_df$donor_id
)


get_data(qc)

qc
```

The `CellQc` class has a few associated plotting functions using the output of `run_cell_qc` and the specified metrics as input

- `violin_plot_sc` method that produces violin plots with outliers highlighted.

```{r}
#| label: qc plots
#| fig-width: 12
#| fig-height: 4
violin_plot_sc(qc)
```

- `joint_plot_sc` method that compares the genes versus UMIs per cell.

```{r}
#| label: qc plots
#| fig-width: 12
#| fig-height: 4
joint_plot_sc(qc)
```

### Filtering cells

We store the outlier flag in the obs table and then set the cells to keep. From
this point on, all downstream methods (HVG selection, PCA, etc.) will only
operate on the retained cells.

```{r}
#| label: set cells to keep
sc_object[["outlier"]] <- qc$combined

cells_to_keep <- qc_df[!qc$combined, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object
```

## Feature selection, PCA and neighbours

With QC done, we move through the standard pipeline: highly variable gene
selection, PCA, and nearest neighbour computation.

```{r}
#| label: hvg pca neighbours
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
```

## Clustering and marker detection

### Leiden clustering

Leiden clustering followed by differential gene expression across all clusters.
You can also run Louvain if you want, but Leiden is thought of to have better
properties, see [Traag, et al.](https://www.nature.com/articles/s41598-019-41695-z).

```{r}
#| label: clustering and markers
sc_object <- find_clusters_sc(sc_object, res = 1, name = "leiden_clusters")

all_markers <- find_all_markers_sc(
  object = sc_object,
  column_of_interest = "leiden_clusters"
)

all_markers[, gene_symbol := ensembl_to_symbol[gene_id]]

head(all_markers[fdr <= 0.05][order(-abs(lfc))])
```

### Fast clustering

In the case of large data sets there is an option to run an accelerated form
of clustering. This runs first k-means clustering (with default `sqrt(N)` 
cells), then kNN on the centroids, followed by Louvain (Leiden is not yet
supported, but on the to-do list) across a set of resolutions. There is also an 
option to run this across several seeds to check for stability of the 
clustering. If you ran the grid search, the membership

```{r}
#| label: fast clustering
fast_cluster_res <- fast_cluster_sc(
  object = sc_object,
  resolutions = c(5, 3, 2, 1.5, 1, 0.5),
  # also return the k-mean clustering
  return_kmeans = TRUE,
  no_seeds = 25L,
  grid_search = TRUE
)

fast_clusted_dt <- get_data(fast_cluster_res)

head(fast_clusted_dt)
```

If you want to explore the k-means memberships or centroids, there are getters
for this.

```{r}
#| label: fast cluster k-means data
centroids <- get_centroids(fast_cluster_res)

kmeans_membership <- get_kmeans_clusters(fast_cluster_res)
```


### Dimensionality reduction

Let us run quickly UMAP and tSNE.

```{r}
#| label: umap and tsne
sc_object <- umap_sc(sc_object)

sc_object <- tsne_sc(
  sc_object,
  perplexity = 10
)
```

We can use `bixverse.plots` package which provides some basic plotting functions such as:

- Plotting the embedding using `embedding_plot_sc` to check out the UMAP embeddings:

```{r}
#| label: umap plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

Let's check out tSNE

```{r}
#| label: tsne plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

- We can also plot a gene expression on the embedding space using `feature_plot_sc`:

```{r}
feature_plot_sc(
  object = sc_object,
  features = "ENSG00000156738",
  feature_labels = c("ENSG00000156738" = "MS4A1"),
  embedding = "umap"
)
```

### Cell type annotation

Define a set of marker genes to perform cell type annotation.

```{r}
cell_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  CD3G = "T cells",
  IL7R = "CD4+ T",
  CD4 = "CD4+ T",
  CD8A = "CD8+ T",
  CD8B = "CD8+ T",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD19 = "B cells",
  CD14 = "CD14+ Mono",
  LYZ = "CD14+ Mono",
  S100A8 = "CD14+ Mono",
  FCGR3A = "CD16+ Mono",
  CDKN1C = "CD16+ Mono",
  GNLY = "NK",
  NKG7 = "NK",
  NCAM1 = "NK",
  FCER1A = "mDC",
  CD1C = "mDC",
  LILRA4 = "pDC",
  CLEC4C = "pDC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)
```

Let's annotate the PBMCs using the scType cell type annotation algorithm by [Ianevski et al. (2022)](https://doi.org/10.1038/s41467-022-28803-w). We first prepare the cell type markers into the right format using `prepare_cell_markers` helper functions. Subsequently calculate the scType score using `calc_sc_type_scores`and summarize on cluster level to get the annotated cell type using `score_clusters`. Finally, we can assign back the cell type annotation into the `sc_object`.

```{r}
cell_markers_dt <- stack(cell_markers) %>%
  as.data.table() %>%
  setnames(., c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var$gene_id[match(gene_symbol, var$gene_symbol)]] %>%
  .[!is.na(gene_id), ]

marker_list <- prepare_cell_markers(sc_object, cell_markers_dt)
sctype_scores <- calc_sc_type_scores(
  object = sc_object,
  cell_marker_list = marker_list
)

```{r}
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "sc_type",
  label_by = "sc_type",
  discrete = F,
  point_alpha = 0.2
)
```

Alternatively, we can also show the expression levels of the cell type markers using the dotplot plotting function `dot_plot_sc` from `bixverse.plots`. 

```{r}
features_vec <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels = ensembl_to_symbol[features_vec]

dot_plot_sc(
  object = sc_object,
  features = features_vec,
  feature_labels = ensembl_to_symbol[features_vec],
  grouping_variable = "leiden_clusters",
  scale_exp = TRUE,
  feature_grouping = cell_markers,
  cluster_groups = T
)
```

### Per-cell gene expression

In addition to the feature plot (`feature_plot_sc`), we can also represent the gene expression using a violin plot 

```{r}
#| label: feature plot
#| fig-width: 10
#| fig-height: 4
genes <- c(
  "CD3D",
  "IL7R",
  "CD8A",
  "MS4A1"
)
stacked_violin_plot_sc(
  sc_object,
  features = ensembl_to_symbol[genes],
  # feature_labels = symbol_to_ensembl[genes],
  grouping_variable = "leiden_clusters"
)
```

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```