---
title: "Analysing PBMCs with bixverse"
vignette: >
  %\VignetteIndexEntry{Analysing PBMCs with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Grant Neilson and Gregor Lueg"
---

## Intro

This vignette walks through a standard single cell analysis on the PBMC3k data
set using `bixverse`. If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_intro.html),
please do so first; this vignette assumes familiarity with how the
`SingleCells` class, on-disk storage and cells-to-keep logic work.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
```

## Loading the data

We start by downloading the PBMC3k data set bundled with the package and
loading it via Cell Ranger-style MTX I/O.

```{r}
#| label: load pbmc3k

pbmc3k_path <- bixverse:::download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  streaming = FALSE,
  .verbose = TRUE
)

sc_object
```

Let's have a quick look at the variable table. The column names from the MTX
files are not always informative, so we rename the gene symbol column to
something sensible and set up mappings between Ensembl IDs and symbols.

```{r}
#| label: var table and renaming

var <- get_sc_var(sc_object)

head(var)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

head(var)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
```

## Quality control

### Gene set proportions

A typical first step is computing the proportion of counts mapping to
mitochondrial and ribosomal genes per cell. These are added directly to the
obs table in the DuckDB.

```{r}
#| label: gene set proportions

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

sc_object[[1:5L]]
```

As we can see we now have MT and Ribo as columns in the obs table.

### MAD outlier detection

We use per-cell QC metrics and MAD-based outlier detection to flag problematic
cells. The `run_cell_qc` function returns a `CellQc` object that carries the
metrics, per-metric outlier calls and a combined outlier vector.

```{r}
#| label: qc outlier detection

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(metrics, directions, threshold = 3)

qc
```

The `CellQc` class has a `plot` method that produces violin plots with outliers
highlighted.

```{r}
#| label: qc plots
#| fig-width: 12
#| fig-height: 4

plots <- plot(qc, qc_df)

plots$log10_lib_size + plots$log10_nnz + plots$MT
```

### Filtering cells

We store the outlier flag in the obs table and then set the cells to keep. From
this point on, all downstream methods (HVG selection, PCA, etc.) will only
operate on the retained cells.

```{r}
#| label: set cells to keep

sc_object[["outlier"]] <- qc$combined

cells_to_keep <- qc_df[!qc$combined, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object
```

## Feature selection, PCA and neighbours

With QC done, we move through the standard pipeline: highly variable gene
selection, PCA, and nearest neighbour computation.

```{r}
#| label: hvg pca neighbours

sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
```

## Clustering and marker detection

Leiden clustering followed by differential gene expression across all clusters.

```{r}
#| label: clustering and markers

sc_object <- find_clusters_sc(sc_object, res = 1.5, name = "leiden_clusters")

all_markers <- find_all_markers_sc(
  object = sc_object,
  column_of_interest = "leiden_clusters"
)

all_markers[, gene_symbol := ensembl_to_symbol[gene_id]]

head(all_markers[fdr <= 0.05][order(-abs(lfc))])
```

## Visualisation

### Dimensionality reduction

Let us run quickly UMAP and tSNE.

```{r}
#| label: umap and tsne

sc_object <- umap_sc(sc_object, knn_method = "annoy")

sc_object <- tsne_sc(sc_object, perplexity = 30, knn_method = "annoy")
```

We can pull the embeddings back into data.tables and plot them (longer term,
the idea will be to provide plotting helpers in `bixverse.plots`). The obs 
columns can be appended directly.

```{r}
#| label: umap plot
#| fig-width: 7
#| fig-height: 5

umap_dt <- as.data.table(
  get_embedding(sc_object, "umap"),
  keep.rownames = "cell_id"
)[, leiden_clusters := sc_object[["leiden_clusters"]]]

ggplot(umap_dt, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = as.factor(leiden_clusters)), size = 0.5) +
  theme_bw() +
  labs(colour = "Leiden cluster")
```

Let's check out tSNE

```{r}
#| label: tsne plot
#| fig-width: 7
#| fig-height: 5

tsne_dt <- as.data.table(
  get_embedding(sc_object, "tsne"),
  keep.rownames = "cell_id"
)[, leiden_clusters := sc_object[["leiden_clusters"]]]

ggplot(tsne_dt, aes(x = tsne_1, y = tsne_2)) +
  geom_point(aes(colour = as.factor(leiden_clusters)), size = 0.5) +
  theme_bw() +
  labs(colour = "Leiden cluster")
```

### Dot plots

Dot plots are a standard way to visualise marker gene expression across
clusters. The `extract_dot_plot_data` method handles the Rust-level extraction,
grouping statistics and optional min-max scaling in one call.

```{r}
#| label: dot plot
#| fig-width: 8
#| fig-height: 6

cell_markers <- c(
  "MS4A1",
  "CD79B",
  "GNLY",
  "CD3D",
  "CD8A",
  "CD4",
  "CD14",
  "FCER1A",
  "FCGR3A",
  "LYZ",
  "PPBP"
)

features <- symbol_to_ensembl[cell_markers]

dot_dt <- extract_dot_plot_data(
  object = sc_object,
  features = features,
  grouping_variable = "leiden_clusters",
  scale_exp = TRUE
)

dot_dt[, gene_symbol := ensembl_to_symbol[as.character(gene)]]
dot_dt[, gene_symbol := factor(gene_symbol, levels = rev(cell_markers))]

ggplot(dot_dt, aes(x = group, y = gene_symbol)) +
  geom_point(aes(size = pct_exp, colour = scaled_exp)) +
  scale_colour_viridis_c() +
  scale_size_continuous(range = c(0, 6)) +
  theme_bw() +
  labs(
    size = "% expressed",
    colour = "Scaled\nexpression",
    x = "Cluster",
    y = ""
  )
```

### Per-cell gene expression

For feature plots or other per-cell visualisations, `extract_gene_expression`
returns a long data.table with one row per cell and columns for each requested
gene, plus any obs metadata you need.

```{r}
#| label: feature plot
#| fig-width: 10
#| fig-height: 4

expr_dt <- extract_gene_expression(
  object = sc_object,
  features = symbol_to_ensembl[c("CD14", "MS4A1", "CD3E")],
  obs_cols = "leiden_clusters"
)

# attach UMAP coordinates
umap_embd <- as.data.table(
  get_embedding(sc_object, "umap"),
  keep.rownames = "cell_id"
)

plot_dt <- merge(expr_dt, umap_embd, by = "cell_id")

# melt to long for faceting
gene_cols <- symbol_to_ensembl[c("CD14", "MS4A1", "CD3E")]

plot_long <- melt(
  plot_dt,
  id.vars = c("cell_id", "umap_1", "umap_2"),
  measure.vars = gene_cols,
  variable.name = "gene",
  value.name = "expression"
)
plot_long[, gene_symbol := ensembl_to_symbol[as.character(gene)]]

ggplot(plot_long, aes(x = umap_1, y = umap_2)) +
  geom_point(aes(colour = expression), size = 0.3) +
  scale_colour_viridis_c() +
  facet_wrap(~gene_symbol) +
  theme_bw() +
  labs(colour = "log1p\nexpression")
```

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```