---
title: "Bixverse R <> Rust interface"
vignette: >
  %\VignetteIndexEntry{Bixverse R <> Rust interface}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Why Rust?

Honest answer? The creation of this package was based on the desire of the
author to play with Rust. To make it useful, the initial playground was stuff
relevant to day-to-day work in bioinformatics and computational biology:
correlations, PCA, other matrix factorisation methods, and so on. It quickly
became apparent that Rust delivered incredible speed gains over R — sometimes
100x — with millions of hypergeometric tests completing in mere seconds. The
results were striking. Memory safety and the fantastic work by the
[rextendr team](https://github.com/extendr/rextendr) made interfacing R and
Rust straightforward and, frankly, quite fun.

### Correlations and covariance

One of the most typical operations in computational biology is computing
correlations and covariances; and these were the first functions implemented,
largely to understand [faer](https://github.com/sarah-ek/faer-rs) and the
Rust/R interface. What becomes apparent very quickly is the following. For
1000 x 1000 matrices on an M1 Max MacBook Pro (using Apple's Accelerate
framework for R BLAS):

**Covariance**

| Implementation | Median (ms, 20 runs) |
|--:|--:|
| base R | 515.73 |
| Rust | 18.66 |

*A 27x increase in speed.*

**Pearson correlation**

| Implementation | Median (ms, 20 runs) |
|--:|--:|
| base R | 512.83 |
| Rust | 20.90 |

*A 24x increase in speed.*

**Spearman correlation**

| Implementation | Median (ms, 20 runs) |
|--:|--:|
| base R | 603.00 |
| Rust | 29.47 |

*A 20x increase in speed.*

The gains are even more pronounced on larger matrices. For a 1000 x 5000
matrix:

**Pearson correlation (1000 x 5000)**

| Implementation | Median (ms, 20 runs) |
|--:|--:|
| base R | 12994.37 |
| Rust | 419.88 |

*A 30x increase in speed.*

Given that matrix algebra underlies a large proportion of computational
bioinformatics workflows, this makes a compelling case for Rust under the
hood. The rest of this vignette shows how to use the exposed Rust functions
directly, with runnable benchmarks you can verify on your own machine.

## Using the Rust functions in more general R code

`bixverse` exposes a number of R/Rust functions that are useful beyond the
higher-level package functionality. This vignette covers:

* Correlations (via `rs_cor()`).
* Covariance (via `rs_covariance()`).
* Covariance to correlation (via `rs_cov2cor()`).
* Distance calculations (via `rs_dist()`).
* Correlations between two matrices (via `rs_cor2()`).
* Mutual information between columns (via `rs_mutual_information()`).
* Set similarities (via `rs_set_similarity_list()` or `rs_set_similarity()`).

These may be useful for your own package development.

::: {.callout-note}
The vignette was built on a GitHub runner which is a 2-core machine with ~8 GB
of memory. The performance differences tend to be (even) more pronounced on 
machines with more cores and computational power.
:::

## Set-up
```{r}
#| label: setup
library(bixverse)
```

Any function prefixed with `rs_` calls directly into Rust and can panic if
misused — check inputs before passing them through. There is a LOT of these
functions in the package, and they are all exported. Why? I want other people
to be able to integrate the code into their own R packages, or if they feel 
brave, they can just use the Rust crate directly, see 
[here](https://crates.io/crates/bixverse-rs).

### Correlations, covariance, and cosine distance

```{r}
#| label: generate random data

no_rows <- 500L
no_cols <- 500L

set.seed(10101L)

random_data <- matrix(
  data = rnorm(no_rows * no_cols),
  ncol = no_cols,
  nrow = no_rows
)
```

#### Pearson's correlation

```{r}
#| label: pearson correlations

r_pearsons_res <- cor(random_data)

rust_pearsons_res <- rs_cor(random_data, spearman = FALSE)

# tiny numerical differences do exist
all.equal(r_pearsons_res, rust_pearsons_res, tolerance = 1e-15)
```

Let's check the speed.

```{r}
#| label: pearson correlations (speed)
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = cor(random_data),
  rust = rs_cor(random_data, spearman = FALSE),
  times = 10L
)
```

#### Spearman's correlation

What about ranked correlations?

```{r}
#| label: spearman correlations

r_spearman_res <- cor(random_data, method = "spearman")

rust_spearman_res <- rs_cor(random_data, spearman = TRUE)

all.equal(r_spearman_res, rust_spearman_res, tolerance = 1e-15)
```

And speed

```{r}
#| label: spearman correlations (speed)
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = cor(random_data, method = "spearman"),
  rust = rs_cor(random_data, spearman = TRUE),
  times = 10L
)
```

#### Covariance

Covariance matrices are also very common...

```{r}
#| label: covariance

r_covar_res <- cov(random_data)

rust_covar_res <- rs_covariance(random_data)

all.equal(r_covar_res, rust_covar_res, tolerance = 1e-15)
```

And speed comparisons

```{r}
#| label: covariance (speed)
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = cov(random_data),
  rust = rs_covariance(random_data),
  times = 10L
)
```

#### Covariance to correlation

```{r}
#| label: covariance to correlation

r_cor_from_covar <- cov2cor(r_covar_res)

rust_cor_from_covar <- rs_cov2cor(rust_covar_res)

all.equal(r_cor_from_covar, rust_cor_from_covar, tolerance = 1e-15)
```

This is where the Rust implementation shows the least improvement — the
operation is already cheap in base R — but the interface remains consistent
with the rest of the `rs_` family.

```{r}
#| label: covariance to correlation (speed)
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = cov2cor(r_covar_res),
  rust = rs_cov2cor(rust_covar_res),
  times = 10L
)
```

#### Correlations between two matrices

So far, all of the functions did pairwise column-based correlations. What about
two correlations against each other?

```{r}
#| label: generate a second random data

no_rows_2 <- 500L
no_cols_2 <- 400L

set.seed(23L)

random_data_2 <- matrix(
  data = rnorm(no_rows_2 * no_cols_2, mean = 2, sd = 2),
  ncol = no_cols_2,
  nrow = no_rows_2
)
```

Let's compare against R:

```{r}
#| label: R vs Rust second correlations
r_cor_2_matrices <- cor(random_data, random_data_2)

rust_cor_2_matrices <- rs_cor2(random_data, random_data_2, spearman = FALSE)

# small precision differences are further propagated here
all.equal(r_cor_2_matrices, rust_cor_2_matrices, tolerance = 1e-14)
```

And speed:

```{r}
#| label: correlations between two matrices (speed)
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = cor(random_data, random_data_2),
  rust = rs_cor2(random_data, random_data_2, spearman = FALSE),
  times = 10L
)
```

### Distance metrics

#### Euclidean distance

Note that `dist()` computes row-wise distances, while `rs_dist()` expects
columns to represent observations — hence the transpose.

```{r}
#| label: euclidean distance

r_euclidean_distance <- as.matrix(
  dist(random_data, method = "euclidean")
)

rs_euclidean_distance <- rs_dist(
  t(random_data),
  distance_type = "euclidean"
)

all.equal(
  r_euclidean_distance,
  rs_euclidean_distance,
  tolerance = 1e-14,
  check.attributes = FALSE
)
```

And speed...

```{r}
#| label: euclidean distance speed
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = dist(random_data, method = "euclidean"),
  rust = rs_dist(random_data, distance_type = "euclidean"),
  times = 10L
)
```

#### Manhattan distance

```{r}
#| label: manhattan distance

r_manhattan_distance <- as.matrix(
  dist(random_data, method = "manhattan")
)

rs_manhattan_distance <- rs_dist(
  t(random_data),
  distance_type = "manhattan"
)

all.equal(
  r_manhattan_distance,
  rs_manhattan_distance,
  tolerance = 1e-14,
  check.attributes = FALSE
)
```

And speed... ?

```{r}
#| label: manhattan distance speed
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = dist(random_data, method = "manhattan"),
  rust = rs_dist(random_data, distance_type = "manhattan"),
  times = 10L
)
```

#### Canberra distance

```{r}
#| label: canberra distance

r_canberra_distance <- as.matrix(
  dist(random_data, method = "canberra")
)

rs_canberra_distance <- rs_dist(
  t(random_data),
  distance_type = "canberra"
)

all.equal(
  r_canberra_distance,
  rs_canberra_distance,
  tolerance = 1e-14,
  check.attributes = FALSE
)
```

Same benchmarks as before...

```{r}
#| label: canberra distance speed
#| eval: !expr requireNamespace("microbenchmark", quietly = TRUE)

microbenchmark::microbenchmark(
  r = dist(random_data, method = "canberra"),
  rust = rs_dist(random_data, distance_type = "canberra"),
  times = 10L
)
```

### Mutual information

```{r}
#| label: generate new smaller data

set.seed(246L)

nrows <- 100
ncols <- 100

mat <- matrix(data = rnorm(nrows * ncols), nrow = nrows, ncol = ncols)
rownames(mat) <- sprintf("sample_%i", 1:nrows)
colnames(mat) <- sprintf("feature_%i", 1:ncols)
```

The Rust version defaults to `sqrt(nrow())` bins when `n_bins = NULL`. Two
discretisation strategies are available.

**Equal width:**

```{r}
#| label: mutual information calculations - equal width
#| eval: !expr requireNamespace("infotheo", quietly = TRUE)

rust_res_mi <- rs_mutual_info(
  mat,
  n_bins = NULL,
  normalise = FALSE,
  strategy = "equal_width"
)
rownames(rust_res_mi) <- colnames(rust_res_mi) <- colnames(mat)

infotheo_res_mi <- infotheo::mutinformation(infotheo::discretize(
  mat,
  disc = "equalwidth",
  nbins = sqrt(nrow(mat))
))

all.equal(rust_res_mi, infotheo_res_mi)
```

```{r}
#| label: speed differences bixverse vs. infotheo (equal width)
#| eval: !expr requireNamespace(c("infotheo", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  infotheo = infotheo::mutinformation(infotheo::discretize(
    mat,
    disc = "equalwidth",
    nbins = sqrt(nrow(mat))
  )),
  rust = rs_mutual_info(
    mat,
    n_bins = NULL,
    normalise = FALSE,
    strategy = "equal_width"
  ),
  times = 10L
)
```

**Equal frequency:**

```{r}
#| label: mutual information calculations - equal frequency
#| eval: !expr requireNamespace("infotheo", quietly = TRUE)

rust_res_mi <- rs_mutual_info(
  mat,
  n_bins = NULL,
  normalise = FALSE,
  strategy = "equal_freq"
)
rownames(rust_res_mi) <- colnames(rust_res_mi) <- colnames(mat)

infotheo_res_mi <- infotheo::mutinformation(infotheo::discretize(
  mat,
  disc = "equalfreq",
  nbins = sqrt(nrow(mat))
))

all.equal(rust_res_mi, infotheo_res_mi)
```

```{r}
#| label: speed differences bixverse vs. infotheo (equal frequency)
#| eval: !expr requireNamespace(c("infotheo", "microbenchmark"), quietly = TRUE)

microbenchmark::microbenchmark(
  infotheo = infotheo::mutinformation(infotheo::discretize(
    mat,
    disc = "equalfreq",
    nbins = sqrt(nrow(mat))
  )),
  rust = rs_mutual_info(
    mat,
    n_bins = NULL,
    normalise = FALSE,
    strategy = "equal_freq"
  ),
  times = 10L
)
```

### Set similarities

`bixverse` also exposes Rust-accelerated set similarity calculations. To
illustrate the performance, below is a comparison against a naive R
implementation of Jaccard similarity across two lists.

```{r}
#| label: R similarity data

jaccard_sim <- function(x, y) {
  length(intersect(x, y)) / length(union(x, y))
}

set.seed(42L)

random_sets_1 <- purrr::map(
  1:1000L,
  ~ paste(sample(LETTERS, sample(1:20, 1)))
)

random_sets_2 <- purrr::map(
  1:500L,
  ~ paste(sample(LETTERS, sample(1:20, 1)))
)
```

Starting with a sequential [purrr](https://purrr.tidyverse.org) approach as
the baseline:
```{r}
#| label: set similarity in R (purrr)
#| eval: !expr requireNamespace(c("tictoc"), quietly = TRUE)

tictoc::tic()

r_results <- purrr::map(
  random_sets_1,
  \(x) {
    purrr::map_dbl(random_sets_2, \(y) jaccard_sim(x, y))
  },
  .progress = TRUE
)

similarity_matrix <- matrix(
  data = unlist(r_results),
  nrow = length(random_sets_1),
  ncol = length(random_sets_2),
  byrow = TRUE
)

tictoc::toc()
```

Parallelising via [furrr](https://furrr.futureverse.org) gives a meaningful
speedup:

```{r}
#| label: set similarity in R (furrr)
#| eval: !expr requireNamespace(c("future", "tictoc", "furrr"), quietly = TRUE)

future::plan(strategy = future::multisession(workers = parallel::detectCores()))

tictoc::tic()

r_results <- furrr::future_map(
  random_sets_1,
  \(x) {
    purrr::map_dbl(random_sets_2, \(y) jaccard_sim(x, y))
  },
  .progress = TRUE
)

similarity_matrix <- matrix(
  data = unlist(r_results),
  nrow = length(random_sets_1),
  ncol = length(random_sets_2),
  byrow = TRUE
)

tictoc::toc()

future::plan(strategy = future::sequential())
```

[mirai](https://mirai.r-lib.org/articles/mirai.html) reduces process
scheduling overhead (somtimes) compared to `future` and does a bit better:

```{r}
#| label: set similarity in R (mirai)
#| eval: !expr requireNamespace(c("tictoc", "mirai"), quietly = TRUE)

mirai::daemons(parallel::detectCores())

tictoc::tic()

r_results_mirai <- mirai::mirai_map(
  random_sets_1,
  \(x, sets_2) {
    purrr::map_dbl(sets_2, \(y) {
      length(intersect(x, y)) / length(union(x, y))
    })
  },
  .args = list(sets_2 = random_sets_2)
)[]

similarity_matrix <- matrix(
  data = unlist(r_results_mirai),
  nrow = length(random_sets_1),
  ncol = length(random_sets_2),
  byrow = TRUE
)

tictoc::toc()

mirai::daemons(0)
```

Depending on your system this optimised parallel R version may be around 4x
faster than the naive sequential baseline. Now compare against the Rust
implementation:

```{r}
#| label: bixverse jaccard similarity function

tictoc::tic()

rust_res <- rs_set_similarity_list2(
  s_1_list = random_sets_1,
  s_2_list = random_sets_2,
  overlap_coefficient = FALSE
)

tictoc::toc()

all.equal(similarity_matrix, rust_res, tolerance = 1e-15)
```

The margin here is not subtle. The functions shown in this vignette are a
representative sample of what is exposed in `bixverse`: there are many more
specialised functions throughout the package worth exploring (if you are
brave and can deal with a panic here and there).