---
title: "Scaling to millions of cells with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Scaling to millions of cells with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette demonstrates how to run a single cell analysis on a data set
containing nearly 1 million cells on a local machine... In this case, a MacBook
Air with 24 GBs (if you think *"IMPOSSIBLE!"*, you would not be wrong coming
from other languages/frameworks). If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the [introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first. The [PBMC3k walkthrough](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html)
covers every step in detail on a small data set; here we focus on what changes
when the cell count moves from thousands to a million.

The data used here is the 1 million cell PBMC data set from Parse Biosciences,
a commonly used reference for showing large single cell analyses, see the
original article
[here](https://support.parsebiosciences.com/hc/en-us/articles/7704577188500-How-to-analyze-a-1-million-cell-data-set-using-Scanpy-and-Harmony).
**The important bit here is that this was run successfully on a MacBook Air with
24 GB (M3)**. And please contrast that with the following line for the analysis
of this data:

<i>To avoid computer crashes and program failures, we recommend you execute the
Python code on a cloud computing platform. The instance you select on the cloud
platform to run this analysis should consist of at least 8 threads and 160 GB of
RAM.</i> - Parse Biosciences

::: {.callout-note}
None of the code chunks below are evaluated. This vignette is a reference for
how one *would* run such an analysis, not a live rendering - the data set is
far too large for CI/CD runners and GitHub would not be happy.
:::

If you want to reproduce this, download the data from Parse Biosciences (or
any other fat data set you have) and go have fun.

```{bash}
#| label: get data
#| eval: false
wget "https://cdn.parsebiosciences.com/1M_PBMC_T1D_Parse.zip"
```

```{r}
#| label: setup
#| eval: false
# main packages
library(bixverse)
library(bixverse.plots)

# GPU-acceleration (experimental)
library(bixverse.gpu)

# Plotting, data manipulation
library(ggplot2)
library(data.table)
```

## Loading the data

Once you have unzipped the data, we can start streaming it into memory and
writing the files to disk.

```{r}
#| label: stream mtx
#| eval: false
data_path <- "~/Desktop/parse_pbmc_1m/"

load_params <- params_sc_mtx_io(
  path_mtx = path.expand(file.path(data_path, "DGE_1M_PBMC.mtx")),
  path_obs = path.expand(file.path(data_path, "cell_metadata_1M_PBMC.csv")),
  path_var = path.expand(file.path(data_path, "all_genes_1M_PBMC.csv")),
  cells_as_rows = TRUE,
  has_hdr = TRUE
)

# We will write everything into the temporary directory
sce <- SingleCells(dir_data = tempdir())

# We enforce level 2 streaming here which will run for a bit longer, but ensure
# your memory never blows up
sce <- load_mtx(object = sce, sc_mtx_io_param = load_params, streaming = 2L)
```

This step is the slowest... A lot of things are actually happening here.

- First, (streaming) scan to understand in how many cells a gene is expressed.
  Only keep genes that are expressed in whatever you set to `min_cells`.
- Second, (streaming) scan, given the genes from the first step, which cells to
  include based on minimum library size and minimum features.
- Third, now that it's clear which cells/genes to include, create quick temp
  files with the cells to keep with the genes already re-indexed. This way we
  avoid loading in the massive MTX file in one go into memory. These files are
  automatically cleaned up at the end and loaded sequentially and written to the
  CSR-type file (while the normalisation is also done on the fly).
- Lastly, load in the cell data in small chunks and do a transpose to a CSC
  style format for easy gene retrieval.

This approach avoids loading the whole data into memory at any point. The DuckDB 
also gets populated with the metadata in the observations (based on what has 
been filtered) and the vars (add another 1 to 2 minutes here). This part -
streaming the data and processing it - is usually what takes the most time. If
you have already loaded in the data and it exists on-disk from a previous
session, skip the streaming step entirely and reconnect.

```{r}
#| label: reconnect
#| eval: false
# the folder is hopefully not temp...
sce <- SingleCells(dir_data = tempdir())

sce <- load_existing(object = sce)
```

## Doublet detection

Before any QC on library sizes and gene-set proportions, we run Scrublet to
flag doublets. Doublets are called per sample via the `group_by` argument,
which is what you want when sample-level differences in capture rates would
otherwise distort a single global score. Results are merged back onto the cell
metadata and the doublets are dropped from the set of cells to keep. The
runs over the individual samples are occuring sequentially to not overwhelm
memory and each run is massively parallelised internally in Rust.

```{r}
#| label: doublet detection
#| eval: false
scrublet_res <- scrublet_sc(object = sce, group_by = "sample")

sce <- add_sc_new_obs(
  object = sce,
  obs_data = get_data(scrublet_res)
)

cells_without_doublets <- sce[[c("doublet", "cell_id")]][
  !(doublet),
  cell_id
]

sce <- set_cells_to_keep(x = sce, cells_to_keep = cells_without_doublets)
```

## Quality control

### Gene set proportions

The same gene set proportion logic from the PBMC3k vignette applies. The
functions automatically recognise if you have more than 100k cells and will
leverage the streaming engine by default in this case. You do have more manual
control via the `streaming = NULL` parameter. If you have a fat memory machine,
you can set this to `FALSE` despite a large data set.

```{r}
#| label: gene set proportions
#| eval: false
var <- get_sc_var(object = sce)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_name), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_name), gene_id]
)

sce <- gene_set_proportions_sc(
  object = sce,
  gene_set_list = gs_of_interest,
  .verbose = TRUE
)
```

### MAD outlier detection

Outlier detection on library size, the number of detected genes, and
mitochondrial proportion. We compute outliers per sample (`groups = sample`),
which matters once batches have non-trivial differences in sequencing depth -
a global threshold would otherwise penalise whole batches rather than
individual cells within them.

```{r}
#| label: qc outlier detection
#| eval: false
qc_df <- sce[[c("cell_id", "sample", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "below",
  log10_nnz = "below",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sce),
  directions = directions,
  threshold = 3,
  groups = qc_df$sample
)

sce <- add_sc_new_obs(sce, obs_data = get_data(qc))
```

I cannot recommend trying to plot the metric distributions naively...
ggplot2 will not be happy at this cell count.

### Filtering cells

```{r}
#| label: set cells to keep
#| eval: false
cells_to_keep <- sce[[c("global_outlier", "cell_id")]][
  !(global_outlier),
  cell_id
]

sce <- set_cells_to_keep(sce, cells_to_keep)
```

## Feature selection and PCA

HVG selection at this scale again requires `streaming = TRUE` (if left as is,
the auto-detection will take care of this and automatically use streaming when
more than 1e5 cells are detected). There are two data passes to avoid 
materialising large data in memory: the first calculates means and standard 
deviations, and the second does the variance stabilisation.

```{r}
#| label: hvg
#| eval: false
sce <- find_hvg_sc(
  object = sce,
  hvg_no = 2000L,
  .verbose = TRUE
)
```

PCA is one of the points where GPU acceleration starts to noticeably matter
(if you want to install the GPU-accelerated versions... If you have some VRAM 
or unified memory to spare it will make a difference).
Both implementations share the same parameter interface, so swapping between
them is a one-line change. The leading components should agree up to sign
between the two, which is worth a quick visual check.

```{r}
#| label: pca cpu
#| eval: false
sce <- calculate_pca_sc(
  object = sce,
  pca_params = params_sc_pca(),
  no_pcs = 32L
)

cpu_res <- get_pca_factors(sce)
cpu_s <- get_pca_singular_val(sce)
```

Let's compare this against the GPU-accelerated version. In this case, the GEMMs
prior to the thin SVD on the massively reduced matrix in the randomised SVD
are run on the GPU. You pay for moving the data, but you should still see
an acceleration compared to the CPU form.

```{r}
#| label: pca gpu
#| eval: false
sce <- calculate_pca_gpu_sc(
  object = sce,
  pca_params = params_sc_pca(),
  no_pcs = 32L,
  .verbose = 2L
)

gpu_res <- get_pca_factors(sce)
gpu_s <- get_pca_singular_val(sce)

# sanity check on the leading component
plot(cpu_res[, 1], gpu_res[, 1])
```

In practice you would pick one. The GPU path becomes the more attractive
option as cell counts grow; on this data set you can feel the difference (if
VRAM permits). You will see some slight differences due to floating operation
differences between CPU and GPU.

## Nearest neighbours (GPU-accelerated)

At nearly 1 million cells, nearest neighbour search is the main computational
bottleneck. The companion package [bixverse.gpu](https://github.com/GregorLueg/bixverse.gpu)
exposes a CAGRA-style GPU-accelerated approximate nearest neighbour search
that makes this quite fast on a laptop with a discrete GPU - thanks to CubeCL
and WGPU, as long as you can fit the data into VRAM (should be feasible with
8 GBs - I think?) this should run quite fast. CPU-side alternatives such as
`"nndescent"` or `"hnsw"` also work but will be considerably slower at this
scale. For empirical reference:

- GPU version on M1 Max with 64 GB takes ~1 minute.
- HNSW somewhere between 3 to 4 minutes.

Likely still comparably much faster if you are used to the more established
single cell methods in terms of CPU speed, but yeah... If you have a GPU,
go use that.

```{r}
#| label: cpu knn
#| eval: false
sce <- find_neighbours_sc(
  object = sce,
  embd_to_use = "pca",
  neighbours_params = params_sc_neighbours(knn = list(knn_method = "hnsw"))
)
```

```{r}
#| label: gpu knn
#| eval: false
sce <- find_neighbours_cagra_sc(
  object = sce,
  embd_to_use = "pca",
  cagra_params = params_sc_cagra(),
  .verbose = TRUE
)
```

## Embeddings before batch correction

### Batch effect quantification

A kBET score on the uncorrected neighbour graph gives a quantitative baseline
to compare against once batch correction is applied...

```{r}
#| label: kbet prior
#| eval: false
kbet_prior <- calculate_kbet_sc(object = sce, batch_column = "sample")

kbet_prior
```

### Plots

Or if you are a visual person, let's check UMAP and tSNE

```{r}
#| label: embeddings prior umap
#| eval: false
sce <- umap_sc(
  sce,
  slot_name = "umap_prior"
)
```

```{r}
#| label: embeddings prior tsne
#| eval: false
sce <- tsne_sc(
  sce,
  slot_name = "tsne_prior",
  # use FFT here! It will make a massive difference
  approx_type = "fft",
  knn_method = "hnsw"
)
```

You might wonder why in the case of tSNE we regenerate the kNN graph. You do not
have to, but the rule-of-thumb is `k_neighbours = perplexity * 3.0`. With the
default perplexity, we would need 30 neighbours, but we only returned 15. So,
we just rerun this. Skip tSNE if you don't need it, but the FFT-accelerated
version is substantially faster than you would expect from tSNE.

```{r}
#| label: plot prior - umap
#| eval: false
embedding_plot_sc(
  sce,
  embedding = "umap_prior",
  colour_by = "sample",
  discrete = TRUE
) +
  labs(
    title = "No batch correction",
    colour = "Batch:"
  )
```

```{r}
#| label: plot prior - tsne
#| eval: false
embedding_plot_sc(
  sce,
  embedding = "tsne_prior",
  colour_by = "sample",
  discrete = TRUE
) +
  labs(
    title = "No batch correction",
    colour = "Batch:"
  )
```

## Batch correction

Okay, there is clearly a big sample batch effect. Let's remove that one. Again
you have two options here... Harmony ([version 2]() - recommended) on CPU or
a GPU-accelerated version. Let's check out the CPU version first...

```{r}
#| label: harmony cpu
#| eval: false
sce <- harmony_v2_sc(object = sce, batch_column = "sample")
```

If you have a GPU at hand, you can also just use the GPU version. Again, you
will observe some slight numerical differences... This is due to the GPU
implementation, but the overall structure of the data will be recovered in the
same way.

```{r}
#| label: harmony gpu
#| eval: false
sce <- harmony_v2_gpu_sc(object = sce, batch_column = "sample")
```

Let's overwrite the kNN graph with the now batch-corrected embedding:

```{r}
#| label: regeneration of the knn graph
#| eval: false
sce <- find_neighbours_cagra_sc(
  object = sce,
  embd_to_use = "harmony_gpu", # or harmony_v2 if you just did the CPU version
  cagra_params = params_sc_cagra(),
  .verbose = TRUE
)
```

### Batch correction quantification

Let's requantify the kBET scores and see if we removed some of the batch 
effects...

```{r}
#| label: kbet post batch correction
#| eval: false
kbet_post_gpu <- calculate_kbet_sc(object = sce, batch_column = "sample")

kbet_post_gpu
```

## Differential gene expression

TODO.

## Key differences from the small-data workflow

The API surface is intentionally identical to the PBMC3k walkthrough. The
meaningful differences when working at the million-cell scale are:

1. **Streaming I/O.** Use `load_mtx(..., streaming = 2L)` (or the equivalent
   on whatever loader fits your input), and the streaming behaviour propagates
   through `gene_set_proportions_sc` and `find_hvg_sc`. This is what makes
   working with this data set on a laptop possible at all.
2. **On-disk persistence.** `load_existing` lets you reconnect to a previously
   streamed data set without re-reading the source. (Works for any data.)
3. **Per-sample QC.** With batches comes the need to compute outliers per
   sample rather than globally.
4. **Doublet detection per sample.** Scrublet is run with `group_by`, which
   matters once capture rates differ between batches.
5. **GPU-accelerated PCA and kNN.** `bixverse.gpu` provides GPU paths for
   PCA, CAGRA-based nearest neighbour search, and Harmony. The CPU versions
   remain highly optimised with SIMD and cache-aware memory layouts, but the
   GPU paths get faster as the data scales.

For context... Running this on my machine (MacBook Pro M1 Max with 64 GB)
took roughly 20 minutes from start to finish, with about half of the time
spent on data I/O. Memory pressure remained green throughout.
