---
title: "A guide to single-cell plotting functions"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{A guide to single-cell plotting functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Liesbeth François"
---

## Intro

This vignette gives an overview of the different plotting functions that are 
available when running a single cell analysis. 

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(bixverse.plots)
library(magrittr)
```

## Preparing the data

We run the standard analysis using the same PBMC3k data set. For more details on 
the data analysis itself, please refer to the 
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html). 
This vignette assumes familiarity with how the `SingleCells` class, on-disk 
storage and cells-to-keep logic work.

```{r}
#| label: prepare-pbmc3k
#| code-fold: TRUE
pbmc3k_path <- download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = TRUE
)

sc_object[["sample"]] <- sample(
  x = c("sample1", "sample2"),
  size = nrow(sc_object),
  replace = TRUE
)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)

scrublet_res <- scrublet_sc(sc_object)

sc_object <- add_sc_new_obs(
  object = sc_object,
  obs_data = get_data(scrublet_res)
)

cells_without_doublets <- sc_object[[c("doublet", "cell_id")]][
  !(doublet),
  cell_id
]

sc_object <- set_cells_to_keep(
  x = sc_object,
  cells_to_keep = cells_without_doublets
)

# Define gene sets whose proportional expression signals low quality
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

# Collect metrics for outlier detection
qc_df <- sc_object[[c("cell_id", "sample", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3,
  groups = qc_df$sample
)
```

## Joint QC plot

`joint_plot_sc()` compares the number of features and number of counts per cell 
similar to `sns.jointplot`.

```{r}
#| label: joint-plot
#| fig-cap: "Joint QC plot showing features versus counts."
joint_plot_sc(qc)
```

## Violin QC Plots 

`violin_plot_sc()` shows the distribution of each QC metric as a violin,
split by donor (or any grouping variable stored in the `qc` object).

```{r}
#| label: violin-plot
#| fig-cap: "Per-donor violin plots for library size, number of non-zero genes, and MT proportion."
violin_plot_sc(qc)
```

Alternatively, we can also provide a datatable generate the violins plots with 
different variables than those considered in the QC.

```{r}
#| label: violin-plot-ribo
obs <- get_sc_obs(sc_object, filtered = TRUE)

violin_plot_sc(
  obs,
  grouping_column = "sample",
  variable = "Ribo",
  show_outlier = TRUE
)
```

## Density QC Plots 

`density_plot_sc()` overlays smoothed density distributions for each group, 
which highlights bimodal distributions or outlier populations that violins can 
obscure.

```{r}
#| label: density-plot
#| fig-cap: "Density plots for each QC metric."
density_plot_sc(qc)
```

# Dimensionality Reduction Plots

After PCA, neighbour-finding, clustering, and UMAP, cells can be visualised on 
any 2D embedding stored in the object.

```{r}
#| label: umap-setup
#| code-fold: TRUE
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
sc_object <- find_clusters_sc(sc_object, res = 0.5, name = "leiden_clusters")
sc_object <- umap_sc(sc_object)

cell_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  CD3G = "T cells",
  IL7R = "CD4+ T",
  CD4 = "CD4+ T",
  CD8A = "CD8+ T",
  CD8B = "CD8+ T",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD19 = "B cells",
  CD14 = "CD14+ Mono",
  LYZ = "CD14+ Mono",
  S100A8 = "CD14+ Mono",
  FCGR3A = "CD16+ Mono",
  CDKN1C = "CD16+ Mono",
  GNLY = "NK",
  NKG7 = "NK",
  NCAM1 = "NK",
  FCER1A = "mDC",
  CD1C = "mDC",
  LILRA4 = "pDC",
  CLEC4C = "pDC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)

cell_markers_dt <- stack(cell_markers) %>%
  as.data.table() %>%
  setnames(., c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var$gene_id[match(gene_symbol, var$gene_symbol)]] %>%
  .[!is.na(gene_id), ]

## Prepare the list of markers
marker_list <- prepare_cell_markers(sc_object, cell_markers_dt)
## Calculate the score for each cell
sctype_scores <- calc_sc_type_scores(
  object = sc_object,
  cell_marker_list = marker_list
)
## Annotate the clusters
cell_type_anno <- score_clusters(
  sctype_scores,
  sc_object[[]][["leiden_clusters"]]
)

obs <- sc_object[[]][, .(cell_idx, leiden_clusters)]
add_sctype <- setNames(
  cell_type_anno$cell_type[match(
    obs$leiden_clusters,
    cell_type_anno$cluster_id
  )],
  nm = obs$cell_idx
)
sc_object[["sc_type"]] <- add_sctype
```

**Colouring by Cluster**

We can plot the UMAP embedding highlighting the leiden clusters. 

```{r}
#| label: embedding-leiden
#| fig-cap: "UMAP coloured by Leiden cluster number."
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

**Colouring by a discrete variable**

Pass any column from the cell metadata to `colour_by`. Use `label_by` to overlay 
text labels at the cluster centroid. These do not need to be the same variables.

```{r}
#| label: embedding-celltype
#| fig-cap: "UMAP coloured by manually annotated cell type. Labels are placed at cluster centroids."
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "sc_type",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

**Colouring by a continuous variable**

Set `discrete = FALSE` to use a continuous colour scale — useful for confidence 
scores, pseudotime, or gene module scores.

```{r}
#| label: embedding-sctype
#| fig-cap: "UMAP coloured by ribosomal content (continuous scale)."
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "Ribo",
  discrete = FALSE,
  point_alpha = 0.2
)
```

**Key arguments:**

| Argument | Description |
|---|---|
| `embedding` | Name of the stored embedding (`"umap"`, `"tsne"`, `"pca`…) |
| `colour_by` | Metadata column used to colour points |
| `label_by` | Metadata column used to label centroids |
| `discrete` | `TRUE` for categorical palettes, `FALSE` for continuous |
| `point_alpha` | Transparency (0–1); reduce for dense datasets |

**Extracting the plotting data**

Alternatively, you can also extract the plotting data directly to make 
customised plots.

```{r}
#| label: embedding-dt
dt <- extract_embedding_data(
  sc_object,
  embedding = "umap",
  obs_cols = c("sc_type", "leiden_clusters", "Ribo")
)
head(dt)
```

# Other 2D embeddings

UMAP is not the only game in town. All the 2D embeddings come from
[manifoldsR](https://github.com/GregorLueg/manifoldsR) and land in the same
object, so `embedding_plot_sc()` works on every one of them.

## Density-preserving embeddings

UMAP and t-SNE throw away local density. A tight, homogeneous population and a
diffuse, heterogeneous one come out at roughly the same size, so reading
anything into the area a cluster takes up on the plot is a mistake. densMAP
and den-SNE fix this with an extra term in the loss that keeps the local radius
of each cell in the embedding correlated with its radius in PCA space. Big blob
now means lots of variation, small tight blob means not much. `lambda` in
`manifoldsR::params_densmap()` and `manifoldsR::params_densne()` sets how hard
that term pulls; `lambda = 0` gives you plain UMAP and t-SNE back.

For the t-SNE flavours we use `approx_type = "fft_3k"`, the three-kernel FFT
optimiser. The FFT options only exist on Unix, on Windows stick to `"bh"`.

```{r}
#| label: density-embeddings
sc_object <- densmap_sc(sc_object, .verbose = FALSE)
sc_object <- tsne_sc(sc_object, approx_type = "fft_3k", .verbose = FALSE)
sc_object <- densne_sc(sc_object, approx_type = "fft_3k", .verbose = FALSE)
```

```{r}
#| label: density-plots
#| fig-width: 10
#| fig-height: 9
#| fig-cap: "UMAP and t-SNE (left) next to their density-preserving versions (right), coloured by cell type."
density_plots <- purrr::map(
  c("umap", "densmap", "tsne", "densne"),
  \(embd) {
    embedding_plot_sc(
      sc_object,
      embedding = embd,
      colour_by = "sc_type",
      discrete = TRUE
    ) +
      ggtitle(embd)
  }
)

patchwork::wrap_plots(density_plots, ncol = 2, guides = "collect")
```

## ForceAtlas2

ForceAtlas2 is a force-directed layout of the neighbour graph: edges pull,
every pair of cells pushes, gravity keeps the lot together. It is what scanpy's
`draw_graph()` does, and a common pick for data with continuous transitions
(think differentiation). `graph = "knn"` lays out the kNN graph the same
way scanpy does. `graph = "snn"` lays out the sNN graph the Leiden clustering
ran on, here started from the UMAP via `init_embd`.

```{r}
#| label: fa2-embeddings
sc_object <- forceatlas2_sc(sc_object, .verbose = FALSE)
sc_object <- forceatlas2_sc(
  sc_object,
  graph = "snn",
  init_embd = "umap",
  slot_name = "fa2_snn",
  .verbose = FALSE
)
```

```{r}
#| label: fa2-plots
#| fig-width: 10
#| fig-height: 4.5
#| fig-cap: "ForceAtlas2 on the kNN graph (left) and on the sNN graph (right), coloured by cell type."
fa2_plots <- purrr::map(
  c("fa2", "fa2_snn"),
  \(embd) {
    embedding_plot_sc(
      sc_object,
      embedding = embd,
      colour_by = "sc_type",
      discrete = TRUE
    ) +
      ggtitle(embd)
  }
)

patchwork::wrap_plots(fa2_plots, ncol = 2, guides = "collect")
```

# Per-Gene Expression plots`

`feature_plot_sc()` projects normalised expression of one or more genes onto the 
embedding.

```{r}
#| label: feature-plot
#| fig-cap: "Expression of MS4A1 on the UMAP. High-expressing cells cluster in expected B cell cluster."
# Select a single feature by Ensembl ID; provide a human-readable label
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]

feature_plot_sc(
  object = sc_object,
  features = features[8],
  feature_labels = feature_labels[8],
  embedding = "umap",
  label_by = "sc_type"
)
```

**Tip:** pass a character vector of Ensembl IDs to `features` and matching
labels to `feature_labels` to produce one panel per gene.

We can also highlight the genes of interest for clearer visualisation.

```{r}
#| label: feature-plot-highlight
#| fig-cap: "Expression of MS4A1 on the UMAP. High-expressing cells cluster in expected B cell cluster."
# Select a single feature by Ensembl ID; provide a human-readable label
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]

feature_plot_sc(
  object = sc_object,
  features = features[8],
  feature_labels = feature_labels[8],
  embedding = "umap",
  label_by = "sc_type",
  highlight_features = TRUE
)
```

**Extracting the plotting data**

Alternatively, you can also extract the plotting data directly to make 
customised plots.

```{r}
#| label: feature-plot-dt
dt <- extract_feature_plot_data(
  sc_object,
  embedding = "umap",
  obs_cols = c("sc_type", "leiden_clusters", "Ribo"),
  features = features[8],
  modality = "rna"
)

head(dt)
```

# Dot Plots 

Dot plots encode two dimensions simultaneously:

- **Dot size** → fraction of cells in the group expressing the gene (≥ 1 count)
- **Dot colour** → mean (scaled) expression among expressing cells

```{r}
#| label: dot-plot-setup
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]
```

```{r}
#| label: dot-plot
#| fig-cap: "Dot plot of canonical cell-type marker genes across Leiden clusters. Gene groups are annotated on the x-axis; expression is scaled per gene."
dot_plot_sc(
  object = sc_object,
  features = features,
  feature_labels = ensembl_to_symbol[features],
  grouping_variable = "leiden_clusters",
  scale_exp = TRUE,
  feature_grouping = cell_markers,
  cluster_groups = TRUE
)
```

**Key arguments:**

| Argument | Description |
|---|---|
| `features` | Ensembl IDs to plot |
| `feature_labels` | Display names shown on the axis |
| `grouping_variable` | Cell metadata column defining rows |
| `scale_exp` | Scale expression per gene (z-score) |
| `feature_grouping` | Named vector mapping gene → group label for x-axis brackets |
| `cluster_groups` | Hierarchically cluster rows by expression profile |

**Extracting the plotting data**


Alternatively, you can also extract the plotting data directly to make 
customised plots.

```{r}
#| label: dot-plot-dt
dt <- extract_dot_plot_data(
  sc_object,
  features = features,
  grouping_variable = "leiden_clusters",
  modality = "rna"
)

head(dt)
```

# Heatmaps

Dot plots summarise each group. Want to see every single cell instead? The
marker heatmap puts genes on the rows and cells on the columns, with a gap
between clusters. Expression is z-scored per gene and clipped at 2.5 by
default, otherwise a few extreme cells eat the whole colour range. These plots
need `bixverse.plots >= "0.2.8"`

```{r}
#| label: heatmap
#| eval: !expr utils::packageVersion("bixverse.plots") >= "0.2.8"
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Heatmap of canonical marker genes across Leiden clusters. Genes are grouped by the cell type they mark, expression is z-scored per gene."
heatmap_plot_sc(
  object = sc_object,
  features = features,
  feature_labels = feature_labels,
  grouping_variable = "leiden_clusters",
  feature_grouping = cell_markers
)
```

Big clusters hog the width and the platelets end up as a sliver.
`max_cells_per_group` subsamples every group, which gives the small ones a fair
share of the plot.

```{r}
#| label: heatmap-subsampled
#| eval: !expr utils::packageVersion("bixverse.plots") >= "0.2.8"
#| fig-width: 9
#| fig-height: 5
#| fig-cap: "Same heatmap grouped by cell type, at most 100 cells per group."
heatmap_plot_sc(
  object = sc_object,
  features = features,
  feature_labels = feature_labels,
  grouping_variable = "sc_type",
  feature_grouping = cell_markers,
  max_cells_per_group = 100L
)
```

The data behind it comes from `extract_gene_expression()`, in case you want to
build your own.

# Stacked Violin Plots

Stacked violins show the per-group expression distribution of several genes in a
compact vertical layout.

```{r}
#| label: stacked-violin
#| fig-cap: "Stacked violin plot of markers across final cell types."
# Take the first five markers
idx <- 1:5

stacked_violin_plot_sc(
  sc_object,
  features = names(feature_labels)[idx],
  feature_labels = feature_labels[idx],
  grouping_variable = "sc_type",
)
```

**Extracting the plotting data**

```{r}
#| label: stacked-violin-dt
dt <- extract_gene_violin_data(
  sc_object,
  features = features,
  grouping_variable = "sc_type",
  modality = "rna"
)

head(dt)
```
