---
title: "Single-cell data I/O in bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Single-cell data I/O in bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

`bixverse` keeps single-cell counts on disk -- a compact Rust binary for the
count matrix and a DuckDB for the `obs` / `var` metadata -- so that millions of
cells can be analysed on a laptop. Ingestion happens once, up front, through
format-specific readers that stream the data into that on-disk layout and apply
the cell/gene QC along the way.

This vignette walks through every reader on a small synthetic data set, so it
runs end-to-end without any downloads. It covers:

- in-memory R objects (`load_r_data`)
- 10x CellRanger MTX, single and multi-sample (`load_mtx`, `load_multi_mtx`)
- AnnData `h5ad`, single and multi-sample (`load_h5ad`, `load_multi_h5ad`)
- 10x CellRanger HDF5 (`load_tenx_h5`)
- checkpointing to and from disk (`save_sc_exp_to_disk`, `load_existing`)

If you have not seen the `SingleCells` class before, read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
vignette first.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(Matrix)
library(magrittr)
```

## A small synthetic data set

We generate a synthetic data set and treat its two halves as two "samples", so
that the multi-sample readers can be demonstrated too. The counts are a
cell-by-gene `dgRMatrix`; `obs` and `var` are `data.table`s.

```{r}
#| label: synthetic data
sim <- generate_single_cell_test_data()

counts <- sim$counts # cells x genes (dgRMatrix)
obs <- sim$obs
var <- sim$var
dim(counts)

# two "samples" for the multi-sample examples
idx_a <- 1:500
idx_b <- 501:1000
counts_a <- as(counts[idx_a, ], "RsparseMatrix")
counts_b <- as(counts[idx_b, ], "RsparseMatrix")

# scratch directories (one on-disk store per loaded object)
base <- file.path(tempdir(), "bixverse_io")
stores <- c(
  "mtx_a", "mtx_b", "store_r", "store_mtx", "store_multi_mtx",
  "store_h5", "store_multi_h5", "store_10x"
)
for (s in stores) dir.create(file.path(base, s), recursive = TRUE, showWarnings = FALSE)

# light QC thresholds, suitable for this small toy data set
qc <- params_sc_min_quality(
  min_unique_genes = 5L,
  min_lib_size = 10L,
  min_cells = 3L
)
```

## In-memory R objects

The simplest entry point: hand the reader a cell-by-gene `dgRMatrix` plus the
`obs` and `var` tables directly.

```{r}
#| label: load r data
sc_r <- SingleCells(dir_data = file.path(base, "store_r"))
sc_r <- load_r_data(
  object = sc_r,
  counts = counts,
  obs = obs,
  var = var,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_r
```

## 10x CellRanger MTX

### Single sample

We write one sample out in CellRanger flat-file form and point an explicit
`params_sc_mtx_io()` at the trio (for real CellRanger output, the convenience
helper `get_cell_ranger_params()` auto-locates `matrix.mtx` / `barcodes.tsv` /
`genes.tsv`).

```{r}
#| label: write + load single mtx
dir_a <- file.path(base, "mtx_a")
# real CellRanger MTX is genes x cells, matching cells_as_rows = FALSE below
write_cellranger_output(dir_a, counts_a, obs[idx_a], var, rows = "genes", .verbose = FALSE)

mtx_params <- params_sc_mtx_io(
  path_mtx = file.path(dir_a, "mat.mtx"),
  path_obs = file.path(dir_a, "barcodes.csv"),
  path_var = file.path(dir_a, "features.csv"),
  cells_as_rows = FALSE,
  has_hdr = TRUE
)

sc_mtx <- SingleCells(dir_data = file.path(base, "store_mtx"))
sc_mtx <- load_mtx(
  object = sc_mtx,
  sc_mtx_io_param = mtx_params,
  sc_qc_param = qc,
  mtx_streaming = FALSE,
  .verbose = FALSE
)
sc_mtx
```

### Multiple samples

`prescan_mtx_dirs()` takes one directory per sample, builds the shared gene
universe across them, and `load_multi_mtx()` then stacks all cells into a single
object with a global gene QC. The originating sample is recorded in the `exp_id`
`obs` column.

```{r}
#| label: write + load multi mtx
dir_b <- file.path(base, "mtx_b")
write_cellranger_output(dir_b, counts_b, obs[idx_b], var, rows = "genes", .verbose = FALSE)

ps_mtx <- prescan_mtx_dirs(
  dirs = c(dir_a, dir_b),
  exp_ids = c("sample_a", "sample_b"),
  has_hdr = TRUE
)

sc_multi_mtx <- SingleCells(dir_data = file.path(base, "store_multi_mtx"))
sc_multi_mtx <- load_multi_mtx(
  object = sc_multi_mtx,
  prescan_result = ps_mtx,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_multi_mtx

# cells per sample
get_sc_obs(sc_multi_mtx)[, .N, by = exp_id]
```

## AnnData h5ad

### Single

`read_h5ad_metadata()` lets you inspect the `obs` / `var` and dimensions without
loading the counts; `load_h5ad()` ingests the matrix.

```{r}
#| label: write + load single h5ad
h5_a <- file.path(base, "sample_a.h5ad")
write_h5ad_sc(h5_a, counts_a, obs[idx_a], var, .verbose = FALSE)

meta <- read_h5ad_metadata(h5_a)
meta$dims

sc_h5 <- SingleCells(dir_data = file.path(base, "store_h5"))
sc_h5 <- load_h5ad(
  object = sc_h5,
  h5_path = h5_a,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_h5
```

### Multiple

`prescan_h5ad_files()` accepts a named vector of paths (the names become the
sample identifiers); `load_multi_h5ad()` mirrors the multi-MTX workflow.

```{r}
#| label: write + load multi h5ad
h5_b <- file.path(base, "sample_b.h5ad")
write_h5ad_sc(h5_b, counts_b, obs[idx_b], var, .verbose = FALSE)

ps_h5 <- prescan_h5ad_files(c(sample_a = h5_a, sample_b = h5_b))

sc_multi_h5 <- SingleCells(dir_data = file.path(base, "store_multi_h5"))
sc_multi_h5 <- load_multi_h5ad(
  object = sc_multi_h5,
  prescan_result = ps_h5,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_multi_h5
```

## 10x CellRanger HDF5

`write_tenx_h5_sc()` expects the counts plus a barcode vector and a `features`
table (with at least `id` and `name`). `load_tenx_h5()` reads the gene-expression
modality (V3 files may also carry e.g. Antibody Capture, which is filtered out
via `feature_type`).

```{r}
#| label: write + load 10x h5
features <- data.table(id = var$gene_id, name = var$ensembl_id)
h5_tenx <- file.path(base, "sample_a_10x.h5")
write_tenx_h5_sc(
  f_path = h5_tenx,
  counts = counts_a,
  barcodes = obs[idx_a]$cell_id,
  features = features,
  version = "v3"
)

sc_10x <- SingleCells(dir_data = file.path(base, "store_10x"))
sc_10x <- load_tenx_h5(
  object = sc_10x,
  h5_path = h5_tenx,
  sc_qc_param = qc,
  feature_type = "Gene Expression",
  .verbose = FALSE
)
sc_10x
```

::: {.callout-note}
Multiple 10x `.h5` files: the streaming Rust combiner is implemented in
`bixverse-rs` ([PR #27](https://github.com/GregorLueg/bixverse-rs/pull/27)) and
will be exposed as `load_multi_10x()` once released; the workflow will mirror
`prescan_h5ad_files()` / `load_multi_h5ad()`.
:::

## Checkpointing to disk

Once an object is loaded, the in-memory mappings can be persisted so a fresh
session can be restored without re-ingesting the counts. The on-disk binary and
DuckDB already live in `dir_data`; `save_sc_exp_to_disk()` adds the lightweight
mapping cache.

```{r}
#| label: save and restore
save_sc_exp_to_disk(sc_h5, type = "rds")

sc_restored <- SingleCells(dir_data = file.path(base, "store_h5"))
sc_restored <- load_existing(sc_restored)
sc_restored
```

## Summary

| Format | Single sample | Multiple samples |
| --- | --- | --- |
| In-memory R | `load_r_data()` | -- |
| 10x MTX | `load_mtx()` | `prescan_mtx_dirs()` + `load_multi_mtx()` |
| AnnData h5ad | `load_h5ad()` | `prescan_h5ad_files()` + `load_multi_h5ad()` |
| 10x HDF5 | `load_tenx_h5()` | *(`load_multi_10x()` -- forthcoming)* |
| Seurat | `load_seurat()` | -- |

All readers share the same `sc_qc_param` QC interface and write into the same
on-disk `SingleCells` layout, so everything downstream (normalisation, HVGs,
PCA, neighbours, clustering, ...) is identical regardless of how the data
came in.
```

