---
title: "Sub-clustering and per-group pipelines"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Sub-clustering and per-group pipelines}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Your first pass of clustering gives you lineages: T cells, B cells, monocytes.
What it does not give you is the structure inside those lineages. The
highly variable genes and the principal components were chosen to separate a
monocyte from a B cell, and on that axis every T cell looks the same. If you
crank up the Leiden resolution on the full object you mostly get the lineages
chopped into arbitrary pieces, not naive versus memory CD4.

The fix is to throw the global feature selection away and redo it on the cells
you actually care about. That is what `SingleCellsSubset` is for. It gives you a
view onto a subset of a [`SingleCells`](https://gregorlueg.github.io/bixverse/reference/SingleCells.html)
object that shares the on-disk counts with the parent, but carries its own
empty cache, so HVG, PCA, neighbours and clusters all get recomputed from
scratch on the subset alone.

This vignette assumes you have read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the [PBMC walkthrough](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html).
The standard chain is not re-explained here.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(bixverse.plots)
library(magrittr)
```

## A coarse first pass

Same PBMC3k data set as the main walkthrough, run through QC, HVG, PCA,
neighbours and Leiden without commentary.

```{r}
#| label: load and process
pbmc3k_path <- download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = get_cell_ranger_params(pbmc3k_path),
  mtx_streaming = FALSE,
  .verbose = FALSE
)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
```

Mitochondrial proportions, MAD outliers, cells to keep.

```{r}
#| label: qc
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = FALSE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

qc <- run_cell_qc(
  metrics = list(
    log10_lib_size = log10(qc_df$lib_size),
    log10_nnz = log10(qc_df$nnz),
    MT = qc_df$MT
  ),
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = c(
    log10_lib_size = "twosided",
    log10_nnz = "twosided",
    MT = "above"
  ),
  threshold = 3
)

sc_object <- set_cells_to_keep(sc_object, qc_df[!qc$combined, cell_id])
```

Now the standard chain. The data is tiny, so exhaustive kNN beats building an
index.

```{r}
#| label: global chain
sc_object <- find_hvg_sc(sc_object, hvg_no = 2000L, .verbose = FALSE)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE,
  .verbose = FALSE
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  .verbose = FALSE
)

sc_object <- find_clusters_sc(sc_object, res = 1, name = "leiden_global")

sc_object <- umap_sc(sc_object, .verbose = FALSE)
```

For the annotation we deliberately stay coarse. No CD4 versus CD8 split, no
naive versus memory. Lineage only, because the fine structure is exactly what
we want the subset to recover.

```{r}
#| label: coarse annotation
coarse_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  CD3G = "T cells",
  TRAC = "T cells",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD19 = "B cells",
  CD14 = "Monocytes",
  LYZ = "Monocytes",
  S100A8 = "Monocytes",
  FCGR3A = "Monocytes",
  GNLY = "NK",
  NKG7 = "NK",
  KLRD1 = "NK",
  FCER1A = "DC",
  CD1C = "DC",
  LILRA4 = "DC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)

coarse_dt <- stack(coarse_markers) %>%
  as.data.table() %>%
  setnames(., c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var$gene_id[match(gene_symbol, var$gene_symbol)]] %>%
  .[!is.na(gene_id), ]

sctype_scores <- calc_sc_type_scores(
  object = sc_object,
  cell_marker_list = prepare_cell_markers(sc_object, coarse_dt)
)

cell_type_anno <- score_clusters(
  sctype_scores,
  sc_object[[]][["leiden_global"]]
)

obs <- get_sc_obs(sc_object, filtered = TRUE)

sc_object[["cell_type"]] <- cell_type_anno$cell_type[match(
  obs$leiden_global,
  cell_type_anno$cluster_id
)]

get_sc_obs(sc_object, filtered = TRUE)[, .N, by = cell_type][order(-N)]
```

```{r}
#| label: global umap
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "cell_type",
  label_by = "cell_type",
  discrete = TRUE
) +
  labs(colour = "Cell type")
```

## Taking a subset

The constructor takes the parent, a column in obs, and the level you want.

```{r}
#| label: make subset
t_cells <- SingleCellsSubset(
  sc_object = sc_object,
  grouping_column = "cell_type",
  group = "T cells"
)

t_cells
```

The print method tells you the important bit: HVG, PCA, kNN and sNN are all
`FALSE`. Nothing was inherited. The subset shares the Rust count connection with
the parent, so no counts were copied and no files were written, but the
`ScCache` is brand new and the parent's HVG selection was dropped on purpose.

Everything else behaves like the parent object. The obs table is filtered to
the group, the var table is untouched, and the usual getters work.

```{r}
#| label: subset getters
dim(t_cells)

head(t_cells)[, .(cell_id, cell_idx, lib_size, cell_type)]
```

### A note on indices

This is the one place where you can shoot yourself in the foot, so it is worth
being explicit. The subset keeps two index spaces around:

- **Subset positions** are 1-indexed rows of the subset's own obs table. This is
  what `[` and `[[` take.
- **Original positions** are the parent's index space. `subset_to_original` and
  the `cell_idx` obs column both hold these, 1-indexed. Everything that reaches
  Rust gets translated into this space, 0-indexed, before the call.

`get_cell_indices()` lets you ask for either.

```{r}
#| label: index spaces
first_three <- get_cell_names(t_cells)[1:3]

# 1-indexed position within the subset
get_cell_indices(t_cells, cell_ids = first_three, rust_index = FALSE)

# 0-indexed position in the parent, which is what Rust sees
get_cell_indices(t_cells, cell_ids = first_three, rust_index = TRUE)
```

The practical upshot: you never translate manually. Hand cell IDs or subset
positions to the getters and let the class deal with it. The `cell_idx` column
is there so you can join results back onto the parent later, which is exactly
what we do further down.

## Re-running the chain

Same functions as on the parent. They dispatch on the subset.

```{r}
#| label: subset chain
t_cells <- find_hvg_sc(t_cells, hvg_no = 2000L, .verbose = FALSE)

t_cells <- calculate_pca_sc(
  object = t_cells,
  no_pcs = 20L,
  sparse_svd = TRUE,
  .verbose = FALSE
)

t_cells <- find_neighbours_sc(
  object = t_cells,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  ),
  .verbose = FALSE
)

t_cells <- find_clusters_sc(t_cells, res = 0.6, name = "t_subcluster")

t_cells <- umap_sc(t_cells, .verbose = FALSE)

t_cells
```

Was any of that worth it? Check how much the feature selection actually moved.

```{r}
#| label: hvg overlap
global_hvg <- get_hvg(sc_object)
subset_hvg <- get_hvg(t_cells)

length(intersect(global_hvg, subset_hvg))
```

Around a thousand of the 2000 genes are different, and the ones that dropped out
are the monocyte and B cell markers that were carrying the global PCA. The
subset PCA is built on genes that vary *within* T cells, which is the whole
point.

```{r}
#| label: subset umap
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  t_cells,
  embedding = "umap",
  colour_by = "t_subcluster",
  label_by = "t_subcluster",
  discrete = TRUE
) +
  labs(colour = "Sub-cluster")
```

## Markers for the sub-clusters

Differential expression dispatches on the subset, so the comparison is
automatically confined to it. Each sub-cluster is tested against the other T
sub-clusters, not against every monocyte in the data set.

```{r}
#| label: subcluster markers
sub_markers <- find_all_markers_sc(
  object = t_cells,
  column_of_interest = "t_subcluster",
  .verbose = FALSE
)

sub_markers[, gene_symbol := ensembl_to_symbol[gene_id]]

sub_markers[fdr <= 0.05][order(grp, -lfc)][,
  head(.SD, 5),
  by = grp,
  .SDcols = c("gene_symbol", "lfc", "fdr")
]
```

S100A4, ANXA1 and IL32 on one side, CCR7, SELL and CD8B on the other. That is
the memory versus naive split, and the global clustering had no way of seeing
it. Cluster 2 is two cells, which is Leiden being Leiden at this resolution.

The expression plots work on the subset too, so you can go straight to a dot
plot without touching the parent.

```{r}
#| label: subcluster dotplot
#| fig-width: 8
#| fig-height: 4
t_features <- setdiff(
  symbol_to_ensembl[c(
    "CCR7",
    "LEF1",
    "SELL",
    "IL7R",
    "S100A4",
    "IL32",
    "CD8A",
    "CD8B",
    "GZMK",
    "NKG7"
  )],
  NA
)

dot_plot_sc(
  object = t_cells,
  features = t_features,
  feature_labels = ensembl_to_symbol[t_features],
  grouping_variable = "t_subcluster",
  scale_exp = TRUE,
  cluster_groups = TRUE
)
```

## Getting the labels back to the parent

So far everything lives in the subset, in memory. To make the labels permanent,
or to plot them on the parent's global UMAP, push them into the parent's DuckDB
joining on `cell_idx`. `add_sc_new_obs()` expects a table that carries the
`is_obs` tag, so we set it by hand.

```{r}
#| label: write back
sub_labels <- get_sc_obs(t_cells)[, .(
  cell_idx,
  t_subcluster = paste0("T_", t_subcluster)
)]

data.table::setattr(sub_labels, "is_obs", TRUE)

sc_object <- add_sc_new_obs(sc_object, sub_labels)

get_sc_obs(sc_object, filtered = TRUE)[, .N, by = t_subcluster][order(t_subcluster)]
```

The join is a left join, so every cell that was not a T cell gets `NA`. That is
the behaviour you want, and the downstream methods respect it.

Note that if you go via `fast_cluster_sc()` instead, the write-back is a
one-liner. That method already returns memberships keyed on `cell_idx` in parent
space, and `get_data()` tags the result for you.

```{r}
#| label: fast cluster route
fast_sub <- fast_cluster_sc(
  object = t_cells,
  resolutions = c(1, 0.5),
  .verbose = FALSE
)

sc_object <- add_sc_new_obs(sc_object, get_data(fast_sub))

head(sc_object)[, .(cell_id, cell_type, t_subcluster, res_1, res_0.5)]
```

## Pipelines

Once you are doing this for more than one cell type, writing out HVG, PCA,
neighbours and clusters by hand five times gets old. `sc_pipeline()` lets you
declare the chain once.

```{r}
#| label: build pipeline
subcluster_pipeline <- sc_pipeline() %>>%
  step_hvg_sc(hvg_no = 2000L, .verbose = FALSE) %>>%
  step_pca_sc(no_pcs = 20L, sparse_svd = TRUE, .verbose = FALSE) %>>%
  step_neighbours_sc(
    neighbours_params = params_sc_neighbours(
      knn = list(knn_method = "exhaustive")
    ),
    .verbose = FALSE
  ) %>>%
  step_clusters_sc(res = 0.6, name = "subcluster")

subcluster_pipeline
```

Pipelines are inert. Nothing runs until you apply one, and the same pipeline
works on a `SingleCells` or a `SingleCellsSubset` because dispatch happens
inside each step.

```{r}
#| label: apply pipeline
b_cells <- SingleCellsSubset(
  sc_object = sc_object,
  grouping_column = "cell_type",
  group = "B cells"
)

b_cells <- apply_pipeline(subcluster_pipeline, b_cells)

b_cells
```

### One pipeline, every group

`apply_pipeline_per_group()` builds the subset for you and returns a named list.

Sizing matters here. PCA with 20 components needs more than 20 cells, so a rare
population that the annotation happened to call will fall over and take the
whole run with it. Restrict the groups explicitly rather than hoping for the
best. On this data set the coarse annotation only returned four lineages and
all of them clear the bar, so nothing gets dropped, but the guard costs nothing.

```{r}
#| label: pick groups
cell_counts <- get_sc_obs(sc_object, filtered = TRUE)[, .N, by = cell_type][
  order(-N)
]

groups_to_run <- cell_counts[N >= 100, cell_type]

groups_to_run
```

```{r}
#| label: per group
per_group <- apply_pipeline_per_group(
  pipeline = subcluster_pipeline,
  object = sc_object,
  group_col = "cell_type",
  groups = groups_to_run
)

summary_dt <- data.table::rbindlist(purrr::imap(per_group, \(x, name) {
  data.table(
    cell_type = name,
    n_cells = dim(x)[1],
    n_subclusters = length(unique(get_sc_obs(x)$subcluster)),
    hvg_shared_with_global = length(intersect(get_hvg(x), global_hvg))
  )
}))

summary_dt[order(-n_cells)]
```

Every lineage picked a different feature set, and the shared fraction with the
global HVGs is nowhere near complete. Running one global PCA and hoping it
resolves everything was never going to work.

Writing all of that back is the same join as before, once per group.

```{r}
#| label: write back per group
all_labels <- data.table::rbindlist(purrr::imap(per_group, \(x, name) {
  get_sc_obs(x)[, .(
    cell_idx,
    fine_label = sprintf("%s_%s", name, subcluster)
  )]
}))

data.table::setattr(all_labels, "is_obs", TRUE)

sc_object <- add_sc_new_obs(sc_object, all_labels)

get_sc_obs(sc_object, filtered = TRUE)[, .N, by = fine_label][order(fine_label)]
```

```{r}
#| label: fine umap
#| fig-width: 8
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "fine_label",
  discrete = TRUE
) +
  labs(colour = "Fine label")
```

## Caveats and what is next

A few sharp edges to be aware of before you build on this.

Metacell generation via `generate_bt_meta_cells_sc()` is still parent-only,
which is a shame given per-sample metacells are the obvious use for
`apply_pipeline_per_group()`. That one needs real work rather than a dispatch
widening, so it is on the list.

Subsets are RNA only. If you hand `modality = "adt"` to a getter you get an
error, so multi-modal work stays on
[`SingleCellsMultiModal`](https://gregorlueg.github.io/bixverse/articles/multi_modal_single_cells.html).

The `setattr(dt, "is_obs", TRUE)` dance for the write-back is a placeholder. A
proper `merge_subset_obs()` helper belongs in the package and is on the list.

Subsets nest conceptually but not in code: you cannot build a
`SingleCellsSubset` from another `SingleCellsSubset`. If you want a third level,
write the labels back to the parent and subset from there.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_pbmc, recursive = TRUE, force = TRUE)
```
