---
title: "Scaling to millions of cells with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Scaling to millions of cells with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg"
---

## Intro

This vignette demonstrates how to run a single cell analysis on a data set
containing nearly 1 million cells on a local machine... In this case, a MacBook 
Air with 24 GBs (if you think *"IMPOSSIBLE!"*, you would not be wrong coming
from other languages/frameworks). If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the [introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first. The [PBMC3k walkthrough](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html)
covers every step in detail on a small data set; here we focus on what changes
when the cell count moves from thousands to a million.

In the orignal version of the vignette, I used a one of the h5ad files from
the Tahoe 100m data. set, but I have decided to update the data set to the 
1 million cell data set from Parse Bioscience as it's an often used data set to 
show case large single cell data analysis, see original article 
[here](https://support.parsebiosciences.com/hc/en-us/articles/7704577188500-How-to-analyze-a-1-million-cell-data-set-using-Scanpy-and-Harmony).
**The important bit here is that this was run successfully on a MacBook Air with
24 GB (M3)**. And please contrast that with the following line for the analysis
of this data:

<i>To avoid computer crashes and program failures, we recommend you execute the 
Python code on a cloud computing platform. The instance you select on the cloud 
platform to run this analysis should consist of at least 8 threads and 160 GB of 
RAM.</i> - Parse Biosciences 

::: {.callout-note}
None of the code chunks below are evaluated. This vignette is a reference for
how one *would* run such an analysis, not a live rendering - the data set is
far too large for CI/CD runners and GitHub would not be happy.
:::

If you want to reproduce this, download the data from Parse Biosciences (or
any other fat data set you have) and go have fun.

```{bash}
#| label: get data
#| eval: false
wget "https://cdn.parsebiosciences.com/1M_PBMC_T1D_Parse.zip"
```

```{r}
#| label: setup
#| eval: false
# main packages
library(bixverse)
library(bixverse.plots)

# GPU-acceleration (experimental)
library(bixverse.gpu)

# Plotting, data manipulation
library(ggplot2)
library(data.table)
```

## Loading the data

Once you have unzipped the data, we can start streaming it into memory and
writing the files to disk.

```{r}
#| label: stream h5ad
#| eval: false
data_path <- "<PATH TO UNZIPPED DATA>"

load_params <- params_sc_mtx_io(
  path_mtx = path.expand(file.path(data_path, "DGE_1M_PBMC.mtx")),
  path_obs = path.expand(file.path(data_path, "cell_metadata_1M_PBMC.csv")),
  path_var = path.expand(file.path(data_path, "all_genes_1M_PBMC.csv")),
  cells_as_rows = TRUE,
  has_hdr = TRUE
)

# We will write everything into the temporary directory
sce <- SingleCells(dir_data = tempdir())

# We enforce level 2 streaming here which will run for a bit longer, but ensure
# your memory never blows up
sce <- load_mtx(object = sce, sc_mtx_io_param = load_params, streaming = 2L)
```

This step is the slowest... A lot of things are actually happening here.

- First, (streaming) scan to understand in how many cells a gene is expressed.
  Only keep genes that are expressed in whatever you set to `min_cells`.
- Second, (streaming) scan, given the genes from the first step, which cells to
  include based on minimum library size and minimum features.
- Third, now that it's clear which cells/genes to include, create quick temp
  files with the cells to keep with the genes already re-indexed. This way we
  avoid loading in the massive MTX file in one go into memory. These files are
  automatically cleaned up at the end and loaded sequentially and written to the
  CSR-type file (while the normalisation is also done on the fly).
- Lastly, load in the cell data in small chunks and do a transpose to a CSC 
  style format for easy gene retrieval.

This approach avoids loading the whole data into memory and any point. The 
DuckDB also gets populated with the metadata in the observations (based on what 
has been filtered) and the vars (add another 1 to 2 minutes here). This part -
streaming the data and processing it - is usually what takes the most time. If 
you have already loaded in the data and it exists on-disk from a previous 
session, skip the streaming step entirely and reconnect. This is the one (very) 
slow part in `bixverse` single cell... To be fair, I have no idea how long it 
would take to load in the full MTX file in  Python or R -- also, I do not want 
to really try...

```{r}
#| label: reconnect
#| eval: false
sc_object <- SingleCells(
  dir_data = dir_files
)

sc_object <- load_existing(object = sc_object)
```

## Quality control

### Gene set proportions

The same gene set proportion logic from the PBMC3k vignette applies. The 
functions automatically recognise if you have more than 100k cells and will
leverage the streaming engine by default in this case. You do have more manual
control via the `streaming = NULL` parameter. If you have a fat memory machine,
you can set this to `FALSE` despite a large data set. 

```{r}
#| label: gene set proportions
#| eval: false
var <- get_sc_var(object = sce)

gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_name), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_name), gene_id]
)

sce <- gene_set_proportions_sc(
  object = sce,
  gene_set_list = gs_of_interest,
  .verbose = TRUE
)

head(sce)
```

### MAD outlier detection

Outlier detection works identically to the small-data case. The QC metrics
table is modest in size (one row per cell, a handful of columns) and fits
comfortably in memory even at millions of cells.

```{r}
#| label: qc outlier detection
#| eval: false
qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)

# not needed anymore
rm(qc_df)

qc
```

I cannot recommend trying to plot this naively... ggplot2 will not be happy.

### Filtering cells

Usual filters.

```{r}
#| label: set cells to keep
#| eval: false
sc_object[["outlier"]] <- qc$combined

cells_to_keep <- qc_df[!qc$combined, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object
```

## Feature selection and PCA

HVG selection at this scale again requires `streaming = TRUE`. There is two
data passes to avoid materialising large data in memory. First pass will 
calculate the means and standard deviations. Second pass will do the variance
stabilisation. PCA uses a sparse SVD, which is automatically selected when cell 
counts exceed 500k but can be set explicitly.

```{r}
#| label: hvg and pca
#| eval: false
sc_object <- find_hvg_sc(
  object = sc_object,
  streaming = TRUE,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE
)
```

## Nearest neighbours (GPU-accelerated)

At nearly 3 million cells, nearest neighbour search is the main computational
bottleneck. The companion package [bixverse.gpu](https://github.com/GregorLueg/bixverse.gpu) 
exposes a CAGRA-style GPU-accelerated approximate nearest neighbour search that 
makes this quite fast on a laptop with a discrete GPU - thanks to CubeCL and 
WGPU as long as you can fit the data into VRAM (should be feasible with 8 GBs - 
I think?) this should run quite fast. CPU-side alternatives such as 
`"nndescent"` or `"hnsw"` also work but will be considerably slower at this 
scale. For (empirical reference):

- GPU version on M1 Max with 64 GB takes ~1 minute.
- HNSW somewhere between 3 to 4 minutes.

Likely still comparably much faster if you are used to the more established 
single cell methods in terms of CPU speed, but yeah... If you have a GPU,
go use that.

```{r}
#| label: gpu knn
#| eval: false
library(bixverse.gpu)

sc_object <- find_neighbours_cagra_sc(
  object = sc_object,
  cagra_params = params_sc_cagra(k_query = 15L, ann_dist = "cosine"),
  extract_knn = FALSE,
  .verbose = TRUE
)
```

## Visualisation

Running UMAP on the shared nearest neighbour graph built from ~2.8 million
cells and 65 million+ edges takes time but does complete on a single machine
(embedding optimisation with parallel Adam optimisation done in <3 minutes for 
me).

```{r}
#| label: umap
#| eval: false
sc_object <- umap_sc(object = sc_object)
```

## Differential gene expression

With metadata stored in the obs table, subsetting cells for a pairwise
comparison is straightforward. Here we compare two drug treatment conditions.
Again, we leverage the streaming here... Instead of holding the full data set
in memory, we just load in the two groups (each roughly 30k cells), do a fast
ranking in Rust and do the DGE. Easy peezy, done in ca. 3 seconds.

```{r}
#| label: dge
#| eval: false
meta_data <- sc_object[[c("cell_id", "cell_name", "drug")]]

cells_grp_a <- meta_data[drug == "AT7519", cell_id]
cells_grp_b <- meta_data[drug == "palbociclib", cell_id]

dge_results <- find_markers_sc(
  object = sc_object,
  cells_1 = cells_grp_a,
  cells_2 = cells_grp_b
)

setorder(dge_results, fdr)

head(dge_results, 25L)
```

## Key differences from the small-data workflow

The API surface is intentionally identical to the PBMC3k walkthrough. The
meaningful differences when working at the million-cell scale are:

1. **Streaming I/O.** Use `stream_h5ad` instead of `load_h5ad`, and set
   `streaming = TRUE` on `gene_set_proportions_sc` and `find_hvg_sc`. This
   is VERY important. This is also what makes this package actually quite cool.
2. **On-disk persistence.** `load_existing` lets you reconnect to a previously
   streamed data set without re-reading the `.h5ad`. (Btw, works for any data.)
3. **Sparse SVD.** Automatically enabled above 500k cells; can be forced
   explicitly via `sparse_svd = TRUE` (slower on small data sets... Modern BLAS
   just eats up small matrix multiplications in `faer`).
4. **GPU-accelerated kNN.** `bixverse.gpu` provides CAGRA-based nearest
   neighbour search (also others) that scales comfortably to millions of cells 
   and makes it run in a minute or so. The CPU versions are still highly 
   optimised with SIMD and memory layouts specifically designed to avoid cache 
   misses.

For context... Running this on my machine took (MacBook Pro M1 Max with 64 GB)
took a grand total of ca. 20 minutes from start to finish, with basically half
of the time being spent on data I/O. Memory pressure remained green through out
the whole thing.