---
title: "A guide to single-cell plotting functions"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{A guide to single-cell plotting functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Liesbeth François"
---

## Intro

This vignette walks through a standard single cell analysis on the PBMC3k data
set using `bixverse`. If you have not read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first; this vignette assumes familiarity with how the
`SingleCells` class, on-disk storage and cells-to-keep logic work.

```{r}
#| label: setup
library(bixverse)
library(ggplot2)
library(data.table)
library(bixverse.plots)
library(magrittr)
```

## Loading the data

We start by downloading the PBMC3k data set bundled with the package and
loading it via Cell Ranger-style MTX I/O.

```{r}
#| label: load pbmc3k
pbmc3k_path <- download_pbmc3k()

tempdir_pbmc <- tempdir()

sc_object <- SingleCells(dir_data = tempdir_pbmc)

mtx_io_params <- get_cell_ranger_params(pbmc3k_path)

sc_object <- load_mtx(
  object = sc_object,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = TRUE
)

sc_object[["sample"]] <- sample(
  x = c("sample1", "sample2"),
  size = nrow(sc_object),
  replace = T
)
```

Let's have a quick look at the variable table. The column names from the MTX
files are not always informative, so we rename the gene symbol column to
something sensible and set up mappings between Ensembl IDs and symbols.

```{r}
#| label: var table and renaming
var <- get_sc_var(sc_object)

head(var)

setnames_sc(
  object = sc_object,
  table = "var",
  old = "column1",
  new = "gene_symbol"
)

var <- get_sc_var(sc_object)

ensembl_to_symbol <- setNames(var$gene_symbol, var$gene_id)
symbol_to_ensembl <- setNames(var$gene_id, var$gene_symbol)
```

# Quality Control Plots

QC metrics are computed with `gene_set_proportions_sc()` and flagged with
`run_cell_qc()`. Three complementary plotting functions then visualise the
results.

```{r qc-setup}
# Define gene sets whose proportional expression signals low quality
gs_of_interest <- list(
  MT = var[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

# Collect metrics for outlier detection
qc_df <- sc_object[[c("cell_id", "sample", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)

directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3,
  groups = qc_df$sample
)
```

## Joint QC Overview — `joint_plot_sc()`

`joint_plot_sc()` arranges all QC panels in a single figure, making it easy to
spot global trends at a glance.

```{r joint-plot, fig.cap="Joint QC plot. Each panel shows one metric; red points indicate MAD-based outliers."}
joint_plot_sc(qc)
```

**When to use:** first-pass assessment of data quality before any filtering.

---

## Violin Plots — `violin_plot_sc()`

`violin_plot_sc()` shows the distribution of each QC metric as a violin,
split by donor (or any grouping variable stored in the `qc` object).

```{r violin-plot, fig.cap="Per-donor violin plots for library size, number of non-zero genes, and MT proportion."}
violin_plot_sc(qc)
```

**Tip:** compare widths across donors to detect batch effects or low-quality
libraries.

---

## Density Plots — `density_plot_sc()`

`density_plot_sc()` overlays smoothed density curves for each group, which
highlights bimodal distributions or outlier populations that violins can obscure.

```{r density-plot, fig.cap="Density plots for each QC metric. Vertical dashed lines mark the MAD thresholds."}
density_plot_sc(qc)
```

---

# Dimensionality Reduction Plots — `embedding_plot_sc()`

After PCA, neighbour-finding, clustering, and UMAP, cells can be visualised
on any 2-D embedding stored in the object.

```{r umap-setup}
#| label: hvg pca neighbours
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L,
  sparse_svd = TRUE
)

# the data is so tiny that exhaustive kNN search is faster than building
# an approximate nearest neighbour index
sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(knn_method = "exhaustive")
  )
)
sc_object <- find_clusters_sc(sc_object, res = 1, name = "leiden_clusters")
sc_object <- umap_sc(sc_object)

cell_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  CD3G = "T cells",
  IL7R = "CD4+ T",
  CD4 = "CD4+ T",
  CD8A = "CD8+ T",
  CD8B = "CD8+ T",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD19 = "B cells",
  CD14 = "CD14+ Mono",
  LYZ = "CD14+ Mono",
  S100A8 = "CD14+ Mono",
  FCGR3A = "CD16+ Mono",
  CDKN1C = "CD16+ Mono",
  GNLY = "NK",
  NKG7 = "NK",
  NCAM1 = "NK",
  FCER1A = "mDC",
  CD1C = "mDC",
  LILRA4 = "pDC",
  CLEC4C = "pDC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)
cell_markers_dt <- stack(cell_markers) %>%
  as.data.table() %>%
  setnames(., c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var$gene_id[match(gene_symbol, var$gene_symbol)]] %>%
  .[!is.na(gene_id), ]

## Prepare the list of markers
marker_list <- prepare_cell_markers(sc_object, cell_markers_dt)
## Calculate the score for each cell
sctype_scores <- calc_sc_type_scores(
  object = sc_object,
  cell_marker_list = marker_list
)
## Annotate the clusters
cell_type_anno <- score_clusters(
  sctype_scores,
  sc_object[[]][["leiden_clusters"]]
)

obs <- sc_object[[]][, .(cell_idx, leiden_clusters)]
add_sctype <- setNames(
  cell_type_anno$cell_type[match(
    obs$leiden_clusters,
    cell_type_anno$cluster_id
  )],
  nm = obs$cell_idx
)
sc_object[["sc_type"]] <- add_sctype
```

## Colouring by Leiden Cluster ID

Showing raw cluster IDs before annotation helps validate the
`group_to_celltype` mapping.

```{r embedding-leiden, fig.cap="UMAP coloured by Leiden cluster number."}
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

## Colouring by a Discrete Variable

Pass any column from the cell metadata to `colour_by`. Use `label_by` to
overlay text labels at the cluster centroid.

```{r embedding-celltype, fig.cap="UMAP coloured by manually annotated cell type. Labels are placed at cluster centroids."}
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "sc_type",
  label_by = "sc_type",
  discrete = TRUE
)
```

## Colouring by a Continuous Variable

Set `discrete = FALSE` to use a continuous colour scale — useful for
confidence scores, pseudotime, or gene module scores.

```{r embedding-sctype, fig.cap="UMAP coloured by scType annotation confidence score (continuous scale)."}
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "Ribo",
  discrete = FALSE,
  point_alpha = 0.2
)
```

**Key arguments:**

| Argument | Description |
|---|---|
| `embedding` | Name of the stored embedding (`"umap"`, `"tsne"`, …) |
| `colour_by` | Metadata column used to colour points |
| `label_by` | Metadata column used to label centroids |
| `discrete` | `TRUE` for categorical palettes, `FALSE` for continuous |
| `point_alpha` | Transparency (0–1); reduce for dense datasets |

---

# Per-Gene Expression — `feature_plot_sc()`

`feature_plot_sc()` projects raw or normalised expression of one or more genes
onto the embedding.

```{r feature-plot, fig.cap="Expression of CAMK2A (excitatory neuron marker) on the UMAP. High-expressing cells cluster in expected Exc neuron regions."}
# Select a single feature by Ensembl ID; provide a human-readable label
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]

feature_plot_sc(
  object = sc_object,
  features = features[8], # CAMK2A
  feature_labels = feature_labels[8],
  embedding = "umap",
  label_by = "sc_type"
)
```

**Tip:** pass a character vector of Ensembl IDs to `features` and matching
labels to `feature_labels` to produce one panel per gene.

```{r feature-plot, fig.cap="Expression of CAMK2A (excitatory neuron marker) on the UMAP. High-expressing cells cluster in expected Exc neuron regions."}
# Select a single feature by Ensembl ID; provide a human-readable label
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]

feature_plot_sc(
  object = sc_object,
  features = features[8],
  feature_labels = feature_labels[8],
  embedding = "umap",
  label_by = "sc_type",
  highlight_features = TRUE
)
```

---

# Dot Plots — `dot_plot_sc()`

Dot plots encode two dimensions simultaneously:

- **Dot size** → fraction of cells in the group expressing the gene (≥ 1 count)
- **Dot colour** → mean (scaled) expression among expressing cells

```{r dot-plot-setup}
features <- setdiff(symbol_to_ensembl[names(cell_markers)], NA)
feature_labels <- ensembl_to_symbol[features]
```

```{r dot-plot, fig.cap="Dot plot of canonical cell-type marker genes across Leiden clusters. Gene groups are annotated on the x-axis; expression is scaled per gene."}
dot_plot_sc(
  object = sc_object,
  features = features,
  feature_labels = ensembl_to_symbol[features],
  grouping_variable = "leiden_clusters",
  scale_exp = TRUE,
  feature_grouping = cell_markers,
  cluster_groups = TRUE
)
```

**Key arguments:**

| Argument | Description |
|---|---|
| `features` | Ensembl IDs to plot |
| `feature_labels` | Display names shown on the axis |
| `grouping_variable` | Cell metadata column defining rows |
| `scale_exp` | Scale expression per gene (z-score) |
| `feature_grouping` | Named vector mapping gene → group label for x-axis brackets |
| `cluster_groups` | Hierarchically cluster rows by expression profile |

---

# Stacked Violin Plots — `stacked_violin_plot_sc()`

Stacked violins show the per-group distribution of several genes in a
compact vertical layout — a convenient complement to the dot plot.

```{r stacked-violin, fig.cap="Stacked violin plot of pan-neuronal markers across final cell types."}
# Take the first five markers
pan_idx <- 1:5

stacked_violin_plot_sc(
  sc_object,
  features = names(feature_labels)[pan_idx],
  feature_labels = feature_labels[pan_idx],
  grouping_variable = "sc_type"
)
```

**When to use:** validating cell-type annotations with a small, curated gene
panel, or presenting figure-ready marker summaries.

---

# Combining Plots with `patchwork`

All `bixverse.plots` functions return standard `ggplot2` objects, so they
compose naturally with `patchwork` or `cowplot`.

```{r patchwork-example, fig.cap="Side-by-side UMAP coloured by cell type (left) and MBP expression (right).", fig.width=12}
library(patchwork)

p1 <- embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "sc_type",
  label_by = "sc_type",
  discrete = TRUE,
  point_alpha = 0.15
)

p2 <- feature_plot_sc(
  object = sc_object,
  label_by = "sc_type",
  features = features[8],
  feature_labels = feature_labels[8],
  embedding = "umap"
)

p1 + p2 + plot_layout(ncol = 2)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
