---
title: "Batch correction with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Batch correction with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Grant Neilson, Gregor Lueg, Liesbeth François"
---

## Intro

Batch effects are systematic technical differences between experimental runs
that can obscure genuine biological signal in single cell data. When datasets
from multiple experiments, technologies, or laboratories are combined, cells
tend to group by their source rather than by cell type. Batch correction methods
aim to remove these technical artefacts whilst preserving biological variation.

`bixverse` provides three batch correction approaches, each with different
trade-offs:

- **fastMNN** ([Haghverdi, et al., 2018](https://doi.org/10.1038/nbt.4091))
identifies mutual nearest neighbours across batches -- pairs of cells from
different batches that are each other's closest match. Correction vectors are
computed from these pairs and applied to align the embeddings. It is
particularly effective when cell type composition differs across batches.

- **Harmony** ([Korsunsky, et al., 2019](https://doi.org/10.1038/s41592-019-0619-0))
operates on PCA embeddings using iterative soft clustering with diversity
penalties. It assigns cells to clusters, estimates batch effects per cluster
via ridge regression, and corrects the embedding. It tends to be fast and works
well across a range of scenarios. Additionally, a version 2 has been released,
see [Patikas, et al., 2026](https://www.biorxiv.org/content/10.64898/2026.03.16.711825v1)
with improvements that increase scalability to larger data sets and 
modifications to the objective functions to not overcorrect across batches
when different cell types are there.

- **BBKNN** ([Polanski, et al., 2020](https://doi.org/10.1093/bioinformatics/btz625))
takes a fundamentally different approach. Rather than correcting an embedding,
it constructs a batch-balanced k-nearest neighbour graph by building separate
neighbour indices per batch. The corrected graph can then be used directly for
clustering and visualisation.

`bixverse` also provides three metrics for assessing batch correction quality:

- **kBET**: tests whether neighbourhood batch proportions match global
proportions (best for embedding-based methods).
- **Batch ASW**: measures batch separation in the embedding space (best for
embedding-based methods).
- **LISI**: measures the effective number of batches per neighbourhood (works
on any kNN graph).

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(data.table)
library(ggplot2)
```

## Preparing the data

### Loading the batches

We use two PBMC datasets (pbmc3k and pbmc4k) as a standard benchmark for
batch correction. These come from the same tissue but differ in sequencing
depth and cell counts, introducing a clear batch effect.

```{r}
#| label: load data
dir_data <- download_pbmc_batches()

tempdir_batch_cor <- file.path(tempdir(), "batch_cor_bixverse")
dir.create(tempdir_batch_cor, showWarnings = FALSE, recursive = TRUE)

h5ad_files <- list.files(dir_data)
h5ad_files <- h5ad_files[grepl(".h5ad", h5ad_files)]
h5ad_paths <- file.path(dir_data, h5ad_files)
names(h5ad_paths) <- c("pbmc3k", "pbmc4k")

h5_tasks <- prescan_h5ad_files(h5_paths = h5ad_paths)

sc_object <- SingleCells(dir_data = tempdir_batch_cor)

sc_object <- load_multi_h5ad(
  object = sc_object,
  prescan_result = h5_tasks,
  .verbose = TRUE
)
```

### Quality control

Before batch correction, we filter out low-quality cells using mitochondrial
gene proportions and library complexity metrics.

```{r}
#| label: qc
var <- get_sc_var(sc_object)

gs_of_interest <- list(
  MT = var[grepl("^MT-", symbol_tenx), gene_id],
  Ribo = var[grepl("^RPS|^RPL", symbol_tenx), gene_id]
)

sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

qc_df <- sc_object[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_object),
  directions = directions,
  threshold = 3
)

qc
```

Let's plot the metrics:

The `CellQc` class has a few associated plotting functions using the output of `run_cell_qc` and the specified metrics as input

- `violin_plot_sc` method that produces violin plots with outliers highlighted.

```{r}
#| label: qc plots
#| fig-width: 12
#| fig-height: 4
violin_plot_sc(qc)
```

- `joint_plot_sc` method that compares the genes versus UMIs per cell.

```{r}
#| label: qc plots
#| fig-width: 12
#| fig-height: 4
joint_plot_sc(qc)
```

Set the cells to keep

```{r}
#| label: apply qc
sc_object[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_object <- set_cells_to_keep(sc_object, cells_to_keep)
```

### Pre-processing

Standard pre-processing: HVG selection, PCA, and neighbour computation.

```{r}
#| label: preprocessing
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours()
)
```

## Comparing the different methods

Let's check out the different methods and how they behave.

### Uncorrected data

Before applying any correction, we can visualise the batch effect and compute
baseline metrics.

```{r}
#| label: uncorrected tsne
sc_object <- tsne_sc(object = sc_object)
```

tSNE plot (tSNE's usually look prettier... It's anyway not a good way to assess
batch effect correction, so at least let's make it pleasant on the eye).

```{r}
#| label: uncorrected plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "Before batch correction",
    colour = "Batch:"
  )
```

Batch correction stats:
```{r}
#| label: uncorrected metrics
kbet_score_prior <- calculate_kbet_sc(sc_object, batch_column = "exp_id")
asw_score_prior <- calculate_batch_asw_sc(sc_object, batch_column = "exp_id")
lisi_score_prior <- calculate_batch_lisi_sc(sc_object, batch_column = "exp_id")

kbet_score_prior
asw_score_prior
lisi_score_prior
```

The high kBET rejection rate and mean chi-square well above the expected value
of 1 confirm a substantial batch effect. LISI scores also indicate very poor
batch matching and the tSNE shows clear separation by batch. 

### fastMNN

fastMNN works on batch-aware HVGs and produces a corrected embedding. It
identifies mutual nearest neighbours across batches to compute correction
vectors. For it to work best, we need the batch-aware HVGs. We will calculate
these, provide them to fastMNN (which regenerates the PCA embedding based on
the batch-aware HVG) and run then the actual algorithm

```{r}
#| label: fastmnn
batch_aware_hvg <- find_hvg_batch_aware_sc(
  object = sc_object,
  batch_column = "exp_id"
)

sc_object <- fast_mnn_sc(
  object = sc_object,
  batch_hvg_genes = batch_aware_hvg$hvg_genes,
  batch_column = "exp_id"
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  embd_to_use = "mnn",
  neighbours_params = params_sc_neighbours()
)
```

```{r}
#| label: fastmnn metrics
kbet_score_post_mnn <- calculate_kbet_sc(
  sc_object,
  batch_column = "exp_id"
)
asw_score_post_mnn <- calculate_batch_asw_sc(
  sc_object,
  embd_to_use = "mnn",
  batch_column = "exp_id"
)
lisi_score_post_mnn <- calculate_batch_lisi_sc(
  sc_object,
  batch_column = "exp_id"
)

kbet_score_post_mnn
asw_score_post_mnn
lisi_score_post_mnn
```

We can see clear improvements across all scores. Lower kBET scores, better mean
AWS scores and LISI scores also closer to 2 (i.e., number of batches).

```{r}
#| label: fastmnn tsne
sc_object <- tsne_sc(object = sc_object, use_knn = TRUE)
```

```{r}
#| label: fastmnn plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "fastMNN batch correction",
    colour = "Batch:"
  )
```

Also, visually the batches get mixed now.

### Harmony

Harmony operates on the PCA embedding directly. The number of clusters is
auto-determined from the dataset size (capped at 100) when left as `NULL`.

The two versions of Harmony are supported. Let's start with the original:

#### Version 1

```{r}
#| label: harmony
sc_object <- harmony_sc(
  object = sc_object,
  batch_column = "exp_id",
  harmony_params = params_sc_harmony()
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  embd_to_use = "harmony",
  neighbours_params = params_sc_neighbours()
)
```

Let's calculate the batch correction-related scores

```{r}
#| label: harmony metrics
kbet_score_post_harmony <- calculate_kbet_sc(
  sc_object,
  batch_column = "exp_id"
)
asw_score_post_harmony <- calculate_batch_asw_sc(
  sc_object,
  embd_to_use = "harmony",
  batch_column = "exp_id"
)
lisi_score_post_harmony <- calculate_batch_lisi_sc(
  sc_object,
  batch_column = "exp_id"
)

kbet_score_post_harmony
asw_score_post_harmony
lisi_score_post_harmony
```

Also here, we observe improvements across the board.

```{r}
#| label: harmony tsne
sc_object <- tsne_sc(object = sc_object, use_knn = TRUE)
```

```{r}
#| label: harmony plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "Harmony batch correction",
    colour = "Batch:"
  )
```

#### Version 2

Compared to v1 of Harmony, several modifications/improvements were implemented
by [Patikas, et al., 2026](https://www.biorxiv.org/content/10.64898/2026.03.16.711825v1),
namely:

- Stabilised diversity penalty
- Batch pruning in ridge regression
- Arrowhead matrix inversion for single-covariate case (making it faster!)
- Dynamic lambda estimation and theta scaling by batch size.

Overall, this makes v2 more amenable and faster on big data sets than v1 (if
you wish to use Harmony, maybe use the version 2 over the v2?).

```{r}
#| label: harmony (v2)
sc_object <- harmony_v2_sc(
  object = sc_object,
  batch_column = "exp_id",
  harmony_params = params_sc_harmony_v2()
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  embd_to_use = "harmony_v2",
  neighbours_params = params_sc_neighbours()
)
```

Let's calculate the batch correction-related scores

```{r}
#| label: harmony (v2) metrics
kbet_score_post_harmony <- calculate_kbet_sc(
  sc_object,
  batch_column = "exp_id"
)
asw_score_post_harmony <- calculate_batch_asw_sc(
  sc_object,
  embd_to_use = "harmony_v2",
  batch_column = "exp_id"
)
lisi_score_post_harmony <- calculate_batch_lisi_sc(
  sc_object,
  batch_column = "exp_id"
)

kbet_score_post_harmony
asw_score_post_harmony
lisi_score_post_harmony
```

```{r}
#| label: harmony (v2) tsne
sc_object <- tsne_sc(object = sc_object, use_knn = TRUE)
```

```{r}
#| label: harmony (v2) plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "Harmony (version 2) batch correction",
    colour = "Batch:"
  )
```

### BBKNN

BBKNN constructs a batch-balanced kNN graph rather than correcting an
embedding. The key parameter is `neighbours_within_batch`, which controls how
many neighbours are found per batch. Setting `no_neighbours_to_keep` below the
total generated neighbours (here 10 from 2 batches = 20 total) ensures the
distance-based filtering introduces genuine variation, making downstream
metrics informative.

Note that kBET and ASW are not well suited for evaluating BBKNN: kBET compares
neighbourhood proportions against global proportions, which BBKNN manipulates
by design, and ASW requires an embedding. LISI on the stored kNN is the most
appropriate metric here.

```{r}
#| label: bbknn
sc_object <- bbknn_sc(
  object = sc_object,
  batch_column = "exp_id",
  no_neighbours_to_keep = 15L,
  bbknn_params = params_sc_bbknn(neighbours_within_batch = 10L)
)
```

```{r}
#| label: bbknn metrics
kbet_score_post_bbknn <- calculate_kbet_sc(
  sc_object,
  batch_column = "exp_id"
)
lisi_score_post_bbknn <- calculate_batch_lisi_sc(
  sc_object,
  batch_column = "exp_id"
)

kbet_score_post_bbknn
lisi_score_post_bbknn
```

```{r}
#| label: bbknn tsne
sc_object <- tsne_sc(object = sc_object, use_knn = TRUE)
```

```{r}
#| label: bbknn plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "tsne",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "BBKNN batch correction",
    colour = "Batch:"
  )
```

## Choosing a method

There is no universally best batch correction method. Some practical guidance:

- **Harmony** is a good default. It is fast, operates on the PCA embedding
directly, and handles multiple batch variables simultaneously. It tends to work
well when batch effects are moderate and cell type composition is broadly
similar across batches. Version 2 is quite an improvement and necessitates less
that the same cells are available across the batches and is even faster when
you provide only one batch effect to regress out.

- **fastMNN** is worth considering when cell type composition differs
substantially across batches. The mutual nearest neighbour approach is less
sensitive to this because it only aligns cells that have a plausible biological
match in another batch.

- **BBKNN** is useful when you want to avoid modifying the embedding at all.
The batch-balanced graph can be passed directly to graph-based clustering
algorithms. It is the lightest-touch approach but provides less control over
the degree of correction.

In practice, running multiple methods and looking at quantitative metrics is the 
most reliable approach. No single metric captures everything -- kBET and ASW
measure different aspects of embedding-based correction, and LISI is most 
informative for graph-based methods. UMAPs and tSNE assessments should be taken
with a big pinch of salt, see 
[Chari, et al.](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011288) 
Admittedly, they do look pretty however.

## Clean up

```{r}
#| label: clean up
unlink(tempdir_batch_cor, recursive = TRUE, force = TRUE)
```