---
title: "Single-cell data I/O in bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Single-cell data I/O in bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Andrew Skelton and Gregor Lueg"
---

## Intro

`bixverse` keeps single-cell counts on disk -- in compact Rust binaries for the
count matrix and a DuckDB for the `obs` / `var` metadata -- so that millions of
cells can be analysed on a laptop. Ingestion happens once, up front, through
format-specific readers that stream the data into that on-disk layout and apply
the basic cell/gene QC and normalisations along the way.

This vignette walks through every reader on a small synthetic data set, so it
runs end-to-end without any downloads. It covers:

- in-memory R objects (`load_r_data`)
- 10x CellRanger MTX, single and multi-sample (`load_mtx`, `load_multi_mtx`)
- AnnData `h5ad`, single and multi-sample (`load_h5ad`, `load_multi_h5ad`)
- 10x CellRanger HDF5, single and multi-sample (`load_tenx_h5`,
  `load_multi_tenx_h5`)
- checkpointing to and from disk (`save_sc_exp_to_disk`, `load_existing`)

If you have not seen the `SingleCells` class before, read the
[design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html)
vignette first.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(Matrix)
library(magrittr)
```

## A small synthetic data set

We generate a synthetic data set and treat its two halves as two "samples", so
that the multi-sample readers can be demonstrated too. The counts are a
cell-by-gene `dgRMatrix`; `obs` and `var` are `data.table`s.

```{r}
#| label: synthetic data
sim <- generate_single_cell_test_data()

counts <- sim$counts # cells x genes (dgRMatrix)
obs <- sim$obs
var <- sim$var
dim(counts)

# two "samples" for the multi-sample examples
idx_a <- 1:500
idx_b <- 501:1000
counts_a <- as(counts[idx_a, ], "RsparseMatrix")
counts_b <- as(counts[idx_b, ], "RsparseMatrix")

# scratch directories (one on-disk store per loaded object)
base <- file.path(tempdir(), "bixverse_io")
stores <- c(
  "mtx_a",
  "mtx_b",
  "mtx_swapped",
  "store_r",
  "store_mtx",
  "store_mtx_swapped",
  "store_multi_mtx",
  "store_h5",
  "store_multi_h5",
  "store_10x",
  "store_multi_10x"
)
for (s in stores) {
  dir.create(file.path(base, s), recursive = TRUE, showWarnings = FALSE)
}

# light QC thresholds, suitable for this small toy data set
qc <- params_sc_min_quality(
  min_unique_genes = 5L,
  min_lib_size = 10L,
  min_cells = 3L
)
```

## In-memory R objects

The simplest entry point: hand the reader a cell-by-gene `dgRMatrix` plus the
`obs` and `var` tables directly.

```{r}
#| label: load r data
sc_r <- SingleCells(dir_data = file.path(base, "store_r"))
sc_r <- load_r_data(
  object = sc_r,
  counts = counts,
  obs = obs,
  var = var,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_r
```

## 10x CellRanger MTX

### Single sample

We write one sample out in CellRanger flat-file form and point an explicit
`params_sc_mtx_io()` at the trio (for real CellRanger output, the convenience
helper `get_cell_ranger_params()` auto-locates `matrix.mtx` / `barcodes.tsv` /
`genes.tsv`).

Two `params_sc_mtx_io()` arguments handle the common deviations from the
strict CellRanger convention:

- `cells_as_rows` -- whether the .mtx has cells on the rows. The CellRanger
  default is `FALSE` (genes are rows). Set to `TRUE` for externally produced
  matrices that are transposed.
- `has_hdr` -- whether the barcodes/features flat files carry a header row.
  CellRanger itself writes headerless TSVs (`has_hdr = FALSE`), but CSV/TSV
  exports from downstream pipelines often carry one. The example below uses
  `has_hdr = TRUE` because `write_cellranger_output()` emits headers.

```{r}
#| label: write + load single mtx
dir_a <- file.path(base, "mtx_a")
# real CellRanger MTX is genes x cells, matching cells_as_rows = FALSE below
write_cellranger_output(
  dir_a,
  counts_a,
  obs[idx_a],
  var,
  rows = "genes",
  .verbose = FALSE
)

mtx_params <- params_sc_mtx_io(
  path_mtx = file.path(dir_a, "mat.mtx"),
  path_obs = file.path(dir_a, "barcodes.csv"),
  path_var = file.path(dir_a, "features.csv"),
  cells_as_rows = FALSE,
  has_hdr = TRUE
)

sc_mtx <- SingleCells(dir_data = file.path(base, "store_mtx"))
sc_mtx <- load_mtx(
  object = sc_mtx,
  sc_mtx_io_param = mtx_params,
  sc_qc_param = qc,
  mtx_streaming = FALSE,
  .verbose = FALSE
)
sc_mtx
```

### Multiple samples

`prescan_mtx_dirs()` takes one directory per sample, builds the shared gene
universe across them, and `load_multi_mtx()` then stacks all cells into a single
object with a global gene QC. The originating sample is recorded in the `exp_id`
`obs` column.

```{r}
#| label: write + load multi mtx
dir_b <- file.path(base, "mtx_b")
write_cellranger_output(
  dir_b,
  counts_b,
  obs[idx_b],
  var,
  rows = "genes",
  .verbose = FALSE
)

ps_mtx <- prescan_mtx_dirs(
  dirs = c(dir_a, dir_b),
  exp_ids = c("sample_a", "sample_b"),
  has_hdr = TRUE
)

sc_multi_mtx <- SingleCells(dir_data = file.path(base, "store_multi_mtx"))
sc_multi_mtx <- load_multi_mtx(
  object = sc_multi_mtx,
  prescan_result = ps_mtx,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_multi_mtx

# cells per sample
get_sc_obs(sc_multi_mtx)[, .N, by = exp_id]
```

### File layout variations

To make the `cells_as_rows` and `has_hdr` knobs concrete, here is the same
sample written with cells on the rows of the .mtx (i.e. transposed relative to
CellRanger). The only thing that changes at load time is `cells_as_rows = TRUE`:

```{r}
#| label: cells-as-rows mtx
dir_swapped <- file.path(base, "mtx_swapped")
write_cellranger_output(
  dir_swapped,
  counts_a,
  obs[idx_a],
  var,
  rows = "cells",
  .verbose = FALSE
)

mtx_params_swapped <- params_sc_mtx_io(
  path_mtx = file.path(dir_swapped, "mat.mtx"),
  path_obs = file.path(dir_swapped, "barcodes.csv"),
  path_var = file.path(dir_swapped, "features.csv"),
  cells_as_rows = TRUE,
  has_hdr = TRUE
)

sc_mtx_swapped <- SingleCells(dir_data = file.path(base, "store_mtx_swapped"))
sc_mtx_swapped <- load_mtx(
  object = sc_mtx_swapped,
  sc_mtx_io_param = mtx_params_swapped,
  sc_qc_param = qc,
  mtx_streaming = FALSE,
  .verbose = FALSE
)
sc_mtx_swapped
```

The resulting on-disk object is identical to the standard-layout single-sample
load above; only the path through the reader differs.

For headerless barcodes/features files (the strict CellRanger convention), set
`has_hdr = FALSE`. The same flag applies uniformly across all input directories
in `prescan_mtx_dirs()` for the multi-sample case.

## AnnData h5ad

### Single

`read_h5ad_metadata()` lets you inspect the `obs` / `var` and dimensions without
loading the counts; `load_h5ad()` ingests the matrix.

```{r}
#| label: write + load single h5ad
h5_a <- file.path(base, "sample_a.h5ad")
write_h5ad_sc(h5_a, counts_a, obs[idx_a], var, .verbose = FALSE)

meta <- read_h5ad_metadata(h5_a)
meta$dims

sc_h5 <- SingleCells(dir_data = file.path(base, "store_h5"))
sc_h5 <- load_h5ad(
  object = sc_h5,
  h5_path = h5_a,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_h5
```

### Multiple

`prescan_h5ad_files()` accepts a named vector of paths (the names become the
sample identifiers); `load_multi_h5ad()` mirrors the multi-MTX workflow.

```{r}
#| label: write + load multi h5ad
h5_b <- file.path(base, "sample_b.h5ad")
write_h5ad_sc(h5_b, counts_b, obs[idx_b], var, .verbose = FALSE)

ps_h5 <- prescan_h5ad_files(c(sample_a = h5_a, sample_b = h5_b))

sc_multi_h5 <- SingleCells(dir_data = file.path(base, "store_multi_h5"))
sc_multi_h5 <- load_multi_h5ad(
  object = sc_multi_h5,
  prescan_result = ps_h5,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_multi_h5
```

## 10x CellRanger HDF5

### Single sample

`write_tenx_h5_sc()` expects the counts plus a barcode vector and a `features`
table (with at least `id` and `name`). `load_tenx_h5()` reads the
gene-expression modality (V3 files may also carry e.g. Antibody Capture, which
is filtered out via `feature_type`).

```{r}
#| label: write + load 10x h5
features <- data.table(id = var$gene_id, name = var$ensembl_id)
h5_tenx <- file.path(base, "sample_a_10x.h5")
write_tenx_h5_sc(
  f_path = h5_tenx,
  counts = counts_a,
  barcodes = obs[idx_a]$cell_id,
  features = features,
  version = "v3"
)

sc_10x <- SingleCells(dir_data = file.path(base, "store_10x"))
sc_10x <- load_tenx_h5(
  object = sc_10x,
  h5_path = h5_tenx,
  sc_qc_param = qc,
  feature_type = "Gene Expression",
  .verbose = FALSE
)
sc_10x
```

### Multiple samples

`prescan_tenx_h5_files()` takes a named vector of `.h5` paths (the names become
the sample identifiers), inspects each file for its CellRanger version (v2 vs
v3), filters to the requested `feature_type`, and builds the global gene
space. `load_multi_tenx_h5()` then mirrors the multi-MTX / multi-h5ad
workflow.

`gene_universe` controls how the feature space is reconciled across files:

- `"intersection"` keeps genes present in **every** input.
- `"union"` keeps genes present in **any** input; cells from a file that lacks
  a given gene get zero counts for it.

Mixed v2 / v3 inputs in the same call are handled transparently, and non-gene
modalities (e.g. Antibody Capture) are silently dropped when `feature_type =
"Gene Expression"`.

```{r}
#| label: write + load multi 10x h5
features_b <- data.table(
  id = var$gene_id,
  name = var$ensembl_id,
  feature_type = "Gene Expression"
)
h5_tenx_b <- file.path(base, "sample_b_10x.h5")
write_tenx_h5_sc(
  f_path = h5_tenx_b,
  counts = counts_b,
  barcodes = obs[idx_b]$cell_id,
  features = features_b,
  version = "v3"
)

ps_10x <- prescan_tenx_h5_files(
  h5_paths = c(sample_a = h5_tenx, sample_b = h5_tenx_b),
  feature_type = "Gene Expression",
  gene_universe = "intersection",
  .verbose = FALSE
)

sc_multi_10x <- SingleCells(dir_data = file.path(base, "store_multi_10x"))
sc_multi_10x <- load_multi_tenx_h5(
  object = sc_multi_10x,
  prescan_result = ps_10x,
  sc_qc_param = qc,
  .verbose = FALSE
)
sc_multi_10x

# cells per sample
get_sc_obs(sc_multi_10x)[, .N, by = exp_id]
```

For multi-modal `.h5` files that carry both Gene Expression and Antibody
Capture features, load the gene side as above and read the ADT side
separately with `read_tenx_h5_adt()`, then attach it to a
`SingleCellsMultiModal` object via `add_adt_counts_sc()`.

## Checkpointing to disk

Once an object is loaded, the in-memory mappings can be persisted so a fresh
session can be restored without re-ingesting the counts. The on-disk binary and
DuckDB already live in `dir_data`; `save_sc_exp_to_disk()` adds the lightweight
mapping cache.

```{r}
#| label: save and restore
save_sc_exp_to_disk(sc_h5, type = "rds")

sc_restored <- SingleCells(dir_data = file.path(base, "store_h5"))
sc_restored <- load_existing(sc_restored)
sc_restored
```

## Summary

| Format | Single sample | Multiple samples |
| --- | --- | --- |
| In-memory R | `load_r_data()` | -- |
| 10x MTX | `load_mtx()` | `prescan_mtx_dirs()` + `load_multi_mtx()` |
| AnnData h5ad | `load_h5ad()` | `prescan_h5ad_files()` + `load_multi_h5ad()` |
| 10x HDF5 | `load_tenx_h5()` | `prescan_tenx_h5_files()` + `load_multi_tenx_h5()` |
| Seurat | `load_seurat()` | -- |

All readers share the same `sc_qc_param` QC interface and write into the same
on-disk `SingleCells` layout, so everything downstream (normalisation, HVGs,
PCA, neighbours, clustering, ...) is identical regardless of how the data
came in.
