---
title: "Spatial data I/O in bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Spatial data I/O in bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

`SpatialSpot` inherits from `SingleCells`, so counts land in the same on-disk
Rust binaries and the same DuckDB, and every reader in the single cell I/O
vignette still applies. What the spatial layer adds is geometry: spot
coordinates, the scale factors that relate them to the slide images, and the
images themselves.

Right now there is exactly one way in, `load_visium()`, and it wants a Space
Ranger output directory. Reading counts and coordinates from anywhere else is
not implemented yet. See the
[design doc](https://gregorlueg.github.io/bixverse/articles/design_spatial.html)
for what else is missing.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(magrittr)
```

Point this at your own Space Ranger output. Everything below runs against a
standard CytAssist FFPE capture area.

```{r}
#| label: paths
visium_dir <- path.expand("~/Desktop/visium_example_data")
list.files(visium_dir)
```

## What Space Ranger writes

Two directories matter.

```{r}
#| label: layout
list.files(file.path(visium_dir, "spatial"))
list.files(file.path(visium_dir, "raw_feature_bc_matrix"))
```

`spatial/` carries the geometry. `tissue_positions.csv` has one row per spot on
the capture area with its array coordinates, its full-resolution pixel
coordinates and an `in_tissue` flag. `scalefactors_json.json` relates
full-resolution pixels to each shipped image. The rest are images.

`*_feature_bc_matrix/` is a plain 10x MTX triplet, exactly as the single cell
readers expect it.

```{r}
#| label: positions
positions <- fread(file.path(visium_dir, "spatial", "tissue_positions.csv"))
head(positions, 3)
positions[, .N, by = in_tissue]
```

Just over a fifth of the 4,992 spots on the capture area sit off tissue. They
carry a handful of counts each and drag every downstream summary around if you
keep them.

## The barcode join

This is the single most important thing the reader does, and it is easy to get
silently wrong.

`barcodes.tsv.gz` is sorted lexicographically. `tissue_positions.csv` is in array
raster order. They contain the same barcodes in different orders.

```{r}
#| label: orders
barcodes <- fread(
  file.path(visium_dir, "raw_feature_bc_matrix", "barcodes.tsv.gz"),
  header = FALSE
)$V1

head(barcodes, 3)
head(positions$barcode, 3)
identical(sort(barcodes), sort(positions$barcode))
identical(barcodes, positions$barcode)
```

Same set, different order. Anything that zips the two together by position
assigns every spot somebody else's coordinates. The result still plots, still
clusters, and is wrong everywhere. `load_visium()` joins on the barcode, and
there is a test that builds a fixture where the two orders are a derangement of
each other, so a positional zip would be wrong for every single spot rather than
accidentally right for a few.

## Loading one sample

```{r}
#| label: load
data_dir <- file.path(tempdir(), "visium_io")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

object <- SpatialSpot(dir_data = data_dir)

object <- load_visium(
  object,
  visium_dir = visium_dir,
  sp_io_param = params_sp_visium_io(),
  sc_qc_param = params_sc_min_quality(),
  exp_id = "example",
  .verbose = TRUE
)

object
```

`params_sp_visium_io()` controls three things.

```{r}
#| label: io params
str(params_sp_visium_io())
```

`matrix_type` picks between `raw_feature_bc_matrix/` and
`filtered_feature_bc_matrix/`; `"auto"` prefers the filtered one and falls back.
`slide_file` registers a user-supplied full-resolution scanner slide, which
Space Ranger does not ship. `in_tissue_only` is the one that changes the numbers.

### in_tissue filtering happens at ingest

With `in_tissue_only = TRUE`, off-tissue spots are resolved to matrix column
indices by barcode and handed to the Rust reader as an allowlist. They never
reach the gene statistics and never get written. So `min_cells` in the QC
parameters sees in-tissue spots only, and `dims[1]` is the in-tissue count rather
than the full capture area.

```{r}
#| label: in tissue
S7::prop(object, "dims")
positions[in_tissue == 1, .N]
```

Turn it off and every spot is written:

```{r}
#| label: all spots
all_spots_dir <- file.path(tempdir(), "visium_io_all")
dir.create(all_spots_dir, recursive = TRUE, showWarnings = FALSE)

all_spots <- load_visium(
  SpatialSpot(dir_data = all_spots_dir),
  visium_dir = visium_dir,
  sp_io_param = params_sp_visium_io(in_tissue_only = FALSE),
  exp_id = "example",
  copy_images = FALSE,
  .verbose = TRUE
)

S7::prop(all_spots, "dims")
```

More spots, and a different gene count, because gene QC now sees the off-tissue
background too.

## The sample record

Each loaded capture area is registered as a `SpatialSample`.

```{r}
#| label: sample
sample <- get_sample(object, "example")
sample
```

It holds the scale factors verbatim from the JSON, plus one derived entry.

```{r}
#| label: scale factors
str(sample$scale_factors)
```

`cytassist_scalef` is not in the file. Space Ranger writes
`regist_target_img_scalef`, which describes the registration *target*, not the
CytAssist image it actually ships. The reader derives the real one from the
dimensions of the file on disk. Use `regist_target_img_scalef` for the CytAssist
frame and your tiles land in the wrong place.

## Coordinates

```{r}
#| label: coords
coords <- get_spatial_coords(object, "example")
head(coords, 3)
dim(coords)
```

Two columns, `x` then `y`, in full-resolution pixels. Rows come back in ascending
`cell_idx`, which is the same order as `get_spot_indices_for_exp()` and the same
order the spatial graph is indexed in. Row `i` of the coordinate matrix pairs
with element `i` of the index vector.

Note the crossing: `x` is `pxl_col_in_fullres` and `y` is `pxl_row_in_fullres`.
Naming in the positions file is row-first, plotting is x-first, and the two are
deliberately swapped here so the coordinate matrix plots the right way up.

```{r}
#| label: coord check
obs <- get_sc_obs(object)
setorderv(obs, "cell_idx")
identical(as.numeric(coords[, "x"]), as.numeric(obs$pxl_col_in_fullres))
```

## Images

Images get copied into `<dir_data>/images/<exp_id>/` at ingest and the sample
stores paths relative to the data directory, so the whole thing stays portable.

```{r}
#| label: image paths
sample$image_paths
```

```{r}
#| label: get image
hires <- get_image(object, "example", "hires")
dim(hires)
```

`get_image()` reads PNG and JPEG. It does not read TIFF, and Space Ranger ships
the CytAssist frame as a TIFF:

```{r}
#| label: cytassist fails
#| error: true
get_image(object, "example", "cytassist")
```

The Rust image backend has no such problem. It reads the header without decoding
the file:

```{r}
#| label: image metadata
cytassist_path <- file.path(
  S7::prop(object, "dir_data"),
  sample$image_paths[["cytassist"]]
)
str(rs_sp_image_metadata(cytassist_path))
```

`n_levels` is 1 and `vendor` is `NA` because this is a plain TIFF rather than a
pyramidal whole-slide image. On a real slide those would report the pyramid
table and the scanner vendor. And `image_features_sp(resolution = "cytassist")`
works on this data even though `get_image()` cannot touch it, which surprises
everyone including me.

With `copy_images = FALSE` nothing gets registered at all, since the class only
stores paths under the data directory:

```{r}
#| label: no images
get_sample(all_spots, "example")$image_paths
```

## Several samples

Multiple capture areas go through a prescan that builds the gene intersection
across inputs, then a single load with global gene QC and sequential spot
indexing.

```{r}
#| label: prescan
multi_dir <- file.path(tempdir(), "visium_io_multi")
dir.create(multi_dir, recursive = TRUE, showWarnings = FALSE)

# the same slide twice, purely so this runs with one download
prescan <- prescan_visium_dirs(
  dirs = c(visium_dir, visium_dir),
  exp_ids = c("slide_a", "slide_b"),
  .verbose = FALSE
)

prescan$universe_size
names(prescan$file_tasks)
```

```{r}
#| label: load multi
multi <- load_multi_visium(
  SpatialSpot(dir_data = multi_dir),
  prescan_result = prescan,
  sc_qc_param = params_sc_min_quality(),
  copy_images = FALSE,
  .verbose = FALSE
)

multi
get_sample_ids(multi)
```

Each input is registered as its own `SpatialSample`, and the obs table carries an
`exp_id` column so everything downstream can split on it.

```{r}
#| label: multi obs
get_sc_obs(multi)[, .N, by = exp_id]
```

### The gene QC differs from the single-sample path

`load_visium()` applies `in_tissue_only` at ingest through a cell allowlist.
`load_multi_visium()` cannot: the multi-file reader has no allowlist, so it
writes every spot and applies the filter afterwards through the `to_keep` flag.

Global gene QC therefore sees the off-tissue counts, and `min_cells` is evaluated
against the full capture areas rather than the in-tissue ones. The two paths give
different gene sets on the same input:

```{r}
#| label: gene counts
S7::prop(object, "dims")[2]
S7::prop(multi, "dims")[2]
```

Not a large difference on this slide, and not something you would notice unless
you looked. Worth knowing if you compare a single-sample run against a
multi-sample one.

## Where to go next

The [worked example](https://gregorlueg.github.io/bixverse/articles/visium_spatial.html)
takes this object through the spatial graph, spatially variable genes,
neighbourhood enrichment and image features.
