---
title: "Reference mapping with Symphony"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Reference mapping with Symphony}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: 
  - name: "Gregor Lueg"
---

## Intro

Once you have a well-annotated reference data set, you usually do not want to
re-do the whole pipeline... QC, HVG selection, PCA, batch correction,
clustering, manual annotation... And that every time a new sample lands on your 
desk. Symphony ([Kang, et al., 2021](https://doi.org/10.1038/s41467-021-25957-x))
addresses this by compressing the reference into a small set of cached terms
that can be used to map new query cells into the same coordinate system in a
single shot, with no need to retain the original reference cells at query
time.

This vignette walks through the full workflow:

1. Build a reference from two PBMC data sets (pbmc3k + pbmc4k), correcting
the batch effect with Harmony v2 and annotating cell types with scType.
2. Compress the reference into a `SymphonyReference` and snapshot the cell 
type labels into it.
3. Load a third PBMC data set (pbmc8k) as the query, project it through the
reference, and transfer labels via kNN majority vote.

If you have not read the 
[batch correction vignette](https://gregorlueg.github.io/bixverse/articles/single_cell_batch_corrections.html)
and the
[introductory vignette](https://gregorlueg.github.io/bixverse/articles/thinking_single_cell.html),
please do so first; this vignette assumes familiarity with both.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(data.table)
library(ggplot2)
library(magrittr)
```

## Building the reference

### Loading the reference batches

We use the same pbmc3k + pbmc4k pair as in the batch correction vignette. Two
batches with the same broad cell type composition and a clear technical
difference between them is exactly the situation Symphony is designed for:
the reference learns the batch correction once, and that correction can then
be applied to future queries without rerunning Harmony.

```{r}
#| label: load reference
#| code-fold: true
dir_data <- download_pbmc_batches()

tempdir_ref <- file.path(tempdir(), "symphony_ref")
dir.create(tempdir_ref, showWarnings = FALSE, recursive = TRUE)

h5ad_files <- list.files(dir_data)
h5ad_files <- h5ad_files[grepl(".h5ad", h5ad_files)]
h5ad_paths <- file.path(dir_data, h5ad_files)
names(h5ad_paths) <- c("pbmc3k", "pbmc4k")

h5_tasks <- prescan_h5ad_files(h5_paths = h5ad_paths)

sc_ref <- SingleCells(dir_data = tempdir_ref)

sc_ref <- load_multi_h5ad(
  object = sc_ref,
  prescan_result = h5_tasks,
  .verbose = TRUE
)
```

### Quality control

Standard QC: mitochondrial gene proportions, library complexity, MAD-based
outlier detection.

```{r}
#| label: reference qc
#| code-fold: true
var_ref <- get_sc_var(sc_ref)

# let's get the gene symbols from one of the h5ad files - in multi file import
# additional columns are dropped from the vars
h5_metadata <- read_h5ad_metadata(h5ad_paths[[1]])

var_ref <- merge(
  var_ref,
  h5_metadata$var[, c("ENSEMBL_ID", "Symbol_TENx")],
  by.x = "gene_id",
  by.y = "ENSEMBL_ID"
)

setnames(var_ref, old = "Symbol_TENx", new = "gene_symbol", skip_absent = TRUE)

gs_of_interest <- list(
  MT = var_ref[grepl("^MT-", gene_symbol), gene_id],
  Ribo = var_ref[grepl("^RPS|^RPL", gene_symbol), gene_id]
)

sc_ref <- gene_set_proportions_sc(
  sc_ref,
  gs_of_interest,
  streaming = FALSE,
  .verbose = TRUE
)

qc_df <- sc_ref[[c("cell_id", "lib_size", "nnz", "MT", "exp_id")]]

metrics <- list(
  log10_lib_size = log10(qc_df$lib_size),
  log10_nnz = log10(qc_df$nnz),
  MT = qc_df$MT
)
directions <- c(
  log10_lib_size = "twosided",
  log10_nnz = "twosided",
  MT = "above"
)

qc <- run_cell_qc(
  metrics = metrics,
  cells_to_keep = get_cells_to_keep(sc_ref),
  directions = directions,
  threshold = 3
)

sc_ref[["outlier"]] <- qc$combined
cells_to_keep <- qc_df[!qc$combined, cell_id]
sc_ref <- set_cells_to_keep(sc_ref, cells_to_keep)
```

### HVGs, PCA, Harmony v2

This block is essentially the Harmony v2 path of the batch correction
vignette: pick HVGs, run PCA, run Harmony v2, build a neighbour graph on the
corrected embedding.

```{r}
#| label: reference preprocessing
sc_ref <- find_hvg_sc(
  object = sc_ref,
  hvg_no = 2000L,
  .verbose = TRUE
)

sc_ref <- calculate_pca_sc(
  object = sc_ref,
  no_pcs = 30L
)

sc_ref <- harmony_v2_sc(
  object = sc_ref,
  batch_column = "exp_id",
  harmony_params = params_sc_harmony_v2()
)

sc_ref <- find_neighbours_sc(
  object = sc_ref,
  embd_to_use = "harmony_v2",
  neighbours_params = params_sc_neighbours()
)
```

We use Harmony v2 throughout. Version 1 is still available via
`harmony_sc()` and `params_sc_harmony()`, but v2 is preferable in most
settings (see the batch correction vignette for the reasoning).

### Clustering and scType annotation

Cluster on the Harmony-corrected graph, then annotate the clusters with
scType using a PBMC marker set. This mirrors the annotation block in the
PBMC vignette.

```{r}
#| label: reference clustering
sc_ref <- find_clusters_sc(sc_ref, res = 0.5, name = "leiden_clusters")

sc_ref <- umap_sc(
  sc_ref,
  slot_name = "umap_harm_v2",
  use_knn = TRUE
)
```

Marker set:

```{r}
#| label: marker set
cell_markers <- c(
  CD3D = "T cells",
  CD3E = "T cells",
  CD3G = "T cells",
  IL7R = "CD4+ T",
  CD4 = "CD4+ T",
  CD8A = "CD8+ T",
  CD8B = "CD8+ T",
  MS4A1 = "B cells",
  CD79A = "B cells",
  CD19 = "B cells",
  CD14 = "CD14+ Mono",
  LYZ = "CD14+ Mono",
  S100A8 = "CD14+ Mono",
  FCGR3A = "CD16+ Mono",
  CDKN1C = "CD16+ Mono",
  GNLY = "NK",
  NKG7 = "NK",
  NCAM1 = "NK",
  FCER1A = "mDC",
  CD1C = "mDC",
  LILRA4 = "pDC",
  CLEC4C = "pDC",
  PPBP = "Platelet",
  PF4 = "Platelet"
)
```

scType scoring and per-cluster assignment:

```{r}
#| label: sctype
cell_markers_dt <- stack(cell_markers) %>%
  as.data.table() %>%
  setnames(., c("values", "ind"), c("cell_type", "gene_symbol")) %>%
  .[, gene_symbol := as.character(gene_symbol)] %>%
  .[, gene_id := var_ref$gene_id[match(gene_symbol, var_ref$gene_symbol)]] %>%
  .[!is.na(gene_id), ]

marker_list <- prepare_cell_markers(sc_ref, cell_markers_dt)

sctype_scores <- calc_sc_type_scores(
  object = sc_ref,
  cell_marker_list = marker_list
)

cell_type_anno <- score_clusters(
  sctype_scores,
  sc_ref[[]][["leiden_clusters"]]
)

obs <- get_sc_obs(sc_ref, filtered = TRUE)[, .(
  cell_idx,
  leiden_clusters
)] %>%
  .[,
    sc_type := cell_type_anno$cell_type[match(
      leiden_clusters,
      cell_type_anno$cluster_id
    )]
  ]

sc_ref[["sc_type"]] <- obs$sc_type
```

Quick sanity check on the reference: batches mixed, cell types separated.

```{r}
#| label: reference batch plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_ref,
  embedding = "umap_harm_v2",
  colour_by = "exp_id",
  label_by = "exp_id",
  discrete = TRUE
) +
  labs(
    title = "Reference (Harmony v2 corrected)",
    colour = "Batch:"
  )
```

```{r}
#| label: reference celltype plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_ref,
  embedding = "umap_harm_v2",
  colour_by = "sc_type",
  discrete = TRUE
) +
  labs(
    title = "Reference cell types (scType)",
    colour = "Cell type:"
  )
```

Let's see how sensible the cell type labels are with some canonical markers

```{r}
#| label: check expression patterns - reference
feature_plot_sc(
  object = sc_ref,
  features = c(
    "ENSG00000167286",
    "ENSG00000153563",
    "ENSG00000177455",
    "ENSG00000170458"
  ),
  feature_labels = c(
    "ENSG00000167286" = "CD3D",
    "ENSG00000153563" = "CD8A",
    "ENSG00000177455" = "CD19",
    "ENSG00000170458" = "CD14"
  ),
  embedding = "umap_harm_v2"
)
```

### Compressing into a SymphonyReference

We now have everything we need: HVG selection, batch annotation, cell type
labels in the obs table. `build_symphony_ref()` re-runs PCA and Harmony
internally using the supplied HVGs, then caches the loadings, centroids and
compression terms (`Nr`, `C`) needed for query mapping. Pass `label_columns`
to snapshot one or more obs columns -- here the scType annotation -- into
the reference for downstream label transfer.

A note on the apparent duplication: yes, indeed, PCA + Harmony are run again 
inside `build_symphony_ref()`, even though we just ran them on the same HVGs in 
the preprocessing block above. This is deliberate. The standard pipeline keeps
its embeddings on the `SingleCells` object and streams from disk; the
Symphony reference keeps a self-contained, in-memory copy of everything it
needs to project queries without the reference cells. Sharing state across
those two representations would tie them together in unhelpful ways. The
cost is one extra PCA + Harmony run at build time while avoiding coupling
between the methods.

```{r}
#| label: build symphony ref
hvg_indices <- get_hvg(sc_ref) + 1L

symphony_ref <- build_symphony_ref(
  object = sc_ref,
  batch_column = "exp_id",
  hvg = hvg_indices,
  harmony_params = params_sc_harmony_v2(),
  pca_params = params_sc_pca(),
  no_pcs = 30L,
  label_columns = "sc_type",
  .verbose = TRUE
)

symphony_ref
```

The print method tells you the Harmony backend, the size of the loadings
and centroid matrices, the batch variables and the stored label columns.

## Mapping a query

### Loading the query

We use pbmc8k as the query. Same tissue, same broad biology, but a
different 10x run -- exactly the kind of "another sample arrives" scenario
Symphony is built for.

```{r}
#| label: load query
#| code-fold: true
# TODO: replace with the proper download function once pbmc8k is on Zenodo.
# pbmc8k_path <- download_pbmc8k()

pbmc8k_path <- "~/Desktop/zenodo_data/pmbc-8k/"

tempdir_query <- file.path(tempdir(), "symphony_query")
dir.create(tempdir_query, showWarnings = FALSE, recursive = TRUE)

mtx_io_params <- get_cell_ranger_params(pbmc8k_path)

sc_query <- SingleCells(dir_data = tempdir_query)

sc_query <- load_mtx(
  object = sc_query,
  sc_mtx_io_param = mtx_io_params,
  mtx_streaming = FALSE,
  .verbose = TRUE
)
```

### Quality control on the query

Same QC as the reference. We are not trying to match library sizes or
anything fancy across data sets -- just remove clearly bad cells.

```{r}
#| label: query qc
#| code-fold: true
var_q <- get_sc_var(sc_query)

gs_q <- list(
  MT = var_q[grepl("^MT-", column1), gene_id],
  Ribo = var_q[grepl("^RPS|^RPL", column1), gene_id]
)

sc_query <- gene_set_proportions_sc(
  sc_query,
  gs_q,
  streaming = FALSE,
  .verbose = TRUE
)

qc_df_q <- sc_query[[c("cell_id", "lib_size", "nnz", "MT")]]

metrics_q <- list(
  log10_lib_size = log10(qc_df_q$lib_size),
  log10_nnz = log10(qc_df_q$nnz),
  MT = qc_df_q$MT
)

qc_q <- run_cell_qc(
  metrics = metrics_q,
  cells_to_keep = get_cells_to_keep(sc_query),
  directions = c(
    log10_lib_size = "twosided",
    log10_nnz = "twosided",
    MT = "above"
  ),
  threshold = 3
)

sc_query[["outlier"]] <- qc_q$combined
sc_query <- set_cells_to_keep(
  sc_query,
  qc_df_q[!qc_q$combined, cell_id]
)
```

### Projecting the query

`map_symphony_query()` does the actual projection: query cells are
standardised against the reference's per-HVG means and SDs, projected
through the reference PCA loadings, and assigned soft cluster weights
against the reference centroids. Those weights then drive the cached MoE
batch correction, putting query cells into the same `z_corr` coordinate
system as the reference.

Genes are matched by name. Missing HVGs are treated as zero columns; the
function reports how many were missing.

Even though pbmc8k is a single batch, we still pass `batch_column = "exp_id"`
so Symphony applies the cached correction relative to the reference batches.
Setting `batch_column = NULL` skips correction entirely (the resulting
`symphony` embedding then equals `symphony_pca`).

```{r}
#| label: map query
sc_query <- map_symphony_query(
  reference = symphony_ref,
  query = sc_query,
  batch_column = NULL,
  params = params_symphony_map(),
  .verbose = TRUE
)
```

The query now carries three embeddings: `"symphony"` (z_corr, the corrected
embedding), `"symphony_pca"` (z_pca, pre-correction), and `"symphony_r"`
(soft cluster assignments, one column per reference centroid).

A neighbour graph and UMAP on the corrected embedding give us something to
plot against:

```{r}
#| label: query umap
sc_query <- find_neighbours_sc(
  object = sc_query,
  embd_to_use = "symphony",
  neighbours_params = params_sc_neighbours()
)

sc_query <- umap_sc(
  sc_query,
  embd_to_use = "symphony",
  slot_name = "umap_symphony",
  use_knn = TRUE
)
```

### Transferring labels

`transfer_labels_symphony()` does a kNN majority vote on the reference's
`z_corr` (the searchable index) using the query's `"symphony"` embedding.
It reads labels directly from the reference's stored `labels` slot -- the
one we populated via `label_columns = "sc_type"` at build time.

Distances on `z_corr` are typically Euclidean; the default `params_sc_knn()`
matches.

```{r}
#| label: transfer labels
predicted <- transfer_labels_symphony(
  reference = symphony_ref,
  query = sc_query,
  label_column = "sc_type",
  knn_params = params_sc_knn(),
  .verbose = TRUE
)

head(predicted)
```

The result is a `data.table` in `get_cells_to_keep(query)` order with a
predicted label and a confidence score (fraction of neighbours agreeing on
the winning class). We can attach both directly to the query obs.

```{r}
#| label: attach predictions
sc_query[[c(
  "predicted_sc_type",
  "sc_type_confidence"
)]] <- list(
  predicted_sc_type = predicted$predicted_sc_type,
  sc_type_confidence = predicted$confidence_sc_type
)
```

### Visualising the result

Query UMAP coloured by transferred cell type:

```{r}
#| label: query celltype plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_query,
  embedding = "umap_symphony",
  colour_by = "predicted_sc_type",
  label_by = "predicted_sc_type",
  discrete = TRUE
) +
  labs(
    title = "pbmc8k - transferred cell types",
    colour = "Cell type:"
  )
```

Same UMAP, but coloured by transfer confidence. Cells in the "interior" of
their cell type usually have confidence 1; cells sitting on cluster
boundaries or in cell types poorly represented in the reference are the
ones to scrutinise.

```{r}
#| label: query confidence plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_query,
  embedding = "umap_symphony",
  colour_by = "sc_type_confidence",
  discrete = FALSE
) +
  labs(
    title = "pbmc8k - kNN transfer confidence",
    colour = "Confidence:"
  )
```

And a per-cell-type confidence distribution -- useful for spotting cell
types where the reference and query disagree systematically:

```{r}
#| label: confidence boxplot
#| fig-width: 7
#| fig-height: 4
conf_dt <- sc_query[[c("predicted_sc_type", "sc_type_confidence")]]

ggplot(conf_dt, aes(x = predicted_sc_type, y = sc_type_confidence)) +
  geom_violin(scale = "width", fill = "grey80") +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.4) +
  labs(
    x = NULL,
    y = "Transfer confidence",
    title = "Per-cell-type kNN agreement on pbmc8k"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
  ylim(0, 1.025)
```

Let's also sanity check the label propagation with some canonical markers:

```{r}
#| label: check expression patterns - query
feature_plot_sc(
  object = sc_query,
  features = c(
    "ENSG00000167286",
    "ENSG00000153563",
    "ENSG00000177455",
    "ENSG00000170458"
  ),
  feature_labels = c(
    "ENSG00000167286" = "CD3D",
    "ENSG00000153563" = "CD8A",
    "ENSG00000177455" = "CD19",
    "ENSG00000170458" = "CD14"
  ),
  embedding = "umap_symphony"
)
```

## Slim references and post-hoc labels

The reference above stores `z_corr`, `z_orig`, the soft cluster assignments
`r` and the labels. For large references that get checked into shared
storage, `z_orig` and `r` are usually dead weight; passing `slim = TRUE` at
build time drops them. `z_corr` is always retained -- it is what kNN label
transfer searches against.

If you want to add or update labels after the reference is already built
(say, you re-ran scType with a different marker set), use
`add_symphony_labels()`. It reads obs columns from a `SingleCells` object
in `cells_to_keep` order and snapshots them into the reference. The
`cells_to_keep` state must match the reference -- the function enforces
this with a hard error if the row count drifts.

```{r}
#| label: slim plus add labels
#| eval: false
slim_ref <- build_symphony_ref(
  object = sc_ref,
  batch_column = "exp_id",
  hvg = hvg_indices,
  harmony_params = params_sc_harmony_v2(),
  pca_params = params_sc_pca(),
  no_pcs = 30L,
  slim = TRUE
)

slim_ref <- add_symphony_labels(
  reference = slim_ref,
  sc_object = sc_ref,
  columns = "sc_type"
)
```

Multiple label columns can be attached at once (`columns = c(...)`),
existing columns are protected unless you pass `overwrite = TRUE`, and
`get_symphony_labels(reference)` returns the stored `data.table` (or `NULL`
if none).

## Clean up

```{r}
#| label: clean up
unlink(tempdir_ref, recursive = TRUE, force = TRUE)
unlink(tempdir_query, recursive = TRUE, force = TRUE)
```