---
title: "Thinking about single cells the bixverse way"
vignette: >
  %\VignetteIndexEntry{Thinking about single cells the bixverse way}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author: "Gregor Lueg and Grant Neilson"
---

## Intro

Before diving into this vignette, it would be really recommended to read
the [design choices](https://gregorlueg.github.io/bixverse/articles/design_single_cell.html) 
to understand **why** single cell has been implemented the way it is in this
package. This will make understanding the downstream text much easier. The aim
of this vignette is to take some small synthetic data and point out some quirks 
due to design choices made.

```{r}
#| label: setup
library(bixverse)
```

### Synthetic data

Let's start with some synthetic data that is powering all the tests in the
package. It's very simple data with 1,000 cells, 100 genes and three very
cleary defined cell groups. Genes 1 to 10 defined cell_type 1, genes 11 to 20
the next cell type and 21 to 30 the last cell type. The remaining genes are
noise genes.

```{r}
#| label: loading in synthetic data

single_cell_test_data <- generate_single_cell_test_data()

str(single_cell_test_data)
```

We have a count matrix with pseudo raw counts, an obs table and a var table. Let
us load this into a `bixverse` SingleCell class and explore what's actually
happening.

### Loading in data

Initialising an experiment works like this. We will not save the data into
any persistent storage, but just leave it as is... If you read the design
docs, you know why you have to provide a path here. If you do not know why,
go back to the docs.

```{r}
#| label: generating the single cell object
test_dir <- file.path(tempdir(), "single_cell_test")
dir.create(test_dir, showWarnings = FALSE, recursive = TRUE)

sc_object <- SingleCells(dir_data = test_dir)

sc_object
```

Is something in the directory?

```{r}
#| label: files pre loading
list.files(test_dir)
```

Nope. But let's now load in the data and see what happens:

```{r}
#| label: load synthetic data to disk

sc_object <- load_r_data(
  object = sc_object,
  counts = single_cell_test_data$counts,
  obs = single_cell_test_data$obs,
  var = single_cell_test_data$var,
  # we let all cells pass for now
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 0L,
    min_lib_size = 0L,
    min_cells = 0L,
    target_size = 1e3 # target size of 1000
  ),
  streaming = FALSE,
  .verbose = TRUE
)

list.files(test_dir)
```

So, what has just happened ... ? Actually quite a lot of things:

1. The R matrix is being pulled into Rust and checked against the number of 
   min_unique_genes, the min_lib_size and in how many cells a given gene needs
   to expressed to be included. Once the genes and cells have been identified
   that pass the minimum threshold (in that order: first gene filtering -> then 
   check the cells against that reduced feature space), the cells are written
   to the first binary file: `counts_cells.bin`. At the same time the counts
   are normalised: division by library size, projected against desired target
   size and log transformed.
2. Rust now transposes under the hood the data that is at the moment very nice
   to read in from a cell-centric level, less so from a gene-centric level. 
   That is why we also have `counts_genes.bin`.
3. Lastly, the observation and variable data is being reduced based on step 1
   and written to the `sc_duckdb.db` and stored there.

If we now print the class:

```{r}
#| label: print the class again

sc_object
```

We can now see the number of cells and genes stored in the object. 

### Interacting with the data

Let's now look into how we interact with the data... We will actually apply
some filters here (the synthetic data is weird, but does the job of assumption
testing).

```{r}
#| label: load synthetic data to disk (more realistic)

sc_object <- load_r_data(
  object = sc_object,
  counts = single_cell_test_data$counts,
  obs = single_cell_test_data$obs,
  var = single_cell_test_data$var,
  # we let all cells pass for now
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 45L,
    min_lib_size = 300L,
    min_cells = 500L, # the data is weird and a lot cells express genes
    target_size = 1e3 # target size of 1000
  ),
  streaming = FALSE,
  .verbose = TRUE
)

sc_object
```

We now have overwritten the files on disk for the filtered versions. 929 cells
and 81 genes pass the threshold. But how do I interact with the data?

#### Understanding mapping and filtering

As data is stored on disk, we need to carefully synchronise state here between
the DuckDB, the binary files and tell the Rust kernel which cells to use and
what to do... There is a bunch of helpers for this. You might have realised in
the print like `To keep n`. This is very important! Let's check this out

```{r}
#| label: cells to keep

get_cells_to_keep(sc_object)[1:10]
```

What is cells to keep? Well, these are the (0-based) indices of cells that shall
be included in the analysis, think HVG selection, PCA, kNN generation, etc. This
is **THE** foot gun of this package. The `get_cells_to_keep()` indices will be
the ones loaded in by Rust. The other cells are not lost, they are still on
disk (a massive difference between filtering in memory and on-disk work!). Also
the DuckDB state will get updated, but let's get to this later. We also have 
other getters.

```{r}
#| label: getters

# get the cell identifiers - filtered for genes to keep; if set to FALSE
# everything will be returned
print(get_cell_names(sc_object, filtered = TRUE)[1:10])

print(get_gene_names(sc_object)[1:10])
```

Let's quickly check out how you access the obs table and then let's exemplify
THE footgun of the paper:

```{r}
#| label: accessing the obs - via `[[]]`

# a more R-native way
sc_object[[]]
```

```{r}
#| label: accessing the obs - subset or rows

# a more R-native way: only first three rows
sc_object[[1:3L]]
```

```{r}
#| label: accessing the obs - column selection

# a more R-native way: specific columns
sc_object[[c("cell_idx", "cell_id", "cell_grp")]]
```

All of these ways above call R primitives. There is also a more general getter
here:

```{r}
#| label: accessing the obs - general getter

get_sc_obs(sc_object, filtered = FALSE)
# this one can retrieve unfiltered cells!
```

Let's pretend we want to do mitochondrial (or other gene set-based removal)

```{r}
#| label: pseudo gene filtering

# pseudo gene sets
gs_of_interest <- list(
  gs_1 = c("gene_001", "gene_002", "gene_003", "gene_004"),
  gs_2 = c("gene_096", "gene_097", "gene_100")
)

# calculate the gene set proportions
sc_object <- gene_set_proportions_sc(
  sc_object,
  gs_of_interest,
  .verbose = TRUE
)
```

If we now look at the obs again:

```{r}
#| label: accessing the obs - new data added

sc_object[[]]
```

We now have two new columns here! That worked well... Let's now do filtering
to really understand what's happening under the hood:

```{r}
#| label: setting new cells to keep

threshold <- 0.05

cells_to_keep <- sc_object[[]][gs_2 < threshold, cell_id]

sc_object <- set_cells_to_keep(sc_object, cells_to_keep)

sc_object
```

The print changed now! We have now told the class that only a subset of 682
cells shall be used in any downstream analyses. This is very important! Most
functions will only return and work on the filtered cells, and you have to 
carefully think and remember what you did when. Also:

```{r}
#| label: differences in obs row numbers

obs_unfiltered <- get_sc_obs(sc_object, filtered = FALSE)

obs_filtered <- get_sc_obs(sc_object, filtered = TRUE)

obs_r_primitive <- sc_object[[]]

cat(sprintf("No of rows unfiltered: %i\n", nrow(obs_unfiltered)))
cat(sprintf("No of rows filtered: %i\n", nrow(obs_filtered)))
cat(sprintf("No of rows from R primitive: %i\n", nrow(obs_r_primitive)))
```

If you understand the concept shown above, you are good to go to analyse 
millions of cells on small compute without having 100's of GB of memory 
available. Let's just talk also about the 2nd big foot gun: **the counts**.

#### Understanding count retrieval

You have actually two files on disk... If you read the design doc you know
why (if you don't, please read it!). Now let's look into what it ACTUALLY 
means... Let's say you are interested in the first 50 cells.

```{r}
#| label: different file formats for the counts

raw_counts_from_cells <- sc_object[
  1:50L,
  ,
  assay = "raw",
  return_format = "cell"
]

print(class(raw_counts_from_cells))

raw_counts_from_genes <- sc_object[
  1:50L,
  ,
  assay = "raw",
  return_format = "gene"
]

print(class(raw_counts_from_genes))
```

What is happening here? In both cases, we are pulling the raw counts. In the
first case, we do it smartly... Rows -> cells, so, we are telling R to tell
Rust to retrieve this as CSR. Which means we ONLY load the cells in that are
needed. The second approach is the stupid way... We are telling R to tell Rust
to load in ALL genes and then filter to the desired cells. The impact of this
can be seen here:

```{r}
#| label: impact of the different retrieval

microbenchmark::microbenchmark(
  the_correct_way = {
    sc_object[
      1:50L,
      ,
      assay = "raw",
      return_format = "cell"
    ]
  },
  the_incorrect_way = {
    sc_object[
      1:50L,
      ,
      assay = "raw",
      return_format = "gene"
    ]
  },
  times = 10L
)
```

The difference seems marginal here, but it WILL bite you if you do this on
large data sets. Also, the memory advantages this approach gives you usually
will disappear if you force the wrong format. Also, another thing to REALLY 
consider... The indices you supply in R should match what the Rust files have
and again you need to THINK about the cells to keep. A nasty example here:

```{r}
#| label: the count getter in more detail

counts_filtered <- get_sc_counts(
  object = sc_object,
  assay = "raw",
  return_format = "gene",
  # get from the R perspective cells 1:10 that were written to disk
  cell_indices = 1:10L,
  # get from the R perspective genes 1:10 that were written to disk
  gene_indices = 1:10L,
  use_cells_to_keep = TRUE
)

counts_unfiltered <- get_sc_counts(
  object = sc_object,
  assay = "raw",
  return_format = "gene",
  # get from the R perspective cells 1:10 that were written to disk
  cell_indices = 1:10L,
  # get from the R perspective genes 1:10 that were written to disk
  gene_indices = 1:10L,
  use_cells_to_keep = FALSE
)

cat(sprintf(
  "Number of cells: %i | Number of genes: %i (filtered version)\n",
  nrow(counts_filtered),
  ncol(counts_filtered)
))
cat(sprintf(
  "Number of cells: %i | Number of genes: %i (unfiltered version)\n",
  nrow(counts_unfiltered),
  ncol(counts_unfiltered)
))
```

Wait, what happened here ... ? Because we set the cells to keep in the step
with the pseudo mitochondrial counts, these indices are now "bad cells". **They
are not filtered, they are just not returned.** You can control this also via
the primitive syntax.
```{r}
#| label: filtering with primitives

filtered_counts <- sc_object[1:10L, 1:10L, use_cells_to_keep = TRUE]

unfiltered_counts <- sc_object[1:10L, 1:10L, use_cells_to_keep = FALSE]

cat(sprintf(
  "Number of cells: %i | Number of genes: %i (filtered version - primitive)\n",
  nrow(counts_filtered),
  ncol(counts_filtered)
))
cat(sprintf(
  "Number of cells: %i | Number of genes: %i (unfiltered version - primitive)\n",
  nrow(counts_unfiltered),
  ncol(counts_unfiltered)
))
```

If you want to avoid thinking about indices altogether (which is recommended),
just use:

```{r}
#| label: using cell and gene identifiers

# just use cell labels and genes directly...
sc_object[
  c("cell_0001", "cell_0003", "cell_0100", "cell_0117"),
  c("gene_001", "gene_002", "gene_003", "gene_081")
]
```

And that's it. If you understand this vignette and the design choices, we have
answered the *why*, the *how* and can move on to the *what* in the other 
vignettes with some example analyses and how to actually use the package to 
analyse single cell experiments.