---
title: "Trajectory inference with Palantir and PAGA"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Trajectory inference with Palantir and PAGA}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Two trajectory methods live in `bixverse`, both operating on the kNN graph you
already built with `find_neighbours_sc()`.

**Palantir** ([Setty et al.,
2019](https://doi.org/10.1038/s41587-019-0068-4)) models differentiation as a
Markov chain over a multiscale diffusion map of the cells. Out come a pseudotime
per cell, the probability of each cell reaching each terminal state, and the
differentiation entropy derived from those probabilities. It needs a start cell
and, optionally, the terminal states.

**PAGA** ([Wolf et al., 2019](https://doi.org/10.1186/s13059-019-1663-x))
abstracts the cell-level graph into a cluster-level one, where the edge weight
between two clusters measures how much more connected they are than a random
null model expects. Its maximum spanning forest is the backbone people usually
draw. PAGA needs a clustering and nothing else.

The data set is the one from the [Palantir sample
notebook](https://github.com/dpeerlab/Palantir/blob/master/notebooks/Palantir_sample_notebook.ipynb):
4142 human CD34+ bone marrow cells across 16106 genes, already filtered. Three
lineages come out of it, erythroid, monocyte and dendritic cell, all seeded from
a haematopoietic stem cell pool.

The two supporting pieces from that notebook are here too. **MAGIC** ([van
Dijk et al., 2018](https://doi.org/10.1016/j.cell.2018.05.061)) smooths counts
over the same kNN graph, which is what makes the marker plots readable.
**Gene trends** fit a landmark Gaussian process per branch, the Mellon-based
estimator the reference uses rather than the legacy GAM. Both are restricted to
a gene subset in `bixverse`, for reasons spelled out where they come up.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(ggplot2)
library(data.table)
library(patchwork)
library(magrittr)
```

## Loading the data

```{r}
#| label: data path
# TODO: swap this for a download_marrow_cd34() helper once the Zenodo record
# is live
marrow_path <- path.expand(
  "~/Desktop/zenodo_data/marrow_sample_scseq_counts.h5ad"
)

read_h5ad_metadata(marrow_path)$dims
```

The `var` index of this file is the gene symbol, so no Ensembl mapping is
needed. Cell barcodes look like `Run4_120703408880541`, which matters later:
the start cell and the terminal states from the reference notebook are given as
barcodes.

```{r}
#| label: load h5ad
sc_object <- SingleCells(dir_data = tempdir())

sc_object <- load_h5ad(
  object = sc_object,
  h5_path = marrow_path,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 100L,
    min_lib_size = 250L,
    min_cells = 10L,
    target_size = 10000
  )
)

sc_object
```

The data ships pre-filtered, so there is no QC step here. On your own data you
would run `run_cell_qc()` and `set_cells_to_keep()` first, as in the
[PBMC vignette](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html).

## Feature selection, PCA and neighbours

Standard pipeline. The one choice worth spelling out is `k = 30` on the kNN
graph: Palantir builds its diffusion kernel from whatever graph is stored on the
object, and 30 neighbours is what the reference uses. The distance metric has to
be euclidean, since that is how `bixverse` decides whether the stored distances
are squared.

`pruning` has to move with `k`. The default of `1/12` is set for the default
`k = 15`, and since the threshold is a share of the neighbourhood, leaving it
alone at `k = 30` prunes twice as hard. Nothing errors, you just get cells with
too few shared neighbours dropping out as one-cell clusters.

```{r}
#| label: hvg pca neighbours
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 1500L
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    pruning = 1 / 24,
    knn = list(k = 30L, ann_dist = "euclidean")
  )
)

sc_object <- find_clusters_sc(
  sc_object,
  res = 0.7,
  name = "leiden_clusters"
)

sc_object <- umap_sc(sc_object)
```

```{r}
#| label: umap clusters
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

The four markers the reference notebook uses: `CD34` for the stem pool, `MPO`
for the myeloid arm, `GATA1` for the erythroid one and `IRF8` for the dendritic
one. Raw, these are sparse and speckled, but the gradients are visible.

```{r}
#| label: marker feature plots
#| fig-width: 9
#| fig-height: 7
markers <- c("CD34", "MPO", "GATA1", "IRF8")

feature_plot_sc(
  object = sc_object,
  features = markers,
  embedding = "umap",
  palette = "spectral"
)
```

## MAGIC imputation

MAGIC builds a row-stochastic operator over the kNN graph and applies it three
times, so every cell becomes a weighted average of its neighbourhood. It is the
same adaptive-bandwidth kernel Palantir uses for its diffusion map, so the two
agree on what a neighbourhood is.

`run_magic_sc()` writes the result onto the object as a cached layer. It takes
a gene list rather than doing everything, and that is deliberate: the output is
dense, and the whole point of the on-disk store is not holding a dense
cells-by-genes matrix. Pick the genes you want to look at.

```{r}
#| label: run magic
sc_object <- run_magic_sc(sc_object, features = markers)

get_magic(sc_object)
```

Every extractor takes `layer = "magic"`, and `feature_plot_sc()` passes it
straight through.

```{r}
#| label: magic feature plots
#| fig-width: 9
#| fig-height: 7
feature_plot_sc(
  object = sc_object,
  features = markers,
  embedding = "umap",
  palette = "spectral",
  layer = "magic"
)
```

That is the picture people know from the paper, and it is worth being clear
about what produced it. Neighbourhoods overlap, so smoothing over them makes
nearby cells resemble each other in every gene at once. Gene-gene correlation
after imputation is inflated, badly, and the inflation is a property of the
graph rather than of the biology. Do not feed these values into Hotspot,
SCENIC, differential correlation or CoReMo: those methods measure exactly the
quantity MAGIC manufactures, and they will report the graph back to you as a
finding. `extract_dot_plot_data()` has no `layer` argument for the same reason.

The layer is tracked like every other cached artefact, so it goes stale when
the graph under it moves.

```{r}
#| label: magic provenance
get_sc_cache_status(sc_object)
```

## PAGA

PAGA is the cheap one, so start there. It reads the kNN indices and a clustering,
ignores the distances entirely, and gives back a clusters-by-clusters graph.

```{r}
#| label: run paga
paga_res <- run_paga_sc(
  sc_object,
  cluster_col = "leiden_clusters"
)
```

`connectivities` is the full abstracted graph, `connectivities_tree` its maximum
spanning forest. Both are symmetric sparse matrices named by cluster, with a
zero diagonal.

```{r}
#| label: paga sizes
paga_res$sizes

round(as.matrix(paga_res$connectivities), 3)
```

`paga_plot_sc()` puts every cluster at the centroid of its cells in an
embedding and draws the graph over them. A free layout of the abstracted graph
would put the nodes somewhere arbitrary, which you then have to relate back to
the UMAP by eye.

```{r}
#| label: paga plot
#| fig-width: 8
#| fig-height: 6
paga_plot_sc(sc_object, paga_res, embedding = "umap")
```

Edge thresholding is not optional. The abstracted graph is close to complete on
real data, so at `threshold = 0` you get a hairball. `tree_only = TRUE` is the
shortcut to the backbone, and it is the readable one on anything above a
handful of clusters.

```{r}
#| label: paga tree plot
#| fig-width: 8
#| fig-height: 6
paga_plot_sc(sc_object, paga_res, embedding = "umap", tree_only = TRUE)
```

If you want the node and edge tables to draw yourself, they come out of
`extract_paga_plot_data()`, which is what the plot is built on.

## Palantir

The start cell and the three terminal states come straight from the reference
notebook. The start cell was picked there for high `CD34` expression; the
terminal states are one cell each in the dendritic, monocyte and erythroid tips.

```{r}
#| label: palantir cells
early_cell <- "Run5_164698952452459"

terminal_states <- c(
  DC = "Run5_131097901611291",
  Mono = "Run5_134936662236454",
  Ery = "Run4_200562869397916"
)
```

`knn = 30` is the neighbour count for the geodesic graph Palantir builds on the
multiscale space, a different knob from the `k` of the input kNN graph, and
`num_waypoints = 500` is the notebook's override of the 1200 default.

Two things do not mirror the reference, and both change what you get.

`use_early_cell_as_start` is `TRUE`, so the early cell you name is the one that
gets used. The reference snaps it to the nearest boundary cell of the diffusion
map instead, and that candidate set is small, at most two cells per multiscale
component. On a branching manifold the snap will cheerfully move a root cell
onto a branch tip and run the whole trajectory backwards.

`n_eigs` is pinned at 10, leaving nine multiscale components. Left at `NULL`
the count comes from the largest eigengap, which on this data picks 3, i.e. two
components. Two components cannot resolve a three-way branch: the erythroid arm
separates fine but the monocyte and dendritic fate probabilities end up
correlated, their branch masks collapse onto each other, and the gene trends
for those two lineages come out nearly identical. There is a section on that
below.

```{r}
#| label: run palantir
palantir_res <- run_palantir_sc(
  sc_object,
  early_cell = early_cell,
  terminal_states = terminal_states,
  palantir_params = params_sc_palantir(
    n_eigs = 10L,
    knn = 30L
  ),
  seed = 42L
)

palantir_res
```

Two things to watch in that print out. `repair_edges` above zero means the kNN
graph was disconnected and had to be bridged, which usually means `k` is too
small for the data. A note about the diffusion eigensolve not converging means
the embedding is under-resolved, and every distance taken on it is suspect;
raise `lanczos_max_restarts` in `params_sc_palantir()` before believing
anything downstream.

### Results back onto the object

`PalantirRes` is keyed by cell name, not position, so match it against the obs
cell names rather than assuming the two are in the same order.

```{r}
#| label: results to obs
obs_cells <- get_cell_names(sc_object, filtered = TRUE)
idx <- match(obs_cells, palantir_res$pseudotime$cell_id)

sc_object[["palantir_pseudotime"]] <- palantir_res$pseudotime$pseudotime[idx]
sc_object[["palantir_entropy"]] <- palantir_res$pseudotime$entropy[idx]
```

The fate probability matrix has one column per terminal state. Because
`terminal_states` was a named vector, those names are already the column names,
so all that is left is the dominant fate per cell.

```{r}
#| label: fate probabilities
fate_probs <- palantir_res$branch_probs[idx, ]

for (lineage in colnames(fate_probs)) {
  sc_object[[paste0("fate_", lineage)]] <- fate_probs[, lineage]
}

sc_object[["palantir_branch"]] <- colnames(fate_probs)[
  max.col(fate_probs, ties.method = "first")
]

head(round(fate_probs, 3))
```

Rows do not necessarily sum to one. Fate probabilities below
`branch_prob_threshold` (0.01 by default) are zeroed without renormalisation,
and cells whose waypoint weight sits on a stranded waypoint lose that mass
entirely.

### Pseudotime and entropy

```{r}
#| label: pseudotime plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_pseudotime",
  palette = "spectral",
  discrete = FALSE
)
```

The same quantity on the abstracted graph is worth a look, because it is the
cheapest check that the two methods agree. PAGA knows nothing about pseudotime,
so if its backbone and the pseudotime gradient point in different directions,
one of them is wrong.

```{r}
#| label: paga pseudotime
#| fig-width: 8
#| fig-height: 6
paga_plot_sc(
  sc_object,
  paga_res,
  embedding = "umap",
  node_colour_by = "palantir_pseudotime",
  palette = "spectral",
  tree_only = TRUE
)
```

Entropy is high where a cell still has several fates open to it and drops to
zero once it commits. On a clean three-lineage data set the stem pool should sit
near `log(3)`.

```{r}
#| label: entropy plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_entropy",
  palette = "spectral",
  discrete = FALSE
)
```

```{r}
#| label: fate plots
#| fig-width: 12
#| fig-height: 4
fate_plots <- lapply(colnames(fate_probs), \(lineage) {
  embedding_plot_sc(
    sc_object,
    embedding = "umap",
    colour_by = paste0("fate_", lineage),
    palette = "spectral",
    discrete = FALSE
  ) +
    labs(title = lineage)
})

wrap_plots(fate_plots, nrow = 1)
```

```{r}
#| label: branch plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_branch",
  label_by = "palantir_branch",
  discrete = TRUE
)
```

### Letting Palantir find the terminal states

Drop `terminal_states` and they get detected from the waypoint Markov chain
instead. Handy when you have no prior, but treat the result as a starting point
rather than an answer: the detection reads extremes of the multiscale space, and
on data sets this size those are not always well determined.

```{r}
#| label: palantir auto
palantir_auto <- run_palantir_sc(
  sc_object,
  early_cell = early_cell,
  palantir_params = params_sc_palantir(
    n_eigs = 10L,
    knn = 30L,
    use_early_cell_as_start = TRUE
  ),
  seed = 42L
)

palantir_auto$terminal_states
```

Where did they land relative to the supplied ones?

```{r}
#| label: auto vs supplied
data.table(
  cell_id = palantir_auto$terminal_states,
  branch = unlist(sc_object[["palantir_branch"]], use.names = FALSE)[
    match(palantir_auto$terminal_states, obs_cells)
  ],
  pseudotime = round(
    palantir_res$pseudotime$pseudotime[
      match(palantir_auto$terminal_states, palantir_res$pseudotime$cell_id)
    ],
    3
  )
)
```

## Gene trends

`run_gene_trends_sc()` does two things. First it assigns cells to branches: for
each fate, the probability threshold is an expanding quantile over the
pseudotime-sorted cells, made monotone with a cumulative maximum, so a fate's
bar can only rise as differentiation proceeds. Then it fits a landmark Gaussian
process with a Matern-5/2 kernel per branch, on a 500 point pseudotime grid.

```{r}
#| label: gene trends
trends <- run_gene_trends_sc(
  sc_object,
  palantir_res = palantir_res,
  features = markers
)

trends
```

`trends$trends` is a long data.table with `branch`, `pseudotime`, `gene` and
`expression`, which is what a facet wants.

```{r}
#| label: trend plot
#| fig-width: 9
#| fig-height: 6
ggplot(
  trends$trends,
  aes(x = pseudotime, y = expression, colour = branch)
) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~gene, scales = "free_y") +
  scale_color_bx() +
  labs(
    x = "Palantir pseudotime",
    y = "Normalised expression",
    colour = "Lineage"
  ) +
  theme_bx()
```

`CD34` falls away from the root, `GATA1` climbs on the erythroid arm, `MPO` on
the monocyte one, `IRF8` on the dendritic one. Set `use_magic = TRUE` to fit
the imputed layer instead, which is what the reference notebook does. Note that
the GP already smooths over pseudotime, so imputing first smooths twice.

`branch_cells` carries the selection itself, which is the branch cell set the
reference notebook plots.

```{r}
#| label: branch cells
purrr::map_int(trends$branch_cells, length)
```

The same trick works with any per-cell quantity. A loess through the entropy
column watches commitment happen along the trajectory.

```{r}
#| label: entropy along pseudotime
#| fig-width: 7
#| fig-height: 4
obs_dt <- get_sc_obs(sc_object, filtered = TRUE)

ggplot(
  obs_dt,
  aes(x = palantir_pseudotime, y = palantir_entropy, colour = palantir_branch)
) +
  geom_point(size = 0.4, alpha = 0.3) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, span = 0.6) +
  scale_color_bx() +
  labs(
    x = "Palantir pseudotime",
    y = "Differentiation entropy",
    colour = "Dominant fate"
  ) +
  theme_bx()
```

## Where this can go wrong

Palantir has more moving parts than most methods in this package, and several of
them fail quietly rather than loudly. In rough order of how often they bite:

- **`pruning` left at its default.** Covered above, and the only one on this
  list that is upstream of Palantir entirely. It costs you clusters, not
  pseudotime, but a one-cell cluster then walks into PAGA and comes back out as
  a fat edge into a dot, because the null model has almost no possible edges to
  normalise against.
- **The start cell.** Covered above. Set `use_early_cell_as_start = FALSE` to
  get the reference behaviour and it can move your root onto a branch tip. If
  pseudotime comes back anti-correlated with what you expect, check `start_cell`
  on the result object first. Note that pseudotime is not pinned to zero at the
  start cell either: the refinement can put the minimum on a neighbour, and a
  start cell far from zero means it disagreed with your anchor.
- **`eigen_converged = FALSE`.** The diffusion eigensolve ran out of restarts,
  so the multiscale embedding is under-resolved. Everything downstream converges
  happily on noise, which is exactly why this is reported. Raise
  `lanczos_max_restarts`.
- **`repair_edges > 0`.** The kNN graph was disconnected and Palantir bridged it
  to carry on. Raise `k` on `find_neighbours_sc()` rather than living with the
  bridges.
- **`stranded_waypoints > 0`.** Waypoints from which no terminal state is
  reachable. Their cells lose fate probability mass, so rows sum below one. A
  large count means the pseudotime-directed pruning cut too hard.
- **`n_eigs`.** The one that bites hardest, and it is silent. Left at `NULL` the
  eigenvector count comes from the largest eigengap, which on this data picks 3,
  leaving two multiscale components for a three-way branch. Nothing errors. What
  happens instead is that two of the three fates stop separating: their
  probabilities correlate, `select_branch_cells()` hands the two branches almost
  the same cells, and their gene trends come out on top of each other. Here the
  monocyte and dendritic masks shared 80% of their cells at the default, with
  exactly one cell unique to the monocyte branch, and pinning `n_eigs` to 10 is
  what fixes it. If two lineages give you the same curve, count the components
  before you blame the estimator.
- **Branch masks are permissive by design.** Even at a sensible `n_eigs`, a cell
  belongs to a branch when its fate probability is near the top of what has been
  seen up to its pseudotime, so the uncommitted trunk sits in every branch at
  once. That is the reference's logic and it is the right one, but it means two
  branches only diverge where their late cells differ. Check `branch_cells`
  overlap before reading much into two curves that hug.
- **The gene trend defaults are prior-dominated.** Pseudotime is min-max scaled
  to `[0, 1]`, so the reference's `length_scale = 1` spans the entire domain and
  its `sigma = 1` sits at roughly the signal scale of log-normalised expression.
  Almost any gene resolves into a smooth monotone or single-peaked curve. That
  is a presentation choice, not inference. Shorten `length_scale` in
  `params_sc_gene_trends()` before believing a bump.
- **Double smoothing.** MAGIC averages over the graph and the GP averages over
  pseudotime. Running both is defensible for a figure and rarely necessary for
  the trend itself. If a feature only shows up with `use_magic = TRUE`, that is
  worth knowing about it.
- **A stale imputed layer.** The layer is stamped against the kNN graph it came
  from, so `get_sc_cache_status()` flags it once the graph or the cell filter
  moves. `run_gene_trends_sc(use_magic = TRUE)` refuses to fit a stale one.
