---
title: "Trajectory inference with Palantir and PAGA"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Trajectory inference with Palantir and PAGA}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

Two trajectory methods live in `bixverse`, both operating on the kNN graph you
already built with `find_neighbours_sc()`.

**Palantir** ([Setty et al.,
2019](https://doi.org/10.1038/s41587-019-0068-4)) models differentiation as a
Markov chain over a multiscale diffusion map of the cells. Out come a pseudotime
per cell, the probability of each cell reaching each terminal state, and the
differentiation entropy derived from those probabilities. It needs a start cell
and, optionally, the terminal states.

**PAGA** ([Wolf et al., 2019](https://doi.org/10.1186/s13059-019-1663-x))
abstracts the cell-level graph into a cluster-level one, where the edge weight
between two clusters measures how much more connected they are than a random
null model expects. Its maximum spanning forest is the backbone people usually
draw. PAGA needs a clustering and nothing else.

The data set is the one from the [Palantir sample
notebook](https://github.com/dpeerlab/Palantir/blob/master/notebooks/Palantir_sample_notebook.ipynb):
4142 human CD34+ bone marrow cells across 16106 genes, already filtered. Three
lineages come out of it, erythroid, monocyte and dendritic cell, all seeded from
a haematopoietic stem cell pool.

Two things from that notebook are not reproduced here. There is no MAGIC
imputation in `bixverse` (and it is a contested step anyway), so the marker gene
plots below are noisier than the published ones. There is also no Mellon-based
gene trend estimator, so the expression-along-pseudotime section uses a plain
loess instead. Everything upstream of that, the diffusion map, the waypoints,
the Markov chain and the fate probabilities, is the same algorithm.

```{r}
#| label: setup
library(bixverse)
library(bixverse.plots)
library(ggplot2)
library(data.table)
library(patchwork)
library(magrittr)
```

## Loading the data

```{r}
#| label: data path
# TODO: swap this for a download_marrow_cd34() helper once the Zenodo record
# is live
marrow_path <- path.expand(
  "~/Desktop/zenodo_data/marrow_sample_scseq_counts.h5ad"
)

read_h5ad_metadata(marrow_path)$dims
```

The `var` index of this file is the gene symbol, so no Ensembl mapping is
needed. Cell barcodes look like `Run4_120703408880541`, which matters later:
the start cell and the terminal states from the reference notebook are given as
barcodes.

```{r}
#| label: load h5ad
sc_object <- SingleCells(dir_data = tempdir())

sc_object <- load_h5ad(
  object = sc_object,
  h5_path = marrow_path,
  sc_qc_param = params_sc_min_quality(
    min_unique_genes = 100L,
    min_lib_size = 250L,
    min_cells = 10L,
    target_size = 10000
  )
)

sc_object
```

The data ships pre-filtered, so there is no QC step here. On your own data you
would run `run_cell_qc()` and `set_cells_to_keep()` first, as in the
[PBMC vignette](https://gregorlueg.github.io/bixverse/articles/pbmc_single_cell.html).

## Feature selection, PCA and neighbours

Standard pipeline. The one choice worth spelling out is `k = 30` on the kNN
graph: Palantir builds its diffusion kernel from whatever graph is stored on the
object, and 30 neighbours is what the reference uses. The distance metric has to
be euclidean, since that is how `bixverse` decides whether the stored distances
are squared.

```{r}
#| label: hvg pca neighbours
sc_object <- find_hvg_sc(
  object = sc_object,
  hvg_no = 2000L
)

sc_object <- calculate_pca_sc(
  object = sc_object,
  no_pcs = 30L
)

sc_object <- find_neighbours_sc(
  object = sc_object,
  neighbours_params = params_sc_neighbours(
    knn = list(k = 30L, ann_dist = "euclidean")
  )
)

sc_object <- find_clusters_sc(
  sc_object,
  res = 0.5,
  name = "leiden_clusters"
)

sc_object <- umap_sc(sc_object)
```

```{r}
#| label: umap clusters
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "leiden_clusters",
  label_by = "leiden_clusters",
  discrete = TRUE
)
```

The four markers the reference notebook uses: `CD34` for the stem pool, `MPO`
for the myeloid arm, `GATA1` for the erythroid one and `IRF8` for the dendritic
one. Without imputation these are sparse and speckled, but the gradients are
visible.

```{r}
#| label: marker feature plots
#| fig-width: 9
#| fig-height: 7
feature_plot_sc(
  object = sc_object,
  features = c("CD34", "MPO", "GATA1", "IRF8"),
  embedding = "umap",
  scale = TRUE
)
```

## PAGA

PAGA is the cheap one, so start there. It reads the kNN indices and a clustering,
ignores the distances entirely, and gives back a clusters-by-clusters graph.

```{r}
#| label: run paga
paga_res <- run_paga_sc(
  sc_object,
  cluster_col = "leiden_clusters"
)

paga_res
```

`connectivities` is the full abstracted graph, `connectivities_tree` its maximum
spanning forest. Both are symmetric sparse matrices named by cluster, with a
zero diagonal.

```{r}
#| label: paga sizes
paga_res$sizes

round(as.matrix(paga_res$connectivities), 3)
```

Hand either matrix to `igraph` to draw it. The forest is the readable one on
anything above a handful of clusters, the full graph keeps the weak edges that
the forest drops.

```{r}
#| label: paga plot
#| fig-width: 10
#| fig-height: 5
paga_full <- igraph::graph_from_adjacency_matrix(
  as.matrix(paga_res$connectivities),
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

paga_tree <- igraph::graph_from_adjacency_matrix(
  as.matrix(paga_res$connectivities_tree),
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

set.seed(42L)
layout <- igraph::layout_with_fr(paga_full)

par(mfrow = c(1, 2))
plot(
  paga_full,
  layout = layout,
  edge.width = igraph::E(paga_full)$weight * 20,
  vertex.size = sqrt(paga_res$sizes$n_cells),
  vertex.color = "grey80",
  main = "Abstracted graph"
)
plot(
  paga_tree,
  layout = layout,
  edge.width = igraph::E(paga_tree)$weight * 20,
  vertex.size = sqrt(paga_res$sizes$n_cells),
  vertex.color = "grey80",
  main = "Maximum spanning forest"
)
par(mfrow = c(1, 1))
```

## Palantir

The start cell and the three terminal states come straight from the reference
notebook. The start cell was picked there for high `CD34` expression; the
terminal states are one cell each in the dendritic, monocyte and erythroid tips.

```{r}
#| label: palantir cells
early_cell <- "Run5_164698952452459"

terminal_states <- c(
  DC = "Run5_131097901611291",
  Mono = "Run5_134936662236454",
  Ery = "Run4_200562869397916"
)
```

`params_sc_palantir()` mirrors the reference defaults. `n_dcs = 5` matches
`run_diffusion_maps(n_components = 5)`, `knn = 30` is the neighbour count for
the geodesic graph Palantir builds on the multiscale space (a different knob
from the `k` of the input kNN graph), and `num_waypoints = 500` is the
notebook's override of the 1200 default.

One default does not mirror the reference. `use_early_cell_as_start` is `TRUE`
here, so the early cell you name is the one that gets used. The reference snaps
it to the nearest boundary cell of the diffusion map instead, and that candidate
set is small, at most two cells per multiscale component. On a branching
manifold the snap will cheerfully move a root cell onto a branch tip and run the
whole trajectory backwards. It is spelled out below rather than left implicit,
since it changes what you get.

```{r}
#| label: run palantir
palantir_res <- run_palantir_sc(
  sc_object,
  early_cell = early_cell,
  terminal_states = terminal_states,
  palantir_params = params_sc_palantir(
    n_dcs = 5L,
    knn = 30L,
    num_waypoints = 500L,
    use_early_cell_as_start = TRUE
  ),
  seed = 42L
)

str(palantir_res)

palantir_res$branch_probs
```

Two things to watch in that print out. `repair_edges` above zero means the kNN
graph was disconnected and had to be bridged, which usually means `k` is too
small for the data. A note about the diffusion eigensolve not converging means
the embedding is under-resolved, and every distance taken on it is suspect;
raise `lanczos_max_restarts` in `params_sc_palantir()` before believing
anything downstream.

### Results back onto the object

`PalantirRes` is keyed by cell name, not position, so match it against the obs
cell names rather than assuming the two are in the same order.

```{r}
#| label: results to obs
obs_cells <- get_cell_names(sc_object, filtered = TRUE)
idx <- match(obs_cells, palantir_res$pseudotime$cell_id)

sc_object[["palantir_pseudotime"]] <- palantir_res$pseudotime$pseudotime[idx]
sc_object[["palantir_entropy"]] <- palantir_res$pseudotime$entropy[idx]
```

The fate probability matrix has one column per terminal state, named by the
terminal cell. Rename to the lineage labels and add the dominant fate per cell.

```{r}
#| label: fate probabilities
fate_probs <- palantir_res$branch_probs[idx, ]
colnames(fate_probs) <- names(terminal_states)[
  match(colnames(fate_probs), terminal_states)
]

for (lineage in colnames(fate_probs)) {
  sc_object[[paste0("fate_", lineage)]] <- fate_probs[, lineage]
}

sc_object[["palantir_branch"]] <- colnames(fate_probs)[
  max.col(fate_probs, ties.method = "first")
]

head(round(fate_probs, 3))
```

Rows do not necessarily sum to one. Fate probabilities below
`branch_prob_threshold` (0.01 by default) are zeroed without renormalisation,
and cells whose waypoint weight sits on a stranded waypoint lose that mass
entirely.

### Pseudotime and entropy

```{r}
#| label: pseudotime plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_pseudotime",
  discrete = FALSE
)
```

Entropy is high where a cell still has several fates open to it and drops to
zero once it commits. On a clean three-lineage data set the stem pool should sit
near `log(3)`.

```{r}
#| label: entropy plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_entropy",
  discrete = FALSE
)
```

```{r}
#| label: fate plots
#| fig-width: 12
#| fig-height: 4
fate_plots <- lapply(colnames(fate_probs), \(lineage) {
  embedding_plot_sc(
    sc_object,
    embedding = "umap",
    colour_by = paste0("fate_", lineage),
    discrete = FALSE
  ) +
    labs(title = lineage)
})

wrap_plots(fate_plots, nrow = 1)
```

```{r}
#| label: branch plot
#| fig-width: 7
#| fig-height: 5
embedding_plot_sc(
  sc_object,
  embedding = "umap",
  colour_by = "palantir_branch",
  label_by = "palantir_branch",
  discrete = TRUE
)
```

### Letting Palantir find the terminal states

Drop `terminal_states` and they get detected from the waypoint Markov chain
instead. Handy when you have no prior, but treat the result as a starting point
rather than an answer: the detection reads extremes of the multiscale space, and
on data sets this size those are not always well determined.

```{r}
#| label: palantir auto
palantir_auto <- run_palantir_sc(
  sc_object,
  early_cell = early_cell,
  palantir_params = params_sc_palantir(
    n_dcs = 5L,
    knn = 30L,
    num_waypoints = 500L,
    use_early_cell_as_start = TRUE
  ),
  seed = 42L
)

palantir_auto$terminal_states
```

Where did they land relative to the supplied ones?

```{r}
#| label: auto vs supplied
data.table(
  cell_id = palantir_auto$terminal_states,
  branch = unlist(sc_object[["palantir_branch"]], use.names = FALSE)[
    match(palantir_auto$terminal_states, obs_cells)
  ],
  pseudotime = round(
    palantir_res$pseudotime$pseudotime[
      match(palantir_auto$terminal_states, palantir_res$pseudotime$cell_id)
    ],
    3
  )
)
```

## Expression along pseudotime

No Mellon here, so this is a loess through the normalised counts rather than a
proper gene trend estimator. It is enough to see the ordering: `CD34` falling
away from the root, `GATA1` climbing on the erythroid arm, `MPO` on the
monocyte one, `IRF8` on the dendritic one.

```{r}
#| label: trend data
markers <- c("CD34", "MPO", "GATA1", "IRF8")

trend_dt <- extract_feature_plot_data(
  object = sc_object,
  features = markers,
  embedding = "umap",
  obs_col = c("palantir_pseudotime", "palantir_branch")
)
```

```{r}
#| label: trend plot
#| fig-width: 9
#| fig-height: 6
ggplot(
  trend_dt,
  aes(x = palantir_pseudotime, y = expression, colour = palantir_branch)
) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE, span = 0.6) +
  facet_wrap(~gene, scales = "free_y") +
  scale_color_bx() +
  labs(
    x = "Palantir pseudotime",
    y = "Normalised expression",
    colour = "Dominant fate"
  ) +
  theme_bx()
```

The same trick works with any per-cell quantity. Swap `expression` for the
entropy column to watch commitment happen along the trajectory.

```{r}
#| label: entropy along pseudotime
#| fig-width: 7
#| fig-height: 4
obs_dt <- get_sc_obs(sc_object, filtered = TRUE)

ggplot(
  obs_dt,
  aes(x = palantir_pseudotime, y = palantir_entropy, colour = palantir_branch)
) +
  geom_point(size = 0.4, alpha = 0.3) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, span = 0.6) +
  scale_color_bx() +
  labs(
    x = "Palantir pseudotime",
    y = "Differentiation entropy",
    colour = "Dominant fate"
  ) +
  theme_bx()
```

## Where this can go wrong

Palantir has more moving parts than most methods in this package, and a couple
of them fail quietly.

- **The start cell.** Covered above. Set `use_early_cell_as_start = FALSE` to
  get the reference behaviour and it can move your root onto a branch tip. If
  pseudotime comes back anti-correlated with what you expect, check `start_cell`
  on the result object first. Note that pseudotime is not pinned to zero at the
  start cell either: the refinement can put the minimum on a neighbour, and a
  start cell far from zero means it disagreed with your anchor.
- **`eigen_converged = FALSE`.** The diffusion eigensolve ran out of restarts,
  so the multiscale embedding is under-resolved. Everything downstream converges
  happily on noise, which is exactly why this is reported. Raise
  `lanczos_max_restarts`.
- **`repair_edges > 0`.** The kNN graph was disconnected and Palantir bridged it
  to carry on. Raise `k` on `find_neighbours_sc()` rather than living with the
  bridges.
- **`stranded_waypoints > 0`.** Waypoints from which no terminal state is
  reachable. Their cells lose fate probability mass, so rows sum below one. A
  large count means the pseudotime-directed pruning cut too hard.
- **`n_eigs`.** Left at `NULL` the eigenvector count comes from the largest
  eigengap. On small or thin manifolds that heuristic keeps too many, and the
  higher harmonics oscillate enough to swamp the pseudotime gaps. Pinning it to
  3, which leaves two multiscale components, is a reasonable first thing to try
  when results look like noise.

## Not covered

MAGIC imputation, Mellon gene trends and the branch cell selection from the
reference notebook all have no equivalent here yet. Gene trends are the one
worth adding: a proper function estimator over pseudotime would make the section
above a real analysis rather than a sanity check. PAGA has no plotting helper
either, hence the raw `igraph` calls, which is an obvious gap to close.
