---
title: "Analysing a Visium slide with bixverse"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Analysing a Visium slide with bixverse}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  - name: "Gregor Lueg"
---

## Intro

A full pass over one Visium capture area: ingest, spatial graph, spatially
variable genes by two methods, clustering, neighbourhood enrichment and per-spot
histology features.

This assumes you have read the
[I/O vignette](https://gregorlueg.github.io/bixverse/articles/spatial_io.html),
and that you are comfortable with how `SingleCells` keeps counts on disk. The
[design doc](https://gregorlueg.github.io/bixverse/articles/design_spatial.html)
covers why the spatial layer looks the way it does, including the bits that are
not finished.

```{r}
#| label: setup
library(bixverse)
library(data.table)
library(ggplot2)
library(magrittr)
```

Point this at your own Space Ranger output.

```{r}
#| label: load
visium_dir <- path.expand("~/Desktop/visium_example_data")

data_dir <- file.path(tempdir(), "visium_worked")
dir.create(data_dir, recursive = TRUE, showWarnings = FALSE)

object <- load_visium(
  SpatialSpot(dir_data = data_dir),
  visium_dir = visium_dir,
  sp_io_param = params_sp_visium_io(),
  sc_qc_param = params_sc_min_quality(),
  exp_id = "example",
  .verbose = FALSE
)

object
```

3,827 in-tissue spots and 18,945 genes after the default QC. Off-tissue spots
were dropped at ingest, so the gene filter never saw them.

## A first look

The coordinates are just obs columns, so plotting is `ggplot2` and nothing else.

```{r}
#| label: qc plot
#| fig-width: 7
#| fig-height: 6
obs <- get_sc_obs(object)

ggplot(obs, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres)) +
  geom_point(aes(colour = log10(lib_size)), size = 0.7) +
  scale_colour_viridis_c(name = expression(log[10] ~ "counts")) +
  coord_fixed() +
  theme_minimal() +
  labs(x = NULL, y = NULL, title = "Library size across the capture area")
```

The `y` axis is negated because pixel rows count downwards from the top of the
image while `ggplot2` counts upwards.

## The spatial graph

Everything except SPARK-X needs to know which spots touch. On Visium the capture
geometry is known exactly, so the graph is a table look-up over the array
coordinates rather than a distance computation.

```{r}
#| label: graph
object <- build_spatial_graph_sp(
  object,
  graph_params = params_sp_graph(layout = "hex", weighting = "binary"),
  .verbose = TRUE
)
```

22,302 directed edges, a median of 6 neighbours per spot, and a median
symmetrised degree of 12. Those last two are different numbers describing the
same thing: `degree` is the row sum of `W + Wᵀ`, so on a reciprocal binary
lattice it is exactly twice the neighbour count. If you want a neighbour count,
count the neighbours.

```{r}
#| label: graph shape
graph <- get_per_sample_spatial_graph(object, "example")
dim(graph)

neighbour_counts <- diff(Matrix::t(graph)@p)
table(neighbour_counts)
```

The spots with fewer than six neighbours are on the tissue boundary, which is
most of the perimeter of any real section.

## Spatially variable genes

### Moran's I

Classic spatial autocorrelation against the graph you just built, tested under
the Cliff-Ord normality null. This is the same statistic and the same null as
`spdep::moran.test(randomisation = FALSE)`.

```{r}
#| label: morans
object <- morans_i_sp(object, .verbose = FALSE)
morans <- get_per_sample_morans_i(object, "example")
morans
```

Under 300 ms for 18,945 genes. `E[I]` and `Var[I]` depend only on the number of
spots and the graph weights, so they get computed once and per-gene work reduces
to a single sparse quadratic form.

```{r}
#| label: morans top
# plain data.table on the way out: `SpMoransRes` sits in front of a data.table
# and `[` carries the class onto anything derived from it, so a slice would
# still print through the result class rather than as the table it now is
morans_dt <- data.table(
  gene_id = morans$gene_id,
  morans_i = morans$morans_i,
  fdr = morans$fdr
)

morans_dt[is.finite(morans_i)][order(-morans_i)][1:10]
```

```{r}
#| label: morans plot
#| fig-width: 7
#| fig-height: 5
plot(morans)
```

Worth a caveat: a gene with zero variance across spots has no defined `I`, comes
back as `NaN`, and the FDR adjustment currently maps it to zero rather than
`NaN`. With the default `min_cells = 10` there are no constant genes here, so it
does not arise. Drop `min_cells` to zero and it will, so filter on
`is.finite(morans_i)` before you read the FDR column.

### How does that compare to Space Ranger?

Space Ranger computes its own Moran's I and ships it as
`spatial_enrichment.csv`. It builds its own neighbour graph, so this is a rank
comparison rather than an equality check.

```{r}
#| label: space ranger
sr <- fread(file.path(visium_dir, "spatial", "spatial_enrichment.csv"))
setnames(sr, c("Feature ID", "I"), c("gene_id", "space_ranger_i"))

joined <- merge(
  morans_dt[, .(gene_id, bixverse_i = morans_i)],
  sr[, .(gene_id, space_ranger_i)],
  by = "gene_id"
)
joined <- joined[is.finite(bixverse_i) & is.finite(space_ranger_i)]

nrow(joined)
cor(joined$bixverse_i, joined$space_ranger_i, method = "spearman")
```

```{r}
#| label: space ranger plot
#| fig-width: 6
#| fig-height: 5
ggplot(joined, aes(x = space_ranger_i, y = bixverse_i)) +
  geom_point(size = 0.4, alpha = 0.3) +
  geom_abline(slope = 1, intercept = 0, colour = "#B2182B") +
  theme_minimal() +
  labs(x = "Space Ranger", y = "bixverse", title = "Moran's I")
```

Rank correlation around 0.99. The points sit off the diagonal because the two
graphs differ, not because the statistic does.

### SPARK-X

Kernel-based and non-parametric. It needs no graph, working straight off the
coordinates with ten kernels by default: five Gaussian, five cosine, bandwidths
taken from quantiles of the pairwise distance distribution.

```{r}
#| label: sparkx
object <- sparkx_sp(object, .verbose = FALSE)
sparkx <- get_per_sample_sparkx(object, "example")
sparkx
```

```{r}
#| label: sparkx plot
#| fig-width: 7
#| fig-height: 4
plot(sparkx)
```

```{r}
#| label: overlap
both <- merge(
  morans_dt[is.finite(morans_i), .(gene_id, moran_fdr = fdr)],
  data.table(gene_id = sparkx$gene_id, sparkx_fdr = sparkx$fdr),
  by = "gene_id"
)

both[, .(
  moran_only = sum(moran_fdr < 0.05 & sparkx_fdr >= 0.05),
  sparkx_only = sum(sparkx_fdr < 0.05 & moran_fdr >= 0.05),
  both = sum(moran_fdr < 0.05 & sparkx_fdr < 0.05)
)]
```

SPARK-X is much the more liberal of the two. It is testing a different null:
Moran's I asks whether neighbouring spots agree, SPARK-X asks whether expression
covaries with position under any of ten smooth kernels, and on real tissue almost
everything expressed does a bit of the latter. Take the intersection when you
want a short list.

```{r}
#| label: top svg plot
#| fig-width: 9
#| fig-height: 4
top_genes <- morans_dt[is.finite(morans_i)][order(-morans_i)][1:2]$gene_id

expr <- extract_gene_expression(
  object,
  features = top_genes,
  obs_cols = c("pxl_col_in_fullres", "pxl_row_in_fullres")
)

plot_dt <- melt(
  expr,
  id.vars = c("cell_id", "pxl_col_in_fullres", "pxl_row_in_fullres"),
  variable.name = "gene",
  value.name = "norm_expr"
)
plot_dt[, x := pxl_col_in_fullres]
plot_dt[, y := -pxl_row_in_fullres]
plot_dt[, value := log1p(norm_expr)]

ggplot(plot_dt, aes(x = x, y = y, colour = value)) +
  geom_point(size = 0.5) +
  scale_colour_viridis_c(name = "log1p(norm)") +
  facet_wrap(~gene) +
  coord_fixed() +
  theme_minimal() +
  labs(x = NULL, y = NULL)
```

## Clustering the spots

`SpatialSpot` inherits from `SingleCells`, so the standard pipeline works
unchanged. Nothing about this step is spatially aware: it clusters on expression
and the spatial coherence you see afterwards is earned, not imposed.

```{r}
#| label: clustering
object <- object %>%
  find_hvg_sc(hvg_no = 2000L, .verbose = FALSE) %>%
  calculate_pca_sc(no_pcs = 30L, .verbose = FALSE) %>%
  find_neighbours_sc(.verbose = FALSE)

object <- find_clusters_sc(object, res = 0.8, name = "spot_cluster")

table(get_sc_obs(object)$spot_cluster)
```

```{r}
#| label: cluster plot
#| fig-width: 7
#| fig-height: 6
obs <- get_sc_obs(object)

ggplot(obs, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres)) +
  geom_point(aes(colour = factor(spot_cluster)), size = 0.7) +
  coord_fixed() +
  theme_minimal() +
  labs(x = NULL, y = NULL, colour = "cluster")
```

## Neighbourhood enrichment

Do those clusters sit next to each other more or less often than chance? The
permutation shuffles the labels and keeps the graph, so the null carries the
spatial structure of the slide rather than destroying it.

```{r}
#| label: nhood
object <- nhood_enrichment_sp(
  object,
  label_col = "spot_cluster",
  nhood_params = params_sp_nhood(n_perm = 1000L),
  seed = 42L,
  .verbose = FALSE
)

nhood <- get_per_sample_nhood_enrichment(object, "example", "spot_cluster")
nhood
```

```{r}
#| label: nhood plot
#| fig-width: 7
#| fig-height: 6
plot(nhood)
```

A strong positive diagonal is what spatially coherent domains look like: a
cluster's spots are overwhelmingly neighboured by its own. The entries worth
reading are the positive off-diagonal ones, which are the compartments that
actually touch.

Results nest by `exp_id` and then by label column, so running this against
pathologist annotations as well keeps both.

## Image features

One tile per spot out of a registered slide image, reduced to 29 numbers: first
order statistics per colour channel, the same after haematoxylin/eosin colour
deconvolution, mean optical density, and Haralick texture off a grey-level
co-occurrence matrix.

```{r}
#| label: image features
object <- image_features_sp(
  object,
  resolution = "hires",
  image_params = params_sp_image(tile_scale = 1.0),
  .verbose = FALSE
)

features <- get_per_sample_image_features(object, "example", "hires")
features
features$feature_names
```

The result joins straight back onto the obs table.

```{r}
#| label: join features
object <- add_sc_new_obs(object, get_data(features))
obs <- get_sc_obs(object)

ggplot(obs, aes(x = pxl_col_in_fullres, y = -pxl_row_in_fullres)) +
  geom_point(aes(colour = haem_mean), size = 0.7) +
  scale_colour_viridis_c(option = "magma", name = "haematoxylin") +
  coord_fixed() +
  theme_minimal() +
  labs(x = NULL, y = NULL)
```

### Read the texture features with suspicion

The spot is 256 px across at full resolution. In the `hires` PNG that is 14.4 px,
and in the CytAssist frame 21.6 px.

```{r}
#| label: spot size
sample <- get_sample(object, "example")
spot_fullres <- sample$scale_factors[["spot_diameter_fullres"]]

data.table(
  resolution = c("lowres", "hires", "cytassist"),
  scale_factor = c(
    sample$scale_factors[["tissue_lowres_scalef"]],
    sample$scale_factors[["tissue_hires_scalef"]],
    sample$scale_factors[["cytassist_scalef"]]
  )
)[, spot_px := round(spot_fullres * scale_factor, 1)][]
```

Colour statistics on a 14 px tile are fine. A grey-level co-occurrence matrix on
one is close to noise: about 200 pixels, quantised to 32 grey levels, and you are
asking for a second-order statistic off that. The `glcm_*` columns come back
finite and they do not mean much.

Texture becomes worth reading once a real scanner slide is registered under
`fullres` through `params_sp_visium_io(slide_file = ...)`. Space Ranger does not
ship one, and this data set offers nothing above 21.6 px.

One more thing that catches people out. The CytAssist frame is a TIFF, and the
Rust image backend reads TIFF while the R side does not:

```{r}
#| label: cytassist
#| error: true
get_image(object, "example", "cytassist")
```

```{r}
#| label: cytassist features
object <- image_features_sp(
  object,
  resolution = "cytassist",
  .verbose = FALSE
)

get_per_sample_image_features(object, "example", "cytassist")
```

Features off an image you cannot load into R. Both statements are true and the
asymmetry is where the two halves happened to land, not a decision anyone made.

## As a pipeline

The spatial steps chain like the single cell ones, which matters once you have
more than one capture area and want the same treatment applied to each.

```{r}
#| label: pipeline
pipeline <- step_spatial_graph_sp(.verbose = FALSE) %>>%
  step_morans_i_sp(.verbose = FALSE) %>>%
  step_sparkx_sp(.verbose = FALSE)

validate_pipeline(pipeline, "SpatialSpotSubset")

subset_obj <- apply_pipeline(
  pipeline,
  SpatialSpotSubset(object, exp_id = "example")
)

get_per_sample_morans_i(subset_obj, "example")
```

`SpatialSpotSubset` carries exactly one capture area and shares the count
connection with its parent rather than copying it, so this is cheap.

## What is not here yet

Per-sample PCA has a slot in the cache and nothing fills it. There is no
spatially aware aggregation into superspots. Ingest wants a Space Ranger
directory, so h5ad from squidpy and image-free counts have no path in. And no
vendor slide format has been through the image code, so `fullres` currently means
generic tiled TIFF and nothing else.
